### 2019-06-10 submission/manuscript.Rmd
Rough in initial stages of manuscript

---
bibliography: references.bib
output:
  pdf_document:
    keep_tex: true
    includes:
      in_header: header.tex
csl: mbio.csl
fontsize: 11pt
geometry: margin=1.0in
---

```{r knitr_settings, tidy=TRUE, eval=TRUE, echo=FALSE, cache=FALSE}
options(tidyverse.quiet = TRUE)

library(tidyverse)
library(knitr)

opts_chunk$set("tidy" = TRUE)
opts_chunk$set("echo" = FALSE)
opts_chunk$set("eval" = TRUE)
opts_chunk$set("warning" = FALSE)
opts_chunk$set("message" = FALSE)
opts_chunk$set("cache" = FALSE)

inline_hook <- function(x){
  print(x)
  if(is.list(x)){
    x <- unlist(x)
  }
  if(is.numeric(x)){
    if(abs(x - round(x)) < .Machine$double.eps^0.5){
      paste(format(x,big.mark=',', digits=0, scientific=FALSE))
    } else {
      paste(format(x,big.mark=',', digits=2, nsmall=2, scientific=FALSE))
    }
  } else {
      paste(x)
  }
}
knitr::knit_hooks$set(inline=inline_hook)
```

\vspace*{10mm}

# Removal of rare sequences from 16S rRNA gene sequence surveys biases the interpretation of community structure data

\vspace{35mm}

Patrick D. Schloss${^1}$${^\dagger}$

\vspace{40mm}

$\dagger$ To whom correspondence should be addressed: \href{mailto:pschloss@umich.edu}{pschloss@umich.edu}

$1$ Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI 48109

\vspace{35mm}

## Observation format

\newpage
\linenumbers

## Abstract
250 words

\newpage

## Importance
150 words

\newpage

Two elements that are held in tension in the analysis of 16S rRNA gene sequence data is how to adequately remove PCR and sequencing artifacts and decrease the granularity of the taxonomic level that is used in the analysis. Previous attempts have included screening for sequencing quality based on quality scores [Kozich/Edgar] followed by a polishing step based on the frequency of the sequences relative to similar sequences [Kozich/Edgar/DeBlur]. Other pipelines model the quality scores and types of errors to cluster sequences directly [Dada2]. As a final step, many pipelines advocate for removing rare sequences from each dataset prior to further clustering of sequences into an operational taxonomic unit (OTU) or phylogenetic group [Knight/Edgar/DeBlur/Dada2]. The assumption of this abundance-based screening approach is that rare sequences are more likely to be artifacts than more abundant sequences. Some pipelines remove all sequences that appear once (i.e. singletons) [XXXX], XXXXXX [XXXXXX], XXXXX [XXXXXXX], or XXXXXX [XXXXXX] times; only the mothur-based pipeline discourages the practice of removing rare sequences. Proponents of this approach point to the ability to reach the desired number of OTUs with data generated from sequencing mock communities when rare sequences are removed. This approach effectively fits the sequence curation approach to data generated from a phylogenetically simple community with an atypical community distribution.

Sequencing of mock communities confirms that artifacts tend to be rare; however, artifacts can also be more abundant and replicate across samples. Furthermore, sequencing of real samples suggests that there are also good sequences that are rare. Ignored in the suggestion that researchers should remove rare sequences from each sample is the common result that the number of sequences generated from each sample may vary by two or three orders of magnitude. Thus a sequence that appears once in a sample with 2,000 sequences is more trustworthy than a sequence that appears once in a sample with 100,000 sequences. Removing such sequences potentially biases the community structure of the sample with fewer sequences. In the current study, I use published sequence data to investigate the effects of removing singletons and other rare sequences on downstream analysis of microbial communities.

***Results.*** 600 words
Sample description
* Logic for selecting studies (Table S1)
* Distribution of sequence counts across samples in each study (Figure S1)

Change in intra-sample variation with removal of rare sequences
* Lose information within each sample as you remove sequences
* Repeat when sequences are randomized across samples to maintain marginal sums

Change in inter-sample variation with removal of rare sequences
* Increased variation between samples as you remove sequences
* Repeat when sequences are randomized across samples to maintain marginal sums

***Conclusion.*** Removing rare sequences decreases the diversity represented by 16S rRNA gene sequence data and increases the variation between samples. Such impacts will hinder the statistical power to differentiate between treatment groups. Instead of removing rare sequences, researchers should focus on optimizing their sequence generation to minimize the amount of PCR and sequencing errors. In addition, samples should be rarefied to a common number of sequences across samples without prior culling of rare sequences. The number of artifacts is correlated to the number of sequences being considered. With this in mind, rarefaction allows one to control for uneven sampling effort and to control for the number of artifacts in the analysis.

Need to treat rare sequences with a grain of salt

\newpage

## Acknowledgements

\newpage

## Materials and Methods

\newpage

## References
25 references

<div id="refs"></div>

\newpage

<!--- \includegraphics{figure_1.png} -->

**Figure 1.**

\newpage

<!--- \includegraphics{figure_2.png} -->

**Figure 2.**

### 2019-06-10 submission/manuscript.tex
Rough in initial stages of manuscript

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\providecommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{}
    \pretitle{\vspace{\droptitle}}
  \posttitle{}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}

\usepackage[none]{hyphenat}

\usepackage{setspace}
\doublespacing
\setlength{\parskip}{1em}

\usepackage{lineno}

\usepackage{pdfpages}

\begin{document}

\vspace*{10mm}

\hypertarget{removal-of-rare-sequences-from-16s-rrna-gene-sequence-surveys-biases-the-interpretation-of-community-structure-data}{%
\section{Removal of rare sequences from 16S rRNA gene sequence surveys
biases the interpretation of community structure
data}\label{removal-of-rare-sequences-from-16s-rrna-gene-sequence-surveys-biases-the-interpretation-of-community-structure-data}}

\vspace{35mm}

Patrick D. Schloss\({^1}\)\({^\dagger}\)

\vspace{40mm}

\(\dagger\) To whom correspondence should be addressed:
\href{mailto:pschloss@umich.edu}{pschloss@umich.edu}

\(1\) Department of Microbiology and Immunology, University of Michigan,
Ann Arbor, MI 48109

\vspace{35mm}

\hypertarget{observation-format}{%
\subsection{Observation format}\label{observation-format}}

\newpage
\linenumbers

\hypertarget{abstract}{%
\subsection{Abstract}\label{abstract}}

\newpage

\hypertarget{importance}{%
\subsection{Importance}\label{importance}}

\newpage

Question: What impact does removing singletons (and other rare
sequences) have on downstream data analysis?

\begin{itemize}
\tightlist
\item
  Many pipelines remove rare sequences prior to clustering sequences
  into OTUs or as part of the generation fo ASVs
\item
  Singletons can be sequences that appear 1-10 times in larger samples
\item
  Seeks to solve the problem of low quality sequence data - more
  singletons with higher error rates
\item
  Likely there are rare sequences that are good - e.g.~high diversity
  samples
\item
  People have overfit their pipelines ot mock community data which have
  unrealistic community distributions (i.e.~uniform abundance or log
  abundance) and phylogenetic diversity
\item
  Better solutions would include:

  \begin{itemize}
  \tightlist
  \item
    Get better sequence data
  \item
    Rarefy data so that the number of errors are controlled for across
    samples
  \end{itemize}
\item
  Not all samples are sequenced evenly, so a singleton doesn't hav the
  same relative abundance across all samples
\end{itemize}

H: Removing rare sequences from each sample will skew the community
distribution to impact the amount of variation in the relative abundance
and alpha and beta diversity values across samples

\newpage

\hypertarget{acknowledgements}{%
\subsection{Acknowledgements}\label{acknowledgements}}

\newpage

\hypertarget{materials-and-methods}{%
\subsection{Materials and Methods}\label{materials-and-methods}}

\newpage

\hypertarget{references}{%
\subsection{References}\label{references}}

\hypertarget{refs}{}

\newpage

\textbf{Figure 1.}

\newpage

\textbf{Figure 2.}

\end{document}

### 2019-06-12 code/alpha.R
Create pipeline for imputing an effect size

@@ -1,0 +1,550 @@
+alpha_files <- c("data/bioethanol/data.100.10.otu.ealpha_diversity", "data/bioethanol/data.54.5.otu.ealpha_diversity",
+"data/bioethanol/data.100.11.otu.ealpha_diversity", "data/bioethanol/data.54.6.otu.ealpha_diversity",
+"data/bioethanol/data.100.1.otu.ealpha_diversity", "data/bioethanol/data.54.7.otu.ealpha_diversity",
+"data/bioethanol/data.100.2.otu.ealpha_diversity", "data/bioethanol/data.54.8.otu.ealpha_diversity",
+"data/bioethanol/data.100.3.otu.ealpha_diversity", "data/bioethanol/data.54.9.otu.ealpha_diversity",
+"data/bioethanol/data.100.4.otu.ealpha_diversity", "data/bioethanol/data.5.4.otu.ealpha_diversity",
+"data/bioethanol/data.100.5.otu.ealpha_diversity", "data/bioethanol/data.55.10.otu.ealpha_diversity",
+"data/bioethanol/data.100.6.otu.ealpha_diversity", "data/bioethanol/data.55.11.otu.ealpha_diversity",
+"data/bioethanol/data.100.7.otu.ealpha_diversity", "data/bioethanol/data.55.1.otu.ealpha_diversity",
+"data/bioethanol/data.100.8.otu.ealpha_diversity", "data/bioethanol/data.55.2.otu.ealpha_diversity",
+"data/bioethanol/data.100.9.otu.ealpha_diversity", "data/bioethanol/data.55.3.otu.ealpha_diversity",
+"data/bioethanol/data.10.10.otu.ealpha_diversity", "data/bioethanol/data.55.4.otu.ealpha_diversity",
+"data/bioethanol/data.10.11.otu.ealpha_diversity", "data/bioethanol/data.55.5.otu.ealpha_diversity",
+"data/bioethanol/data.10.1.otu.ealpha_diversity", "data/bioethanol/data.55.6.otu.ealpha_diversity",
+"data/bioethanol/data.10.2.otu.ealpha_diversity", "data/bioethanol/data.55.7.otu.ealpha_diversity",
+"data/bioethanol/data.10.3.otu.ealpha_diversity", "data/bioethanol/data.55.8.otu.ealpha_diversity",
[…]

### 2019-06-12 code/amova_analysis.R

@@ -1,0 +1,30 @@
+library(tidyverse)
+
+input <- commandArgs(trailingOnly=TRUE)
+
+amova_files <- input[-length(input)]
+output_file_name <- input[length(input)]
+
+print(amova_files)
+
+get_summary <- function(amova_file){
+
+	seed <- str_replace(amova_file, ".*/data\\.(\\d*)\\..*", "\\1") %>% as.integer()
+	prune <- str_replace(amova_file, ".*/data\\.\\d*\\.(\\d*).*", "\\1") %>% as.integer()
+	clustering <- str_replace(amova_file, ".*/data\\.\\d*\\.\\d*\\.(.*)\\..amova", "\\1")
+
+	p_value <- scan(amova_file, what=character(), sep="\n", quiet=TRUE) %>% str_subset(., pattern="p-value: ") %>% str_replace(., "p-value: ", "") %>% str_replace(., "\\*", "")
+
+	p_value <- ifelse(str_detect(p_value, "<"), 0.000, as.numeric(p_value))
+
+	tibble(prune=prune, seed=seed, clustering=clustering, p_value = p_value)
+
+}
+
+
+map_df(amova_files, get_summary) %>%
+	mutate(sig = p_value < 0.05) %>%
+	group_by(clustering, prune) %>%
+	summarize(frac_sig = mean(sig)) %>%
+	ungroup() %>%
+	write_tsv(output_file_name)

### 2019-06-12 code/effect_prune.R

@@ -1,0 +1,86 @@
+library(tidyverse)
+library(data.table)
+
+input <- commandArgs(trailingOnly=TRUE)
+path <- input[1] # e.g. path <- 'data/mice/'
+seed <- as.numeric(input[2]) # e.g. seed <- 1
+min_class <- as.numeric(input[3]) # e.g. min_class <- 2
+fraction <- as.numeric(input[4]) # e.g. fraction <- 0.9
+
+set.seed(seed)	#	set the random number generator seed so multiple runs generate the same
+								#	randomization
+
+
+#	Here we read in the count_table, clean it up a bit, gather it, and remove those sequence/samples
+# combinations that are zero.
+orig_count <- fread(paste0(path, "/data.count_table"), colClasses=c(Representative_Sequence="character")) %>%
+	melt(id.vars=c("Representative_Sequence"), variable.name="group", variable.factor=F, value.name="n_seqs") %>%
+  filter(group != "total") %>%
+	rename(sequences=Representative_Sequence) %>%
+	filter(n_seqs != 0)
+
+# If there are groups that need to be removed (see data.remove_accnos) then let's remove them...
+remove_groups <- scan(paste0(path, "/data.remove_accnos"), what="character", sep="-", quiet=TRUE)
+
+if(length(remove_groups) > 0){
+  orig_count <- orig_count %>% filter(!group %in% remove_groups)
+}
+
+
+full_dataset <- orig_count %>% group_by(sequences) %>% summarize(n_seqs = sum(n_seqs))
+sub_dataset <- sample_frac(full_dataset, fraction)
+
+# full_join(full_dataset, sub_dataset, by="sequences") %>%
+# 	mutate(n_seqs.y = replace_na(n_seqs.y, 0),
+# 				rel_abund.x =n_seqs.x/sum(n_seqs.x),
+# 				rel_abund.y = n_seqs.y/sum(n_seqs.y)
+# 			) %>%
+# 	group_by(sequences) %>%
+# 	mutate(min_rel_abund = min(rel_abund.x, rel_abund.y)) %>%
+# 	ungroup() %>%
+# 	summarize(shannon.x = -sum(rel_abund.x * log(rel_abund.x)),
+# 						shannon.y = -sum(rel_abund.y * ifelse(rel_abund.y != 0, log(rel_abund.y), 0)),
+# 						richness.x = sum(n_seqs.x != 0),
+# 						richness.y = sum(n_seqs.y != 0),
+# 						bc = 1-sum(min_rel_abund)
+# 					)
+
+
[…]

### 2019-06-12 code/get_beta_diversity_matrix.sh

@@ -1,0 +1,11 @@
+SHARED=$1
+
+mothur "#set.seed(seed=19760620);
+dist.shared(shared=$SHARED, calc=braycurtis, subsample=T, iters=100, processors=1)"
+
+INDIST=`echo $SHARED | sed -e "s/.shared/braycurtis*.ave.dist/"`
+OUTDIST=`echo $SHARED | sed -e "s/shared/beta_matrix/"`
+
+mv $INDIST $OUTDIST
+rm -f `echo $INDIST | sed -e "s/ave/std/"`
+rm -f `echo $INDIST | sed -e "s/ave.//"`

### 2019-06-12 code/prune_rand_otu_shared.R

@@ -15,10 +15,10 @@
 library(tidyverse)
 
 input <- commandArgs(trailingOnly=TRUE)
-rand_pruned_file <- input[1] #e.g. rand_pruned_file <-'data/bioethanol/data.1.1.rand_pruned_groups'
+rand_pruned_file <- input[1] #e.g. rand_pruned_file <-'data/bioethanol/data.1.1.effect_pruned_groups'
 otu_mapping_file <- input[2] #e.g. otu_mapping_file <- 'data/bioethanol/data.otu_seq.map'
 
-shared_file <- str_replace(rand_pruned_file, "rand_pruned_groups", "otu.rshared")
+shared_file <- str_replace(rand_pruned_file, "([^.])[^.]*_pruned_groups", "otu.\\1shared")
 
 
 # load the mapping file that indicates the number of times each pre-clustered sequence showed up in
@@ -43,4 +43,3 @@ inner_join(group_seqs_count_mapping, otu_seq_otu_mapping, by="sequences") %>%
 	rename(Group=group) %>%
 	select(label, Group, numOtus, everything()) %>%
 	write_tsv(shared_file)
-

### 2019-06-12 code/prune_rand_pc_shared.R

@@ -2,8 +2,8 @@ library(tidyverse)
 library(data.table)
 
 input <- commandArgs(trailingOnly=TRUE)
-pc_mapping_file <- input[1] # e.g. pc_mapping_file <- 'data/bioethanol/data.1.1.rand_pruned_groups'
-count_file <- str_replace(pc_mapping_file, "rand_pruned_groups", "pc.temp_count")
+pc_mapping_file <- input[1] # e.g. pc_mapping_file <- 'data/bioethanol/data.1.1.effect_pruned_groups'
+count_file <- str_replace(pc_mapping_file, "([^.])[^.]*_pruned_groups", "pc.\\1temp_count")
 
 # read in the mapping file and generate column names that are formatted like Otu0001, Otu0002, etc.
 # and then join these back into the original dataframe and spread the columns and format to be
@@ -19,4 +19,4 @@ system(paste0("mothur '#make.shared(count=", count_file, ", label=pc)'"))
 
 unlink(str_replace(count_file, "temp_count", "map"))
 unlink(count_file)
-file.rename(str_replace(count_file, "temp_count", "shared"), str_replace(count_file, "temp_count", "rshared"))
+file.rename(str_replace(count_file, ".temp_count", "shared"), str_replace(count_file, "temp_count", "shared"))

### 2019-06-12 code/run_amova.sh

@@ -1,0 +1,9 @@
+DIST=$1
+DESIGN=$2
+
+mothur "#amova(phylip=$DIST, design=$DESIGN)"
+
+OUTPUT=`echo $DIST | sed -e "s/.beta_matrix/amova/"`
+AMOVA=`echo $DIST | sed -e "s/beta_matrix/amova/"`
+
+mv $OUTPUT $AMOVA

### 2019-06-12 code/run_wilcox.R

@@ -1,0 +1,47 @@
+library(tidyverse)
+library(broom)
+
+input <- commandArgs(trailingOnly=TRUE)
+alpha_files <- input
+# source("code/alpha.R")
+# alpha_files <- list.files('data/bioethanol', pattern="ealpha_diversity", full.names=T)
+
+output_file_name <- paste0(dirname(alpha_files[1]), "/data.effect.alpha_summary")
+
+
+get_summary <- function(alpha_file){
+
+	design_file <- str_replace(alpha_file, "[^.]*\\.ealpha_diversity", "design")
+
+	seed <- str_replace(alpha_file, ".*/data\\.(\\d*)\\..*", "\\1") %>% as.integer()
+	prune <- str_replace(alpha_file, ".*/data\\.\\d*\\.(\\d*).*", "\\1") %>% as.integer()
+	clustering <- str_replace(alpha_file, ".*/data\\.\\d*\\.\\d*\\.(.*)\\.ealpha_diversity", "\\1")
+
+	read_tsv(alpha_file,
+					col_types=cols(label=col_character(), group=col_character(), method=col_character())) %>%
+		filter(method == "ave") %>%
+		select(group, sobs, shannon, invsimpson) %>%
+		inner_join(.,
+							read_tsv(design_file, col_names=c("group", "grouping"), col_types="cc"),
+							by="group") %>%
+		gather(sobs, shannon, invsimpson, key="metric", value="value") %>%
+		group_by(metric) %>%
+		nest() %>%
+		mutate(test = map(data, ~tidy(wilcox.test(.x$value~.x$grouping, exact=FALSE))),
+					median_A = map(data, ~.x %>% filter(grouping=="A") %>% summarize(median_A=median(value))),
+					median_B = map(data, ~.x %>% filter(grouping=="B") %>% summarize(median_B=median(value))),
+				) %>%
+		select(metric, median_A, median_B, test) %>%
+		unnest() %>%
+		mutate(seed = seed, prune=prune, method=clustering) %>%
+		select(prune, seed, method, metric, median_A, median_B, p.value)
+
+}
+
+
+map_df(alpha_files, get_summary) %>%
+	mutate(sig = p.value < 0.05) %>%
+	group_by(method, metric, prune) %>%
+	summarize(frac_sig = mean(sig)) %>%
+	ungroup() %>%
+	write_tsv(output_file_name)

### 2019-11-05 code/bffect_prune.R
Complete simulations

@@ -2,10 +2,14 @@ library(tidyverse)
 library(data.table)
 
 input <- commandArgs(trailingOnly=TRUE)
-path <- input[1] # e.g. path <- 'data/mice/'
+path <- input[1] # e.g. path <- 'data/bioethanol/'
 seed <- as.numeric(input[2]) # e.g. seed <- 1
-min_class <- as.numeric(input[3]) # e.g. min_class <- 2
-fraction <- as.numeric(input[4]) # e.g. fraction <- 0.9
+min_class <- as.numeric(input[3]) # e.g. min_class <- 1
+perturb_factor <- as.numeric(input[4]) # e.g. perturb_factor <- 0.1  # becomes 1.1
+perturb_fraction <- as.numeric(input[5]) # e.g. perturb_fraction <- 0.2
+
+# the effect size shrinks with increasing value of min_class. perhaps take original community
+# structures and mutes/amplifies selected OTUs to preserve intra treatment variation
 
 set.seed(seed)	#	set the random number generator seed so multiple runs generate the same
 								#	randomization
@@ -13,7 +17,8 @@ set.seed(seed)	#	set the random number generator seed so multiple runs generate 
 
 #	Here we read in the count_table, clean it up a bit, gather it, and remove those sequence/samples
 # combinations that are zero.
-orig_count <- fread(paste0(path, "/data.count_table"), colClasses=c(Representative_Sequence="character")) %>%
+orig_count <- fread(paste0(path, "/data.count_table"),
+										colClasses=c(Representative_Sequence="character")) %>%
 	melt(id.vars=c("Representative_Sequence"), variable.name="group", variable.factor=F, value.name="n_seqs") %>%
   filter(group != "total") %>%
 	rename(sequences=Representative_Sequence) %>%
@@ -25,26 +30,6 @@ remove_groups <- scan(paste0(path, "/data.remove_accnos"), what="character", sep
 if(length(remove_groups) > 0){
   orig_count <- orig_count %>% filter(!group %in% remove_groups)
 }
-
-
-full_dataset <- orig_count %>% group_by(sequences) %>% summarize(n_seqs = sum(n_seqs))
-sub_dataset <- sample_frac(full_dataset, fraction)
-
[…]

### 2019-11-05 code/get_beta_diversity_data.R

@@ -1,0 +1,31 @@
+# This script takes in a phylip-formatted distance matrix and converts it to a three column format,
+# which is then outputted as a beta_diversity tsv
+
+library(stringr)
+
+d_matrix_filename <- commandArgs(trailingOnly=TRUE)	#d_matrix_filename <- "data.1.1.otu.rbeta_matrix"
+d_diversity_filename <- str_replace(d_matrix_filename, "matrix", "diversity")
+
+d_matrix <- scan(d_matrix_filename, sep="", what=character(), quiet=TRUE)
+
+n_seqs <- as.numeric(d_matrix[1])
+d_matrix <- d_matrix[-1]
+
+sample_indices <- c(1, rep(0,(n_seqs-1)))
+for(i in 1:(n_seqs-1)){
+	sample_indices[i+1] <- sample_indices[i] + i
+}
+
+samples <- d_matrix[ sample_indices ]
+d_matrix <- d_matrix[-sample_indices ]
+d_diversity <- data.frame(row=NA, column=NA, distance=NA)
+
+counter <- 1
+for(i in 2:n_seqs){
+	for(j in 1:(i-1)){
+		d_diversity[counter,] <- c(row=samples[i], column=samples[j], distance=d_matrix[counter])
+		counter <- counter+1
+	}
+}
+
+write.table(d_diversity, file=d_diversity_filename, row.names=FALSE, col.names=FALSE, quote=FALSE, sep="\t")

### 2019-11-05 code/get_beta_diversity_data.sh

@@ -1,11 +1,0 @@
-SHARED=$1
-
-mothur "#set.seed(seed=19760620);
-dist.shared(shared=$SHARED, calc=braycurtis, subsample=T, iters=100, processors=1, output=column)"
-
-INDIST=`echo $SHARED | sed -e "s/.shared/braycurtis*.ave.dist/"`
-OUTDIST=`echo $SHARED | sed -e "s/shared/beta_diversity/"`
-
-mv $INDIST $OUTDIST
-rm -f `echo $INDIST | sed -e "s/ave/std/"`
-rm -f `echo $INDIST | sed -e "s/ave.//"`

### 2019-11-05 code/randomize_prune.R

@@ -27,7 +27,8 @@ set.seed(seed)	#	set the random number generator seed so multiple runs generate 
 
 #	Here we read in the count_table, clean it up a bit, gather it, and remove those sequence/samples
 # combinations that are zero.
-orig_count <- fread(paste0(path, "/data.count_table"), colClasses=c(Representative_Sequence="character")) %>%
+orig_count <- fread(paste0(path, "/data.count_table"),
+ 										colClasses=c(Representative_Sequence="character")) %>%
   as_tibble() %>%
   select(-total) %>%
 	rename(sequences=Representative_Sequence) %>%
@@ -56,3 +57,11 @@ randomize_prune_count <-
 	ungroup() %>%
 	filter(n_seqs >= min_class) %>%
 	write_tsv(paste0(path, "/data.", seed, ".", min_class, ".rand_pruned_groups"))
+
+
+# Output randomized design file
+orig_count %>%
+	select(group) %>%
+	unique() %>%
+	mutate(grouping = sample(rep(c("A", "B"), length.out=nrow(.)))) %>%
+	write_tsv(paste0(path, "/data.", seed, ".", min_class, ".rdesign"), col_names=F)

### 2019-11-05 code/run_wilcox.R

@@ -2,46 +2,50 @@ library(tidyverse)
 library(broom)
 
 input <- commandArgs(trailingOnly=TRUE)
-alpha_files <- input
+alpha_files <- input[-length(input)]
 # source("code/alpha.R")
 # alpha_files <- list.files('data/bioethanol', pattern="ealpha_diversity", full.names=T)
 
-output_file_name <- paste0(dirname(alpha_files[1]), "/data.effect.alpha_summary")
+#take outputfile name from command line
+output_file_name <- input[length(input)]
+# output_file_name <- paste0(dirname(alpha_files[1]), "/data.effect.alpha_summary")
 
 
 get_summary <- function(alpha_file){
 
-	design_file <- str_replace(alpha_file, "[^.]*\\.ealpha_diversity", "design")
+  sim_type <- str_replace(alpha_file, ".*\\.(.)alpha_diversity", "\\1")
+  design_file <- str_replace(alpha_file, "[^.]*\\..alpha_diversity", paste0(sim_type, "design"))
 
-	seed <- str_replace(alpha_file, ".*/data\\.(\\d*)\\..*", "\\1") %>% as.integer()
-	prune <- str_replace(alpha_file, ".*/data\\.\\d*\\.(\\d*).*", "\\1") %>% as.integer()
-	clustering <- str_replace(alpha_file, ".*/data\\.\\d*\\.\\d*\\.(.*)\\.ealpha_diversity", "\\1")
+  seed <- str_replace(alpha_file, ".*/data\\.(\\d*)\\..*", "\\1") %>% as.integer()
+  prune <- str_replace(alpha_file, ".*/data\\.\\d*\\.(\\d*).*", "\\1") %>% as.integer()
+  clustering <- str_replace(alpha_file, ".*/data\\.\\d*\\.\\d*\\.(.*)\\..alpha_diversity", "\\1")
 
-	read_tsv(alpha_file,
-					col_types=cols(label=col_character(), group=col_character(), method=col_character())) %>%
-		filter(method == "ave") %>%
-		select(group, sobs, shannon, invsimpson) %>%
-		inner_join(.,
-							read_tsv(design_file, col_names=c("group", "grouping"), col_types="cc"),
-							by="group") %>%
-		gather(sobs, shannon, invsimpson, key="metric", value="value") %>%
-		group_by(metric) %>%
-		nest() %>%
-		mutate(test = map(data, ~tidy(wilcox.test(.x$value~.x$grouping, exact=FALSE))),
[…]

### 2019-11-05 submission/manuscript.Rmd
Generate data and viz for first results paragraph

@@
 * Logic for selecting studies (Table S1)
+
+Description / N Samples / N Reads / Median / IQR/Range / Reference
+
 * Distribution of sequence counts across samples in each study (Figure S1)
@@
 * Repeat when sequences are randomized across samples to maintain marginal sums
+* Loss of statistical power when rare sequences are removed/no impact on type 1 errors
 

### 2019-11-05 code/get_sample_summary_statistics.R

@@ -1,0 +1,32 @@
+library(tidyverse)
+
+summarize_data <- function(environment) {
+
+	removed_samples <- paste0("data/", environment, "/data.remove_accnos") %>%
+		read_lines(.) %>%
+		str_split("-") %>%
+		unlist() %>%
+		enframe(., name=NULL, value="sample")
+
+	paste0("data/", environment, "/data.count.summary") %>%
+		read_tsv(col_names=c("sample", "n_seqs"), col_types=cols(sample=col_character(), n_seqs=col_integer())) %>%
+		anti_join(., removed_samples, by="sample") %>%
+		summarize(n_samples=n(), total_seqs=sum(n_seqs), mean=mean(n_seqs),
+			min=min(n_seqs), l_quartile = quantile(n_seqs, 0.25), median=median(n_seqs),
+			u_quartile = quantile(n_seqs, 0.75), max=max(n_seqs),
+			n_removed = removed_samples %>% filter(sample != "") %>% nrow(),
+			fold_difference = max/min)
+}
+
+
+tibble(directory=c("bioethanol", "human", "lake", "marine", "mice", "rainforest",
+									"rice", "seagrass", "sediment", "soil", "stream"),
+			nice_name=c("Bioethanol", "Human", "Lake", "Marine", "Mice", "Rainforest",
+									"Rice", "Seagrass", "Sediment", "Soil", "Stream"),
+			reference=rep("", 11)
+		) %>%
+		group_by(directory) %>%
+		mutate(summary_data = map(directory, summarize_data)) %>%
+		unnest(summary_data) %>%
+		ungroup() %>%
+		write_tsv("data/process/study_summary_statistics.tsv")

### 2019-11-05 code/plot_seqs_per_sample.R

@@ -1,0 +1,52 @@
+library(tidyverse)
+
+set.seed(19760620)
+
+get_removed_samples <- function(environment) {
+
+	paste0("data/", environment, "/data.remove_accnos") %>%
+		read_lines(.) %>%
+		str_split("-") %>%
+		unlist() %>%
+		enframe(., name=NULL, value="sample") %>%
+		mutate(study = environment)
+
+}
+
+
+studies <- c("bioethanol", "human", "lake", "marine", "mice", "rainforest",
+									"rice", "seagrass", "sediment", "soil", "stream")
+
+nice_studies <- c("Bioethanol", "Human", "Lake", "Marine", "Mice", "Rainforest",
+						"Rice", "Seagrass", "Sediment", "Soil", "Stream")
+names(nice_studies) <- studies
+
+removed_samples <- studies %>%
+	map_df(., get_removed_samples) %>%
+	filter(sample != "") %>%
+	mutate(status = "removed")
+
+
+counts_data <- tibble(study=studies,
+			file_name=paste0("data/", study, "/data.count.summary")) %>%
+	group_by(file_name) %>%
+	nest() %>%
+	mutate(count_data = map(file_name, ~read_tsv(file=.x, col_names=c("sample", "n_seqs"), col_types=cols(sample=col_character(), n_seqs=col_integer())))) %>%
+	ungroup() %>%
+	select(-file_name) %>%
+	unnest(cols=c("data", "count_data")) %>%
+	full_join(., removed_samples, by=c("study", "sample")) %>%
+	mutate(status = replace_na(status, "kept")) %>%
+	mutate(study = factor(study, sort(studies, decreasing=T)))
+
+ggplot(counts_data, aes(y=n_seqs, x=study)) +
+	geom_vline(xintercept=seq(1:11), col="gray") +
+	geom_jitter(width=0.2, pch=19, alpha=0.33, aes(color=status)) +
+	scale_y_log10(limits=c(0.9, 1e6), breaks=c(1, 10, 100, 1000, 10000, 100000, 1e6), labels=c("1", "10", "100", "1,000", "10,000", "100,000", "1,000,000")) +
+	scale_x_discrete(breaks=studies, labels=nice_studies[studies]) +
+	scale_color_manual(name=NULL, values=c("black", "red"), breaks=c("kept", "removed"), labels=c("Kept", "Removed"))+
+	coord_flip() +
[…]

### 2020-09-15 code/get_skew_design.R
Add skewed distribution analysis

@@ -1,0 +1,14 @@
+library(tidyverse)
+
+rprune_groups_file <- commandArgs(trailingOnly=TRUE)
+# rprune_groups_file <- "data/mice/data.1.1.rand_pruned_groups"
+
+skewed_design_file <- str_replace(rprune_groups_file, "rand_pruned_groups", "sdesign")
+# design_file <- "data/mice/data.1.1.sdesign"
+
+read_tsv(rprune_groups_file, col_types=cols(.default=col_character(), n_seqs=col_integer())) %>%
+	group_by(group) %>%
+	summarize(count = sum(n_seqs)) %>%
+	mutate(grouping = ifelse(count < median(count), "small", "large")) %>%
+	select(group, grouping) %>%
+	write_tsv(skewed_design_file, col_names=F)

### 2020-09-15 submission/manuscript.Rmd
First pass through Results

--- 2019-06-10 submission/manuscript.Rmd
@@
 csl: mbio.csl
-fontsize: 11pt
 geometry: margin=1.0in
 ---
+
+## word choice: sample vs. community
 
@@
 library(knitr)
+suppressPackageStartupMessages(library(here))
 
@@
 
-inline_hook <- function(x){
-  print(x)
+inline_hook <- function(x, digits=2){
+
   if(is.list(x)){
@@
   if(is.numeric(x)){
-    if(abs(x - round(x)) < .Machine$double.eps^0.5){
-      paste(format(x,big.mark=',', digits=0, scientific=FALSE))
-    } else {
-      paste(format(x,big.mark=',', digits=2, nsmall=2, scientific=FALSE))
-    }
+      paste(format(x,big.mark=',', digits=digits, nsmall=digits, scientific=FALSE))
   } else {
@@
 knitr::knit_hooks$set(inline=inline_hook)
+
+oxford_comma <- function(x, digits=2) {
+
+	x <- map_chr(x, inline_hook, digits=digits)
+
+	if(length(x) < 2){
+		x
+	} else if(length(x) == 2){
+		paste(x, collapse = " and ")
+	} else {
+		paste(paste(x[-length(x)], collapse=", "), x[length(x)], sep=", and ")
+	}
+}
+
 ```
-
-\vspace*{10mm}
 
@@
 
-Patrick D. Schloss${^1}$${^\dagger}$
+Patrick D. Schloss${^\dagger}$
 
@@
 
-$\dagger$ To whom correspondence should be addressed: \href{mailto:pschloss@umich.edu}{pschloss@umich.edu}
+$\dagger$ To whom correspondence should be addressed:
 
-$1$ Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI 48109
+\href{mailto:pschloss@umich.edu}{pschloss@umich.edu}
+
+Department of Microbiology and Immunology
+University of Michigan
+Ann Arbor, MI 48109
 
@@
 
-## Observation format
+## Research article format
 
@@
 
-Two elements that are held in tension in the analysis of 16S rRNA gene sequence data is how to adequately remove PCR and sequencing artifacts and decrease the granularity of the taxonomic level that is used in the analysis.
-Previous attempts have included screening for sequencing quality based on quality scores [Kozich/Edgar] followed by a polishing step based on the frequency of the sequences relative to similar sequences [Kozich/Edgar/DeBlur].
+```{r datasets}
+table_1 <- read_tsv(here('data/process/study_summary_statistics.tsv'))
+
+n_sample_range <- table_1 %>%
+	pull(n_samples) %>%
+	range()
+
+n_med_sequence_range <- table_1 %>%
+	pull(median) %>%
+	round() %>%
+	range()
+
+low_fold_samples <- table_1 %>%
+	filter(fold_difference < 2) %>%
+	pull(directory)
+
+high_fold_range <- table_1 %>%
+	filter(fold_difference >= 2) %>%
+	pull(fold_difference) %>%
+	range() %>%
+	round(digits=1)
+```
+
+**EDIT: Two elements that are held in tension in the analysis of 16S rRNA gene sequence data is how to adequately remove PCR and sequencing artifacts and decrease the granularity of the taxonomic level that is used in the analysis.** Previous attempts have included screening for sequencing quality based on quality scores [Kozich/Edgar] followed by a polishing step based on the frequency of the sequences relative to similar sequences [Kozich/Edgar/DeBlur].
 Other pipelines model the quality scores and types of errors to cluster sequences directly [Dada2].
-As a final step, many pipelines advocate for removing rare sequences from each dataset prior to further clustering of sequences into an operational taxonomic unit (OTU) or phylogenetic group [Knight/Edgar/DeBlur/Dada2].
-The assumption of this abundance-based screening approach is that rare sequences are more likely to be artifacts than more abundant sequences.
-Some pipelines remove all sequences that appear once (i.e.
-singletons) [XXXX], XXXXXX [XXXXXX], XXXXX [XXXXXXX], or XXXXXX [XXXXXX] times; only the mothur-based pipeline discourages the practice of removing rare sequences.
-Proponents of this approach point to the ability to reach the desired number of OTUs with data generated from sequencing mock communities when rare sequences are removed.
-This approach effectively fits the sequence curation approach to data generated from a phylogenetically simple community with an atypical community distribution.
+But, as a final step many pipelines advocate for removing rare sequences from each dataset prior to outputting the sequence data as amplicon sequence variants (ASVs) [Knight/Edgar/DeBlur/Dada2].
+ASVs are often clustered further to generate operational taxonomic units or phylotypes.
+Some pipelines remove all ASVs that appear once (i.e.
+singletons) [XXXX], XXXXXX [XXXXXX], XXXXX [XXXXXXX], or XXXXXX [XXXXXX] times prior to further clustering or making ecological comparisions.
+Notably, the mothur-based pipeline discourages the practice of removing rare sequences.
 
-Sequencing of mock communities confirms that artifacts tend to be rare; however, artifacts can also be more abundant and replicate across samples.
-Furthermore, sequencing of real samples suggests that there are also good sequences that are rare.
-Ignored in the suggestion that researchers should remove rare sequences from each sample is the common result that the number of sequences generated from each sample may vary by two or three orders of magnitude.
-Thus a sequence that appears once in a sample with 2,000 sequences is more trustworthy than a sequence that appears once in a sample with 100,000 sequences.
-Removing such sequences potentially biases the community structure of the sample with fewer sequences.
-In the current study, I use published sequence data to investigate the effects of removing singletons and other rare sequences on downstream analysis of microbial communities.
+The abundance-based screening approach assumes that rare ASVs are more likely to be artifacts than more abundant ASVs.
+Sequencing of mock communities confirms that artifacts tend to be rare.
+Proponents of abundance-based screening point to their ability to obtain the correct number of ASVs, OTUs, or phylogypes with data generated from sequencing mock communities when rare ASVs are removed.
+However, this approach effectively overfits the curation pipeline to data generated from a phylogenetically simple community with an atypical community distribution that is often sequenced to a depth that is not achieved with biological samples.
+It is necessary to think more deeply about the practice of abundance-based screening.
 
-***Results.*** 600 words
-Sample description
-* Logic for selecting studies (Table S1)
-* Distribution of sequence counts across samples in each study (Figure S1)
+The minimum abundance thresholds that have been proscribed are developed and applied without regard for the total number of sequences generated from each sample.
+Ignored in their recommendations is the common experience that the number of sequences generated from each sample may vary by two or three orders of magnitude.
+An ASV that appears once in a sample with 2,000 sequences is more trustworthy than an ASV that appears once in a sample with 100,000 sequences since it has a 50-fold higher relative abundance.
+But, according to the pipeline recommendations, they are treated as being equally trustworthy.
+Rather than removing rare ASVs, the approach taken by the mothur pipeline applies the classical ecological approach of rarefaction.
+Each sample is rarefied to the same sequencing depth so that the number of artifacts that appears in each sample is controlled.
 
-Change in intra-sample variation with removal of rare sequences
-* Lose information within each sample as you remove sequences
-* Repeat when sequences are randomized across samples to maintain marginal sums
+Experience sequencing biological samples demonstrates that there are good ASVs that may have an abundance below the proscribed threshold.
+For example, the abundance of an ASV may be below the threshold in some samples or time points and above the threshold in others.
+However, rarity, both in terms of prevalence and incidence, is an important ecological concept.
+Removing rare ASVs likely hinders one's ability to ability to make inferences about the dynamics and nature of the populations that rare ASVs represent.
+Furthermore, removing ASVs whose abundances are below the proscribed threshold also potentially biases the community structure of the samples.
 
-Change in inter-sample variation with removal of rare sequences
-* Increased variation between samples as you remove sequences
-* Repeat when sequences are randomized across samples to maintain marginal sums
+In the current study, I use published sequence data from `r nrow(table_1)` studies to investigate the nature of rare ASVs (i.e.
+those that appear 10 or fewer times) and the effects of removing them on downstream analysis of microbial communities.
+The analysis was also performed using traditional operational taxonomic units, where ASVs subjected to abundance-based screening were clustered such that the ASVs within an OTU were no more than 3% different from each other.
+The results reject the assumptions built into abundance-based screening and highlight the problems inherent in removing rare ASVs.
+
+***Results.***
+
+**Datasets.** I collected `r nrow(table_1)` publicly available datasets that used the Illumina MiSeq platform to sequence the V4 region of the 16S rRNA gene from a variety of environments (Table 1).
+To insure the highest possible data quality, datasets were limited to those where the 500 cycle v2 MiSeq chemistry was used to sequence the amplicons.
+The paired 250 nt reads resulted in near complete 2-fold sequencing coverage of every nucleotide in the ca.
+250 nt-long region.
+This region and sequencing platform were selected because previous work has shown that a standard data analysis pipeline in mothur results in a sequencing error rate below 0.02% [@Kozich2012].
+All sequence data were obtained from the Sequence Read Archive and processed using a standard mothur-based sequencing pipeline that resulted in ASVs as generated by the pre.cluster algorithm [@Kozich2012; @Schloss2009].
+After removing poor quality and chimeric ASVs and samples that had uncharacteristically low number of sequences for the dataset, these datasets included between `r oxford_comma(n_sample_range)` samples.
+The median number of sequences for each dataset ranged between `r oxford_comma(n_med_sequence_range)` (Table 1, Figure S1).
+Strikingly, aside from the relatively small `r oxford_comma(low_fold_samples)` datasets, the difference between the sample with the fewest sequences and the sample with the most sequences for each dataset varied between `r oxford_comma(high_fold_range, digits=1)`-fold (Table 1).
+
+```{r}
+
+frac_lost <- read_tsv(here('data/process/sequence_loss_table_raw.tsv')) %>%
+	filter(min_freq == 1) %>%
+	group_by(dataset) %>%
+	summarize(median=median(fraction_lost))
+
+fraction_singletons_range <- frac_lost %>% arrange(median) %>% slice(1, nrow(.))
+percent_singletons_range_vals <- fraction_singletons_range %>% pull(median) * 100
+percent_singletons_range_names <- fraction_singletons_range %>% pull(dataset)
+
+cor_n <- read_tsv(here('data/process/sequence_loss_table_cor.tsv')) %>%
+ 	filter(min_freq == 1) %>%
+	select(dataset, cor_n_estimate, cor_n_p.value) %>%
+	mutate(sig = cor_n_p.value < 0.05)
+
+cor_n_nosig <- cor_n %>% filter(!sig) %>% pull(dataset)
+
+cor_n_sig <- cor_n %>%
+	filter(sig) %>%
+	select(dataset, cor_n_estimate) %>%
+	arrange(desc(cor_n_estimate)) %>%
+	slice(1, nrow(.))
+
+cor_n_sig_values <- cor_n_sig %>% pull(cor_n_estimate)
+cor_n_sig_names <- cor_n_sig %>% pull(dataset)
+
+frac_coverage <- read_tsv(here('data/process/sequence_coverage_table_raw.tsv')) %>%
+	filter(min_freq == 1) %>%
+	group_by(dataset) %>%
+	summarize(median=median(coverage))
+
+high_coverage <- frac_coverage %>% filter(median > 0.5) %>% pull(dataset)
+
+coverage_n <- read_tsv(here('data/process/sequence_coverage_table_cor.tsv')) %>%
+	filter(min_freq == 1) %>%
+	select(dataset, cor_n_estimate, cor_n_p.value) %>%
+	mutate(sig = cor_n_p.value < 0.05)
+
+coverage_n_nosig <- coverage_n %>% filter(!sig) %>% pull(dataset)
+
+coverage_n_sig <- coverage_n %>%
+	filter(sig) %>%
+	select(dataset, cor_n_estimate) %>%
+	arrange(desc(cor_n_estimate)) %>%
+	slice(1, nrow(.))
+
+coverage_n_sig_values <- coverage_n_sig %>% pull(cor_n_estimate)
+coverage_n_sig_names <- coverage_n_sig %>% pull(dataset)
+```
+
+**The nature of singletons.** Removal of rare ASVs is commonly justified as a method of removing ASVs that are artifacts.
+If such ASVs are artifacts, then one would expect the number of singleton ASVs to accumulate with sequencing depth.
+Contrary to this expectation, the median percentage of sequences that were discarded when singleton ASVs were removed from each dataset varied between `r oxford_comma(percent_singletons_range_vals)`% (`r oxford_comma(percent_singletons_range_names)`).
+In addition, with the exception of the samples from the `r oxford_comma(cor_n_nosig)` datasets (Spearman correlation, P>0.05), the fraction of singleton ASVs in the sample was negatively correlated with the number of sequences in each sample with a range between `r oxford_comma(cor_n_sig_values)` (`r oxford_comma(cor_n_sig_names)`).
+This showed that with additional sequencing, the probability of seeing rare ASVs multiple times was greater than the probability of generating an artifact.
+This suggests that the rare ASVs are not as likely to be artifacts as previously thought.
+Furthermore, if rare ASVs were artifacts, then one would not expect to find them in other samples from the same dataset.
+In fact, singleton ASVs from samples with fewer sequences were often found in samples with more sequences.
+At least 50% of the singleton ASVs found in the samples from the `r oxford_comma(high_coverage)` datasets were found in another sample from the same dataset.
+Considering the likelihood of finding an ASV duplicated in another sample is confounded by the number of samples and inter-sample diversity, the high coverage of singleton ASVs in these datasets was remarkable.
+The correlation between the number of sequences in a sample and the fraction of that sample's singleton ASVs that were covered by another sample in the dataset was significant and negative for `r nrow(coverage_n) - length(coverage_n_nosig)` of the datasets ranging between `r oxford_comma(coverage_n_sig_values)` for the `r oxford_comma(coverage_n_sig_names)` datasets, respectively.
+The negative correlation indicated that the singleton ASVs in the smaller samples were more likely to be covered by ASVs in the larger samples.
+Among the three datasets without a significant correlation (Spearman correlation, P>0.05), the marine and soil datasets had the fewest samples in our collection and the stream dataset already had a high level of coverage regardless of the number of sequences.
+Contrary to the common motivation for removing rare ASVs, these results indicate that this practice disproportionately impacts samples with fewer sequences and likely removes a large number of non-artifact ASVs.
+
+```{r}
+ointra <- read_tsv(here("data/process/ointra_analysis.tsv"),
+		col_types = cols(.default=col_double(),
+					dataset=col_character(), method=col_character(), group=col_character())
+	) %>%
+	group_by(dataset, method, min_class) %>%
+	summarize(mean_s_frac = 100*(1-mean(s_yes / s_no)),
+		mean_h_frac = 100*(1-mean(h_yes / h_no))) %>%
+	ungroup()
+
+asv_s_two <- ointra %>% filter(min_class == 2, method== "pc") %>% arrange(mean_s_frac)	%>% slice(1, nrow(.)) %>% pull(mean_s_frac)
+
+asv_s_eleven <- ointra %>% filter(min_class == 11, method== "pc") %>% arrange(mean_s_frac)	%>% slice(1, nrow(.)) %>% pull(mean_s_frac)
+
+asv_h_two <- ointra %>% filter(min_class == 2, method== "pc") %>% arrange(mean_h_frac)	%>% slice(1, nrow(.)) %>% pull(mean_h_frac)
+
+asv_h_eleven <- ointra %>% filter(min_class == 11, method== "pc") %>% arrange(mean_h_frac)	%>% slice(1, nrow(.)) %>% pull(mean_h_frac)
+```
+
+**How do communities change when rare ASVs are removed?** Removing rare ASVs will reduce the number of observed ASVs and proportionally increase the relative abundance of the remaining ASVs.
+The result was expected to be a loss of information contained within each sample.
+To quantify the effect of removing rare ASVs on the information contained within each sample, I varied the minimum abundance threshold to simulate removing ASVs of varying rarity from each sample.
+The number of ASVs in each sample decreased by between `r oxford_comma(asv_s_two, 1)`% when removing those ASVs that only appeared once and by between `r oxford_comma(asv_s_eleven, 1)`% when removing those that appeared ten or fewer times from each sample (Figure 1).
+Similarly, the Shannon diversity decreased by between `r oxford_comma(asv_h_two, 1)`% when removing ASVs that only appeared once and by between `r oxford_comma(asv_s_eleven, 1)`% when removing ASVs that appeared ten or fewer times from each sample (Figure 1).
+Next, I assigned the ASVs to OTUs to assess the impact of removing rare ASVs on higher level taxonomic groupings that are commonly used in microbial ecology studies.
+Although pooling similar ASVs into OTUs reduced the impact of removing the rare ASVs relative to the ASV-based analysis, the minimum abundance threshold still decreased the number of observed OTUs and the diversity decreased relative to the original community (**Figure 1**).
+In contrast to the number of ASVs or OTUs and their diversity, the Kullback–Leibler divergence compares the relative abundance of each ASV or OTU between representations of the community.
+I calculated the Kullback–Leibler divergence from the full to pruned communities when rare ASVs were removed.
+As the threshold for removing ASVs increased, the amount of information lost also increased for both ASVs and OTUs (Figure 1).
+The relative loss of information was generally lower for OTUs than than it was for ASVs.
+Removing rare ASVs, regardless of abundance threshold, had profound impacts on the representation of the communities.
+
+**Removing treatment group effects from community data.** Because treatment effects often affect a sample's diversity and inter-sample variation, I generated null distributions for each study by randomizing the number of times each ASV was observed in each sample such that the total number of sequences in each sample and the total number of times each ASV was observed across all samples in the study was the same as was originally observed.
+This effectively made every community in a study a statistical sample of the study-wide composite community distribution.
+For example, after this procedure, the 490 samples from the human dataset would be expected to have the same number of ASVs and diversity and one would not expect to find treatment-based effects between the samples.
+Because of the risk of bias if only one representation of the null distribution was generated, I generated 100 randomized datasets for each study.
+The same trends between removing rare ASVs and the number of ASVs, sample diversity, and information loss that were identified using the observed community community distribution data were also identified with the data from the null distribution (**Figure S2**).
+The null distribution data were used in the remainder of the study.
+
+```{r}
+r_cov_alpha <- read_tsv(here("data/process/ralpha_analysis.tsv"),
+		col_types = cols(.default=col_double(),
+					dataset=col_character(), method=col_character())
+	) %>% select(dataset, method, prune, starts_with("cv_")) %>%
+	filter(prune == 2 | prune ==1) %>%
+	pivot_wider(id_cols=c(dataset, method), names_from=prune, values_from=starts_with("cv_")) %>%
+	mutate(cv_sobs = cv_sobs_2/cv_sobs_1,
+			cv_shannon = cv_shannon_2/cv_shannon_1
+		) %>%
+	select(dataset, method, cv_sobs, cv_shannon)
+
+pc_cov_alpha <- r_cov_alpha %>% filter(method == "pc")
+pc_cov_sobs <- pc_cov_alpha %>% arrange(cv_sobs) %>% slice(1, nrow(.)) %>% pull(cv_sobs)
+pc_cov_shannon <- pc_cov_alpha %>% arrange(cv_shannon) %>% slice(1, nrow(.)) %>% pull(cv_shannon)
+
+otu_cov_alpha <- r_cov_alpha %>% filter(method == "otu")
+otu_cov_sobs <- otu_cov_alpha %>% arrange(cv_sobs) %>% slice(1, nrow(.)) %>% pull(cv_sobs)
+otu_cov_shannon <- otu_cov_alpha %>% arrange(cv_shannon) %>% slice(1, nrow(.)) %>% pull(cv_shannon)
+
+pc_v_otu_alpha <- r_cov_alpha %>% pivot_wider(dataset, names_from=method, values_from=c(cv_sobs, cv_shannon)) %>% mutate(sobs = 100*(1-cv_sobs_otu/cv_sobs_pc), shannon = 100*(1-cv_shannon_otu/cv_shannon_pc))
+
+pc_v_otu_sobs <- range(pc_v_otu_alpha$sobs)
+pc_v_otu_shannon <- range(pc_v_otu_alpha$shannon)
+
+r_cov_bc <- read_tsv(here("data/process/rbeta_analysis.tsv"),
+		col_types = cols(.default=col_double(),
+					dataset=col_character(), method=col_character())) %>%
+		filter(prune == 2 | prune ==1) %>%
+		pivot_wider(id_cols=c(dataset, method), names_from=prune, values_from=starts_with("sd_")) %>%
+		mutate(cv_diff = `2`/`1`) %>%
+		select(dataset, method, cv_diff)
+
+pc_cov_bc <- r_cov_bc %>% filter(method == "pc") %>% arrange(cv_diff) %>% slice(1, nrow(.)) %>% pull(cv_diff)
+
+otu_cov_bc <- r_cov_bc %>% filter(method == "otu") %>% arrange(cv_diff) %>% slice(1, nrow(.)) %>% pull(cv_diff)
+
+```
+
+**How do the relationships between communities change when they are screened by removing rare ASVs.** Considering the loss of the number of ASVs, diversity, and information when a community has its rarest ASVs removed, it seemed likely that the relationship between communities would be altered.
+To assess the impact of removing rare ASVs on measures of alpha diversity between samples I calculated the coefficients of variation (COV, i.e.
+standard deviation divided by the mean) for the number of ASVs and diversity for each study at multiple abundance thresholds.
+The COV for the number of ASVs across the studies after removing singletons were between `r oxford_comma(pc_cov_sobs, 1)`-times larger than they were without removing singleton ASVs.
+Similarly, the COVs for the diversity of ASVs were between `r oxford_comma(pc_cov_shannon, 1)`-times larger when singletons were removed than when they were not removed.To assess the impact of removing rare ASVs on measures of beta diversity between samples I calculated the COV of the Bray-Curtis distances between samples within the same study at multiple abundance thresholds.
+The COV between Bray-Curtis distances within a study when singletons were removed was between `r oxford_comma(pc_cov_bc, 1)`-times larger than when they were not removed.
+Increasing the minimum abundance threshold increased the COV between samples when using metrics of alpha and beta diversity.
+When ASVs were clustered into OTUs the difference in the COV was less than it was for the ASVs (**Figure 2**).
+These results indicate that removing rare ASVs increases the dissimilarity between samples, which could have a significant impact on the statistical power to detect differences between treatment groups.
+
+```{r}
+b_alpha <- read_tsv(here("data/process/bffect_alpha_analysis.tsv"),
+		col_types = cols(.default=col_double(),
+					dataset=col_character(), method=col_character(), metric=col_character())) %>%
+	filter(metric != "invsimpson", dataset != "marine") %>%
+	filter(prune == 2 | prune ==1) %>%
+	pivot_wider(id_cols=c(dataset, metric, method), names_from=c(prune), values_from=frac_sig) %>%
+	mutate(pp_drop = `1` - `2`,
+		percentage_drop = 100*(1-`2` / `1`)) %>%
+	arrange(percentage_drop)
+
+b_base_pc_sobs <- b_alpha %>% filter(metric == "sobs", method == "pc") %>% select(`1`) %>% range()
+b_base_pc_shannon <- b_alpha %>% filter(metric == "shannon", method == "pc") %>% select(`1`) %>% range()
+
+b_base_otu_sobs <- b_alpha %>% filter(metric == "sobs", method == "otu") %>% select(`1`) %>% range()
+b_base_otu_shannon <- b_alpha %>% filter(metric == "shannon", method == "otu") %>% select(`1`) %>% range()
+
+b_pc_sobs <- b_alpha %>% filter(metric == "sobs", method == "pc") %>% slice(1, n()) %>% pull(percentage_drop)
+
+b_pc_shannon <- b_alpha %>% filter(metric == "shannon", method == "pc") %>% slice(1, n()) %>% pull(percentage_drop)
+
+b_beta <- read_tsv(here("data/process/bffect_beta_analysis.tsv"),
+		col_types = cols(.default=col_double(),
+					dataset=col_character(), clustering=col_character())) %>%
+	filter(dataset != "marine") %>%
+	filter(prune==11 | prune == 2 | prune ==1) %>%
+	pivot_wider(id_cols=c(dataset, clustering), names_from=c(prune), values_from=frac_sig) %>%
+	mutate(p_drop_singletons = 100*(1-`2` / `1`),
+					p_drop_tens = 100*(1-`11` / `1`)) %>%
+	arrange(p_drop_singletons)
+
+b_base_beta <- b_beta %>% filter(clustering == "pc") %>% select(`1`) %>% range()
+
+n_unfazed_singletons <- b_beta %>% filter(clustering=="pc" & p_drop_singletons <= 0) %>% nrow()
+
+fazed_singletons <- b_beta %>% filter(clustering=="pc" & p_drop_singletons > 0) %>% select(p_drop_singletons) %>% range()
+
+n_unfazed_tentons <- b_beta %>% filter(clustering=="pc" & p_drop_tens <= 10) %>% nrow()
+stopifnot(n_unfazed_tentons == 0)
+
+fazed_tentons <- b_beta %>% filter(clustering=="pc" & p_drop_tens > 10) %>% select(p_drop_tens) %>% range()
+```
+
+**How does the increased inter-sample variation impact the ability to detect differences between samples?** To test the effect of increased inter-sample variation, I randomly assigned samples to one of two treatment groups.
+In the first treatment group, communities were randomly sampled from the null distribution as described above.
+For the second treatment group, I increased the abundance of 10% of the ASVs in the pooled study distribution by 5%.
+I randomly generated 100 simulated sets of treatment groups and samples.
+I then tested the ability to detect a difference between the two treatment groups using alpha and beta diversity metrics.
+The fraction of significant tests was a measurement of the statistical power to detect the difference between the treatment groups.
+Across all of the metrics I tested, the marine dataset yielded no simulated sets that were statistically significant, which was likely due to the small number of samples in the study (N=7); the marine dataset is not included in the following analysis.
+Among the remaining datasets, the power to detect a difference in the observed number of ASVs ranged between `r oxford_comma(b_base_pc_sobs, 2)` and between `r oxford_comma(b_base_pc_shannon, 2)` to detect a difference in diversity when using a Wilcox test (**Figure X**).
+When singleton ASVs were removed, the power to detect a difference in the number of ASVs dropped by between `r oxford_comma(b_pc_sobs, 1)`% and by between `r oxford_comma(b_pc_shannon, 1)`% to detect a difference in diversity.
+The effect of removing rare ASVs on the number of OTUs and their diversity was similar.
+A popular alternative approach calculates the dissimilarity between communities based on the abundance of each ASV or OTU.
+We used the Bray-Curtis dissimilarity index to compare the simulated communities within each dataset and calculated the power to detect differences between the two simulated treatment groups using the analysis of molecular variance (also called PERMANOVA).
+Without removing rare sequences, the power to detect a difference between the two simulated treatment groups varied between `r oxford_comma(b_base_beta, 2)`.
+Aside from `r n_unfazed_singletons` datasets, the power to detect differences dropped by between `r oxford_comma(fazed_singletons, 1)`% when singletons were removed.
+However, when ASVs that occurred 10 or fewer times were removed from each sample, the power to detect differences dropped by `r oxford_comma(fazed_tentons, 1)`%; similar results were observed when ASVs were clustered into OTUs.
+Removing rare ASVs reduced the ability to detect simulated treatment effects using metrics commonly used to compare microbial communities.
+
+```{r}
+type_one_null_alpha_summary <-
+  read_tsv(here("data/process/rffect_alpha_analysis.tsv"),
+		col_types = cols(.default=col_double(),
+					dataset=col_character(), method=col_character(), metric=col_character())) %>%
+	filter(dataset != "marine" & metric != "invsimpson") %>%
+  filter((metric == "sobs" | metric == "shannon")) %>%
+	group_by(prune, method, metric) %>%
+	summarize(p=mean(frac_sig), p_sd = sd(frac_sig)) %>%
+	ungroup()
+
+type_one_null_beta_summary <- read_tsv(here("data/process/rffect_beta_analysis.tsv"),
+		col_types = cols(.default=col_double(),
+					dataset=col_character(), clustering=col_character())) %>%
+	filter(dataset != "marine") %>%
+	group_by(prune, clustering) %>%
+	summarize(p=mean(frac_sig), p_sd = sd(frac_sig)) %>%
+  ungroup()
+
+type_one_skew_alpha_summary <-
+  read_tsv(here("data/process/sffect_alpha_analysis.tsv"),
+		col_types = cols(.default=col_double(),
+					dataset=col_character(), method=col_character(), metric=col_character())) %>%
+	filter(dataset != "marine" & metric != "invsimpson") %>%
+  filter((metric == "sobs" | metric == "shannon")) %>%
+  group_by(prune, method, metric) %>%
+  summarize(p=mean(frac_sig), p_sd = sd(frac_sig), min_p=min(frac_sig)) %>%
+  ungroup()
+
+type_one_skew_beta_summary <- read_tsv(here("data/process/sffect_beta_analysis.tsv"),
+		col_types = cols(.default=col_double(),
+					dataset=col_character(), clustering=col_character())) %>%
+	filter(dataset != "marine") %>%
+	group_by(prune, clustering) %>%
+	summarize(p=mean(frac_sig), p_sd = sd(frac_sig), min_p=min(frac_sig)) %>%
+  ungroup()
+
+```
+
+**How does the removing of rare ASVs impact the probability of falsely detecting a difference between treatment groups?** Observing reduced ability to detect differences between communities when rare ASVs were removed from each sample, I next asked whether removing rare ASVs could lead to falsely claiming that a treatment effect had a significant effect on community diversity and structure.
+First, for each dataset, I sampled sequences from the null distribution and randomly assigned each sample to one of two treatment groups and determined the number of observed ASVs and OTUs and their Shannon diversity index.
+Testing at an experiment-wise error rate of 0.05, I expected 5% of the iterations for each dataset to yield a significant test result.
+Indeed, there was no evidence that removing rare ASVs resulted in an inflated experiment-wise error rate.
+The average fraction of significant tests did not meaningfully vary from 0.05 across the minimum abundance threshold, dataset, metric of describing sample alpha-diversity, or whether the abundance of ASVs or OTUs were used (**Figure XA**).
+Similarly, the average fraction of significant tests did not meaningfully vary from 0.05 when using analysis of molecular variance to compare communities using Bray-Curtis distances.
+Second, I again sampled sequences from the null distribution, but assigned samples to one of two treatment groups based on the number of sequences in each sample.
+The samples with fewer than the median number of sequences for the dataset were assigned to one group and those with more than the median were assigned to the other.
+This exaggerated bias has been observed in comparisons of the lung and oral microbiota because of the larger nubmer of non-specific amplicons that can be sequenced from lung samples relative to those in the oral cavity [**REF**].
+When rare sequences were not removed, the fraction of significant tests did not differ from 5% for comparing the number of observed taxa (using ASVs or OTUs), their shannon diversity, or Bray-Curtis distances.
+However, when rare taxa of any frequency were removed, the probability of falsely detecing a difference as signifiant increased with the definition of rarity (**Figure XB**).
+Across the datasets, the average fraction the average fraction of falsely detecting a difference was when only singletons were removed was greater than 90%.
+If there is any relationship between the number of sequences and the treatment group, the risk of falsely rejecting the null hypothesis is inflated when researchers use the strategy of removing rare sequences.
+The most conservative approach is to not remove low abundance sequences.
 
@@
 
+* sequencing pipeline description
+
 \newpage
@@
 ## References
-25 references
 

### 2020-09-15 submission/manuscript.tex
First pass through Results

@@
-\documentclass[11pt,]{article}
+\documentclass[]{article}
 \usepackage{lmodern}
@@
     \usepackage{mathspec}
+    \usepackage{xltxtra,xunicode}
   \else
@@
   \fi
-  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
+  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
+  \newcommand{\euro}{€}
 \fi
@@
 \usepackage[margin=1.0in]{geometry}
-\usepackage{hyperref}
-\hypersetup{unicode=true,
-            pdfborder={0 0 0},
-            breaklinks=true}
-\urlstyle{same}  % don't use monospace font for urls
-\usepackage{graphicx,grffile}
+\usepackage{graphicx}
 \makeatletter
@@
 \setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-\IfFileExists{parskip.sty}{%
-\usepackage{parskip}
-}{% else
+\ifxetex
+  \usepackage[setpagesize=false, % page size defined by xetex
+              unicode=false, % unicode breaks when used with xetex
+              xetex]{hyperref}
+\else
+  \usepackage[unicode=true]{hyperref}
+\fi
+\hypersetup{breaklinks=true,
+            bookmarks=true,
+            pdfauthor={},
+            pdftitle={},
+            colorlinks=true,
+            citecolor=blue,
+            urlcolor=blue,
+            linkcolor=magenta,
+            pdfborder={0 0 0}}
+\urlstyle{same}  % don't use monospace font for urls
 \setlength{\parindent}{0pt}
 \setlength{\parskip}{6pt plus 2pt minus 1pt}
-}
 \setlength{\emergencystretch}{3em}  % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 \setcounter{secnumdepth}{0}
-% Redefines (sub)paragraphs to behave more like sections
-\ifx\paragraph\undefined\else
-\let\oldparagraph\paragraph
-\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-\let\oldsubparagraph\subparagraph
-\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
 
@@
 
-\vspace*{10mm}
+\subsection{word choice: sample
+vs.~community}\label{word-choice-sample-vs.community}
 
-\hypertarget{removal-of-rare-sequences-from-16s-rrna-gene-sequence-surveys-biases-the-interpretation-of-community-structure-data}{%
 \section{Removal of rare sequences from 16S rRNA gene sequence surveys
 biases the interpretation of community structure
-data}\label{removal-of-rare-sequences-from-16s-rrna-gene-sequence-surveys-biases-the-interpretation-of-community-structure-data}}
+data}\label{removal-of-rare-sequences-from-16s-rrna-gene-sequence-surveys-biases-the-interpretation-of-community-structure-data}
 
@@
 
-Patrick D. Schloss\({^1}\)\({^\dagger}\)
+Patrick D. Schloss${^\dagger}$
 
@@
 
-\(\dagger\) To whom correspondence should be addressed:
+$\dagger$ To whom correspondence should be addressed:
+
 \href{mailto:pschloss@umich.edu}{pschloss@umich.edu}
 
-\(1\) Department of Microbiology and Immunology, University of Michigan,
-Ann Arbor, MI 48109
+Department of Microbiology and Immunology University of Michigan Ann
+Arbor, MI 48109
 
@@
 
-\hypertarget{observation-format}{%
-\subsection{Observation format}\label{observation-format}}
+\subsection{Research article format}\label{research-article-format}
 
@@
 
-\hypertarget{abstract}{%
-\subsection{Abstract}\label{abstract}}
+\subsection{Abstract}\label{abstract}
+
+250 words
 
@@
 
-\hypertarget{importance}{%
-\subsection{Importance}\label{importance}}
+\subsection{Importance}\label{importance}
+
+150 words
 
@@
 
-Question: What impact does removing singletons (and other rare
-sequences) have on downstream data analysis?
+\textbf{EDIT: Two elements that are held in tension in the analysis of
+16S rRNA gene sequence data is how to adequately remove PCR and
+sequencing artifacts and decrease the granularity of the taxonomic level
+that is used in the analysis.} Previous attempts have included screening
+for sequencing quality based on quality scores {[}Kozich/Edgar{]}
+followed by a polishing step based on the frequency of the sequences
+relative to similar sequences {[}Kozich/Edgar/DeBlur{]}. Other pipelines
+model the quality scores and types of errors to cluster sequences
+directly {[}Dada2{]}. But, as a final step many pipelines advocate for
+removing rare sequences from each dataset prior to outputting the
+sequence data as amplicon sequence variants (ASVs)
+{[}Knight/Edgar/DeBlur/Dada2{]}. ASVs are often clustered further to
+generate operational taxonomic units or phylotypes. Some pipelines
+remove all ASVs that appear once (i.e.~singletons) {[}XXXX{]}, XXXXXX
+{[}XXXXXX{]}, XXXXX {[}XXXXXXX{]}, or XXXXXX {[}XXXXXX{]} times prior to
+further clustering or making ecological comparisions. Notably, the
+mothur-based pipeline discourages the practice of removing rare
+sequences.
 
-\begin{itemize}
-\tightlist
-\item
-  Many pipelines remove rare sequences prior to clustering sequences
-  into OTUs or as part of the generation fo ASVs
-\item
-  Singletons can be sequences that appear 1-10 times in larger samples
-\item
-  Seeks to solve the problem of low quality sequence data - more
-  singletons with higher error rates
-\item
-  Likely there are rare sequences that are good - e.g.~high diversity
-  samples
-\item
-  People have overfit their pipelines ot mock community data which have
-  unrealistic community distributions (i.e.~uniform abundance or log
-  abundance) and phylogenetic diversity
-\item
-  Better solutions would include:
+The abundance-based screening approach assumes that rare ASVs are more
+likely to be artifacts than more abundant ASVs. Sequencing of mock
+communities confirms that artifacts tend to be rare. Proponents of
+abundance-based screening point to their ability to obtain the correct
+number of ASVs, OTUs, or phylogypes with data generated from sequencing
+mock communities when rare ASVs are removed. However, this approach
+effectively overfits the curation pipeline to data generated from a
+phylogenetically simple community with an atypical community
+distribution that is often sequenced to a depth that is not achieved
+with biological samples. It is necessary to think more deeply about the
+practice of abundance-based screening.
 
-  \begin{itemize}
-  \tightlist
-  \item
-    Get better sequence data
-  \item
-    Rarefy data so that the number of errors are controlled for across
-    samples
-  \end{itemize}
-\item
-  Not all samples are sequenced evenly, so a singleton doesn't hav the
-  same relative abundance across all samples
-\end{itemize}
+The minimum abundance thresholds that have been proscribed are developed
+and applied without regard for the total number of sequences generated
+from each sample. Ignored in their recommendations is the common
+experience that the number of sequences generated from each sample may
+vary by two or three orders of magnitude. An ASV that appears once in a
+sample with 2,000 sequences is more trustworthy than an ASV that appears
+once in a sample with 100,000 sequences since it has a 50-fold higher
+relative abundance. But, according to the pipeline recommendations, they
+are treated as being equally trustworthy. Rather than removing rare
+ASVs, the approach taken by the mothur pipeline applies the classical
+ecological approach of rarefaction. Each sample is rarefied to the same
+sequencing depth so that the number of artifacts that appears in each
+sample is controlled.
 
-H: Removing rare sequences from each sample will skew the community
-distribution to impact the amount of variation in the relative abundance
-and alpha and beta diversity values across samples
+Experience sequencing biological samples demonstrates that there are
+good ASVs that may have an abundance below the proscribed threshold. For
+example, the abundance of an ASV may be below the threshold in some
+samples or time points and above the threshold in others. However,
+rarity, both in terms of prevalence and incidence, is an important
+ecological concept. Removing rare ASVs likely hinders one's ability to
+ability to make inferences about the dynamics and nature of the
+populations that rare ASVs represent. Furthermore, removing ASVs whose
+abundances are below the proscribed threshold also potentially biases
+the community structure of the samples.
+
+In the current study, I use published sequence data from 12 studies to
+investigate the nature of rare ASVs (i.e.~those that appear 10 or fewer
+times) and the effects of removing them on downstream analysis of
+microbial communities. The analysis was also performed using traditional
+operational taxonomic units, where ASVs subjected to abundance-based
+screening were clustered such that the ASVs within an OTU were no more
+than 3\% different from each other. The results reject the assumptions
+built into abundance-based screening and highlight the problems inherent
+in removing rare ASVs.
+
+\textbf{\emph{Results.}}
+
+\textbf{Datasets.} I collected 12 publicly available datasets that used
+the Illumina MiSeq platform to sequence the V4 region of the 16S rRNA
+gene from a variety of environments (Table 1). To insure the highest
+possible data quality, datasets were limited to those where the 500
+cycle v2 MiSeq chemistry was used to sequence the amplicons. The paired
+250 nt reads resulted in near complete 2-fold sequencing coverage of
+every nucleotide in the ca. 250 nt-long region. This region and
+sequencing platform were selected because previous work has shown that a
+standard data analysis pipeline in mothur results in a sequencing error
+rate below 0.02\% (\textbf{???}). All sequence data were obtained from
+the Sequence Read Archive and processed using a standard mothur-based
+sequencing pipeline that resulted in ASVs as generated by the
+pre.cluster algorithm (\textbf{???}, 1). After removing poor quality and
+chimeric ASVs and samples that had uncharacteristically low number of
+sequences for the dataset, these datasets included between 7.00 and
+490.00 samples. The median number of sequences for each dataset ranged
+between 6,477.00 and 193,464.00 (Table 1, Figure S1). Strikingly, aside
+from the relatively small marine and soil datasets, the difference
+between the sample with the fewest sequences and the sample with the
+most sequences for each dataset varied between 7.4 and 96.6-fold (Table
+1).
+
+\textbf{The nature of singletons.} Removal of rare ASVs is commonly
+justified as a method of removing ASVs that are artifacts. If such ASVs
+are artifacts, then one would expect the number of singleton ASVs to
+accumulate with sequencing depth. Contrary to this expectation, the
+median percentage of sequences that were discarded when singleton ASVs
+were removed from each dataset varied between 0.42 and 22.23\%
+(bioethanol and seagrass). In addition, with the exception of the
+samples from the marine and sediment datasets (Spearman correlation,
+P\textgreater{}0.05), the fraction of singleton ASVs in the sample was
+negatively correlated with the number of sequences in each sample with a
+range between -0.27 and -0.87 (rice and bioethanol). This showed that
+with additional sequencing, the probability of seeing rare ASVs multiple
+times was greater than the probability of generating an artifact. This
+suggests that the rare ASVs are not as likely to be artifacts as
+previously thought. Furthermore, if rare ASVs were artifacts, then one
+would not expect to find them in other samples from the same dataset. In
+fact, singleton ASVs from samples with fewer sequences were often found
+in samples with more sequences. At least 50\% of the singleton ASVs
+found in the samples from the mice, rice, seagrass, and stream datasets
+were found in another sample from the same dataset. Considering the
+likelihood of finding an ASV duplicated in another sample is confounded
+by the number of samples and inter-sample diversity, the high coverage
+of singleton ASVs in these datasets was remarkable. The correlation
+between the number of sequences in a sample and the fraction of that
+sample's singleton ASVs that were covered by another sample in the
+dataset was significant and negative for 9 of the datasets ranging
+between -0.31 and -0.84 for the rice and seagrass datasets,
+respectively. The negative correlation indicated that the singleton ASVs
+in the smaller samples were more likely to be covered by ASVs in the
+larger samples. Among the three datasets without a significant
+correlation (Spearman correlation, P\textgreater{}0.05), the marine and
+soil datasets had the fewest samples in our collection and the stream
+dataset already had a high level of coverage regardless of the number of
+sequences. Contrary to the common motivation for removing rare ASVs,
+these results indicate that this practice disproportionately impacts
+samples with fewer sequences and likely removes a large number of
+non-artifact ASVs.
+
+\textbf{How do communities change when rare ASVs are removed?} Removing
+rare ASVs will reduce the number of observed ASVs and proportionally
+increase the relative abundance of the remaining ASVs. The result was
+expected to be a loss of information contained within each sample. To
+quantify the effect of removing rare ASVs on the information contained
+within each sample, I varied the minimum abundance threshold to simulate
+removing ASVs of varying rarity from each sample. The number of ASVs in
+each sample decreased by between 34.4 and 86.2\% when removing those
+ASVs that only appeared once and by between 76.0 and 95.6\% when
+removing those that appeared ten or fewer times from each sample (Figure
+1). Similarly, the Shannon diversity decreased by between 1.8 and 15.9\%
+when removing ASVs that only appeared once and by between 76.0 and
+95.6\% when removing ASVs that appeared ten or fewer times from each
+sample (Figure 1). Next, I assigned the ASVs to OTUs to assess the
+impact of removing rare ASVs on higher level taxonomic groupings that
+are commonly used in microbial ecology studies. Although pooling similar
+ASVs into OTUs reduced the impact of removing the rare ASVs relative to
+the ASV-based analysis, the minimum abundance threshold still decreased
+the number of observed OTUs and the diversity decreased relative to the
+original community (\textbf{Figure 1}). In contrast to the number of
+ASVs or OTUs and their diversity, the Kullback--Leibler divergence
+compares the relative abundance of each ASV or OTU between
+representations of the community. I calculated the Kullback--Leibler
+divergence from the full to pruned communities when rare ASVs were
+removed. As the threshold for removing ASVs increased, the amount of
+information lost also increased for both ASVs and OTUs (Figure 1). The
+relative loss of information was generally lower for OTUs than than it
+was for ASVs. Removing rare ASVs, regardless of abundance threshold, had
+profound impacts on the representation of the communities.
+
+\textbf{Removing treatment group effects from community data.} Because
+treatment effects often affect a sample's diversity and inter-sample
+variation, I generated null distributions for each study by randomizing
+the number of times each ASV was observed in each sample such that the
+total number of sequences in each sample and the total number of times
+each ASV was observed across all samples in the study was the same as
+was originally observed. This effectively made every community in a
+study a statistical sample of the study-wide composite community
+distribution. For example, after this procedure, the 490 samples from
+the human dataset would be expected to have the same number of ASVs and
+diversity and one would not expect to find treatment-based effects
+between the samples. Because of the risk of bias if only one
+representation of the null distribution was generated, I generated 100
+randomized datasets for each study. The same trends between removing
+rare ASVs and the number of ASVs, sample diversity, and information loss
+that were identified using the observed community community distribution
+data were also identified with the data from the null distribution
+(\textbf{Figure S2}). The null distribution data were used in the
+remainder of the study.
+
+\textbf{How do the relationships between communities change when they
+are screened by removing rare ASVs.} Considering the loss of the number
+of ASVs, diversity, and information when a community has its rarest ASVs
+removed, it seemed likely that the relationship between communities
+would be altered. To assess the impact of removing rare ASVs on measures
+of alpha diversity between samples I calculated the coefficients of
+variation (COV, i.e.~standard deviation divided by the mean) for the
+number of ASVs and diversity for each study at multiple abundance
+thresholds. The COV for the number of ASVs across the studies after
+removing singletons were between 3.6 and 32.7-times larger than they
+were without removing singleton ASVs. Similarly, the COVs for the
+diversity of ASVs were between 1.8 and 20.4-times larger when singletons
+were removed than when they were not removed.To assess the impact of
+removing rare ASVs on measures of beta diversity between samples I
+calculated the COV of the Bray-Curtis distances between samples within
+the same study at multiple abundance thresholds. The COV between
+Bray-Curtis distances within a study when singletons were removed was
+between 1.3 and 18.6-times larger than when they were not removed.
+Increasing the minimum abundance threshold increased the COV between
+samples when using metrics of alpha and beta diversity. When ASVs were
+clustered into OTUs the difference in the COV was less than it was for
+the ASVs (\textbf{Figure 2}). These results indicate that removing rare
+ASVs increases the dissimilarity between samples, which could have a
+significant impact on the statistical power to detect differences
+between treatment groups.
+
+\textbf{How does the increased inter-sample variation impact the ability
+to detect differences between samples?} To test the effect of increased
+inter-sample variation, I randomly assigned samples to one of two
+treatment groups. In the first treatment group, communities were
+randomly sampled from the null distribution as described above. For the
+second treatment group, I increased the abundance of 10\% of the ASVs in
+the pooled study distribution by 5\%. I randomly generated 100 simulated
+sets of treatment groups and samples. I then tested the ability to
+detect a difference between the two treatment groups using alpha and
+beta diversity metrics. The fraction of significant tests was a
+measurement of the statistical power to detect the difference between
+the treatment groups. Across all of the metrics I tested, the marine
+dataset yielded no simulated sets that were statistically significant,
+which was likely due to the small number of samples in the study (N=7);
+the marine dataset is not included in the following analysis. Among the
+remaining datasets, the power to detect a difference in the observed
+number of ASVs ranged between 0.10 and 0.49 and between 0.10 and 0.53 to
+detect a difference in diversity when using a Wilcox test
+(\textbf{Figure X}). When singleton ASVs were removed, the power to
+detect a difference in the number of ASVs dropped by between 27.3 and
+92.9\% and by between 40.0 and 93.3\% to detect a difference in
+diversity. The effect of removing rare ASVs on the number of OTUs and
+their diversity was similar. A popular alternative approach calculates
+the dissimilarity between communities based on the abundance of each ASV
+or OTU. We used the Bray-Curtis dissimilarity index to compare the
+simulated communities within each dataset and calculated the power to
+detect differences between the two simulated treatment groups using the
+analysis of molecular variance (also called PERMANOVA). Without removing
+rare sequences, the power to detect a difference between the two
+simulated treatment groups varied between 0.41 and 1.00. Aside from 3
+datasets, the power to detect differences dropped by between 6.5 and
+64.0\% when singletons were removed. However, when ASVs that occurred 10
+or fewer times were removed from each sample, the power to detect
+differences dropped by 12.0 and 97.2\%; similar results were observed
+when ASVs were clustered into OTUs. Removing rare ASVs reduced the
+ability to detect simulated treatment effects using metrics commonly
+used to compare microbial communities.
+
+\textbf{How does the removing of rare ASVs impact the probability of
+falsely detecting a difference between treatment groups?} Observing
+reduced ability to detect differences between communities when rare ASVs
+were removed from each sample, I next asked whether removing rare ASVs
+could lead to falsely claiming that a treatment effect had a significant
+effect on community diversity and structure. First, for each dataset, I
+sampled sequences from the null distribution and randomly assigned each
+sample to one of two treatment groups and determined the number of
+observed ASVs and OTUs and their Shannon diversity index. Testing at an
+experiment-wise error rate of 0.05, I expected 5\% of the iterations for
+each dataset to yield a significant test result. Indeed, there was no
+evidence that removing rare ASVs resulted in an inflated experiment-wise
+error rate. The average fraction of significant tests did not
+meaningfully vary from 0.05 across the minimum abundance threshold,
+dataset, metric of describing sample alpha-diversity, or whether the
+abundance of ASVs or OTUs were used (\textbf{Figure XA}). Similarly, the
+average fraction of significant tests did not meaningfully vary from
+0.05 when using analysis of molecular variance to compare communities
+using Bray-Curtis distances. Second, I again sampled sequences from the
+null distribution, but assigned samples to one of two treatment groups
+based on the number of sequences in each sample. The samples with fewer
+than the median number of sequences for the dataset were assigned to one
+group and those with more than the median were assigned to the other.
+This exaggerated bias has been observed in comparisons of the lung and
+oral microbiota because of the larger nubmer of non-specific amplicons
+that can be sequenced from lung samples relative to those in the oral
+cavity {[}\textbf{REF}{]}. When rare sequences were not removed, the
+fraction of significant tests did not differ from 5\% for comparing the
+number of observed taxa (using ASVs or OTUs), their shannon diversity,
+or Bray-Curtis distances. However, when rare taxa of any frequency were
+removed, the probability of falsely detecing a difference as signifiant
+increased with the definition of rarity (\textbf{Figure XB}). Across the
+datasets, the average fraction the average fraction of falsely detecting
+a difference was when only singletons were removed was greater than
+90\%. If there is any relationship between the number of sequences and
+the treatment group, the risk of falsely rejecting the null hypothesis
+is inflated when researchers use the strategy of removing rare
+sequences. The most conservative approach is to not remove low abundance
+sequences.
+
+\textbf{\emph{Conclusion.}} Removing rare sequences decreases the
+diversity represented by 16S rRNA gene sequence data and increases the
+variation between samples. Such impacts will hinder the statistical
+power to differentiate between treatment groups. Instead of removing
+rare sequences, researchers should focus on optimizing their sequence
+generation to minimize the amount of PCR and sequencing errors. In
+addition, samples should be rarefied to a common number of sequences
+across samples without prior culling of rare sequences. The number of
+artifacts is correlated to the number of sequences being considered.
+With this in mind, rarefaction allows one to control for uneven sampling
+effort and to control for the number of artifacts in the analysis.
+
+Need to treat rare sequences with a grain of salt
 
@@
 
-\hypertarget{acknowledgements}{%
-\subsection{Acknowledgements}\label{acknowledgements}}
+\subsection{Acknowledgements}\label{acknowledgements}
 
@@
 
-\hypertarget{materials-and-methods}{%
-\subsection{Materials and Methods}\label{materials-and-methods}}
+\subsection{Materials and Methods}\label{materials-and-methods}
+
+\begin{itemize}
+\itemsep1pt\parskip0pt\parsep0pt
+\item
+  sequencing pipeline description
+\end{itemize}
 
@@
 
-\hypertarget{references}{%
-\subsection{References}\label{references}}
-
-\hypertarget{refs}{}
+\subsection{References}\label{references}
 
@@
 
+1. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
+\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
+\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
+\textbf{Thallinger GG}, \textbf{Horn DJV}, \textbf{Weber CF}. 2009.
+Introducing mothur: Open-source, platform-independent,
+community-supported software for describing and comparing microbial
+communities. Applied and Environmental Microbiology
+\textbf{75}:7537--7541.
+doi:\href{http://dx.doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.
+
 \end{document}

### 2020-09-17 submission/manuscript.Rmd
Add Table 1 and move packages to YAML

@@
     keep_tex: true
-    includes:
-      in_header: header.tex
+header-includes:
+ - \usepackage{booktabs}
+ - \usepackage{longtable}
+ - \usepackage{array}
+ - \usepackage{multirow}
+ - \usepackage{wrapfig}
+ - \usepackage{float}
+ - \usepackage{colortbl}
+ - \usepackage{pdflscape}
+ - \usepackage{tabu}
+ - \usepackage{threeparttable}
+ - \usepackage{threeparttablex}
+ - \usepackage[normalem]{ulem}
+ - \usepackage{makecell}
+ - \usepackage{setspace}
+ - \doublespacing
+ - \usepackage[left]{lineno}
+ - \linenumbers
+ - \modulolinenumbers
+ - \usepackage{helvet} % Helvetica font
+ - \renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
+ - \usepackage[T1]{fontenc}
 csl: mbio.csl
@@
 library(tidyverse)
+library(glue)
+library(kableExtra)
 library(knitr)
@@
 
+**Table 1.
+Summary of studies used in the analysis.** For all studies, the number of sequences used from each study was rarefied to the smallest sample size.
+A graphical represenation of the distribution of sample sizes for each study and the samples that were removed from each study are provided in Figure S1.
+
+```{r}
+read_tsv(here('data/process/study_summary_statistics.tsv')) %>%
+  arrange(nice_name) %>%
+  mutate(name_ref = glue("{nice_name} ({reference})"),
+    total_seqs=format(total_seqs, big.mark=",", trim=TRUE),
+    median=format(as.integer(median), big.mark=",", trim=TRUE),
+    min=format(min, big.mark=",", trim=TRUE),
+    max=format(max, big.mark=",", trim=TRUE),
+    range = glue("{min}-{max}"),
+    n_samples = n_samples,
+    fold_difference = round(fold_difference, 1)) %>%
+  select(name_ref, n_samples, total_seqs, median, range, fold_difference) %>%
+  kable("latex", booktabs=TRUE, escape=F, align="lrrrrr", linesep="",
+		col.names = linebreak(c("Study (Ref)", "Samples", "Total\nsequences", "Median\nsequences",
+			"Range of\nsequences", "Fold-difference\nbetween largest\nand smallest sample"), align="c")
+	) %>%
+	row_spec(0, bold=T)
+```
+
+\newpage
+
 <!--- \includegraphics{figure_1.png} -->
@@
 **Figure 2.**
+v

### 2020-09-17 submission/manuscript.tex
Add Table 1 and move packages to YAML

@@
 \documentclass[]{article}
+\usepackage[T1]{fontenc}
 \usepackage{lmodern}
@@
 \usepackage{fixltx2e} % provides \textsubscript
+% use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
   \usepackage[utf8]{inputenc}
@@
 \fi
-% use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 % use microtype if available
-\IfFileExists{microtype.sty}{%
-\usepackage{microtype}
-\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
+\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
 \usepackage[margin=1.0in]{geometry}
 \usepackage{graphicx}
+% Redefine \includegraphics so that, unless explicit options are
+% given, the image width will not exceed the width of the page.
+% Images get their normal width if they fit onto the page, but
+% are scaled down if they would overflow the margins.
 \makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\def\ScaleIfNeeded{%
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
 \makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\let\Oldincludegraphics\includegraphics
+{%
+ \catcode`\@=11\relax%
+ \gdef\includegraphics{\@ifnextchar[{\Oldincludegraphics}{\Oldincludegraphics[width=\ScaleIfNeeded]}}%
+}%
 \ifxetex
@@
 
-%%% Use protect on footnotes to avoid problems with footnotes in titles
-\let\rmarkdownfootnote\footnote%
-\def\footnote{\protect\rmarkdownfootnote}
-
-%%% Change title format to be more compact
-\usepackage{titling}
-
-% Create subtitle command for use in maketitle
-\providecommand{\subtitle}[1]{
-  \posttitle{
-    \begin{center}\large#1\end{center}
-    }
-}
-
-\setlength{\droptitle}{-2em}
-
-  \title{}
-    \pretitle{\vspace{\droptitle}}
-  \posttitle{}
 \author{}
-    \preauthor{}\postauthor{}
-    \date{}
-    \predate{}\postdate{}
-
+\date{\vspace{-2.5em}}
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{setspace}
+\doublespacing
+\usepackage[left]{lineno}
+\linenumbers
+\modulolinenumbers
 \usepackage{helvet} % Helvetica font
@@
 \usepackage[T1]{fontenc}
-
-\usepackage[none]{hyphenat}
-
-\usepackage{setspace}
-\doublespacing
-\setlength{\parskip}{1em}
-
-\usepackage{lineno}
-
-\usepackage{pdfpages}
 
@@
 
+\textbf{Table 1. Summary of studies used in the analysis.} For all
+studies, the number of sequences used from each study was rarefied to
+the smallest sample size. A graphical represenation of the distribution
+of sample sizes for each study and the samples that were removed from
+each study are provided in Figure S1.
+
+\begin{tabular}{lrrrrr}
+\toprule
+\textbf{Study (Ref)} & \textbf{Samples} & \textbf{\makecell[c]{Total\\sequences}} & \textbf{\makecell[c]{Median\\sequences}} & \textbf{\makecell[c]{Range of\\sequences}} & \textbf{\makecell[c]{Fold-difference\\between largest\\and smallest sample}}\\
+\midrule
+Bioethanol (NA) & 95 & 3,972,943 & 16,015 & 3,688-356,136 & 96.6\\
+Human (NA) & 490 & 20,909,768 & 32,505 & 10,523-430,415 & 40.9\\
+Lake (NA) & 52 & 3,169,868 & 69,041 & 15,347-112,871 & 7.4\\
+Marine (NA) & 7 & 1,391,396 & 193,464 & 133,516-254,060 & 1.9\\
+Mice (NA) & 348 & 2,813,747 & 6,477 & 1,804-30,565 & 16.9\\
+Peromyscus (NA) & 111 & 1,555,545 & 12,446 & 4,464-33,644 & 7.5\\
+Rainforest (NA) & 69 & 946,295 & 11,561 & 4,932-37,767 & 7.7\\
+Rice (NA) & 490 & 22,591,168 & 43,216 & 2,776-193,464 & 69.7\\
+Seagrass (NA) & 286 & 4,130,454 & 13,567 & 1,803-45,191 & 25.1\\
+Sediment (NA) & 58 & 1,154,174 & 17,584 & 7,685-68,321 & 8.9\\
+Soil (NA) & 18 & 956,656 & 51,844 & 47,806-59,956 & 1.3\\
+Stream (NA) & 201 & 21,162,574 & 90,159 & 9,175-390,964 & 42.6\\
+\bottomrule
+\end{tabular}
+
+\newpage
+
 \textbf{Figure 1.}
@@
 
-\textbf{Figure 2.}
+\textbf{Figure 2.} v
 

### 2020-10-15 code/plot_seqs_per_sample.R
Output tiff with lzw compression

@@ -49,4 +49,4 @@ ggplot(counts_data, aes(y=n_seqs, x=study)) +
 	labs(y="Number of sequences per sample", x=NULL) +
 	theme_classic()
 
-ggsave("results/figures/seqs_per_sample.pdf", width=7, height=4)
+ggsave("results/figures/seqs_per_sample.tiff", width=7, height=4, compression="lzw")

### 2020-10-15 submission/manuscript.Rmd
Create figure showing quality of singletons

@@
 Contrary to this expectation, the median percentage of sequences that were discarded when singleton ASVs were removed from each dataset varied between `r oxford_comma(percent_singletons_range_vals)`% (`r oxford_comma(percent_singletons_range_names)`).
-In addition, with the exception of the samples from the `r oxford_comma(cor_n_nosig)` datasets (Spearman correlation, P>0.05), the fraction of singleton ASVs in the sample was negatively correlated with the number of sequences in each sample with a range between `r oxford_comma(cor_n_sig_values)` (`r oxford_comma(cor_n_sig_names)`).
+In addition, with the exception of the samples from the `r oxford_comma(cor_n_nosig)` datasets (Spearman correlation, P>0.05), the fraction of singleton ASVs in the sample was negatively correlated with the number of sequences in each sample with a range between `r oxford_comma(cor_n_sig_values)` (`r oxford_comma(cor_n_sig_names)`) (Figure 1A).
 This showed that with additional sequencing, the probability of seeing rare ASVs multiple times was greater than the probability of generating an artifact.
@@
 In fact, singleton ASVs from samples with fewer sequences were often found in samples with more sequences.
-At least 50% of the singleton ASVs found in the samples from the `r oxford_comma(high_coverage)` datasets were found in another sample from the same dataset.
+At least 50% of the singleton ASVs found in the samples from the `r oxford_comma(high_coverage)` datasets were found in another sample from the same dataset (Figure 1B).
 Considering the likelihood of finding an ASV duplicated in another sample is confounded by the number of samples and inter-sample diversity, the high coverage of singleton ASVs in these datasets was remarkable.
-The correlation between the number of sequences in a sample and the fraction of that sample's singleton ASVs that were covered by another sample in the dataset was significant and negative for `r nrow(coverage_n) - length(coverage_n_nosig)` of the datasets ranging between `r oxford_comma(coverage_n_sig_values)` for the `r oxford_comma(coverage_n_sig_names)` datasets, respectively.
+The correlation between the number of sequences in a sample and the fraction of that sample's singleton ASVs that were covered by another sample in the dataset was significant and negative for `r nrow(coverage_n) - length(coverage_n_nosig)` of the datasets ranging between `r oxford_comma(coverage_n_sig_values)` for the `r oxford_comma(coverage_n_sig_names)` datasets, respectively (Figure 1C).
 The negative correlation indicated that the singleton ASVs in the smaller samples were more likely to be covered by ASVs in the larger samples.
@@
 To quantify the effect of removing rare ASVs on the information contained within each sample, I varied the minimum abundance threshold to simulate removing ASVs of varying rarity from each sample.
-The number of ASVs in each sample decreased by between `r oxford_comma(asv_s_two, 1)`% when removing those ASVs that only appeared once and by between `r oxford_comma(asv_s_eleven, 1)`% when removing those that appeared ten or fewer times from each sample (Figure 1).
-Similarly, the Shannon diversity decreased by between `r oxford_comma(asv_h_two, 1)`% when removing ASVs that only appeared once and by between `r oxford_comma(asv_s_eleven, 1)`% when removing ASVs that appeared ten or fewer times from each sample (Figure 1).
+The number of ASVs in each sample decreased by between `r oxford_comma(asv_s_two, 1)`% when removing those ASVs that only appeared once and by between `r oxford_comma(asv_s_eleven, 1)`% when removing those that appeared ten or fewer times from each sample (Figure X).
+Similarly, the Shannon diversity decreased by between `r oxford_comma(asv_h_two, 1)`% when removing ASVs that only appeared once and by between `r oxford_comma(asv_s_eleven, 1)`% when removing ASVs that appeared ten or fewer times from each sample (Figure X).
 Next, I assigned the ASVs to OTUs to assess the impact of removing rare ASVs on higher level taxonomic groupings that are commonly used in microbial ecology studies.
@@
 I calculated the Kullback–Leibler divergence from the full to pruned communities when rare ASVs were removed.
-As the threshold for removing ASVs increased, the amount of information lost also increased for both ASVs and OTUs (Figure 1).
+As the threshold for removing ASVs increased, the amount of information lost also increased for both ASVs and OTUs (Figure X).
 The relative loss of information was generally lower for OTUs than than it was for ASVs.

### 2020-10-15 code/plot_correlation_coverage.R

@@ -1,0 +1,88 @@
+library(tidyverse)
+library(cowplot)
+
+studies <- c("bioethanol", "human", "lake", "marine", "mice", "peromyscus",
+							"rainforest", "rice", "seagrass", "sediment", "soil",
+							"stream") %>%
+						sort(decreasing=T)
+
+nice_studies <- c("Bioethanol", "Human", "Lake", "Marine", "Mice", "Peromyscus",
+									"Rainforest", "Rice", "Seagrass", "Sediment", "Soil",
+									"Stream") %>%
+								sort(decreasing=T)
+names(nice_studies) <- studies
+
+methods <- str_wrap(c(
+	"Correlation between the number of sequences and the number of singletons in a sample",
+	"Correlation between the number of sequences and the fraction of singletons shared across samples",
+	"Fraction of singletons shared across samples"), width=35)
+names(methods) <- c("cor_loss", "cor_coverage", "coverage")
+
+correlations <- bind_rows(
+		cor_loss=read_tsv('data/process/sequence_loss_table_cor.tsv'),
+		cor_coverage=read_tsv('data/process/sequence_coverage_table_cor.tsv'),
+		.id="comparison"
+	) %>%
+ 	filter(min_freq == 1) %>%
+  mutate(sig = if_else(cor_n_p.value < 0.05, NA_character_, "not\nsignificant"),
+    value = if_else(is.na(sig), cor_n_estimate, NA_real_)
+	) %>%
+	select(comparison, dataset, value, sig)
+
+coverage <- read_tsv('data/process/sequence_coverage_table_raw.tsv') %>%
+	filter(min_freq == 1) %>%
+	group_by(dataset) %>%
+	summarize(
+		value=median(coverage),
+		lci = quantile(coverage, prob=0.025),
+		uci = quantile(coverage, prob=0.975),
+		.groups="drop"
+	) %>%
+	mutate(comparison = "coverage")
+
+limits <- tibble(
+	comparison = c("cor_loss", "cor_loss", "cor_coverage", "cor_coverage", "coverage", "coverage"),
+	value = c(-1, 0, -1, 0, 0, 1),
+	dataset = c("zlower", "zupper", "zlower", "zupper", "zlower", "zupper")
+)
+
+panel_label <- tibble(
+	comparison = factor(c("cor_loss", "cor_coverage", "coverage"),
[…]

### 2020-10-20 code/plot_correlation_coverage.R
Fix panel labels

@@ -54,7 +54,7 @@ panel_label <- tibble(
 	y=c(12.5, 12.5, 12.5)
 )
 
-bind_rows(correlations, coverage, limits) %>%
+figure <- bind_rows(correlations, coverage, limits) %>%
 	mutate(comparison = factor(comparison,
  		levels=c("cor_loss", "coverage", "cor_coverage")),
 		dataset = factor(dataset, levels=studies)	) %>%
@@ -78,11 +78,16 @@ bind_rows(correlations, coverage, limits) %>%
 		axis.text.x=element_text(size=7),
 		line=element_line(lineend="round"),
 		plot.margin=margin(t=10)
-	) +
-	geom_text(data=panel_label,
-		aes(x=x, y=y, label=label),
-		fontface="bold",
-		size=8,
-		nudge_y=0.25, nudge_x=0.02)
+	)
 
-ggsave("results/figures/correlation_coverage.tiff", width=7, height=4.25, compression = "lzw")
+ggdraw(figure) +
+	draw_plot_label(
+		label=c("A", "B", "C"),
+		x=c(0.1, 0.4, 0.7),
+		y=c(1.02, 1.02, 1.02),
+		size = 20,
+		fontface="bold"
+	)
+
+
+ggsave("results/figures/correlation_coverage.tiff", width=6.87, height=4.25, compression = "lzw")

### 2020-10-20 submission/manuscript.Rmd
Create Figures 2 and S2

@@
 To quantify the effect of removing rare ASVs on the information contained within each sample, I varied the minimum abundance threshold to simulate removing ASVs of varying rarity from each sample.
-The number of ASVs in each sample decreased by between `r oxford_comma(asv_s_two, 1)`% when removing those ASVs that only appeared once and by between `r oxford_comma(asv_s_eleven, 1)`% when removing those that appeared ten or fewer times from each sample (Figure X).
-Similarly, the Shannon diversity decreased by between `r oxford_comma(asv_h_two, 1)`% when removing ASVs that only appeared once and by between `r oxford_comma(asv_s_eleven, 1)`% when removing ASVs that appeared ten or fewer times from each sample (Figure X).
+The number of ASVs in each sample decreased by between `r oxford_comma(asv_s_two, 1)`% when removing those ASVs that only appeared once and by between `r oxford_comma(asv_s_eleven, 1)`% when removing those that appeared ten or fewer times from each sample (Figure 2A).
+Similarly, the Shannon diversity decreased by between `r oxford_comma(asv_h_two, 1)`% when removing ASVs that only appeared once and by between `r oxford_comma(asv_s_eleven, 1)`% when removing ASVs that appeared ten or fewer times from each sample (Figure 2B).
 Next, I assigned the ASVs to OTUs to assess the impact of removing rare ASVs on higher level taxonomic groupings that are commonly used in microbial ecology studies.
-Although pooling similar ASVs into OTUs reduced the impact of removing the rare ASVs relative to the ASV-based analysis, the minimum abundance threshold still decreased the number of observed OTUs and the diversity decreased relative to the original community (**Figure 1**).
+Although pooling similar ASVs into OTUs reduced the impact of removing the rare ASVs relative to the ASV-based analysis, the minimum abundance threshold still decreased the number of observed OTUs and the diversity decreased relative to the original community (**Figure S2AB**).
 In contrast to the number of ASVs or OTUs and their diversity, the Kullback–Leibler divergence compares the relative abundance of each ASV or OTU between representations of the community.
 I calculated the Kullback–Leibler divergence from the full to pruned communities when rare ASVs were removed.
-As the threshold for removing ASVs increased, the amount of information lost also increased for both ASVs and OTUs (Figure X).
+As the threshold for removing ASVs increased, the amount of information lost also increased for both ASVs and OTUs (Figure 2C and Figure S2C).
 The relative loss of information was generally lower for OTUs than than it was for ASVs.

### 2020-10-20 code/plot_loss_of_information.R

@@ -1,0 +1,94 @@
+library(tidyverse)
+library(cowplot)
+
+studies <- c("bioethanol", "human", "lake", "marine", "mice", "peromyscus",
+							"rainforest", "rice", "seagrass", "sediment", "soil",
+							"stream")
+
+nice_studies <- studies %>% stringr::str_to_title()
+
+
+loss <- read_tsv("data/process/ointra_analysis.tsv",
+  	col_types = cols(.default=col_double(),
+      dataset = col_character(),
+      method=col_character(),
+      group=col_character()
+    )
+  ) %>%
+  group_by(dataset, method, min_class) %>%
+  summarize(mean_s_drop = 100*(1-mean(s_yes / s_no)),
+    mean_h_drop = 100*(1-mean(h_yes / h_no)),
+    mean_kl_drop = mean(kl),
+    .groups="drop"
+  )
+
+
+plot_data <- function(subset_method, subset_string) {
+
+  subset_data <- loss %>%
+    filter(method == subset_method) %>%
+    pivot_longer(starts_with("mean"), names_to="metric", values_to="value") %>%
+    mutate(dataset = factor(dataset, levels=studies, labels=nice_studies)) %>%
+    mutate(metric = factor(metric,
+      levels=c("mean_s_drop", "mean_h_drop", "mean_kl_drop"),
+      labels=c(
+        paste0("Reduction in number\nof ", subset_string, "s (%)"),
+        paste0("Reduction in Shannon\ndiversity using ", subset_string, "s (%)"),
+        paste0("Kullbac-Leibler divergence\nusing ", subset_string, "s"))
+      )
+    )
+
+  subset_data_points <- subset_data %>%
+      filter(min_class %in% c(2, 5, 8, 11))
+
+  figure <- ggplot(
+			subset_data,
+			aes(x=min_class, y=value, group=dataset, color=dataset)
+		) +
+    geom_line() +
+    geom_point(
+      data=subset_data_points,
+      mapping=aes(x=min_class, y=value, color=dataset, shape=dataset, fill=dataset)
+    ) +
+    facet_wrap(facet="metric", ncol=1, scales="free_y", strip.position="left") +
+    labs(y=NULL, x="Smallest number of sequences per ASV") +
+    scale_y_continuous(limits=c(0, NA)) +
+    scale_x_continuous(breaks=2:11) +
[…]

### 2020-10-20 submission/manuscript.Rmd
Create Figure S3, rename files

@@
 Because of the risk of bias if only one representation of the null distribution was generated, I generated 100 randomized datasets for each study.
-The same trends between removing rare ASVs and the number of ASVs, sample diversity, and information loss that were identified using the observed community community distribution data were also identified with the data from the null distribution (**Figure S2**).
-The null distribution data were used in the remainder of the study.
+The trends between removing rare ASVs and the number of ASVs or OTUs, sample diversity, and information loss that were identified using the observed community community distribution data were also identified with the data from the null distribution; however, the losses were larger when using the null distribution data (**Figure S3**).
+The null distribution data were used in the remainder of the study to minimize the risk of bias.
 

### 2020-10-20 code/plot_loss_of_information_obs.R

@@ -32,9 +32,9 @@ plot_data <- function(subset_method, subset_string) {
     mutate(metric = factor(metric,
       levels=c("mean_s_drop", "mean_h_drop", "mean_kl_drop"),
       labels=c(
-        paste0("Reduction in number\nof ", subset_string, "s (%)"),
-        paste0("Reduction in Shannon\ndiversity using ", subset_string, "s (%)"),
-        paste0("Kullbac-Leibler divergence\nusing ", subset_string, "s"))
+        paste0("Reduction in\nrichness (%)"),
+        paste0("Reduction in\nShannon diversity (%)"),
+        paste0("Kullbac-Leibler\ndivergence"))
       )
     )
 
@@ -82,7 +82,11 @@ plot_data <- function(subset_method, subset_string) {
 		)
 
   ggsave(
-    paste0("results/figures/", tolower(subset_string), "_loss_of_information.tiff"),
+    paste0(
+			"results/figures/",
+			tolower(subset_string),
+			"_loss_of_information_obs.tiff"
+		),
     width=6.87,
     height=6,
     compression = "lzw"

### 2020-10-20 code/plot_loss_of_information_random.R

@@ -1,0 +1,85 @@
+library(tidyverse)
+library(cowplot)
+
+studies <- c("bioethanol", "human", "lake", "marine", "mice", "peromyscus",
+							"rainforest", "rice", "seagrass", "sediment", "soil",
+							"stream")
+
+nice_studies <- studies %>% stringr::str_to_title()
+
+
+loss <- read_tsv("data/process/rintra_analysis.tsv",
+  	col_types = cols(.default=col_double(),
+      dataset = col_character(),
+      method=col_character(),
+      group=col_character()
+    )
+  ) %>%
+  group_by(dataset, method, min_class) %>%
+  summarize(mean_s_drop = 100*(1-mean(s_yes / s_no)),
+    mean_h_drop = 100*(1-mean(h_yes / h_no)),
+    mean_kl_drop = mean(kl),
+    .groups="drop"
+  )
+
+
+subset_data <- loss %>%
+  pivot_longer(starts_with("mean"), names_to="metric", values_to="value") %>%
+  filter(is.finite(value)) %>%
+  mutate(dataset = factor(dataset, levels=studies, labels=nice_studies)) %>%
+  mutate(metric = factor(metric,
+    levels=c("mean_s_drop", "mean_h_drop", "mean_kl_drop"),
+    labels=c(
+      paste0("Reduction in\nrichness (%)"),
+      paste0("Reduction in\nShannon diversity (%)"),
+      paste0("Kullbac-Leibler\ndivergence"))
+    )
+  ) %>%
+  mutate(method = factor(method,
+    levels=c("pc", "otu"),
+    labels=c("ASVs", "OTUs")
+    )
+  )
+
+subset_data_points <- subset_data %>%
+    filter(min_class %in% c(2, 5, 8, 11))
+
+ggplot(
+		subset_data,
+		aes(x=min_class, y=value, group=dataset, color=dataset)
+	) +
+  geom_line() +
+  geom_point(
+    data=subset_data_points,
+    mapping=aes(x=min_class, y=value, color=dataset, shape=dataset, fill=dataset)
+  ) +
+  facet_grid(metric~method, scales="free_y", switch="y") +
+  labs(y=NULL, x="Smallest number of sequences per ASV") +
+  scale_y_continuous(limits=c(0, NA)) +
+  scale_x_continuous(breaks=2:11) +
+  scale_color_manual(name=NULL,
[…]

### 2020-10-20 submission/manuscript.Rmd
Create Figures 3 and S4

@@
 otu_cov_bc <- r_cov_bc %>% filter(method == "otu") %>% arrange(cv_diff) %>% slice(1, nrow(.)) %>% pull(cv_diff)
-
 ```
@@
 standard deviation divided by the mean) for the number of ASVs and diversity for each study at multiple abundance thresholds.
-The COV for the number of ASVs across the studies after removing singletons were between `r oxford_comma(pc_cov_sobs, 1)`-times larger than they were without removing singleton ASVs.
-Similarly, the COVs for the diversity of ASVs were between `r oxford_comma(pc_cov_shannon, 1)`-times larger when singletons were removed than when they were not removed.To assess the impact of removing rare ASVs on measures of beta diversity between samples I calculated the COV of the Bray-Curtis distances between samples within the same study at multiple abundance thresholds.
-The COV between Bray-Curtis distances within a study when singletons were removed was between `r oxford_comma(pc_cov_bc, 1)`-times larger than when they were not removed.
+The COV for the number of ASVs across the studies after removing singletons were between `r oxford_comma(pc_cov_sobs, 1)`-times larger than they were without removing singleton ASVs (Figure 3A).
+Similarly, the COVs for the diversity of ASVs were between `r oxford_comma(pc_cov_shannon, 1)`-times larger when singletons were removed than when they were not removed (Figure 3B).
+To assess the impact of removing rare ASVs on measures of beta diversity between samples I calculated the COV of the Bray-Curtis distances between samples within the same study at multiple abundance thresholds.
+The COV between Bray-Curtis distances within a study when singletons were removed was between `r oxford_comma(pc_cov_bc, 1)`-times larger than when they were not removed (Figure 3C).
 Increasing the minimum abundance threshold increased the COV between samples when using metrics of alpha and beta diversity.
-When ASVs were clustered into OTUs the difference in the COV was less than it was for the ASVs (**Figure 2**).
+When ASVs were clustered into OTUs the difference in the COV was less than it was for the ASVs (Figure S4).
 These results indicate that removing rare ASVs increases the dissimilarity between samples, which could have a significant impact on the statistical power to detect differences between treatment groups.

### 2020-10-20 code/plot_inter_sample_variation.R

@@ -1,0 +1,109 @@
+library(tidyverse)
+library(cowplot)
+
+studies <- c("bioethanol", "human", "lake", "marine", "mice", "peromyscus",
+							"rainforest", "rice", "seagrass", "sediment", "soil",
+							"stream")
+
+nice_studies <- studies %>% stringr::str_to_title()
+
+
+r_cov_alpha <- read_tsv("data/process/ralpha_analysis.tsv",
+		col_types = cols(.default=col_double(),
+      dataset=col_character(), method=col_character())
+    ) %>%
+    select(dataset, method, prune, cv_sobs, cv_shannon)
+
+r_cov_bc <- read_tsv("data/process/rbeta_analysis.tsv",
+		col_types = cols(.default=col_double(),
+					dataset=col_character(), method=col_character())) %>%
+    select(dataset, method, prune, sd_mean) %>%
+    rename(cv_braycurtis = sd_mean)
+
+cov_data <- inner_join(r_cov_alpha, r_cov_bc, by=c("dataset", "method", "prune")) %>%
+  pivot_longer(
+    starts_with("cv_"),
+    names_to="metric",
+    names_prefix="cv_",
+    values_to="cv"
+  ) %>%
+  rename(min_class = prune) %>%
+  mutate(cv = 100 * cv)
+
+
+plot_data <- function(subset_method, subset_string) {
+
+  subset_data <- cov_data %>%
+    filter(method == subset_method) %>%
+    mutate(metric = factor(
+      metric,
+      levels=c("sobs", "shannon", "braycurtis"),
+      labels=c(paste0("Richness"),
+          paste0("Shannon diversity"),
+          paste0("Bray-Curtis distances")
+        )
+      )
+    ) %>%
+    mutate(dataset = factor(dataset, levels=studies, labels=nice_studies))
+
+
+  subset_data_points <- subset_data %>%
+      filter(min_class %in% c(2, 5, 8, 11))
+
+  figure <- ggplot(
+        data=subset_data,
+        aes(x=min_class, y=cv, group=dataset, color=dataset)
+      ) +
+      geom_line() +
+      geom_point(
+        data=subset_data_points,
+        mapping=aes(x=min_class, y=cv, color=dataset, shape=dataset, fill=dataset)
+      ) +
+      facet_wrap(facet="metric", ncol=1, strip.position = "left", scales="free_y") +
[…]

### 2020-10-20 submission/manuscript.Rmd
Create Figures 4 and S5

@@
 The fraction of significant tests was a measurement of the statistical power to detect the difference between the treatment groups.
+
+This doesn't seem to agree with the figure...
 Across all of the metrics I tested, the marine dataset yielded no simulated sets that were statistically significant, which was likely due to the small number of samples in the study (N=7); the marine dataset is not included in the following analysis.
-Among the remaining datasets, the power to detect a difference in the observed number of ASVs ranged between `r oxford_comma(b_base_pc_sobs, 2)` and between `r oxford_comma(b_base_pc_shannon, 2)` to detect a difference in diversity when using a Wilcox test (**Figure X**).
-When singleton ASVs were removed, the power to detect a difference in the number of ASVs dropped by between `r oxford_comma(b_pc_sobs, 1)`% and by between `r oxford_comma(b_pc_shannon, 1)`% to detect a difference in diversity.
-The effect of removing rare ASVs on the number of OTUs and their diversity was similar.
+
+Among the remaining datasets, the power to detect a difference in the observed number of ASVs ranged between `r oxford_comma(b_base_pc_sobs, 2)` and between `r oxford_comma(b_base_pc_shannon, 2)` to detect a difference in diversity when using a Wilcox test (Figure 4A).
+When singleton ASVs were removed, the power to detect a difference in the diversity of ASVs dropped by between `r oxford_comma(b_pc_sobs, 1)`% and by between `r oxford_comma(b_pc_shannon, 1)`% (Figure 4B).
+The effect of removing rare ASVs on the number of OTUs and their diversity was similar (Figure S5AB).
 A popular alternative approach calculates the dissimilarity between communities based on the abundance of each ASV or OTU.
-We used the Bray-Curtis dissimilarity index to compare the simulated communities within each dataset and calculated the power to detect differences between the two simulated treatment groups using the analysis of molecular variance (also called PERMANOVA).
+We used the Bray-Curtis dissimilarity index to compare the simulated communities within each dataset and calculated the power to detect differences between the two simulated treatment groups using the analysis of molecular variance (also called PERMANOVA) (Figure 4C and S5C).
 Without removing rare sequences, the power to detect a difference between the two simulated treatment groups varied between `r oxford_comma(b_base_beta, 2)`.

### 2020-10-20 code/plot_power.R

@@ -1,0 +1,113 @@
+library(tidyverse)
+library(cowplot)
+
+studies <- c("bioethanol", "human", "lake", "marine", "mice", "peromyscus",
+							"rainforest", "rice", "seagrass", "sediment", "soil",
+							"stream")
+
+nice_studies <- studies %>% stringr::str_to_title()
+
+
+b_alpha <- read_tsv(
+		"data/process/bffect_alpha_analysis.tsv",
+		col_types = cols(.default=col_double(),
+					dataset=col_character(), method=col_character(), metric=col_character()
+				)
+	) %>%
+	filter(metric != "invsimpson") %>%
+	pivot_wider(names_from=metric, values_from=frac_sig)
+
+b_beta <- read_tsv(
+		"data/process/bffect_beta_analysis.tsv",
+		col_types = cols(.default=col_double(),
+					dataset=col_character(), clustering=col_character()
+				)
+	) %>%
+	rename(braycurtis=frac_sig, method = clustering)
+
+cov_data <- inner_join(b_alpha, b_beta, by=c("dataset", "method", "prune")) %>%
+  pivot_longer(
+    cols=c(shannon, sobs, braycurtis),
+    names_to="metric",
+    values_to="percent_sig"
+  ) %>%
+  rename(min_class = prune) %>%
+  mutate(percent_sig = 100 * percent_sig)
+
+
+plot_data <- function(subset_method, subset_string) {
+
+  subset_data <- cov_data %>%
+    filter(method == subset_method) %>%
+    mutate(metric = factor(
+      metric,
+      levels=c("sobs", "shannon", "braycurtis"),
+      labels=c(paste0("Richness"),
+          paste0("Shannon diversity"),
+          paste0("Bray-Curtis distances")
+        )
+      )
+    ) %>%
+    mutate(dataset = factor(dataset, levels=studies, labels=nice_studies))
+
+
+  subset_data_points <- subset_data %>%
+      filter(min_class %in% c(1, 3, 5, 7, 9, 11))
+
+  figure <- ggplot(
+        data=subset_data,
+        aes(x=min_class, y=percent_sig, group=dataset, color=dataset)
+      ) +
+      geom_line() +
+      geom_point(
+        data=subset_data_points,
+        mapping=aes(x=min_class, y=percent_sig, color=dataset, shape=dataset, fill=dataset)
+      ) +
[…]

### 2020-10-20 submission/manuscript.Rmd
Create Figures 5 and S6

@@
 Indeed, there was no evidence that removing rare ASVs resulted in an inflated experiment-wise error rate.
-The average fraction of significant tests did not meaningfully vary from 0.05 across the minimum abundance threshold, dataset, metric of describing sample alpha-diversity, or whether the abundance of ASVs or OTUs were used (**Figure XA**).
-Similarly, the average fraction of significant tests did not meaningfully vary from 0.05 when using analysis of molecular variance to compare communities using Bray-Curtis distances.
+The average fraction of significant tests did not meaningfully vary from 0.05 across the minimum abundance threshold, dataset, metric of describing sample alpha-diversity, or whether the abundance of ASVs or OTUs were used (Figure 5A and S6A).
+Similarly, the average fraction of significant tests did not meaningfully vary from 0.05 when using analysis of molecular variance to compare communities using Bray-Curtis distances (Figure 5A and S6A).
 Second, I again sampled sequences from the null distribution, but assigned samples to one of two treatment groups based on the number of sequences in each sample.
@@
 This exaggerated bias has been observed in comparisons of the lung and oral microbiota because of the larger nubmer of non-specific amplicons that can be sequenced from lung samples relative to those in the oral cavity [**REF**].
-When rare sequences were not removed, the fraction of significant tests did not differ from 5% for comparing the number of observed taxa (using ASVs or OTUs), their shannon diversity, or Bray-Curtis distances.
-However, when rare taxa of any frequency were removed, the probability of falsely detecing a difference as signifiant increased with the definition of rarity (**Figure XB**).
+When rare sequences were not removed, the fraction of significant tests did not differ from 5% for comparing the number of observed taxa (using ASVs or OTUs), their shannon diversity, or Bray-Curtis distances (Figure 5B and S6B).
+However, when rare taxa of any frequency were removed, the probability of falsely detecing a difference as signifiant increased with the definition of rarity (Figure 5B and S6B).
+
 Across the datasets, the average fraction the average fraction of falsely detecting a difference was when only singletons were removed was greater than 90%.
+
 If there is any relationship between the number of sequences and the treatment group, the risk of falsely rejecting the null hypothesis is inflated when researchers use the strategy of removing rare sequences.

### 2020-10-20 code/plot_type1.R

@@ -1,0 +1,131 @@
+library(tidyverse)
+library(cowplot)
+
+studies <- c("bioethanol", "human", "lake", "marine", "mice", "peromyscus",
+							"rainforest", "rice", "seagrass", "sediment", "soil",
+							"stream")
+
+nice_studies <- studies %>% stringr::str_to_title()
+
+null_alpha <- read_tsv("data/process/rffect_alpha_analysis.tsv",
+		col_types = cols(.default=col_double(),
+					dataset=col_character(), method=col_character(), metric=col_character()
+        )
+      ) %>%
+      filter((metric == "sobs" | metric == "shannon"))
+
+null_beta <- read_tsv("data/process/rffect_beta_analysis.tsv",
+		col_types = cols(.default=col_double(),
+					dataset=col_character(), clustering=col_character()
+        )
+      ) %>%
+      mutate(metric = "braycurtis") %>%
+      rename(method = clustering)
+
+null <- bind_rows(null_alpha, null_beta)
+
+skew_alpha <- read_tsv("data/process/sffect_alpha_analysis.tsv",
+		col_types = cols(.default=col_double(),
+					dataset=col_character(), method=col_character(), metric=col_character()
+        )
+      ) %>%
+  filter((metric == "sobs" | metric == "shannon"))
+
+skew_beta <- read_tsv("data/process/sffect_beta_analysis.tsv",
+		col_types = cols(.default=col_double(),
+					dataset=col_character(), clustering=col_character())) %>%
+  mutate(metric = "braycurtis") %>%
+  rename(method = clustering)
+
+skew <- bind_rows(skew_alpha, skew_beta)
+
+type_one <- bind_rows(null=null, skew=skew, .id="simulation") %>%
+  rename(min_class = prune, percent_sig = frac_sig) %>%
+  mutate(percent_sig = 100 * percent_sig)
+
+subset_method <- "pc"
+subset_string <- "ASV"
+
+
+plot_data <- function(subset_method, subset_string) {
+
+  subset_data <- type_one %>%
+    filter(method == subset_method) %>%
+    mutate(metric = factor(
+      metric,
+      levels=c("sobs", "shannon", "braycurtis"),
+      labels=c(paste0("Richness"),
+          paste0("Shannon diversity"),
+          paste0("Bray-Curtis distances")
+        )
+      )
+    ) %>%
+    mutate(
[…]

### 2020-10-20 code/plot_inter_sample_variation.R
Increase frequency of plotting symbol

@@ -48,7 +48,7 @@ plot_data <- function(subset_method, subset_string) {
 
 
   subset_data_points <- subset_data %>%
-      filter(min_class %in% c(2, 5, 8, 11))
+      filter(min_class %in% c(1, 3, 5, 7, 9, 11))
 
   figure <- ggplot(
         data=subset_data,

### 2020-10-27 submission/manuscript.Rmd
Edit Results section

@@
 
-## word choice: sample vs. community
-
 ```{r knitr_settings, tidy=TRUE, eval=TRUE, echo=FALSE, cache=FALSE}
@@
 library(tidyverse)
-library(glue)
-library(kableExtra)
+suppressPackageStartupMessages(library(glue))
+suppressPackageStartupMessages(library(kableExtra))
 library(knitr)
@@
 ## Abstract
+
+**word choice: sample vs. community**
+
 250 words
@@
 	pull(n_samples) %>%
-	range()
+	range() %>% as.integer()
 
@@
 	pull(median) %>%
-	round() %>%
-	range()
+	range() %>% as.integer()
 
@@
 Sequencing of mock communities confirms that artifacts tend to be rare.
-Proponents of abundance-based screening point to their ability to obtain the correct number of ASVs, OTUs, or phylogypes with data generated from sequencing mock communities when rare ASVs are removed.
+Proponents of abundance-based screening point to their ability to obtain the correct number of ASVs, OTUs, or phylotypes with data generated from sequencing mock communities when rare ASVs are removed.
 However, this approach effectively overfits the curation pipeline to data generated from a phylogenetically simple community with an atypical community distribution that is often sequenced to a depth that is not achieved with biological samples.
@@
 
-***Results.***
+## Results
 
@@
 250 nt-long region.
-This region and sequencing platform were selected because previous work has shown that a standard data analysis pipeline in mothur results in a sequencing error rate below 0.02% [@Kozich2012].
-All sequence data were obtained from the Sequence Read Archive and processed using a standard mothur-based sequencing pipeline that resulted in ASVs as generated by the pre.cluster algorithm [@Kozich2012; @Schloss2009].
-After removing poor quality and chimeric ASVs and samples that had uncharacteristically low number of sequences for the dataset, these datasets included between `r oxford_comma(n_sample_range)` samples.
-The median number of sequences for each dataset ranged between `r oxford_comma(n_med_sequence_range)` (Table 1, Figure S1).
-Strikingly, aside from the relatively small `r oxford_comma(low_fold_samples)` datasets, the difference between the sample with the fewest sequences and the sample with the most sequences for each dataset varied between `r oxford_comma(high_fold_range, digits=1)`-fold (Table 1).
+This region and sequencing platform were selected because previous work has shown that a standard data analysis pipeline in mothur results in a sequencing error rate below 0.02% [@Kozich2013].
+All sequence data were obtained from the Sequence Read Archive and processed using a standard mothur-based sequencing pipeline that resulted in ASVs as generated by the pre.cluster algorithm [@Kozich2013; @Schloss2009].
+After removing poor quality and chimeric ASVs and samples that had uncharacteristically low number of sequences for the dataset, these datasets included between `r oxford_comma(n_sample_range)` samples (Figure S1).
+The median number of sequences for each dataset ranged between `r oxford_comma(n_med_sequence_range)` (Table 1).
+Strikingly, aside from the relatively small `r oxford_comma(low_fold_samples)` datasets, the difference between the sample with the fewest sequences and the sample with the most sequences for each dataset varied by between `r oxford_comma(high_fold_range, digits=1)`-fold (Table 1).
 
@@
 Contrary to this expectation, the median percentage of sequences that were discarded when singleton ASVs were removed from each dataset varied between `r oxford_comma(percent_singletons_range_vals)`% (`r oxford_comma(percent_singletons_range_names)`).
-In addition, with the exception of the samples from the `r oxford_comma(cor_n_nosig)` datasets (Spearman correlation, P>0.05), the fraction of singleton ASVs in the sample was negatively correlated with the number of sequences in each sample with a range between `r oxford_comma(cor_n_sig_values)` (`r oxford_comma(cor_n_sig_names)`) (Figure 1A).
-This showed that with additional sequencing, the probability of seeing rare ASVs multiple times was greater than the probability of generating an artifact.
-This suggests that the rare ASVs are not as likely to be artifacts as previously thought.
-Furthermore, if rare ASVs were artifacts, then one would not expect to find them in other samples from the same dataset.
+In addition, with the exception of the samples from the `r oxford_comma(cor_n_nosig)` datasets (Spearman correlation, P>0.05), the fraction of singleton ASVs in samples was negatively correlated with the number of sequences in each sample with a range between `r oxford_comma(cor_n_sig_values)` (`r oxford_comma(cor_n_sig_names)`) (Figure 1A).
+This showed that with additional sequencing, the probability of seeing singleton ASVs in multiple samples was greater than the probability of generating an artifact.
+This suggests that the singleton ASVs are not as likely to be artifacts as previously thought.
+Furthermore, if singleton ASVs were artifacts, then one would not expect to find them in other samples from the same dataset.
 In fact, singleton ASVs from samples with fewer sequences were often found in samples with more sequences.
@@
 Among the three datasets without a significant correlation (Spearman correlation, P>0.05), the marine and soil datasets had the fewest samples in our collection and the stream dataset already had a high level of coverage regardless of the number of sequences.
-Contrary to the common motivation for removing rare ASVs, these results indicate that this practice disproportionately impacts samples with fewer sequences and likely removes a large number of non-artifact ASVs.
+Contrary to the common motivation for removing rare ASVs, these results indicate that this practice disproportionately impacts samples with fewer sequences and likely removes more non-artifact ASVs than those that are artifacts.
 
@@
 
-**How do communities change when rare ASVs are removed?** Removing rare ASVs will reduce the number of observed ASVs and proportionally increase the relative abundance of the remaining ASVs.
+**The impact of removing rare ASVs on the information represented in each sample.** Removing rare ASVs will reduce the richness of ASVs and proportionally increase the relative abundance of the remaining ASVs.
 The result was expected to be a loss of information contained within each sample.
 To quantify the effect of removing rare ASVs on the information contained within each sample, I varied the minimum abundance threshold to simulate removing ASVs of varying rarity from each sample.
-The number of ASVs in each sample decreased by between `r oxford_comma(asv_s_two, 1)`% when removing those ASVs that only appeared once and by between `r oxford_comma(asv_s_eleven, 1)`% when removing those that appeared ten or fewer times from each sample (Figure 2A).
+The richness of ASVs in each sample (i.e.
+the number of ASVs) decreased by between `r oxford_comma(asv_s_two, 1)`% when removing those ASVs that only appeared once and by between `r oxford_comma(asv_s_eleven, 1)`% when removing those that appeared ten or fewer times from each sample (Figure 2A).
 Similarly, the Shannon diversity decreased by between `r oxford_comma(asv_h_two, 1)`% when removing ASVs that only appeared once and by between `r oxford_comma(asv_s_eleven, 1)`% when removing ASVs that appeared ten or fewer times from each sample (Figure 2B).
 Next, I assigned the ASVs to OTUs to assess the impact of removing rare ASVs on higher level taxonomic groupings that are commonly used in microbial ecology studies.
-Although pooling similar ASVs into OTUs reduced the impact of removing the rare ASVs relative to the ASV-based analysis, the minimum abundance threshold still decreased the number of observed OTUs and the diversity decreased relative to the original community (**Figure S2AB**).
-In contrast to the number of ASVs or OTUs and their diversity, the Kullback–Leibler divergence compares the relative abundance of each ASV or OTU between representations of the community.
+Although pooling similar ASVs into OTUs reduced the impact of removing the rare ASVs relative to the ASV-based analysis, the minimum abundance threshold still decreased the richness of OTUs and the diversity decreased relative to the full community (**Figure S2AB**).
+In contrast to the richness and diversity measurements, the Kullback–Leibler divergence compares the relative abundance of specific ASVs or OTUs between representations of the community.
 I calculated the Kullback–Leibler divergence from the full to pruned communities when rare ASVs were removed.
@@
 This effectively made every community in a study a statistical sample of the study-wide composite community distribution.
-For example, after this procedure, the 490 samples from the human dataset would be expected to have the same number of ASVs and diversity and one would not expect to find treatment-based effects between the samples.
+For example, after this procedure, the 490 samples from the human dataset would be expected to have the same richness and diversity of ASVs and one would not expect to find treatment-based effects between the samples.
 Because of the risk of bias if only one representation of the null distribution was generated, I generated 100 randomized datasets for each study.
-The trends between removing rare ASVs and the number of ASVs or OTUs, sample diversity, and information loss that were identified using the observed community community distribution data were also identified with the data from the null distribution; however, the losses were larger when using the null distribution data (**Figure S3**).
+The trends between removing rare ASVs and the richness, diversity, and information loss that were identified using the observed community community distribution data were also identified with the data from the null distribution; however, the losses were larger when using the null distribution data (**Figure S3**).
 The null distribution data were used in the remainder of the study to minimize the risk of bias.
@@
 
-**How do the relationships between communities change when they are screened by removing rare ASVs.** Considering the loss of the number of ASVs, diversity, and information when a community has its rarest ASVs removed, it seemed likely that the relationship between communities would be altered.
-To assess the impact of removing rare ASVs on measures of alpha diversity between samples I calculated the coefficients of variation (COV, i.e.
-standard deviation divided by the mean) for the number of ASVs and diversity for each study at multiple abundance thresholds.
-The COV for the number of ASVs across the studies after removing singletons were between `r oxford_comma(pc_cov_sobs, 1)`-times larger than they were without removing singleton ASVs (Figure 3A).
+**The impact of removing rare ASVs on the information represented between samples.** Considering the loss of richness, diversity, and information when a community has its rarest ASVs removed, it seemed likely that the relationship between communities would also be altered.
+To assess the impact of removing rare ASVs on measures of alpha diversity between samples I calculated the coefficients of variation (COVs, i.e.
+standard deviation divided by the mean) for richness and diversity for each study at multiple abundance thresholds.
+The COVs for richness of ASVs across the studies after removing singletons were between `r oxford_comma(pc_cov_sobs, 1)`-times larger than they were without removing singleton ASVs (Figure 3A).
 Similarly, the COVs for the diversity of ASVs were between `r oxford_comma(pc_cov_shannon, 1)`-times larger when singletons were removed than when they were not removed (Figure 3B).
-To assess the impact of removing rare ASVs on measures of beta diversity between samples I calculated the COV of the Bray-Curtis distances between samples within the same study at multiple abundance thresholds.
-The COV between Bray-Curtis distances within a study when singletons were removed was between `r oxford_comma(pc_cov_bc, 1)`-times larger than when they were not removed (Figure 3C).
-Increasing the minimum abundance threshold increased the COV between samples when using metrics of alpha and beta diversity.
-When ASVs were clustered into OTUs the difference in the COV was less than it was for the ASVs (Figure S4).
+To assess the impact of removing rare ASVs on measures of beta diversity between samples I calculated the COVs of the Bray-Curtis distances between samples within the same study at multiple abundance thresholds.
+The COVs between Bray-Curtis distances within a study when singletons were removed was between `r oxford_comma(pc_cov_bc, 1)`-times larger than when they were not removed (Figure 3C).
+Increasing the minimum abundance threshold increased the COVs between samples when using metrics of alpha and beta diversity.
+When ASVs were clustered into OTUs the difference in COVs was less than it was for the ASVs (Figure S4).
 These results indicate that removing rare ASVs increases the dissimilarity between samples, which could have a significant impact on the statistical power to detect differences between treatment groups.
@@
 					dataset=col_character(), clustering=col_character())) %>%
-	filter(dataset != "marine") %>%
+	# filter(dataset != "marine") %>%
 	filter(prune==11 | prune == 2 | prune ==1) %>%
@@
 
-**How does the increased inter-sample variation impact the ability to detect differences between samples?** To test the effect of increased inter-sample variation, I randomly assigned samples to one of two treatment groups.
+**The impact of removing rare ASVs on the ability to detect statistically significant differences between treatment groups.** To test the effect of increased inter-sample variation, I randomly assigned samples to one of two treatment groups.
 In the first treatment group, communities were randomly sampled from the null distribution as described above.
@@
 The fraction of significant tests was a measurement of the statistical power to detect the difference between the treatment groups.
-
-This doesn't seem to agree with the figure...
-Across all of the metrics I tested, the marine dataset yielded no simulated sets that were statistically significant, which was likely due to the small number of samples in the study (N=7); the marine dataset is not included in the following analysis.
-
-Among the remaining datasets, the power to detect a difference in the observed number of ASVs ranged between `r oxford_comma(b_base_pc_sobs, 2)` and between `r oxford_comma(b_base_pc_shannon, 2)` to detect a difference in diversity when using a Wilcox test (Figure 4A).
+When considering the differences in richness and diversity, the marine dataset yielded no simulated sets that were statistically significant, which was likely due to the small number of samples in the study (N=7).
+Among the remaining datasets, the power to detect a difference in the richness of ASVs ranged between `r oxford_comma(b_base_pc_sobs, 2)` and between `r oxford_comma(b_base_pc_shannon, 2)` to detect a difference in diversity when using a Wilcox test (Figure 4A).
 When singleton ASVs were removed, the power to detect a difference in the diversity of ASVs dropped by between `r oxford_comma(b_pc_sobs, 1)`% and by between `r oxford_comma(b_pc_shannon, 1)`% (Figure 4B).
-The effect of removing rare ASVs on the number of OTUs and their diversity was similar (Figure S5AB).
-A popular alternative approach calculates the dissimilarity between communities based on the abundance of each ASV or OTU.
-We used the Bray-Curtis dissimilarity index to compare the simulated communities within each dataset and calculated the power to detect differences between the two simulated treatment groups using the analysis of molecular variance (also called PERMANOVA) (Figure 4C and S5C).
+The effect of removing rare ASVs on the richness of OTUs and their diversity was similar (Figure S5AB).
+I used the Bray-Curtis dissimilarity index to compare the simulated communities within each dataset and calculated the power to detect differences between the two simulated treatment groups using the analysis of molecular variance (also called PERMANOVA) (Figure 4C and S5C).
 Without removing rare sequences, the power to detect a difference between the two simulated treatment groups varied between `r oxford_comma(b_base_beta, 2)`.
@@
 	summarize(p=mean(frac_sig), p_sd = sd(frac_sig), min_p=min(frac_sig)) %>%
-  ungroup()
+  ungroup() %>%
+  mutate(metric = "braycurtis") %>%
+  rename(method = clustering)
+
+min_ave_p <-
+  bind_rows(type_one_skew_alpha_summary, type_one_skew_beta_summary) %>%
+  filter(prune > 1) %>%
+  summarize(min_p = min(p)) %>%
+  pull(min_p)
 
@@
 
-**How does the removing of rare ASVs impact the probability of falsely detecting a difference between treatment groups?** Observing reduced ability to detect differences between communities when rare ASVs were removed from each sample, I next asked whether removing rare ASVs could lead to falsely claiming that a treatment effect had a significant effect on community diversity and structure.
-First, for each dataset, I sampled sequences from the null distribution and randomly assigned each sample to one of two treatment groups and determined the number of observed ASVs and OTUs and their Shannon diversity index.
+**The impact of removing rare ASVs on the probability of falsely detecting a difference between treatment groups.** Observing reduced ability to detect differences between communities when rare ASVs were removed from each sample, I next asked whether removing rare ASVs could lead to falsely claiming that a treatment effect had a significant effect on community diversity and structure.
+First, I sampled sequences from the null distribution for each dataset and randomly assigned each sample to one of two treatment groups and determined the richness and diversity of ASVs and OTUs.
 Testing at an experiment-wise error rate of 0.05, I expected 5% of the iterations for each dataset to yield a significant test result.
@@
 The samples with fewer than the median number of sequences for the dataset were assigned to one group and those with more than the median were assigned to the other.
-This exaggerated bias has been observed in comparisons of the lung and oral microbiota because of the larger nubmer of non-specific amplicons that can be sequenced from lung samples relative to those in the oral cavity [**REF**].
-When rare sequences were not removed, the fraction of significant tests did not differ from 5% for comparing the number of observed taxa (using ASVs or OTUs), their shannon diversity, or Bray-Curtis distances (Figure 5B and S6B).
+This exaggerated bias has been observed in comparisons of the lung and oral microbiota because of the larger number of non-specific amplicons that can be sequenced from lung samples relative to those in the oral cavity leading to a significant difference in sequencing depth between treatment groups [**REF**].
+When rare sequences were not removed, the fraction of significant tests did not differ from 5% for comparing the richness, their diversity, or Bray-Curtis distances (Figure 5B and S6B).
 However, when rare taxa of any frequency were removed, the probability of falsely detecing a difference as signifiant increased with the definition of rarity (Figure 5B and S6B).
-
-Across the datasets, the average fraction the average fraction of falsely detecting a difference was when only singletons were removed was greater than 90%.
-
+Not including the small marine dataset, the average fraction the average fraction of falsely detecting a difference across datasets when only singletons were removed was `r 100*min_ave_p`%.
 If there is any relationship between the number of sequences and the treatment group, the risk of falsely rejecting the null hypothesis is inflated when researchers use the strategy of removing rare sequences.

### 2020-10-27 submission/manuscript.tex
Edit Results section

@@
 \usepackage[T1]{fontenc}
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
 
 \begin{document}
-
-\subsection{word choice: sample
-vs.~community}\label{word-choice-sample-vs.community}
 
@@
 \subsection{Abstract}\label{abstract}
+
+\textbf{word choice: sample vs.~community}
 
@@
 abundance-based screening point to their ability to obtain the correct
-number of ASVs, OTUs, or phylogypes with data generated from sequencing
+number of ASVs, OTUs, or phylotypes with data generated from sequencing
 mock communities when rare ASVs are removed. However, this approach
@@
 
-\textbf{\emph{Results.}}
+\subsection{Results}\label{results}
 
@@
 standard data analysis pipeline in mothur results in a sequencing error
-rate below 0.02\% (\textbf{???}). All sequence data were obtained from
-the Sequence Read Archive and processed using a standard mothur-based
-sequencing pipeline that resulted in ASVs as generated by the
-pre.cluster algorithm (\textbf{???}, 1). After removing poor quality and
-chimeric ASVs and samples that had uncharacteristically low number of
-sequences for the dataset, these datasets included between 7.00 and
-490.00 samples. The median number of sequences for each dataset ranged
-between 6,477.00 and 193,464.00 (Table 1, Figure S1). Strikingly, aside
-from the relatively small marine and soil datasets, the difference
-between the sample with the fewest sequences and the sample with the
-most sequences for each dataset varied between 7.4 and 96.6-fold (Table
-1).
+rate below 0.02\% (1). All sequence data were obtained from the Sequence
+Read Archive and processed using a standard mothur-based sequencing
+pipeline that resulted in ASVs as generated by the pre.cluster algorithm
+(1, 2). After removing poor quality and chimeric ASVs and samples that
+had uncharacteristically low number of sequences for the dataset, these
+datasets included between 7 and 490 samples (Figure S1). The median
+number of sequences for each dataset ranged between 6,477 and 193,464
+(Table 1). Strikingly, aside from the relatively small marine and soil
+datasets, the difference between the sample with the fewest sequences
+and the sample with the most sequences for each dataset varied by
+between 7.4 and 96.6-fold (Table 1).
 
@@
 samples from the marine and sediment datasets (Spearman correlation,
-P\textgreater{}0.05), the fraction of singleton ASVs in the sample was
+P\textgreater{}0.05), the fraction of singleton ASVs in samples was
 negatively correlated with the number of sequences in each sample with a
-range between -0.27 and -0.87 (rice and bioethanol). This showed that
-with additional sequencing, the probability of seeing rare ASVs multiple
-times was greater than the probability of generating an artifact. This
-suggests that the rare ASVs are not as likely to be artifacts as
-previously thought. Furthermore, if rare ASVs were artifacts, then one
-would not expect to find them in other samples from the same dataset. In
-fact, singleton ASVs from samples with fewer sequences were often found
-in samples with more sequences. At least 50\% of the singleton ASVs
-found in the samples from the mice, rice, seagrass, and stream datasets
-were found in another sample from the same dataset. Considering the
-likelihood of finding an ASV duplicated in another sample is confounded
-by the number of samples and inter-sample diversity, the high coverage
-of singleton ASVs in these datasets was remarkable. The correlation
-between the number of sequences in a sample and the fraction of that
-sample's singleton ASVs that were covered by another sample in the
-dataset was significant and negative for 9 of the datasets ranging
-between -0.31 and -0.84 for the rice and seagrass datasets,
-respectively. The negative correlation indicated that the singleton ASVs
-in the smaller samples were more likely to be covered by ASVs in the
-larger samples. Among the three datasets without a significant
-correlation (Spearman correlation, P\textgreater{}0.05), the marine and
-soil datasets had the fewest samples in our collection and the stream
-dataset already had a high level of coverage regardless of the number of
-sequences. Contrary to the common motivation for removing rare ASVs,
-these results indicate that this practice disproportionately impacts
-samples with fewer sequences and likely removes a large number of
-non-artifact ASVs.
+range between -0.27 and -0.87 (rice and bioethanol) (Figure 1A). This
+showed that with additional sequencing, the probability of seeing
+singleton ASVs in multiple samples was greater than the probability of
+generating an artifact. This suggests that the singleton ASVs are not as
+likely to be artifacts as previously thought. Furthermore, if singleton
+ASVs were artifacts, then one would not expect to find them in other
+samples from the same dataset. In fact, singleton ASVs from samples with
+fewer sequences were often found in samples with more sequences. At
+least 50\% of the singleton ASVs found in the samples from the mice,
+rice, seagrass, and stream datasets were found in another sample from
+the same dataset (Figure 1B). Considering the likelihood of finding an
+ASV duplicated in another sample is confounded by the number of samples
+and inter-sample diversity, the high coverage of singleton ASVs in these
+datasets was remarkable. The correlation between the number of sequences
+in a sample and the fraction of that sample's singleton ASVs that were
+covered by another sample in the dataset was significant and negative
+for 9 of the datasets ranging between -0.31 and -0.84 for the rice and
+seagrass datasets, respectively (Figure 1C). The negative correlation
+indicated that the singleton ASVs in the smaller samples were more
+likely to be covered by ASVs in the larger samples. Among the three
+datasets without a significant correlation (Spearman correlation,
+P\textgreater{}0.05), the marine and soil datasets had the fewest
+samples in our collection and the stream dataset already had a high
+level of coverage regardless of the number of sequences. Contrary to the
+common motivation for removing rare ASVs, these results indicate that
+this practice disproportionately impacts samples with fewer sequences
+and likely removes more non-artifact ASVs than those that are artifacts.
 
-\textbf{How do communities change when rare ASVs are removed?} Removing
-rare ASVs will reduce the number of observed ASVs and proportionally
-increase the relative abundance of the remaining ASVs. The result was
-expected to be a loss of information contained within each sample. To
-quantify the effect of removing rare ASVs on the information contained
-within each sample, I varied the minimum abundance threshold to simulate
-removing ASVs of varying rarity from each sample. The number of ASVs in
-each sample decreased by between 34.4 and 86.2\% when removing those
-ASVs that only appeared once and by between 76.0 and 95.6\% when
-removing those that appeared ten or fewer times from each sample (Figure
-1). Similarly, the Shannon diversity decreased by between 1.8 and 15.9\%
-when removing ASVs that only appeared once and by between 76.0 and
-95.6\% when removing ASVs that appeared ten or fewer times from each
-sample (Figure 1). Next, I assigned the ASVs to OTUs to assess the
-impact of removing rare ASVs on higher level taxonomic groupings that
-are commonly used in microbial ecology studies. Although pooling similar
-ASVs into OTUs reduced the impact of removing the rare ASVs relative to
-the ASV-based analysis, the minimum abundance threshold still decreased
-the number of observed OTUs and the diversity decreased relative to the
-original community (\textbf{Figure 1}). In contrast to the number of
-ASVs or OTUs and their diversity, the Kullback--Leibler divergence
-compares the relative abundance of each ASV or OTU between
-representations of the community. I calculated the Kullback--Leibler
-divergence from the full to pruned communities when rare ASVs were
-removed. As the threshold for removing ASVs increased, the amount of
-information lost also increased for both ASVs and OTUs (Figure 1). The
-relative loss of information was generally lower for OTUs than than it
-was for ASVs. Removing rare ASVs, regardless of abundance threshold, had
-profound impacts on the representation of the communities.
+\textbf{The impact of removing rare ASVs on the information represented
+in each sample.} Removing rare ASVs will reduce the richness of ASVs and
+proportionally increase the relative abundance of the remaining ASVs.
+The result was expected to be a loss of information contained within
+each sample. To quantify the effect of removing rare ASVs on the
+information contained within each sample, I varied the minimum abundance
+threshold to simulate removing ASVs of varying rarity from each sample.
+The richness of ASVs in each sample (i.e.~the number of ASVs) decreased
+by between 34.4 and 86.2\% when removing those ASVs that only appeared
+once and by between 76.0 and 95.6\% when removing those that appeared
+ten or fewer times from each sample (Figure 2A). Similarly, the Shannon
+diversity decreased by between 1.8 and 15.9\% when removing ASVs that
+only appeared once and by between 76.0 and 95.6\% when removing ASVs
+that appeared ten or fewer times from each sample (Figure 2B). Next, I
+assigned the ASVs to OTUs to assess the impact of removing rare ASVs on
+higher level taxonomic groupings that are commonly used in microbial
+ecology studies. Although pooling similar ASVs into OTUs reduced the
+impact of removing the rare ASVs relative to the ASV-based analysis, the
+minimum abundance threshold still decreased the richness of OTUs and the
+diversity decreased relative to the full community (\textbf{Figure
+S2AB}). In contrast to the richness and diversity measurements, the
+Kullback--Leibler divergence compares the relative abundance of specific
+ASVs or OTUs between representations of the community. I calculated the
+Kullback--Leibler divergence from the full to pruned communities when
+rare ASVs were removed. As the threshold for removing ASVs increased,
+the amount of information lost also increased for both ASVs and OTUs
+(Figure 2C and Figure S2C). The relative loss of information was
+generally lower for OTUs than than it was for ASVs. Removing rare ASVs,
+regardless of abundance threshold, had profound impacts on the
+representation of the communities.
 
@@
 distribution. For example, after this procedure, the 490 samples from
-the human dataset would be expected to have the same number of ASVs and
-diversity and one would not expect to find treatment-based effects
-between the samples. Because of the risk of bias if only one
+the human dataset would be expected to have the same richness and
+diversity of ASVs and one would not expect to find treatment-based
+effects between the samples. Because of the risk of bias if only one
 representation of the null distribution was generated, I generated 100
-randomized datasets for each study. The same trends between removing
-rare ASVs and the number of ASVs, sample diversity, and information loss
-that were identified using the observed community community distribution
-data were also identified with the data from the null distribution
-(\textbf{Figure S2}). The null distribution data were used in the
-remainder of the study.
+randomized datasets for each study. The trends between removing rare
+ASVs and the richness, diversity, and information loss that were
+identified using the observed community community distribution data were
+also identified with the data from the null distribution; however, the
+losses were larger when using the null distribution data (\textbf{Figure
+S3}). The null distribution data were used in the remainder of the study
+to minimize the risk of bias.
 
-\textbf{How do the relationships between communities change when they
-are screened by removing rare ASVs.} Considering the loss of the number
-of ASVs, diversity, and information when a community has its rarest ASVs
-removed, it seemed likely that the relationship between communities
-would be altered. To assess the impact of removing rare ASVs on measures
-of alpha diversity between samples I calculated the coefficients of
-variation (COV, i.e.~standard deviation divided by the mean) for the
-number of ASVs and diversity for each study at multiple abundance
-thresholds. The COV for the number of ASVs across the studies after
-removing singletons were between 3.6 and 32.7-times larger than they
-were without removing singleton ASVs. Similarly, the COVs for the
-diversity of ASVs were between 1.8 and 20.4-times larger when singletons
-were removed than when they were not removed.To assess the impact of
+\textbf{The impact of removing rare ASVs on the information represented
+between samples.} Considering the loss of richness, diversity, and
+information when a community has its rarest ASVs removed, it seemed
+likely that the relationship between communities would also be altered.
+To assess the impact of removing rare ASVs on measures of alpha
+diversity between samples I calculated the coefficients of variation
+(COVs, i.e.~standard deviation divided by the mean) for richness and
+diversity for each study at multiple abundance thresholds. The COVs for
+richness of ASVs across the studies after removing singletons were
+between 3.6 and 32.7-times larger than they were without removing
+singleton ASVs (Figure 3A). Similarly, the COVs for the diversity of
+ASVs were between 1.8 and 20.4-times larger when singletons were removed
+than when they were not removed (Figure 3B). To assess the impact of
 removing rare ASVs on measures of beta diversity between samples I
-calculated the COV of the Bray-Curtis distances between samples within
-the same study at multiple abundance thresholds. The COV between
+calculated the COVs of the Bray-Curtis distances between samples within
+the same study at multiple abundance thresholds. The COVs between
 Bray-Curtis distances within a study when singletons were removed was
-between 1.3 and 18.6-times larger than when they were not removed.
-Increasing the minimum abundance threshold increased the COV between
-samples when using metrics of alpha and beta diversity. When ASVs were
-clustered into OTUs the difference in the COV was less than it was for
-the ASVs (\textbf{Figure 2}). These results indicate that removing rare
-ASVs increases the dissimilarity between samples, which could have a
-significant impact on the statistical power to detect differences
+between 1.3 and 18.6-times larger than when they were not removed
+(Figure 3C). Increasing the minimum abundance threshold increased the
+COVs between samples when using metrics of alpha and beta diversity.
+When ASVs were clustered into OTUs the difference in COVs was less than
+it was for the ASVs (Figure S4). These results indicate that removing
+rare ASVs increases the dissimilarity between samples, which could have
+a significant impact on the statistical power to detect differences
 between treatment groups.
 
-\textbf{How does the increased inter-sample variation impact the ability
-to detect differences between samples?} To test the effect of increased
-inter-sample variation, I randomly assigned samples to one of two
-treatment groups. In the first treatment group, communities were
-randomly sampled from the null distribution as described above. For the
-second treatment group, I increased the abundance of 10\% of the ASVs in
-the pooled study distribution by 5\%. I randomly generated 100 simulated
-sets of treatment groups and samples. I then tested the ability to
-detect a difference between the two treatment groups using alpha and
-beta diversity metrics. The fraction of significant tests was a
-measurement of the statistical power to detect the difference between
-the treatment groups. Across all of the metrics I tested, the marine
-dataset yielded no simulated sets that were statistically significant,
-which was likely due to the small number of samples in the study (N=7);
-the marine dataset is not included in the following analysis. Among the
-remaining datasets, the power to detect a difference in the observed
-number of ASVs ranged between 0.10 and 0.49 and between 0.10 and 0.53 to
-detect a difference in diversity when using a Wilcox test
-(\textbf{Figure X}). When singleton ASVs were removed, the power to
-detect a difference in the number of ASVs dropped by between 27.3 and
-92.9\% and by between 40.0 and 93.3\% to detect a difference in
-diversity. The effect of removing rare ASVs on the number of OTUs and
-their diversity was similar. A popular alternative approach calculates
-the dissimilarity between communities based on the abundance of each ASV
-or OTU. We used the Bray-Curtis dissimilarity index to compare the
-simulated communities within each dataset and calculated the power to
-detect differences between the two simulated treatment groups using the
-analysis of molecular variance (also called PERMANOVA). Without removing
-rare sequences, the power to detect a difference between the two
-simulated treatment groups varied between 0.41 and 1.00. Aside from 3
-datasets, the power to detect differences dropped by between 6.5 and
+\textbf{The impact of removing rare ASVs on the ability to detect
+statistically significant differences between treatment groups.} To test
+the effect of increased inter-sample variation, I randomly assigned
+samples to one of two treatment groups. In the first treatment group,
+communities were randomly sampled from the null distribution as
+described above. For the second treatment group, I increased the
+abundance of 10\% of the ASVs in the pooled study distribution by 5\%. I
+randomly generated 100 simulated sets of treatment groups and samples. I
+then tested the ability to detect a difference between the two treatment
+groups using alpha and beta diversity metrics. The fraction of
+significant tests was a measurement of the statistical power to detect
+the difference between the treatment groups. When considering the
+differences in richness and diversity, the marine dataset yielded no
+simulated sets that were statistically significant, which was likely due
+to the small number of samples in the study (N=7). Among the remaining
+datasets, the power to detect a difference in the richness of ASVs
+ranged between 0.10 and 0.49 and between 0.10 and 0.53 to detect a
+difference in diversity when using a Wilcox test (Figure 4A). When
+singleton ASVs were removed, the power to detect a difference in the
+diversity of ASVs dropped by between 27.3 and 92.9\% and by between 40.0
+and 93.3\% (Figure 4B). The effect of removing rare ASVs on the richness
+of OTUs and their diversity was similar (Figure S5AB). I used the
+Bray-Curtis dissimilarity index to compare the simulated communities
+within each dataset and calculated the power to detect differences
+between the two simulated treatment groups using the analysis of
+molecular variance (also called PERMANOVA) (Figure 4C and S5C). Without
+removing rare sequences, the power to detect a difference between the
+two simulated treatment groups varied between 0.41 and 1.00. Aside from
+3 datasets, the power to detect differences dropped by between 6.5 and
 64.0\% when singletons were removed. However, when ASVs that occurred 10
@@
 
-\textbf{How does the removing of rare ASVs impact the probability of
-falsely detecting a difference between treatment groups?} Observing
-reduced ability to detect differences between communities when rare ASVs
-were removed from each sample, I next asked whether removing rare ASVs
-could lead to falsely claiming that a treatment effect had a significant
-effect on community diversity and structure. First, for each dataset, I
-sampled sequences from the null distribution and randomly assigned each
-sample to one of two treatment groups and determined the number of
-observed ASVs and OTUs and their Shannon diversity index. Testing at an
-experiment-wise error rate of 0.05, I expected 5\% of the iterations for
-each dataset to yield a significant test result. Indeed, there was no
-evidence that removing rare ASVs resulted in an inflated experiment-wise
-error rate. The average fraction of significant tests did not
-meaningfully vary from 0.05 across the minimum abundance threshold,
-dataset, metric of describing sample alpha-diversity, or whether the
-abundance of ASVs or OTUs were used (\textbf{Figure XA}). Similarly, the
+\textbf{The impact of removing rare ASVs on the probability of falsely
+detecting a difference between treatment groups.} Observing reduced
+ability to detect differences between communities when rare ASVs were
+removed from each sample, I next asked whether removing rare ASVs could
+lead to falsely claiming that a treatment effect had a significant
+effect on community diversity and structure. First, I sampled sequences
+from the null distribution for each dataset and randomly assigned each
+sample to one of two treatment groups and determined the richness and
+diversity of ASVs and OTUs. Testing at an experiment-wise error rate of
+0.05, I expected 5\% of the iterations for each dataset to yield a
+significant test result. Indeed, there was no evidence that removing
+rare ASVs resulted in an inflated experiment-wise error rate. The
 average fraction of significant tests did not meaningfully vary from
-0.05 when using analysis of molecular variance to compare communities
-using Bray-Curtis distances. Second, I again sampled sequences from the
-null distribution, but assigned samples to one of two treatment groups
-based on the number of sequences in each sample. The samples with fewer
-than the median number of sequences for the dataset were assigned to one
-group and those with more than the median were assigned to the other.
-This exaggerated bias has been observed in comparisons of the lung and
-oral microbiota because of the larger nubmer of non-specific amplicons
-that can be sequenced from lung samples relative to those in the oral
-cavity {[}\textbf{REF}{]}. When rare sequences were not removed, the
-fraction of significant tests did not differ from 5\% for comparing the
-number of observed taxa (using ASVs or OTUs), their shannon diversity,
-or Bray-Curtis distances. However, when rare taxa of any frequency were
+0.05 across the minimum abundance threshold, dataset, metric of
+describing sample alpha-diversity, or whether the abundance of ASVs or
+OTUs were used (Figure 5A and S6A). Similarly, the average fraction of
+significant tests did not meaningfully vary from 0.05 when using
+analysis of molecular variance to compare communities using Bray-Curtis
+distances (Figure 5A and S6A). Second, I again sampled sequences from
+the null distribution, but assigned samples to one of two treatment
+groups based on the number of sequences in each sample. The samples with
+fewer than the median number of sequences for the dataset were assigned
+to one group and those with more than the median were assigned to the
+other. This exaggerated bias has been observed in comparisons of the
+lung and oral microbiota because of the larger number of non-specific
+amplicons that can be sequenced from lung samples relative to those in
+the oral cavity leading to a significant difference in sequencing depth
+between treatment groups {[}\textbf{REF}{]}. When rare sequences were
+not removed, the fraction of significant tests did not differ from 5\%
+for comparing the richness, their diversity, or Bray-Curtis distances
+(Figure 5B and S6B). However, when rare taxa of any frequency were
 removed, the probability of falsely detecing a difference as signifiant
-increased with the definition of rarity (\textbf{Figure XB}). Across the
-datasets, the average fraction the average fraction of falsely detecting
-a difference was when only singletons were removed was greater than
-90\%. If there is any relationship between the number of sequences and
-the treatment group, the risk of falsely rejecting the null hypothesis
-is inflated when researchers use the strategy of removing rare
-sequences. The most conservative approach is to not remove low abundance
-sequences.
+increased with the definition of rarity (Figure 5B and S6B). Not
+including the small marine dataset, the average fraction the average
+fraction of falsely detecting a difference across datasets when only
+singletons were removed was 92.45\%. If there is any relationship
+between the number of sequences and the treatment group, the risk of
+falsely rejecting the null hypothesis is inflated when researchers use
+the strategy of removing rare sequences. The most conservative approach
+is to not remove low abundance sequences.
 
@@
 
-1. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+1. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
+\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
+dual-index sequencing strategy and curation pipeline for analyzing
+amplicon sequence data on the MiSeq Illumina sequencing platform.
+Applied and environmental microbiology \textbf{79}:5112--5120.
+
+2. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},

### 2020-10-27 submission/manuscript.Rmd
Edit Results section

@@
 }
-
 ```
@@
 
-## Abstract
-
-**word choice: sample vs. community**
-
-250 words
-
-\newpage
-
-## Importance
-150 words
-
-\newpage
-
 ```{r datasets}
@@
 ```
+
+## Abstract
+
+**word choice: sample vs. community**
+
+250 words
+
+\newpage
+
+## Importance
+150 words
+
+\newpage
 
@@
 
-asv_s_two <- ointra %>% filter(min_class == 2, method== "pc") %>% arrange(mean_s_frac)	%>% slice(1, nrow(.)) %>% pull(mean_s_frac)
+asv_s_two <- ointra %>% filter(min_class == 2, method== "pc") %>% arrange(mean_s_frac)	%>% slice(1, nrow(.))
 
-asv_s_eleven <- ointra %>% filter(min_class == 11, method== "pc") %>% arrange(mean_s_frac)	%>% slice(1, nrow(.)) %>% pull(mean_s_frac)
+asv_s_eleven <- ointra %>% filter(min_class == 11, method== "pc") %>% arrange(mean_s_frac)	%>% slice(1, nrow(.))
 
-asv_h_two <- ointra %>% filter(min_class == 2, method== "pc") %>% arrange(mean_h_frac)	%>% slice(1, nrow(.)) %>% pull(mean_h_frac)
+asv_h_two <- ointra %>% filter(min_class == 2, method== "pc") %>% arrange(mean_h_frac)	%>% slice(1, nrow(.))
 
-asv_h_eleven <- ointra %>% filter(min_class == 11, method== "pc") %>% arrange(mean_h_frac)	%>% slice(1, nrow(.)) %>% pull(mean_h_frac)
+asv_h_eleven <- ointra %>% filter(min_class == 11, method== "pc") %>% arrange(mean_h_frac)	%>% slice(1, nrow(.))
 ```
 
-**The impact of removing rare ASVs on the information represented in each sample.** Removing rare ASVs will reduce the richness of ASVs and proportionally increase the relative abundance of the remaining ASVs.
-The result was expected to be a loss of information contained within each sample.
+**The impact of removing rare ASVs on the information represented in each sample.** Removing rare ASVs will reduce the richness of ASVs (i.e.
+the number of ASVs per sample) and increase the relative abundance of the remaining ASVs.
 To quantify the effect of removing rare ASVs on the information contained within each sample, I varied the minimum abundance threshold to simulate removing ASVs of varying rarity from each sample.
-The richness of ASVs in each sample (i.e.
-the number of ASVs) decreased by between `r oxford_comma(asv_s_two, 1)`% when removing those ASVs that only appeared once and by between `r oxford_comma(asv_s_eleven, 1)`% when removing those that appeared ten or fewer times from each sample (Figure 2A).
-Similarly, the Shannon diversity decreased by between `r oxford_comma(asv_h_two, 1)`% when removing ASVs that only appeared once and by between `r oxford_comma(asv_s_eleven, 1)`% when removing ASVs that appeared ten or fewer times from each sample (Figure 2B).
-Next, I assigned the ASVs to OTUs to assess the impact of removing rare ASVs on higher level taxonomic groupings that are commonly used in microbial ecology studies.
-Although pooling similar ASVs into OTUs reduced the impact of removing the rare ASVs relative to the ASV-based analysis, the minimum abundance threshold still decreased the richness of OTUs and the diversity decreased relative to the full community (**Figure S2AB**).
+The richness of ASVs in each sample decreased by between `r oxford_comma(pull(asv_s_two, mean_s_frac), 1)`% (`r oxford_comma(pull(asv_s_two, dataset))`) when removing those ASVs that only appeared once and by between `r oxford_comma(pull(asv_s_eleven, mean_s_frac), 1)`% (`r oxford_comma(pull(asv_s_eleven, dataset))`) when removing those that appeared ten or fewer times from each sample (Figure 2A).
+Similarly, the Shannon diversity decreased by between `r oxford_comma(pull(asv_h_two, mean_h_frac), 1)`% (`r oxford_comma(pull(asv_h_two, dataset))`) when removing ASVs that only appeared once and by between `r oxford_comma(pull(asv_h_eleven, mean_h_frac), 1)`% (`r oxford_comma(pull(asv_h_eleven, dataset))`) when removing ASVs that appeared ten or fewer times from each sample (Figure 2B).
+Next, I assigned the ASVs to OTUs, which were defined as a group of ASVs that were more than 97% similar to each other to assess the impact of removing rare ASVs on higher level taxonomic groupings that are commonly used in microbial ecology studies.
+Although pooling similar ASVs into OTUs reduced the impact of removing the rare ASVs relative to the ASV-based analysis, the minimum abundance threshold still decreased the richness of OTUs and the diversity decreased relative to the full community (Figure S2AB).
 In contrast to the richness and diversity measurements, the Kullback–Leibler divergence compares the relative abundance of specific ASVs or OTUs between representations of the community.
-I calculated the Kullback–Leibler divergence from the full to pruned communities when rare ASVs were removed.
+I calculated the Kullback–Leibler divergence between the full communities and those where rare ASVs were removed.
 As the threshold for removing ASVs increased, the amount of information lost also increased for both ASVs and OTUs (Figure 2C and Figure S2C).
-The relative loss of information was generally lower for OTUs than than it was for ASVs.
+The relative loss of information was generally smaller for OTUs than than it was for ASVs.
 Removing rare ASVs, regardless of abundance threshold, had profound impacts on the representation of the communities.
 
-**Removing treatment group effects from community data.** Because treatment effects often affect a sample's diversity and inter-sample variation, I generated null distributions for each study by randomizing the number of times each ASV was observed in each sample such that the total number of sequences in each sample and the total number of times each ASV was observed across all samples in the study was the same as was originally observed.
+**Removing treatment group effects from community data.** Because treatment effects often affect a sample's diversity and inter-sample variation, I generated null distributions for each study by randomizing, without replacement, the number of times each ASV was observed in each sample such that the total number of sequences in each sample and the total number of times each ASV was observed across all samples in the study was the same as was originally observed.
 This effectively made every community in a study a statistical sample of the study-wide composite community distribution.
@@
 Because of the risk of bias if only one representation of the null distribution was generated, I generated 100 randomized datasets for each study.
-The trends between removing rare ASVs and the richness, diversity, and information loss that were identified using the observed community community distribution data were also identified with the data from the null distribution; however, the losses were larger when using the null distribution data (**Figure S3**).
+The trends between removing rare ASVs and the richness, diversity, and information loss that were identified using the observed community community distribution data were also identified with the data from the null distribution; however, the losses were larger when using the null distribution data (Figure S3).
 The null distribution data were used in the remainder of the study to minimize the risk of bias.
@@
 pc_cov_alpha <- r_cov_alpha %>% filter(method == "pc")
-pc_cov_sobs <- pc_cov_alpha %>% arrange(cv_sobs) %>% slice(1, nrow(.)) %>% pull(cv_sobs)
-pc_cov_shannon <- pc_cov_alpha %>% arrange(cv_shannon) %>% slice(1, nrow(.)) %>% pull(cv_shannon)
-
-otu_cov_alpha <- r_cov_alpha %>% filter(method == "otu")
-otu_cov_sobs <- otu_cov_alpha %>% arrange(cv_sobs) %>% slice(1, nrow(.)) %>% pull(cv_sobs)
-otu_cov_shannon <- otu_cov_alpha %>% arrange(cv_shannon) %>% slice(1, nrow(.)) %>% pull(cv_shannon)
-
-pc_v_otu_alpha <- r_cov_alpha %>% pivot_wider(dataset, names_from=method, values_from=c(cv_sobs, cv_shannon)) %>% mutate(sobs = 100*(1-cv_sobs_otu/cv_sobs_pc), shannon = 100*(1-cv_shannon_otu/cv_shannon_pc))
-
-pc_v_otu_sobs <- range(pc_v_otu_alpha$sobs)
-pc_v_otu_shannon <- range(pc_v_otu_alpha$shannon)
+pc_cov_sobs <- pc_cov_alpha %>% arrange(cv_sobs) %>% slice(1, nrow(.))
+pc_cov_shannon <- pc_cov_alpha %>% arrange(cv_shannon) %>% slice(1, nrow(.))
 
@@
 
-pc_cov_bc <- r_cov_bc %>% filter(method == "pc") %>% arrange(cv_diff) %>% slice(1, nrow(.)) %>% pull(cv_diff)
-
-otu_cov_bc <- r_cov_bc %>% filter(method == "otu") %>% arrange(cv_diff) %>% slice(1, nrow(.)) %>% pull(cv_diff)
+pc_cov_bc <- r_cov_bc %>% filter(method == "pc") %>% arrange(cv_diff) %>% slice(1, nrow(.))
+otu_cov_bc <- r_cov_bc %>% filter(method == "otu") %>% arrange(cv_diff) %>% slice(1, nrow(.))
 ```
@@
 To assess the impact of removing rare ASVs on measures of alpha diversity between samples I calculated the coefficients of variation (COVs, i.e.
-standard deviation divided by the mean) for richness and diversity for each study at multiple abundance thresholds.
-The COVs for richness of ASVs across the studies after removing singletons were between `r oxford_comma(pc_cov_sobs, 1)`-times larger than they were without removing singleton ASVs (Figure 3A).
-Similarly, the COVs for the diversity of ASVs were between `r oxford_comma(pc_cov_shannon, 1)`-times larger when singletons were removed than when they were not removed (Figure 3B).
-To assess the impact of removing rare ASVs on measures of beta diversity between samples I calculated the COVs of the Bray-Curtis distances between samples within the same study at multiple abundance thresholds.
-The COVs between Bray-Curtis distances within a study when singletons were removed was between `r oxford_comma(pc_cov_bc, 1)`-times larger than when they were not removed (Figure 3C).
-Increasing the minimum abundance threshold increased the COVs between samples when using metrics of alpha and beta diversity.
+the standard deviation divided by the mean) for richness and diversity for each study at multiple abundance thresholds.
+The COVs for the richness of ASVs across the studies after removing singletons were between `r oxford_comma(pull(pc_cov_sobs, cv_sobs), 1)`-times larger than they were without removing singleton ASVs (`r oxford_comma(pull(pc_cov_sobs, dataset))`; Figure 3A).
+Similarly, the COVs for the diversity of ASVs were between `r oxford_comma(pull(pc_cov_shannon,cv_shannon), 1)`-times larger when singletons were removed than when they were not removed (`r oxford_comma(pull(pc_cov_shannon, dataset))`; Figure 3B).
+To assess the impact of removing rare ASVs on measures of beta diversity between samples, I calculated the COVs of the Bray-Curtis distances between samples within the same study at multiple abundance thresholds.
+The COVs between Bray-Curtis distances within a study when singletons were removed was between `r oxford_comma(pull(pc_cov_bc, cv_diff), 1)`-times larger than when they were not removed (`r oxford_comma(pull(pc_cov_bc, dataset))`; Figure 3C).
 When ASVs were clustered into OTUs the difference in COVs was less than it was for the ASVs (Figure S4).
@@
 
-b_base_pc_sobs <- b_alpha %>% filter(metric == "sobs", method == "pc") %>% select(`1`) %>% range()
-b_base_pc_shannon <- b_alpha %>% filter(metric == "shannon", method == "pc") %>% select(`1`) %>% range()
+b_base_pc_sobs <- b_alpha %>% filter(metric == "sobs", method == "pc") %>% arrange(`1`) %>% slice(1, nrow(.))
+b_base_pc_shannon <- b_alpha %>% filter(metric == "shannon", method == "pc") %>% arrange(`1`) %>% slice(1, nrow(.))
 
-b_base_otu_sobs <- b_alpha %>% filter(metric == "sobs", method == "otu") %>% select(`1`) %>% range()
-b_base_otu_shannon <- b_alpha %>% filter(metric == "shannon", method == "otu") %>% select(`1`) %>% range()
-
-b_pc_sobs <- b_alpha %>% filter(metric == "sobs", method == "pc") %>% slice(1, n()) %>% pull(percentage_drop)
-
-b_pc_shannon <- b_alpha %>% filter(metric == "shannon", method == "pc") %>% slice(1, n()) %>% pull(percentage_drop)
+b_pc_sobs <- b_alpha %>% filter(metric == "sobs", method == "pc") %>% slice(1, n())
+b_pc_shannon <- b_alpha %>% filter(metric == "shannon", method == "pc") %>% slice(1, n())
 
@@
 
-b_base_beta <- b_beta %>% filter(clustering == "pc") %>% select(`1`) %>% range()
+b_base_beta <- b_beta %>% filter(clustering == "pc") %>% arrange(`1`)
 
-n_unfazed_singletons <- b_beta %>% filter(clustering=="pc" & p_drop_singletons <= 0) %>% nrow()
+unfazed_singletons <- b_beta %>% filter(clustering=="pc" & p_drop_singletons <= 0) %>% arrange(dataset) %>% pull(dataset)
 
-fazed_singletons <- b_beta %>% filter(clustering=="pc" & p_drop_singletons > 0) %>% select(p_drop_singletons) %>% range()
+fazed_singletons <- b_beta %>% filter(clustering=="pc" & p_drop_singletons > 0) %>% arrange(p_drop_singletons) %>% slice(1, nrow(.))
 
@@
 
-fazed_tentons <- b_beta %>% filter(clustering=="pc" & p_drop_tens > 10) %>% select(p_drop_tens) %>% range()
+fazed_tentons <- b_beta %>% filter(clustering=="pc" & p_drop_tens > 10) %>% arrange(p_drop_tens) %>% slice(1, nrow(.))
 ```
@@
 In the first treatment group, communities were randomly sampled from the null distribution as described above.
-For the second treatment group, I increased the abundance of 10% of the ASVs in the pooled study distribution by 5%.
+For the second treatment group, I randomly selected 10% of the ASVs in the pooled study distribution to increase their abundance by 5%.
 I randomly generated 100 simulated sets of treatment groups and samples.
@@
 When considering the differences in richness and diversity, the marine dataset yielded no simulated sets that were statistically significant, which was likely due to the small number of samples in the study (N=7).
-Among the remaining datasets, the power to detect a difference in the richness of ASVs ranged between `r oxford_comma(b_base_pc_sobs, 2)` and between `r oxford_comma(b_base_pc_shannon, 2)` to detect a difference in diversity when using a Wilcox test (Figure 4A).
-When singleton ASVs were removed, the power to detect a difference in the diversity of ASVs dropped by between `r oxford_comma(b_pc_sobs, 1)`% and by between `r oxford_comma(b_pc_shannon, 1)`% (Figure 4B).
+Among the remaining datasets, the power to detect a difference in the richness of ASVs ranged between `r oxford_comma(pull(b_base_pc_sobs,'1'), 2)` (`r oxford_comma(pull(b_base_pc_sobs,dataset))`) and between `r oxford_comma(pull(b_base_pc_shannon, '1'), 2)` (`r oxford_comma(pull(b_base_pc_shannon, dataset))`) to detect a difference in diversity when using a Wilcox test (Figure 4A).
+When singleton ASVs were removed, the power to detect a difference in the diversity of ASVs dropped by between `r oxford_comma(pull(b_pc_sobs, percentage_drop), 1)`% (`r oxford_comma(pull(b_pc_sobs, dataset))`) and by between `r oxford_comma(pull(b_pc_shannon, percentage_drop), 1)`% (`r oxford_comma(pull(b_pc_shannon, dataset))`; Figure 4B).
 The effect of removing rare ASVs on the richness of OTUs and their diversity was similar (Figure S5AB).
 I used the Bray-Curtis dissimilarity index to compare the simulated communities within each dataset and calculated the power to detect differences between the two simulated treatment groups using the analysis of molecular variance (also called PERMANOVA) (Figure 4C and S5C).
-Without removing rare sequences, the power to detect a difference between the two simulated treatment groups varied between `r oxford_comma(b_base_beta, 2)`.
-Aside from `r n_unfazed_singletons` datasets, the power to detect differences dropped by between `r oxford_comma(fazed_singletons, 1)`% when singletons were removed.
-However, when ASVs that occurred 10 or fewer times were removed from each sample, the power to detect differences dropped by `r oxford_comma(fazed_tentons, 1)`%; similar results were observed when ASVs were clustered into OTUs.
+Without removing rare sequences, the power to detect a difference between the two simulated treatment groups varied between `r oxford_comma(pull(b_base_beta, '1') %>% range(), 2)` (`r slice(b_base_beta, 1) %>% pull(dataset)` and `r oxford_comma(slice_max(b_base_beta, '1') %>% pull(dataset))`).
+Aside from the `r oxford_comma(unfazed_singletons)` datasets, the power to detect differences dropped by between `r oxford_comma(pull(fazed_singletons,p_drop_singletons), 1)`% (`r oxford_comma(pull(fazed_singletons, dataset))`) when singletons were removed.
+However, when ASVs that occurred 10 or fewer times were removed from each sample, the power to detect differences dropped by `r oxford_comma(pull(fazed_tentons, p_drop_tens), 1)`% (`r oxford_comma(pull(fazed_tentons, dataset))`); similar results were observed when ASVs were clustered into OTUs.
 Removing rare ASVs reduced the ability to detect simulated treatment effects using metrics commonly used to compare microbial communities.
@@
 
-**The impact of removing rare ASVs on the probability of falsely detecting a difference between treatment groups.** Observing reduced ability to detect differences between communities when rare ASVs were removed from each sample, I next asked whether removing rare ASVs could lead to falsely claiming that a treatment effect had a significant effect on community diversity and structure.
+**The impact of removing rare ASVs on the probability of falsely detecting a difference between treatment groups.** I next asked whether removing rare ASVs could lead to falsely claiming that a treatment effect had a significant effect on community diversity and structure.
 First, I sampled sequences from the null distribution for each dataset and randomly assigned each sample to one of two treatment groups and determined the richness and diversity of ASVs and OTUs.
@@
 The samples with fewer than the median number of sequences for the dataset were assigned to one group and those with more than the median were assigned to the other.
-This exaggerated bias has been observed in comparisons of the lung and oral microbiota because of the larger number of non-specific amplicons that can be sequenced from lung samples relative to those in the oral cavity leading to a significant difference in sequencing depth between treatment groups [**REF**].
+This exaggerated bias has been observed in comparisons of the lung and oral microbiota because of the larger number of non-specific amplicons that can be sequenced from lung samples relative to those in the oral cavity leading to a significant difference in sequencing depth between treatment groups [].
 When rare sequences were not removed, the fraction of significant tests did not differ from 5% for comparing the richness, their diversity, or Bray-Curtis distances (Figure 5B and S6B).
@@
 
-***Conclusion.*** Removing rare sequences decreases the diversity represented by 16S rRNA gene sequence data and increases the variation between samples.
+## Discussion
+
+Removing rare sequences decreases the diversity represented by 16S rRNA gene sequence data and increases the variation between samples.
 Such impacts will hinder the statistical power to differentiate between treatment groups.
@@
 
-## Acknowledgements
-
-\newpage
-
 ## Materials and Methods
@@
 
-\newpage
-
-## References
-
-<div id="refs"></div>
+## Acknowledgements
 
@@
 ```{r}
-read_tsv(here('data/process/study_summary_statistics.tsv')) %>%
+table_1 %>%
   arrange(nice_name) %>%
-  mutate(name_ref = glue("{nice_name} ({reference})"),
-    total_seqs=format(total_seqs, big.mark=",", trim=TRUE),
+  mutate(total_seqs=format(total_seqs, big.mark=",", trim=TRUE),
     median=format(as.integer(median), big.mark=",", trim=TRUE),
@@
     fold_difference = round(fold_difference, 1)) %>%
-  select(name_ref, n_samples, total_seqs, median, range, fold_difference) %>%
+  select(nice_name, n_samples, total_seqs, median, range, fold_difference) %>%
   kable("latex", booktabs=TRUE, escape=F, align="lrrrrr", linesep="",
-		col.names = linebreak(c("Study (Ref)", "Samples", "Total\nsequences", "Median\nsequences",
+		col.names = linebreak(c("Study", "Samples", "Total\nsequences", "Median\nsequences",
 			"Range of\nsequences", "Fold-difference\nbetween largest\nand smallest sample"), align="c")
@@
 **Figure 2.**
-v
+
+\newpage
+
+## References
+
+<div id="refs"></div>

### 2020-10-27 submission/manuscript.tex
Edit Results section

@@
 \textbf{The impact of removing rare ASVs on the information represented
-in each sample.} Removing rare ASVs will reduce the richness of ASVs and
-proportionally increase the relative abundance of the remaining ASVs.
-The result was expected to be a loss of information contained within
-each sample. To quantify the effect of removing rare ASVs on the
-information contained within each sample, I varied the minimum abundance
-threshold to simulate removing ASVs of varying rarity from each sample.
-The richness of ASVs in each sample (i.e.~the number of ASVs) decreased
-by between 34.4 and 86.2\% when removing those ASVs that only appeared
-once and by between 76.0 and 95.6\% when removing those that appeared
-ten or fewer times from each sample (Figure 2A). Similarly, the Shannon
-diversity decreased by between 1.8 and 15.9\% when removing ASVs that
-only appeared once and by between 76.0 and 95.6\% when removing ASVs
-that appeared ten or fewer times from each sample (Figure 2B). Next, I
-assigned the ASVs to OTUs to assess the impact of removing rare ASVs on
-higher level taxonomic groupings that are commonly used in microbial
-ecology studies. Although pooling similar ASVs into OTUs reduced the
-impact of removing the rare ASVs relative to the ASV-based analysis, the
-minimum abundance threshold still decreased the richness of OTUs and the
-diversity decreased relative to the full community (\textbf{Figure
-S2AB}). In contrast to the richness and diversity measurements, the
-Kullback--Leibler divergence compares the relative abundance of specific
-ASVs or OTUs between representations of the community. I calculated the
-Kullback--Leibler divergence from the full to pruned communities when
-rare ASVs were removed. As the threshold for removing ASVs increased,
-the amount of information lost also increased for both ASVs and OTUs
-(Figure 2C and Figure S2C). The relative loss of information was
-generally lower for OTUs than than it was for ASVs. Removing rare ASVs,
-regardless of abundance threshold, had profound impacts on the
-representation of the communities.
+in each sample.} Removing rare ASVs will reduce the richness of ASVs
+(i.e.~the number of ASVs per sample) and increase the relative abundance
+of the remaining ASVs. To quantify the effect of removing rare ASVs on
+the information contained within each sample, I varied the minimum
+abundance threshold to simulate removing ASVs of varying rarity from
+each sample. The richness of ASVs in each sample decreased by between
+34.4 and 86.2\% (peromyscus and soil) when removing those ASVs that only
+appeared once and by between 76.0 and 95.6\% (sediment and soil) when
+removing those that appeared ten or fewer times from each sample (Figure
+2A). Similarly, the Shannon diversity decreased by between 1.8 and
+15.9\% (human and soil) when removing ASVs that only appeared once and
+by between 5.4 and 35.4\% (human and seagrass) when removing ASVs that
+appeared ten or fewer times from each sample (Figure 2B). Next, I
+assigned the ASVs to OTUs, which were defined as a group of ASVs that
+were more than 97\% similar to each other to assess the impact of
+removing rare ASVs on higher level taxonomic groupings that are commonly
+used in microbial ecology studies. Although pooling similar ASVs into
+OTUs reduced the impact of removing the rare ASVs relative to the
+ASV-based analysis, the minimum abundance threshold still decreased the
+richness of OTUs and the diversity decreased relative to the full
+community (Figure S2AB). In contrast to the richness and diversity
+measurements, the Kullback--Leibler divergence compares the relative
+abundance of specific ASVs or OTUs between representations of the
+community. I calculated the Kullback--Leibler divergence between the
+full communities and those where rare ASVs were removed. As the
+threshold for removing ASVs increased, the amount of information lost
+also increased for both ASVs and OTUs (Figure 2C and Figure S2C). The
+relative loss of information was generally smaller for OTUs than than it
+was for ASVs. Removing rare ASVs, regardless of abundance threshold, had
+profound impacts on the representation of the communities.
 
@@
 treatment effects often affect a sample's diversity and inter-sample
-variation, I generated null distributions for each study by randomizing
-the number of times each ASV was observed in each sample such that the
-total number of sequences in each sample and the total number of times
-each ASV was observed across all samples in the study was the same as
-was originally observed. This effectively made every community in a
-study a statistical sample of the study-wide composite community
-distribution. For example, after this procedure, the 490 samples from
-the human dataset would be expected to have the same richness and
-diversity of ASVs and one would not expect to find treatment-based
-effects between the samples. Because of the risk of bias if only one
-representation of the null distribution was generated, I generated 100
-randomized datasets for each study. The trends between removing rare
-ASVs and the richness, diversity, and information loss that were
-identified using the observed community community distribution data were
-also identified with the data from the null distribution; however, the
-losses were larger when using the null distribution data (\textbf{Figure
-S3}). The null distribution data were used in the remainder of the study
-to minimize the risk of bias.
+variation, I generated null distributions for each study by randomizing,
+without replacement, the number of times each ASV was observed in each
+sample such that the total number of sequences in each sample and the
+total number of times each ASV was observed across all samples in the
+study was the same as was originally observed. This effectively made
+every community in a study a statistical sample of the study-wide
+composite community distribution. For example, after this procedure, the
+490 samples from the human dataset would be expected to have the same
+richness and diversity of ASVs and one would not expect to find
+treatment-based effects between the samples. Because of the risk of bias
+if only one representation of the null distribution was generated, I
+generated 100 randomized datasets for each study. The trends between
+removing rare ASVs and the richness, diversity, and information loss
+that were identified using the observed community community distribution
+data were also identified with the data from the null distribution;
+however, the losses were larger when using the null distribution data
+(Figure S3). The null distribution data were used in the remainder of
+the study to minimize the risk of bias.
 
@@
 diversity between samples I calculated the coefficients of variation
-(COVs, i.e.~standard deviation divided by the mean) for richness and
+(COVs, i.e.~the standard deviation divided by the mean) for richness and
 diversity for each study at multiple abundance thresholds. The COVs for
-richness of ASVs across the studies after removing singletons were
+the richness of ASVs across the studies after removing singletons were
 between 3.6 and 32.7-times larger than they were without removing
-singleton ASVs (Figure 3A). Similarly, the COVs for the diversity of
-ASVs were between 1.8 and 20.4-times larger when singletons were removed
-than when they were not removed (Figure 3B). To assess the impact of
-removing rare ASVs on measures of beta diversity between samples I
-calculated the COVs of the Bray-Curtis distances between samples within
-the same study at multiple abundance thresholds. The COVs between
-Bray-Curtis distances within a study when singletons were removed was
-between 1.3 and 18.6-times larger than when they were not removed
-(Figure 3C). Increasing the minimum abundance threshold increased the
-COVs between samples when using metrics of alpha and beta diversity.
-When ASVs were clustered into OTUs the difference in COVs was less than
-it was for the ASVs (Figure S4). These results indicate that removing
-rare ASVs increases the dissimilarity between samples, which could have
-a significant impact on the statistical power to detect differences
-between treatment groups.
+singleton ASVs (mice and stream; Figure 3A). Similarly, the COVs for the
+diversity of ASVs were between 1.8 and 20.4-times larger when singletons
+were removed than when they were not removed (mice and rice; Figure 3B).
+To assess the impact of removing rare ASVs on measures of beta diversity
+between samples, I calculated the COVs of the Bray-Curtis distances
+between samples within the same study at multiple abundance thresholds.
+The COVs between Bray-Curtis distances within a study when singletons
+were removed was between 1.3 and 18.6-times larger than when they were
+not removed (mice and stream; Figure 3C). When ASVs were clustered into
+OTUs the difference in COVs was less than it was for the ASVs (Figure
+S4). These results indicate that removing rare ASVs increases the
+dissimilarity between samples, which could have a significant impact on
+the statistical power to detect differences between treatment groups.
 
@@
 communities were randomly sampled from the null distribution as
-described above. For the second treatment group, I increased the
-abundance of 10\% of the ASVs in the pooled study distribution by 5\%. I
-randomly generated 100 simulated sets of treatment groups and samples. I
-then tested the ability to detect a difference between the two treatment
-groups using alpha and beta diversity metrics. The fraction of
-significant tests was a measurement of the statistical power to detect
-the difference between the treatment groups. When considering the
-differences in richness and diversity, the marine dataset yielded no
-simulated sets that were statistically significant, which was likely due
-to the small number of samples in the study (N=7). Among the remaining
-datasets, the power to detect a difference in the richness of ASVs
-ranged between 0.10 and 0.49 and between 0.10 and 0.53 to detect a
-difference in diversity when using a Wilcox test (Figure 4A). When
-singleton ASVs were removed, the power to detect a difference in the
-diversity of ASVs dropped by between 27.3 and 92.9\% and by between 40.0
-and 93.3\% (Figure 4B). The effect of removing rare ASVs on the richness
-of OTUs and their diversity was similar (Figure S5AB). I used the
-Bray-Curtis dissimilarity index to compare the simulated communities
-within each dataset and calculated the power to detect differences
-between the two simulated treatment groups using the analysis of
-molecular variance (also called PERMANOVA) (Figure 4C and S5C). Without
-removing rare sequences, the power to detect a difference between the
-two simulated treatment groups varied between 0.41 and 1.00. Aside from
-3 datasets, the power to detect differences dropped by between 6.5 and
-64.0\% when singletons were removed. However, when ASVs that occurred 10
-or fewer times were removed from each sample, the power to detect
-differences dropped by 12.0 and 97.2\%; similar results were observed
-when ASVs were clustered into OTUs. Removing rare ASVs reduced the
-ability to detect simulated treatment effects using metrics commonly
-used to compare microbial communities.
+described above. For the second treatment group, I randomly selected
+10\% of the ASVs in the pooled study distribution to increase their
+abundance by 5\%. I randomly generated 100 simulated sets of treatment
+groups and samples. I then tested the ability to detect a difference
+between the two treatment groups using alpha and beta diversity metrics.
+The fraction of significant tests was a measurement of the statistical
+power to detect the difference between the treatment groups. When
+considering the differences in richness and diversity, the marine
+dataset yielded no simulated sets that were statistically significant,
+which was likely due to the small number of samples in the study (N=7).
+Among the remaining datasets, the power to detect a difference in the
+richness of ASVs ranged between 0.10 and 0.49 (sediment and stream) and
+between 0.10 and 0.53 (rainforest and stream) to detect a difference in
+diversity when using a Wilcox test (Figure 4A). When singleton ASVs were
+removed, the power to detect a difference in the diversity of ASVs
+dropped by between 27.3 and 92.9\% (bioethanol and soil) and by between
+40.0 and 93.3\% (rainforest and soil; Figure 4B). The effect of removing
+rare ASVs on the richness of OTUs and their diversity was similar
+(Figure S5AB). I used the Bray-Curtis dissimilarity index to compare the
+simulated communities within each dataset and calculated the power to
+detect differences between the two simulated treatment groups using the
+analysis of molecular variance (also called PERMANOVA) (Figure 4C and
+S5C). Without removing rare sequences, the power to detect a difference
+between the two simulated treatment groups varied between 0.41 and 1.00
+(rainforest and rainforest). Aside from the bioethanol, human, and mice
+datasets, the power to detect differences dropped by between 6.5 and
+64.0\% (soil and rice) when singletons were removed. However, when ASVs
+that occurred 10 or fewer times were removed from each sample, the power
+to detect differences dropped by 12.0 and 97.2\% (human and peromyscus);
+similar results were observed when ASVs were clustered into OTUs.
+Removing rare ASVs reduced the ability to detect simulated treatment
+effects using metrics commonly used to compare microbial communities.
 
 \textbf{The impact of removing rare ASVs on the probability of falsely
-detecting a difference between treatment groups.} Observing reduced
-ability to detect differences between communities when rare ASVs were
-removed from each sample, I next asked whether removing rare ASVs could
-lead to falsely claiming that a treatment effect had a significant
-effect on community diversity and structure. First, I sampled sequences
-from the null distribution for each dataset and randomly assigned each
-sample to one of two treatment groups and determined the richness and
-diversity of ASVs and OTUs. Testing at an experiment-wise error rate of
-0.05, I expected 5\% of the iterations for each dataset to yield a
-significant test result. Indeed, there was no evidence that removing
-rare ASVs resulted in an inflated experiment-wise error rate. The
+detecting a difference between treatment groups.} I next asked whether
+removing rare ASVs could lead to falsely claiming that a treatment
+effect had a significant effect on community diversity and structure.
+First, I sampled sequences from the null distribution for each dataset
+and randomly assigned each sample to one of two treatment groups and
+determined the richness and diversity of ASVs and OTUs. Testing at an
+experiment-wise error rate of 0.05, I expected 5\% of the iterations for
+each dataset to yield a significant test result. Indeed, there was no
+evidence that removing rare ASVs resulted in an inflated experiment-wise
+error rate. The average fraction of significant tests did not
+meaningfully vary from 0.05 across the minimum abundance threshold,
+dataset, metric of describing sample alpha-diversity, or whether the
+abundance of ASVs or OTUs were used (Figure 5A and S6A). Similarly, the
 average fraction of significant tests did not meaningfully vary from
-0.05 across the minimum abundance threshold, dataset, metric of
-describing sample alpha-diversity, or whether the abundance of ASVs or
-OTUs were used (Figure 5A and S6A). Similarly, the average fraction of
-significant tests did not meaningfully vary from 0.05 when using
-analysis of molecular variance to compare communities using Bray-Curtis
-distances (Figure 5A and S6A). Second, I again sampled sequences from
-the null distribution, but assigned samples to one of two treatment
-groups based on the number of sequences in each sample. The samples with
-fewer than the median number of sequences for the dataset were assigned
-to one group and those with more than the median were assigned to the
-other. This exaggerated bias has been observed in comparisons of the
-lung and oral microbiota because of the larger number of non-specific
-amplicons that can be sequenced from lung samples relative to those in
-the oral cavity leading to a significant difference in sequencing depth
-between treatment groups {[}\textbf{REF}{]}. When rare sequences were
-not removed, the fraction of significant tests did not differ from 5\%
-for comparing the richness, their diversity, or Bray-Curtis distances
-(Figure 5B and S6B). However, when rare taxa of any frequency were
-removed, the probability of falsely detecing a difference as signifiant
-increased with the definition of rarity (Figure 5B and S6B). Not
-including the small marine dataset, the average fraction the average
+0.05 when using analysis of molecular variance to compare communities
+using Bray-Curtis distances (Figure 5A and S6A). Second, I again sampled
+sequences from the null distribution, but assigned samples to one of two
+treatment groups based on the number of sequences in each sample. The
+samples with fewer than the median number of sequences for the dataset
+were assigned to one group and those with more than the median were
+assigned to the other. This exaggerated bias has been observed in
+comparisons of the lung and oral microbiota because of the larger number
+of non-specific amplicons that can be sequenced from lung samples
+relative to those in the oral cavity leading to a significant difference
+in sequencing depth between treatment groups {[}{]}. When rare sequences
+were not removed, the fraction of significant tests did not differ from
+5\% for comparing the richness, their diversity, or Bray-Curtis
+distances (Figure 5B and S6B). However, when rare taxa of any frequency
+were removed, the probability of falsely detecing a difference as
+signifiant increased with the definition of rarity (Figure 5B and S6B).
+Not including the small marine dataset, the average fraction the average
 fraction of falsely detecting a difference across datasets when only
@@
 
-\textbf{\emph{Conclusion.}} Removing rare sequences decreases the
-diversity represented by 16S rRNA gene sequence data and increases the
-variation between samples. Such impacts will hinder the statistical
-power to differentiate between treatment groups. Instead of removing
-rare sequences, researchers should focus on optimizing their sequence
-generation to minimize the amount of PCR and sequencing errors. In
-addition, samples should be rarefied to a common number of sequences
-across samples without prior culling of rare sequences. The number of
-artifacts is correlated to the number of sequences being considered.
-With this in mind, rarefaction allows one to control for uneven sampling
-effort and to control for the number of artifacts in the analysis.
+\subsection{Discussion}\label{discussion}
+
+Removing rare sequences decreases the diversity represented by 16S rRNA
+gene sequence data and increases the variation between samples. Such
+impacts will hinder the statistical power to differentiate between
+treatment groups. Instead of removing rare sequences, researchers should
+focus on optimizing their sequence generation to minimize the amount of
+PCR and sequencing errors. In addition, samples should be rarefied to a
+common number of sequences across samples without prior culling of rare
+sequences. The number of artifacts is correlated to the number of
+sequences being considered. With this in mind, rarefaction allows one to
+control for uneven sampling effort and to control for the number of
+artifacts in the analysis.
 
 Need to treat rare sequences with a grain of salt
-
-\newpage
-
-\subsection{Acknowledgements}\label{acknowledgements}
 
@@
 
-\newpage
-
-\subsection{References}\label{references}
+\subsection{Acknowledgements}\label{acknowledgements}
 
@@
 \toprule
-\textbf{Study (Ref)} & \textbf{Samples} & \textbf{\makecell[c]{Total\\sequences}} & \textbf{\makecell[c]{Median\\sequences}} & \textbf{\makecell[c]{Range of\\sequences}} & \textbf{\makecell[c]{Fold-difference\\between largest\\and smallest sample}}\\
+\textbf{Study} & \textbf{Samples} & \textbf{\makecell[c]{Total\\sequences}} & \textbf{\makecell[c]{Median\\sequences}} & \textbf{\makecell[c]{Range of\\sequences}} & \textbf{\makecell[c]{Fold-difference\\between largest\\and smallest sample}}\\
 \midrule
-Bioethanol (NA) & 95 & 3,972,943 & 16,015 & 3,688-356,136 & 96.6\\
-Human (NA) & 490 & 20,909,768 & 32,505 & 10,523-430,415 & 40.9\\
-Lake (NA) & 52 & 3,169,868 & 69,041 & 15,347-112,871 & 7.4\\
-Marine (NA) & 7 & 1,391,396 & 193,464 & 133,516-254,060 & 1.9\\
-Mice (NA) & 348 & 2,813,747 & 6,477 & 1,804-30,565 & 16.9\\
-Peromyscus (NA) & 111 & 1,555,545 & 12,446 & 4,464-33,644 & 7.5\\
-Rainforest (NA) & 69 & 946,295 & 11,561 & 4,932-37,767 & 7.7\\
-Rice (NA) & 490 & 22,591,168 & 43,216 & 2,776-193,464 & 69.7\\
-Seagrass (NA) & 286 & 4,130,454 & 13,567 & 1,803-45,191 & 25.1\\
-Sediment (NA) & 58 & 1,154,174 & 17,584 & 7,685-68,321 & 8.9\\
-Soil (NA) & 18 & 956,656 & 51,844 & 47,806-59,956 & 1.3\\
-Stream (NA) & 201 & 21,162,574 & 90,159 & 9,175-390,964 & 42.6\\
+Bioethanol & 95 & 3,972,943 & 16,015 & 3,688-356,136 & 96.6\\
+Human & 490 & 20,909,768 & 32,505 & 10,523-430,415 & 40.9\\
+Lake & 52 & 3,169,868 & 69,041 & 15,347-112,871 & 7.4\\
+Marine & 7 & 1,391,396 & 193,464 & 133,516-254,060 & 1.9\\
+Mice & 348 & 2,813,747 & 6,477 & 1,804-30,565 & 16.9\\
+Peromyscus & 111 & 1,555,545 & 12,446 & 4,464-33,644 & 7.5\\
+Rainforest & 69 & 946,295 & 11,561 & 4,932-37,767 & 7.7\\
+Rice & 490 & 22,591,168 & 43,216 & 2,776-193,464 & 69.7\\
+Seagrass & 286 & 4,130,454 & 13,567 & 1,803-45,191 & 25.1\\
+Sediment & 58 & 1,154,174 & 17,584 & 7,685-68,321 & 8.9\\
+Soil & 18 & 956,656 & 51,844 & 47,806-59,956 & 1.3\\
+Stream & 201 & 21,162,574 & 90,159 & 9,175-390,964 & 42.6\\
 \bottomrule
@@
 
-\textbf{Figure 2.} v
+\textbf{Figure 2.}
+
+\newpage
+
+\subsection{References}\label{references}
 

### 2020-10-29 submission/manuscript.Rmd
Write Discussion

@@
 The samples with fewer than the median number of sequences for the dataset were assigned to one group and those with more than the median were assigned to the other.
-This exaggerated bias has been observed in comparisons of the lung and oral microbiota because of the larger number of non-specific amplicons that can be sequenced from lung samples relative to those in the oral cavity leading to a significant difference in sequencing depth between treatment groups [].
+This exaggerated bias has been observed in comparisons of the lung and oral microbiota because of the larger number of non-specific amplicons that can be sequenced from lung samples relative to those in the oral cavity leading to a significant difference in sequencing depth between treatment groups [@Morris2013].
 When rare sequences were not removed, the fraction of significant tests did not differ from 5% for comparing the richness, their diversity, or Bray-Curtis distances (Figure 5B and S6B).
@@
 
+### Big points
+* Rare sequences are likely to be covered in other samples, suggesting they are rare; removing them biases against samples with fewer sequences
+* Removing rare sequences reduces the diversity within samples and changes the shape of the communities
+* The results include
+  - increases the differences between samples; this makes it harder to detect differences between samples when those differences actually exist
+  - If there is confounding between the number of reads per sample and the experimental grouping, then removing rare sequences increases the probability of falsely detecting a difference
+
 Removing rare sequences decreases the diversity represented by 16S rRNA gene sequence data and increases the variation between samples.
 Such impacts will hinder the statistical power to differentiate between treatment groups.
-Instead of removing rare sequences, researchers should focus on optimizing their sequence generation to minimize the amount of PCR and sequencing errors.
+
+### What to do
+* Removal of rare sequences is largely a response to optimizing the number of reads and length of sequences over their quality
+* Researchers should instead focus on optimizing their sequence generation to minimize the amount of PCR and sequencing errors.
+We know that incomplete overlap of sequence reads significantly increases the sequencing error rate.
+Didn't test on crappy data
+* Rarefy data to control for uneven sampling of communities and PCR or sequencing artifacts
+
 In addition, samples should be rarefied to a common number of sequences across samples without prior culling of rare sequences.
@@
 
-Need to treat rare sequences with a grain of salt
+### Should we implicitly trust rare sequences?
+* No.
+  - Need to treat rare sequences with a grain of salt, but don't discard
+  - Need to use other methods to validate results observed using rare taxa
+* For community metrics
+  - rare sequencs are an important component and removing them shifts the community distribution.
+  - metrics need to be treated on a relative rather than absolute basis
+
+### Definition of rarity
+* Idea of a blanket removal of reads below a threshold is stupid anyway
+* Microbial communities contain numerically rare, temporarily rare, and spatially rare taxa.
+* Removing those populations fundamentally changes how we think of microbial communities
 

### 2020-10-30 submission/manuscript.tex

@@
 relative to those in the oral cavity leading to a significant difference
-in sequencing depth between treatment groups {[}{]}. When rare sequences
+in sequencing depth between treatment groups (3). When rare sequences
 were not removed, the fraction of significant tests did not differ from
@@
 
+\subsubsection{Big points}\label{big-points}
+
+\begin{itemize}
+\itemsep1pt\parskip0pt\parsep0pt
+\item
+  Rare sequences are likely to be covered in other samples, suggesting
+  they are rare; removing them biases against samples with fewer
+  sequences
+\item
+  Removing rare sequences reduces the diversity within samples and
+  changes the shape of the communities
+\item
+  The results include
+\item
+  increases the differences between samples; this makes it harder to
+  detect differences between samples when those differences actually
+  exist
+\item
+  If there is confounding between the number of reads per sample and the
+  experimental grouping, then removing rare sequences increases the
+  probability of falsely detecting a difference
+\end{itemize}
+
 Removing rare sequences decreases the diversity represented by 16S rRNA
@@
 impacts will hinder the statistical power to differentiate between
-treatment groups. Instead of removing rare sequences, researchers should
-focus on optimizing their sequence generation to minimize the amount of
-PCR and sequencing errors. In addition, samples should be rarefied to a
-common number of sequences across samples without prior culling of rare
-sequences. The number of artifacts is correlated to the number of
-sequences being considered. With this in mind, rarefaction allows one to
-control for uneven sampling effort and to control for the number of
-artifacts in the analysis.
+treatment groups.
 
-Need to treat rare sequences with a grain of salt
+\subsubsection{What to do}\label{what-to-do}
+
+\begin{itemize}
+\itemsep1pt\parskip0pt\parsep0pt
+\item
+  Removal of rare sequences is largely a response to optimizing the
+  number of reads and length of sequences over their quality
+\item
+  Researchers should instead focus on optimizing their sequence
+  generation to minimize the amount of PCR and sequencing errors. We
+  know that incomplete overlap of sequence reads significantly increases
+  the sequencing error rate. Didn't test on crappy data
+\item
+  Rarefy data to control for uneven sampling of communities and PCR or
+  sequencing artifacts
+\end{itemize}
+
+In addition, samples should be rarefied to a common number of sequences
+across samples without prior culling of rare sequences. The number of
+artifacts is correlated to the number of sequences being considered.
+With this in mind, rarefaction allows one to control for uneven sampling
+effort and to control for the number of artifacts in the analysis.
+
+\subsubsection{Should we implicitly trust rare
+sequences?}\label{should-we-implicitly-trust-rare-sequences}
+
+\begin{itemize}
+\itemsep1pt\parskip0pt\parsep0pt
+\item
+  No.
+\item
+  Need to treat rare sequences with a grain of salt, but don't discard
+\item
+  Need to use other methods to validate results observed using rare taxa
+\item
+  For community metrics
+\item
+  rare sequencs are an important component and removing them shifts the
+  community distribution.
+\item
+  metrics need to be treated on a relative rather than absolute basis
+\end{itemize}
+
+\subsubsection{Definition of rarity}\label{definition-of-rarity}
+
+\begin{itemize}
+\itemsep1pt\parskip0pt\parsep0pt
+\item
+  Idea of a blanket removal of reads below a threshold is stupid anyway
+\item
+  Microbial communities contain numerically rare, temporarily rare, and
+  spatially rare taxa.
+\item
+  Removing those populations fundamentally changes how we think of
+  microbial communities
+\end{itemize}
 
@@
 
+3. \textbf{Morris A}, \textbf{Beck JM}, \textbf{Schloss PD},
+\textbf{Campbell TB}, \textbf{Crothers K}, \textbf{Curtis JL},
+\textbf{Flores SC}, \textbf{Fontenot AP}, \textbf{Ghedin E},
+\textbf{Huang L}, \textbf{Jablonski K}, \textbf{Kleerup E},
+\textbf{Lynch SV}, \textbf{Sodergren E}, \textbf{Twigg H}, \textbf{Young
+VB}, \textbf{Bassis CM}, \textbf{Venkataraman A}, \textbf{Schmidt TM},
+\textbf{Weinstock GM}. 2013. Comparison of the respiratory microbiome in
+healthy nonsmokers and smokers. American Journal of Respiratory and
+Critical Care Medicine \textbf{187}:1067--1075.
+doi:\href{http://dx.doi.org/10.1164/rccm.201210-1913oc}{10.1164/rccm.201210-1913oc}.
+
 \end{document}

### 2020-10-30 submission/manuscript.Rmd
Write and edit Discussion

--- 2020-10-27 submission/manuscript.Rmd
@@
     keep_tex: true
+csl: mbio.csl
+geometry: margin=1.0in
 header-includes:
@@
  - \usepackage[T1]{fontenc}
-csl: mbio.csl
-geometry: margin=1.0in
 ---
@@
 \newpage
-\linenumbers
+<!-- \linenumbers -->
 
@@
 b_base_pc_shannon <- b_alpha %>% filter(metric == "shannon", method == "pc") %>% arrange(`1`) %>% slice(1, nrow(.))
-
 b_pc_sobs <- b_alpha %>% filter(metric == "sobs", method == "pc") %>% slice(1, n())
@@
   pull(min_p)
-
 ```
@@
 
-Removing rare sequences decreases the diversity represented by 16S rRNA gene sequence data and increases the variation between samples.
-Such impacts will hinder the statistical power to differentiate between treatment groups.
-Instead of removing rare sequences, researchers should focus on optimizing their sequence generation to minimize the amount of PCR and sequencing errors.
-In addition, samples should be rarefied to a common number of sequences across samples without prior culling of rare sequences.
-The number of artifacts is correlated to the number of sequences being considered.
-With this in mind, rarefaction allows one to control for uneven sampling effort and to control for the number of artifacts in the analysis.
+Removing rare sequences from 16S rRNA gene sequence data is a common practice that is used as a heuristic to help remove residual PCR artifacts and low quality sequences.
+In this analysis, I have shown that rare sequences are more common in samples with shallow sequencing than in those with deep sequencing and that rare sequences are frequently observed in multiple samples.
+These observations suggest that many of the sequences being removed are actually good sequences.
+Becuase rarity is often defined by a fixed number of observations per sample (e.g.
+sequences that only appear once in a sample, regardless of the size of the sample), removing rare sequences has a disproportionate impact on samples with fewer sequences.
+Removing rare sequences resulted in a reduction in the alpha diversity and a pronounced change in the structure of individual samples.
+The effect was an increase in the differences observed between samples, which made it more difficult to detect differences between treatment groups when differences actually existed.
+Furthermore, if the number of reads per sample was confounded with the treatment groups, then removing rare sequences increased the probability of falsely detecting a difference between the treatment groups.
+The practice of removing rare sequences from samples should be stopped.
 
-Need to treat rare sequences with a grain of salt
+The practice of removing rare sequences from samples seems to be a response to researchers prioritizing the number of reads and length of sequences over their quality [@Bokulich2013].
+Previous work has shown that assembly of fully overlapping sequence reads results in the lowest sequencing error rates [@Kozich2013].
+The studies highlighted in this analysis sequenced the V4 region of the 16S rRNA gene using the Illumina MiSeq sequencing platform with their v2 chemistry.
+The resulting data consists of two 250 nt reads that span a region that is about 250 nt long.
+In contrast, there has been a movement to sequence longer regions with similar chemistry resulting in less overlap between the sequencing reads [@Fadrosh2014; @Holm2019].
+Alternatively, others have prioritized increasing the number of sequences per sample by sequencing the V4 region, but with paired or single reads that are 150 nt long [@Thompson2017; @McDonald2018].
+Both practices result in a significantly higher error rate for the resulting assmebly.
+Instead, researchers should prioritize the quality over the quantity and length of their data.
+For these reasons, this analysis did not use lower quality data generated by alternative methods
+
+The impacts of removing rare sequences on the representation of communities are caused by the uneven impact of applying a single abundance threshold across all samples.
+The number of sequences per sample in the datasets highlighted in this analysis varied by 100-fold (Figure S1); I suspect the range is actually larger since researchers may have opted against depositing samples with fewer reads into the SRA.
+If the range in read coverage for a study is 100-fold, a singleton in a small sample would have had a comparable relative abundance as a sequence that appeared 100 times in a more densely sequenced sample.
+However, it would have been removed from the smaller sample and not the larger sample.
+Thus, applying a single abundance threshold disproportionately impacted the samples with fewer sequences.
+This seems to contradict the purpose of removing rare sequences since a singleton in the smaller sample is more reliable than the singleton in the larger sample.
+A superior approach to removing rare sequences is to use rarefaction to conrol for uneven sampling.
+
+Even with the best sequencing approaches, PCR artifacts and sequencing errors persist.
+This may account for some sequences not being observed other samples.
+Although this lack of inter-sample coverage could have been due to treatment effects and natural variation, it is likely that some portion of the sequences that were unique to samples were artifacts or contained errors.
+Researchers are encouraged to use rarefaction to control for uneven sampling and the presence of spurious sequences.
+Previous work sequencing mock communities has shown that the number of spurious sequences increases with sequencing depth [@Kozich2013; @Schloss2011].
+By rarefying data to a common number of sequences per sample, the number of spurious sequences can be controlled.
+As shown in the data I presented, which was rarefied to a common number of reads per sample within a dataset, when rare sequences were not removed the power to detect differences was the highest and the false discovery rate was the expected 5% (Figures 5 and S6).
+
+In addition to considerations of how to control for the presence of spurious sequences, researchers also need to be mindful of how to interpret the results of their work.
+Because every dataset will contain residual sequences that are spurious, measures of richness and diversity should be made on a relative basis.
+For example, pronouncements that communities from an environment or treatment contain a specific number of taxa are problematic.
+Instead, we should limit ourselves to indicating that samples from one treatment group has more taxa than another without using absolute values of richness.
+Furthermore, we must take caution in interpreting rare taxa.
+Although the data from the studies highlighted here suggest that most rare sequences are not spurious, it is likely that some are.
+Therefore, researchers must approach rare sequences with more skepticism than more abundant sequences.
+Reseachers should seek out other methods to confirm inferences that they make about rare sequences.
+This is a standard that should be applied regardless of their abundance [@Schloss2018].
+
+How to curate and interpret rare sequences has been a significant challenge since microbial ecologists transitioned away from Sanger sequencing of samples [@Sogin2006; @Huse2010; @Reeder2009].
+Although the extent of the "rare biosphere" is still an open question, it is important to appreciate the importance of rare populations in all communities.
+Populations can be numerically rare but ubuiquitous or abundant and limited in their geographic range.
+Alternatively, they can be numerically rare but temporally common or abundant but present infrequently.
+Removing sequences from any of these settings will limit our ability to study the role of such populations or the processes that drive their patchy distributions that are so common with microbial communities [@Reid2011].
 
@@
 ## Acknowledgements
+
+\newpage
+
+## References
+
+<div id="refs"></div>
 
@@
 **Figure 2.**
-
-\newpage
-
-## References
-
-<div id="refs"></div>

### 2020-10-30 submission/manuscript.tex
Write and edit Discussion

--- 2020-10-27 submission/manuscript.tex
@@
 \newpage
-\linenumbers
 
@@
 
-Removing rare sequences decreases the diversity represented by 16S rRNA
-gene sequence data and increases the variation between samples. Such
-impacts will hinder the statistical power to differentiate between
-treatment groups. Instead of removing rare sequences, researchers should
-focus on optimizing their sequence generation to minimize the amount of
-PCR and sequencing errors. In addition, samples should be rarefied to a
-common number of sequences across samples without prior culling of rare
-sequences. The number of artifacts is correlated to the number of
-sequences being considered. With this in mind, rarefaction allows one to
-control for uneven sampling effort and to control for the number of
-artifacts in the analysis.
+Removing rare sequences from 16S rRNA gene sequence data is a common
+practice that is used as a heuristic to help remove residual PCR
+artifacts and low quality sequences. In this analysis, I have shown that
+rare sequences are more common in samples with shallow sequencing than
+in those with deep sequencing and that rare sequences are frequently
+observed in multiple samples. These observations suggest that many of
+the sequences being removed are actually good sequences. Becuase rarity
+is often defined by a fixed number of observations per sample
+(e.g.~sequences that only appear once in a sample, regardless of the
+size of the sample), removing rare sequences has a disproportionate
+impact on samples with fewer sequences. Removing rare sequences resulted
+in a reduction in the alpha diversity and a pronounced change in the
+structure of individual samples. The effect was an increase in the
+differences observed between samples, which made it more difficult to
+detect differences between treatment groups when differences actually
+existed. Furthermore, if the number of reads per sample was confounded
+with the treatment groups, then removing rare sequences increased the
+probability of falsely detecting a difference between the treatment
+groups. The practice of removing rare sequences from samples should be
+stopped.
 
-Need to treat rare sequences with a grain of salt
+The practice of removing rare sequences from samples seems to be a
+response to researchers prioritizing the number of reads and length of
+sequences over their quality (4). Previous work has shown that assembly
+of fully overlapping sequence reads results in the lowest sequencing
+error rates (1). The studies highlighted in this analysis sequenced the
+V4 region of the 16S rRNA gene using the Illumina MiSeq sequencing
+platform with their v2 chemistry. The resulting data consists of two 250
+nt reads that span a region that is about 250 nt long. In contrast,
+there has been a movement to sequence longer regions with similar
+chemistry resulting in less overlap between the sequencing reads (5, 6).
+Alternatively, others have prioritized increasing the number of
+sequences per sample by sequencing the V4 region, but with paired or
+single reads that are 150 nt long (7, 8). Both practices result in a
+significantly higher error rate for the resulting assmebly. Instead,
+researchers should prioritize the quality over the quantity and length
+of their data. For these reasons, this analysis did not use lower
+quality data generated by alternative methods
+
+The impacts of removing rare sequences on the representation of
+communities are caused by the uneven impact of applying a single
+abundance threshold across all samples. The number of sequences per
+sample in the datasets highlighted in this analysis varied by 100-fold
+(Figure S1); I suspect the range is actually larger since researchers
+may have opted against depositing samples with fewer reads into the SRA.
+If the range in read coverage for a study is 100-fold, a singleton in a
+small sample would have had a comparable relative abundance as a
+sequence that appeared 100 times in a more densely sequenced sample.
+However, it would have been removed from the smaller sample and not the
+larger sample. Thus, applying a single abundance threshold
+disproportionately impacted the samples with fewer sequences. This seems
+to contradict the purpose of removing rare sequences since a singleton
+in the smaller sample is more reliable than the singleton in the larger
+sample. A superior approach to removing rare sequences is to use
+rarefaction to conrol for uneven sampling.
+
+Even with the best sequencing approaches, PCR artifacts and sequencing
+errors persist. This may account for some sequences not being observed
+other samples. Although this lack of inter-sample coverage could have
+been due to treatment effects and natural variation, it is likely that
+some portion of the sequences that were unique to samples were artifacts
+or contained errors. Researchers are encouraged to use rarefaction to
+control for uneven sampling and the presence of spurious sequences.
+Previous work sequencing mock communities has shown that the number of
+spurious sequences increases with sequencing depth (1, 9). By rarefying
+data to a common number of sequences per sample, the number of spurious
+sequences can be controlled. As shown in the data I presented, which was
+rarefied to a common number of reads per sample within a dataset, when
+rare sequences were not removed the power to detect differences was the
+highest and the false discovery rate was the expected 5\% (Figures 5 and
+S6).
+
+In addition to considerations of how to control for the presence of
+spurious sequences, researchers also need to be mindful of how to
+interpret the results of their work. Because every dataset will contain
+residual sequences that are spurious, measures of richness and diversity
+should be made on a relative basis. For example, pronouncements that
+communities from an environment or treatment contain a specific number
+of taxa are problematic. Instead, we should limit ourselves to
+indicating that samples from one treatment group has more taxa than
+another without using absolute values of richness. Furthermore, we must
+take caution in interpreting rare taxa. Although the data from the
+studies highlighted here suggest that most rare sequences are not
+spurious, it is likely that some are. Therefore, researchers must
+approach rare sequences with more skepticism than more abundant
+sequences. Reseachers should seek out other methods to confirm
+inferences that they make about rare sequences. This is a standard that
+should be applied regardless of their abundance (10).
+
+How to curate and interpret rare sequences has been a significant
+challenge since microbial ecologists transitioned away from Sanger
+sequencing of samples (11--13). Although the extent of the ``rare
+biosphere'' is still an open question, it is important to appreciate the
+importance of rare populations in all communities. Populations can be
+numerically rare but ubuiquitous or abundant and limited in their
+geographic range. Alternatively, they can be numerically rare but
+temporally common or abundant but present infrequently. Removing
+sequences from any of these settings will limit our ability to study the
+role of such populations or the processes that drive their patchy
+distributions that are so common with microbial communities (14).
 
@@
 \subsection{Acknowledgements}\label{acknowledgements}
+
+\newpage
+
+\subsection{References}\label{references}
 
@@
 
-\newpage
-
-\subsection{References}\label{references}
-
 1. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
@@
 
+3. \textbf{Morris A}, \textbf{Beck JM}, \textbf{Schloss PD},
+\textbf{Campbell TB}, \textbf{Crothers K}, \textbf{Curtis JL},
+\textbf{Flores SC}, \textbf{Fontenot AP}, \textbf{Ghedin E},
+\textbf{Huang L}, \textbf{Jablonski K}, \textbf{Kleerup E},
+\textbf{Lynch SV}, \textbf{Sodergren E}, \textbf{Twigg H}, \textbf{Young
+VB}, \textbf{Bassis CM}, \textbf{Venkataraman A}, \textbf{Schmidt TM},
+\textbf{Weinstock GM}. 2013. Comparison of the respiratory microbiome in
+healthy nonsmokers and smokers. American Journal of Respiratory and
+Critical Care Medicine \textbf{187}:1067--1075.
+doi:\href{http://dx.doi.org/10.1164/rccm.201210-1913oc}{10.1164/rccm.201210-1913oc}.
+
+4. \textbf{Bokulich NA}, \textbf{Subramanian S}, \textbf{Faith JJ},
+\textbf{Gevers D}, \textbf{Gordon JI}, \textbf{Knight R}, \textbf{Mills
+DA}, \textbf{Caporaso JG}. 2013. Quality-filtering vastly improves
+diversity estimates from illumina amplicon sequencing. Nature Methods
+\textbf{10}:57--59.
+doi:\href{http://dx.doi.org/10.1038/nmeth.2276}{10.1038/nmeth.2276}.
+
+5. \textbf{Fadrosh DW}, \textbf{Ma B}, \textbf{Gajer P},
+\textbf{Sengamalay N}, \textbf{Ott S}, \textbf{Brotman RM},
+\textbf{Ravel J}. 2014. An improved dual-indexing approach for
+multiplexed 16S rRNA gene sequencing on the illumina MiSeq platform.
+Microbiome \textbf{2}:6.
+doi:\href{http://dx.doi.org/10.1186/2049-2618-2-6}{10.1186/2049-2618-2-6}.
+
+6. \textbf{Holm JB}, \textbf{Humphrys MS}, \textbf{Robinson CK},
+\textbf{Settles ML}, \textbf{Ott S}, \textbf{Fu L}, \textbf{Yang H},
+\textbf{Gajer P}, \textbf{He X}, \textbf{McComb E}, \textbf{Gravitt PE},
+\textbf{Ghanem KG}, \textbf{Brotman RM}, \textbf{Ravel J}. 2019.
+Ultrahigh-throughput multiplexing and sequencing of $>$500-base-pair
+amplicon regions on the Illumina HiSeq 2500 platform. mSystems
+\textbf{4}.
+doi:\href{http://dx.doi.org/10.1128/msystems.00029-19}{10.1128/msystems.00029-19}.
+
+7. \textbf{Thompson LR}, \textbf{Jon G. Sanders}, \textbf{McDonald D},
+\textbf{Amir A}, \textbf{Ladau J}, \textbf{Locey KJ}, \textbf{Prill RJ},
+\textbf{Tripathi A}, \textbf{Gibbons SM}, \textbf{Ackermann G},
+\textbf{Navas-Molina JA}, \textbf{Janssen S}, \textbf{Kopylova E},
+\textbf{V{á}zquez-Baeza Y}, \textbf{Gonz{á}lez A}, \textbf{Morton JT},
+\textbf{Mirarab S}, \textbf{Xu ZZ}, \textbf{Jiang L}, \textbf{Haroon
+MF}, \textbf{Kanbar J}, \textbf{Zhu Q}, \textbf{Song SJ},
+\textbf{Kosciolek T}, \textbf{Bokulich NA}, \textbf{Lefler J},
+\textbf{Brislawn CJ}, \textbf{Humphrey G}, \textbf{Owens SM},
+\textbf{Hampton-Marcell J}, \textbf{Berg-Lyons D}, \textbf{McKenzie V},
+\textbf{Fierer N}, \textbf{Fuhrman JA}, \textbf{Clauset A},
+\textbf{Stevens RL}, \textbf{Shade A}, \textbf{Pollard KS},
+\textbf{Goodwin KD}, \textbf{Jansson JK}, \textbf{Gilbert JA},
+\textbf{Knight R}. 2017. A communal catalogue reveals earth's multiscale
+microbial diversity. Nature \textbf{551}:457--463.
+doi:\href{http://dx.doi.org/10.1038/nature24621}{10.1038/nature24621}.
+
+8. \textbf{McDonald D}, \textbf{Hyde E}, \textbf{Debelius JW},
+\textbf{Morton JT}, \textbf{Gonzalez A}, \textbf{Ackermann G},
+\textbf{Aksenov AA}, \textbf{Behsaz B}, \textbf{Brennan C}, \textbf{Chen
+Y}, \textbf{Goldasich LD}, \textbf{Dorrestein PC}, \textbf{Dunn RR},
+\textbf{Fahimipour AK}, \textbf{Gaffney J}, \textbf{Gilbert JA},
+\textbf{Gogul G}, \textbf{Green JL}, \textbf{Hugenholtz P},
+\textbf{Humphrey G}, \textbf{Huttenhower C}, \textbf{Jackson MA},
+\textbf{Janssen S}, \textbf{Jeste DV}, \textbf{Jiang L}, \textbf{Kelley
+ST}, \textbf{Knights D}, \textbf{Kosciolek T}, \textbf{Ladau J},
+\textbf{Leach J}, \textbf{Marotz C}, \textbf{Meleshko D}, \textbf{Melnik
+AV}, \textbf{Metcalf JL}, \textbf{Mohimani H}, \textbf{Montassier E},
+\textbf{Navas-Molina J}, \textbf{Nguyen TT}, \textbf{Peddada S},
+\textbf{Pevzner P}, \textbf{Pollard KS}, \textbf{Rahnavard G},
+\textbf{Robbins-Pianka A}, \textbf{Sangwan N}, \textbf{Shorenstein J},
+\textbf{Smarr L}, \textbf{Song SJ}, \textbf{Spector T}, \textbf{Swafford
+AD}, \textbf{Thackray VG}, \textbf{Thompson LR}, \textbf{Tripathi A},
+\textbf{V{á}zquez-Baeza Y}, \textbf{Vrbanac A}, \textbf{Wischmeyer P},
+\textbf{Wolfe E}, \textbf{Zhu Q}, \textbf{Knight R}, \textbf{Mann AE},
+\textbf{Amir A}, \textbf{Frazier A}, \textbf{Martino C},
+\textbf{Lebrilla C}, \textbf{Lozupone C}, \textbf{Lewis CM},
+\textbf{Raison C}, \textbf{Zhang C}, \textbf{Lauber CL},
+\textbf{Warinner C}, \textbf{Lowry CA}, \textbf{Callewaert C},
+\textbf{Bloss C}, \textbf{Willner D}, \textbf{Galzerani DD},
+\textbf{Gonzalez DJ}, \textbf{Mills DA}, \textbf{Chopra D},
+\textbf{Gevers D}, \textbf{Berg-Lyons D}, \textbf{Sears DD},
+\textbf{Wendel D}, \textbf{Lovelace E}, \textbf{Pierce E},
+\textbf{TerAvest E}, \textbf{Bolyen E}, \textbf{Bushman FD}, \textbf{Wu
+GD}, \textbf{Church GM}, \textbf{Saxe G}, \textbf{Holscher HD},
+\textbf{Ugrina I}, \textbf{German JB}, \textbf{Caporaso JG},
+\textbf{Wozniak JM}, \textbf{Kerr J}, \textbf{Ravel J}, \textbf{Lewis
+JD}, \textbf{Suchodolski JS}, \textbf{Jansson JK},
+\textbf{Hampton-Marcell JT}, \textbf{Bobe J}, \textbf{Raes J},
+\textbf{Chase JH}, \textbf{Eisen JA}, \textbf{Monk J}, \textbf{Clemente
+JC}, \textbf{Petrosino J}, \textbf{Goodrich J}, \textbf{Gauglitz J},
+\textbf{Jacobs J}, \textbf{Zengler K}, \textbf{Swanson KS},
+\textbf{Lewis K}, \textbf{Mayer K}, \textbf{Bittinger K}, \textbf{Dillon
+L}, \textbf{Zaramela LS}, \textbf{Schriml LM}, \textbf{Dominguez-Bello
+MG}, \textbf{Jankowska MM}, \textbf{Blaser M}, \textbf{Pirrung M},
+\textbf{Minson M}, \textbf{Kurisu M}, \textbf{Ajami N}, \textbf{Gottel
+NR}, \textbf{Chia N}, \textbf{Fierer N}, \textbf{White O}, \textbf{Cani
+PD}, \textbf{Gajer P}, \textbf{Strandwitz P}, \textbf{Kashyap P},
+\textbf{Dutton R}, \textbf{Park RS}, \textbf{Xavier RJ}, \textbf{Mills
+RH}, \textbf{Krajmalnik-Brown R}, \textbf{Ley R}, \textbf{Owens SM},
+\textbf{Klemmer S}, \textbf{Matamoros S}, \textbf{Mirarab S},
+\textbf{Moorman S}, \textbf{Holmes S}, \textbf{Schwartz T},
+\textbf{Eshoo-Anton TW}, \textbf{Vigers T}, \textbf{Pandey V},
+\textbf{Treuren WV}, \textbf{Fang X}, \textbf{Xu ZZ}, \textbf{Jarmusch
+A}, \textbf{Geier J}, \textbf{Reeve N}, \textbf{Silva R},
+\textbf{Kopylova E}, \textbf{Nguyen D}, \textbf{Sanders K},
+\textbf{Benitez RAS}, \textbf{Heale AC}, \textbf{Abramson M},
+\textbf{Waldispühl J}, \textbf{Butyaev A}, \textbf{Drogaris C},
+\textbf{Nazarova E}, \textbf{Ball M}, \textbf{Gunderson B}. 2018.
+American gut: an open platform for citizen science microbiome research.
+mSystems \textbf{3}.
+doi:\href{http://dx.doi.org/10.1128/msystems.00031-18}{10.1128/msystems.00031-18}.
+
+9. \textbf{Schloss PD}, \textbf{Gevers D}, \textbf{Westcott SL}. 2011.
+Reducing the effects of PCR amplification and sequencing artifacts on
+16S rRNA-based studies. PLoS ONE \textbf{6}:e27310.
+doi:\href{http://dx.doi.org/10.1371/journal.pone.0027310}{10.1371/journal.pone.0027310}.
+
+10. \textbf{Schloss PD}. 2018. Identifying and overcoming threats to
+reproducibility, replicability, robustness, and generalizability in
+microbiome research. mBio \textbf{9}.
+doi:\href{http://dx.doi.org/10.1128/mbio.00525-18}{10.1128/mbio.00525-18}.
+
+11. \textbf{Sogin ML}, \textbf{Morrison HG}, \textbf{Huber JA},
+\textbf{Welch DM}, \textbf{Huse SM}, \textbf{Neal PR}, \textbf{Arrieta
+JM}, \textbf{Herndl GJ}. 2006. Microbial diversity in the deep sea and
+the underexplored ``rare biosphere''. Proceedings of the National
+Academy of Sciences \textbf{103}:12115--12120.
+doi:\href{http://dx.doi.org/10.1073/pnas.0605127103}{10.1073/pnas.0605127103}.
+
+12. \textbf{Huse SM}, \textbf{Welch DM}, \textbf{Morrison HG},
+\textbf{Sogin ML}. 2010. Ironing out the wrinkles in the rare biosphere
+through improved OTU clustering. Environmental Microbiology
+\textbf{12}:1889--1898.
+doi:\href{http://dx.doi.org/10.1111/j.1462-2920.2010.02193.x}{10.1111/j.1462-2920.2010.02193.x}.
+
+13. \textbf{Reeder J}, \textbf{Knight R}. 2009. The rare biosphere: a
+reality check. Nature Methods \textbf{6}:636--637.
+doi:\href{http://dx.doi.org/10.1038/nmeth0909-636}{10.1038/nmeth0909-636}.
+
+14. \textbf{Reid A}, \textbf{Buckley M}. 2011. The rare biosphere.
+
 \end{document}

### 2020-11-05 submission/manuscript.Rmd
Figure legends and figure edits

@@
 
-<!--- \includegraphics{figure_1.png} -->
+\includegraphics{figure_1.png}
 
-**Figure 1.**
+**Figure 1.
+Singletons are more common in samples with fewer seqeunces and tend to be shared with samples having more sequences.** For each of the 12 datasets, Spearman correlation coefficients were calcualted between the number of sequences in each sample and the number of singletons in the sample (A) and the fraction of its singletons that were shared with another sample (C).
+Those correlations that were not statisically significant had a P-value greater than 0.05.
+The faction of singletons shared across samples (B) were calculated for each dataset.
+The median value is shown with a solid circle and the 95% confidence interval is indicated by the solid line.
 
@@
 
-<!--- \includegraphics{figure_2.png} -->
+\includegraphics{figure_2.png}
 
-**Figure 2.**
+**Figure 2.
+Removing rare sequences from samples alters their representation of alpha-diversity using amplicon sequence variants (ASVs).** The average difference in the richness (A), Shannon diversity (B), and Kullback-Leiber divergence (C) for each sample within a dataset was calculated between the original community structures relative to applying different minimum abundance thresholds.
+
+\newpage
+
+\includegraphics{figure_3.png}
+
+**Figure 3.
+Removing rare sequences from samples increases the inter-sample variation for amplicon sequence variants (ASVs).** The coefficient of variation in richness (A), Shannon diversity (B), and Bray-Curtis distances (C) for each dataset was calculated using the null distributed samples for each dataset with varying minimum abundance thresholds.
+
+\newpage
+
+\includegraphics{figure_4.png}
+
+**Figure 4.
+Removing rare sequences from samples reduces the statistical power to detect differences between empirically generated treatment groups when using amplicon sequence variants (ASVs).** The fraction of significant tests comparing the richness (A) and Shannon diversity (B) using a Wilcox test and Bray-Curtis distances (C) using analysis of molecular variance for each dataset was calculated using empirically generated treatment groups containing equal numbers of samples for each dataset with varying minimum abundance thresholds.
+For each dataset and minimum abundance threshold, 100 randomizations were peformed.
+
+\newpage
+
+\includegraphics{figure_5.png}
+
+**Figure 5.
+Removing rare sequences does not impact the false detection rate unless the number of sequences per sample is confounded with the treatment groups when using amplicon sequence variants (ASVs).** The fraction of significant tests comparing the richness and Shannon diversity using a Wilcox test and Bray-Curtis distances using analysis of molecular variance for each dataset was calculated.
+Empirically generated treatment groups were generated containing equal numbers of samples where the samples represented a null distribution.
+In one simulation the samples were randomly assigned to a treatment group (A) and in the other the samples were assigned based on the number of sequences in each sample (B).
+For each dataset and minimum abundance threshold, 100 randomizations were peformed.
+
+\newpage
+
+\includegraphics{figure_s1.png}
+
+**Figure S1.
+Distribution of the number of sequences per sample in the 12 datasets included in this study.** A different minimum number of sequences per sample threshold was applied to each dataset based on identifying natural breaks in the distribution of the number of sequences per sample.
+
+\newpage
+
+\includegraphics{figure_s2.png}
+
+**Figure S2.
+Removing rare sequences from samples alters their representation of alpha-diversity using operational taxonomic units (OTUs).** The average difference in the richness (A), Shannon diversity (B), and Kullback-Leiber divergence (C) for each sample within a dataset was calculated between the original community structures relative to applying different minimum abundance thresholds.
+
+\newpage
+
+\includegraphics{figure_s3.png}
+
+**Figure S3.
+Removing rare sequences from samples alters their representation of alpha-diversity when regenerating samples using a null distribution for each dataset.** The average difference in the richness, Shannon diversity, and Kullback-Leiber divergence for each sample within a dataset was calculated between the original community structures relative to applying different minimum abundance thresholds.
+Some values of Kullback-Leiber divergence are missing because undefined values were calculated due to the removal of rare sequences.
+Data are shown for amplicon sequence variants (ASVs) and operational taxonomic units (OTUs)
+
+\newpage
+
+\includegraphics{figure_s4.png}
+
+**Figure S4.
+Removing rare sequences from samples increases the inter-sample variation for operational taxonomic units (OTUs).** The coefficient of variation in richness (A), Shannon diversity (B), and Bray-Curtis distances (C) for each dataset was calculated using the null distributed samples for each dataset with varying minimum abundance thresholds.
+
+\newpage
+
+\includegraphics{figure_s5.png}
+
+**Figure S5.
+Removing rare sequences from samples reduces the statistical power to detect differences between empirically generated treatment groups when using operational taxonomic units (OTUs).** The fraction of significant tests comparing the richness (A) and Shannon diversity (B) using a Wilcox test and Bray-Curtis distances (C) using analysis of molecular variance for each dataset was calculated using empirically generated treatment groups containing equal numbers of samples for each dataset with varying minimum abundance thresholds.
+For each dataset and minimum abundance threshold, 100 randomizations were peformed.
+
+\newpage
+
+\includegraphics{figure_s6.png}
+
+**Figure S6.
+Removing rare sequences does not impact the false detection rate unless the number of sequences per sample is confounded with the treatment groups when using operational taxonomic units (OTUs).** The fraction of significant tests comparing the richness and Shannon diversity using a Wilcox test and Bray-Curtis distances using analysis of molecular variance for each dataset was calculated.
+Empirically generated treatment groups were generated containing equal numbers of samples where the samples represented a null distribution.
+In one simulation the samples were randomly assigned to a treatment group (A) and in the other the samples were assigned based on the number of sequences in each sample (B).
+For each dataset and minimum abundance threshold, 100 randomizations were peformed.

### 2020-11-05 submission/manuscript.tex
Figure legends and figure edits

@@
-\documentclass[]{article}
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+%
+\documentclass[
+]{article}
+\usepackage{lmodern}
+\usepackage{amsmath}
+\usepackage{ifxetex,ifluatex}
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
   \usepackage[T1]{fontenc}
-\usepackage{lmodern}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
-% use upquote if available, for straight quotes in verbatim environments
+  \usepackage[utf8]{inputenc}
+  \usepackage{textcomp} % provide euro and other symbols
+  \usepackage{amssymb}
+\else % if luatex or xetex
+  \usepackage{unicode-math}
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+% Use upquote if available, for straight quotes in verbatim environments
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[utf8]{inputenc}
-\else % if luatex or xelatex
-  \ifxetex
-    \usepackage{mathspec}
-    \usepackage{xltxtra,xunicode}
-  \else
-    \usepackage{fontspec}
-  \fi
-  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
-  \newcommand{\euro}{€}
-\fi
-% use microtype if available
-\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
+\IfFileExists{microtype.sty}{% use microtype if available
+  \usepackage[]{microtype}
+  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\hypersetup{
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
+\urlstyle{same} % disable monospaced font for URLs
 \usepackage[margin=1.0in]{geometry}
 \usepackage{graphicx}
-% Redefine \includegraphics so that, unless explicit options are
-% given, the image width will not exceed the width of the page.
-% Images get their normal width if they fit onto the page, but
-% are scaled down if they would overflow the margins.
 \makeatletter
-\def\ScaleIfNeeded{%
-  \ifdim\Gin@nat@width>\linewidth
-    \linewidth
-  \else
-    \Gin@nat@width
-  \fi
-}
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
 \makeatother
-\let\Oldincludegraphics\includegraphics
-{%
- \catcode`\@=11\relax%
- \gdef\includegraphics{\@ifnextchar[{\Oldincludegraphics}{\Oldincludegraphics[width=\ScaleIfNeeded]}}%
-}%
-\ifxetex
-  \usepackage[setpagesize=false, % page size defined by xetex
-              unicode=false, % unicode breaks when used with xetex
-              xetex]{hyperref}
-\else
-  \usepackage[unicode=true]{hyperref}
-\fi
-\hypersetup{breaklinks=true,
-            bookmarks=true,
-            pdfauthor={},
-            pdftitle={},
-            colorlinks=true,
-            citecolor=blue,
-            urlcolor=blue,
-            linkcolor=magenta,
-            pdfborder={0 0 0}}
-\urlstyle{same}  % don't use monospace font for urls
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
 \setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{0}
-
-\author{}
-\date{\vspace{-2.5em}}
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 \usepackage{booktabs}
@@
 \usepackage{makecell}
+\ifluatex
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newenvironment{CSLReferences}[3] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
+  % set entry spacing
+  \ifnum #2 > 0
+  \setlength{\parskip}{#2\baselineskip}
+  \fi
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+
+\author{}
+\date{\vspace{-2.5em}}
 
@@
 
+\hypertarget{removal-of-rare-sequences-from-16s-rrna-gene-sequence-surveys-biases-the-interpretation-of-community-structure-data}{%
 \section{Removal of rare sequences from 16S rRNA gene sequence surveys
 biases the interpretation of community structure
-data}\label{removal-of-rare-sequences-from-16s-rrna-gene-sequence-surveys-biases-the-interpretation-of-community-structure-data}
+data}\label{removal-of-rare-sequences-from-16s-rrna-gene-sequence-surveys-biases-the-interpretation-of-community-structure-data}}
 
@@
 
-Patrick D. Schloss${^\dagger}$
+Patrick D. Schloss\({^\dagger}\)
 
@@
 
-$\dagger$ To whom correspondence should be addressed:
+\(\dagger\) To whom correspondence should be addressed:
 
@@
 
-\subsection{Research article format}\label{research-article-format}
+\hypertarget{research-article-format}{%
+\subsection{Research article format}\label{research-article-format}}
 
@@
 
-\subsection{Abstract}\label{abstract}
+\hypertarget{abstract}{%
+\subsection{Abstract}\label{abstract}}
 
@@
 
-\subsection{Importance}\label{importance}
+\hypertarget{importance}{%
+\subsection{Importance}\label{importance}}
 
@@
 
-\subsection{Results}\label{results}
+\hypertarget{results}{%
+\subsection{Results}\label{results}}
 
@@
 samples from the marine and sediment datasets (Spearman correlation,
-P\textgreater{}0.05), the fraction of singleton ASVs in samples was
+P\textgreater0.05), the fraction of singleton ASVs in samples was
 negatively correlated with the number of sequences in each sample with a
@@
 datasets without a significant correlation (Spearman correlation,
-P\textgreater{}0.05), the marine and soil datasets had the fewest
-samples in our collection and the stream dataset already had a high
-level of coverage regardless of the number of sequences. Contrary to the
-common motivation for removing rare ASVs, these results indicate that
-this practice disproportionately impacts samples with fewer sequences
-and likely removes more non-artifact ASVs than those that are artifacts.
+P\textgreater0.05), the marine and soil datasets had the fewest samples
+in our collection and the stream dataset already had a high level of
+coverage regardless of the number of sequences. Contrary to the common
+motivation for removing rare ASVs, these results indicate that this
+practice disproportionately impacts samples with fewer sequences and
+likely removes more non-artifact ASVs than those that are artifacts.
 
@@
 
-\subsection{Discussion}\label{discussion}
+\hypertarget{discussion}{%
+\subsection{Discussion}\label{discussion}}
 
@@
 
-\subsection{Materials and Methods}\label{materials-and-methods}
+\hypertarget{materials-and-methods}{%
+\subsection{Materials and Methods}\label{materials-and-methods}}
 
 \begin{itemize}
-\itemsep1pt\parskip0pt\parsep0pt
+\tightlist
 \item
@@
 
-\subsection{Acknowledgements}\label{acknowledgements}
+\hypertarget{acknowledgements}{%
+\subsection{Acknowledgements}\label{acknowledgements}}
 
@@
 
-\subsection{References}\label{references}
+\hypertarget{references}{%
+\subsection{References}\label{references}}
 
-\newpage
+\hypertarget{refs}{}
+\begin{CSLReferences}{0}{0}
+\leavevmode\hypertarget{ref-Kozich2013}{}%
+\CSLLeftMargin{1. }
+\CSLRightInline{\textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter
+NT}, \textbf{Highlander SK}, \textbf{Schloss PD}. 2013. {Development of
+a dual-index sequencing strategy and curation pipeline for analyzing
+amplicon sequence data on the MiSeq Illumina sequencing platform}.
+Applied and environmental microbiology \textbf{79}:5112--5120.}
 
-\textbf{Table 1. Summary of studies used in the analysis.} For all
-studies, the number of sequences used from each study was rarefied to
-the smallest sample size. A graphical represenation of the distribution
-of sample sizes for each study and the samples that were removed from
-each study are provided in Figure S1.
-
-\begin{tabular}{lrrrrr}
-\toprule
-\textbf{Study} & \textbf{Samples} & \textbf{\makecell[c]{Total\\sequences}} & \textbf{\makecell[c]{Median\\sequences}} & \textbf{\makecell[c]{Range of\\sequences}} & \textbf{\makecell[c]{Fold-difference\\between largest\\and smallest sample}}\\
-\midrule
-Bioethanol & 95 & 3,972,943 & 16,015 & 3,688-356,136 & 96.6\\
-Human & 490 & 20,909,768 & 32,505 & 10,523-430,415 & 40.9\\
-Lake & 52 & 3,169,868 & 69,041 & 15,347-112,871 & 7.4\\
-Marine & 7 & 1,391,396 & 193,464 & 133,516-254,060 & 1.9\\
-Mice & 348 & 2,813,747 & 6,477 & 1,804-30,565 & 16.9\\
-Peromyscus & 111 & 1,555,545 & 12,446 & 4,464-33,644 & 7.5\\
-Rainforest & 69 & 946,295 & 11,561 & 4,932-37,767 & 7.7\\
-Rice & 490 & 22,591,168 & 43,216 & 2,776-193,464 & 69.7\\
-Seagrass & 286 & 4,130,454 & 13,567 & 1,803-45,191 & 25.1\\
-Sediment & 58 & 1,154,174 & 17,584 & 7,685-68,321 & 8.9\\
-Soil & 18 & 956,656 & 51,844 & 47,806-59,956 & 1.3\\
-Stream & 201 & 21,162,574 & 90,159 & 9,175-390,964 & 42.6\\
-\bottomrule
-\end{tabular}
-
-\newpage
-
-\textbf{Figure 1.}
-
-\newpage
-
-\textbf{Figure 2.}
-
-1. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
-\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
-dual-index sequencing strategy and curation pipeline for analyzing
-amplicon sequence data on the MiSeq Illumina sequencing platform.
-Applied and environmental microbiology \textbf{79}:5112--5120.
-
-2. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
-\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
-\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
-\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
-\textbf{Thallinger GG}, \textbf{Horn DJV}, \textbf{Weber CF}. 2009.
+\leavevmode\hypertarget{ref-Schloss2009}{}%
+\CSLLeftMargin{2. }
+\CSLRightInline{\textbf{Schloss PD}, \textbf{Westcott SL},
+\textbf{Ryabin T}, \textbf{Hall JR}, \textbf{Hartmann M},
+\textbf{Hollister EB}, \textbf{Lesniewski RA}, \textbf{Oakley BB},
+\textbf{Parks DH}, \textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres
+B}, \textbf{Thallinger GG}, \textbf{Horn DJV}, \textbf{Weber CF}. 2009.
 Introducing mothur: Open-source, platform-independent,
@@
 \textbf{75}:7537--7541.
-doi:\href{http://dx.doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.
+doi:\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.}
 
-3. \textbf{Morris A}, \textbf{Beck JM}, \textbf{Schloss PD},
-\textbf{Campbell TB}, \textbf{Crothers K}, \textbf{Curtis JL},
+\leavevmode\hypertarget{ref-Morris2013}{}%
+\CSLLeftMargin{3. }
+\CSLRightInline{\textbf{Morris A}, \textbf{Beck JM}, \textbf{Schloss
+PD}, \textbf{Campbell TB}, \textbf{Crothers K}, \textbf{Curtis JL},
 \textbf{Flores SC}, \textbf{Fontenot AP}, \textbf{Ghedin E},
@@
 Critical Care Medicine \textbf{187}:1067--1075.
-doi:\href{http://dx.doi.org/10.1164/rccm.201210-1913oc}{10.1164/rccm.201210-1913oc}.
+doi:\href{https://doi.org/10.1164/rccm.201210-1913oc}{10.1164/rccm.201210-1913oc}.}
 
-4. \textbf{Bokulich NA}, \textbf{Subramanian S}, \textbf{Faith JJ},
-\textbf{Gevers D}, \textbf{Gordon JI}, \textbf{Knight R}, \textbf{Mills
-DA}, \textbf{Caporaso JG}. 2013. Quality-filtering vastly improves
-diversity estimates from illumina amplicon sequencing. Nature Methods
-\textbf{10}:57--59.
-doi:\href{http://dx.doi.org/10.1038/nmeth.2276}{10.1038/nmeth.2276}.
+\leavevmode\hypertarget{ref-Bokulich2013}{}%
+\CSLLeftMargin{4. }
+\CSLRightInline{\textbf{Bokulich NA}, \textbf{Subramanian S},
+\textbf{Faith JJ}, \textbf{Gevers D}, \textbf{Gordon JI}, \textbf{Knight
+R}, \textbf{Mills DA}, \textbf{Caporaso JG}. 2013. Quality-filtering
+vastly improves diversity estimates from illumina amplicon sequencing.
+Nature Methods \textbf{10}:57--59.
+doi:\href{https://doi.org/10.1038/nmeth.2276}{10.1038/nmeth.2276}.}
 
-5. \textbf{Fadrosh DW}, \textbf{Ma B}, \textbf{Gajer P},
+\leavevmode\hypertarget{ref-Fadrosh2014}{}%
+\CSLLeftMargin{5. }
+\CSLRightInline{\textbf{Fadrosh DW}, \textbf{Ma B}, \textbf{Gajer P},
 \textbf{Sengamalay N}, \textbf{Ott S}, \textbf{Brotman RM},
 \textbf{Ravel J}. 2014. An improved dual-indexing approach for
-multiplexed 16S rRNA gene sequencing on the illumina MiSeq platform.
+multiplexed 16S {rRNA} gene sequencing on the illumina {MiSeq} platform.
 Microbiome \textbf{2}:6.
-doi:\href{http://dx.doi.org/10.1186/2049-2618-2-6}{10.1186/2049-2618-2-6}.
+doi:\href{https://doi.org/10.1186/2049-2618-2-6}{10.1186/2049-2618-2-6}.}
 
-6. \textbf{Holm JB}, \textbf{Humphrys MS}, \textbf{Robinson CK},
-\textbf{Settles ML}, \textbf{Ott S}, \textbf{Fu L}, \textbf{Yang H},
-\textbf{Gajer P}, \textbf{He X}, \textbf{McComb E}, \textbf{Gravitt PE},
-\textbf{Ghanem KG}, \textbf{Brotman RM}, \textbf{Ravel J}. 2019.
-Ultrahigh-throughput multiplexing and sequencing of $>$500-base-pair
-amplicon regions on the Illumina HiSeq 2500 platform. mSystems
+\leavevmode\hypertarget{ref-Holm2019}{}%
+\CSLLeftMargin{6. }
+\CSLRightInline{\textbf{Holm JB}, \textbf{Humphrys MS}, \textbf{Robinson
+CK}, \textbf{Settles ML}, \textbf{Ott S}, \textbf{Fu L}, \textbf{Yang
+H}, \textbf{Gajer P}, \textbf{He X}, \textbf{McComb E}, \textbf{Gravitt
+PE}, \textbf{Ghanem KG}, \textbf{Brotman RM}, \textbf{Ravel J}. 2019.
+Ultrahigh-throughput multiplexing and sequencing of \(>\)500-base-pair
+amplicon regions on the {Illumina} {HiSeq} 2500 platform. {mSystems}
 \textbf{4}.
-doi:\href{http://dx.doi.org/10.1128/msystems.00029-19}{10.1128/msystems.00029-19}.
+doi:\href{https://doi.org/10.1128/msystems.00029-19}{10.1128/msystems.00029-19}.}
 
-7. \textbf{Thompson LR}, \textbf{Jon G. Sanders}, \textbf{McDonald D},
-\textbf{Amir A}, \textbf{Ladau J}, \textbf{Locey KJ}, \textbf{Prill RJ},
-\textbf{Tripathi A}, \textbf{Gibbons SM}, \textbf{Ackermann G},
-\textbf{Navas-Molina JA}, \textbf{Janssen S}, \textbf{Kopylova E},
-\textbf{V{á}zquez-Baeza Y}, \textbf{Gonz{á}lez A}, \textbf{Morton JT},
-\textbf{Mirarab S}, \textbf{Xu ZZ}, \textbf{Jiang L}, \textbf{Haroon
-MF}, \textbf{Kanbar J}, \textbf{Zhu Q}, \textbf{Song SJ},
-\textbf{Kosciolek T}, \textbf{Bokulich NA}, \textbf{Lefler J},
+\leavevmode\hypertarget{ref-Thompson2017}{}%
+\CSLLeftMargin{7. }
+\CSLRightInline{\textbf{Thompson LR}, \textbf{Jon G. Sanders and},
+\textbf{McDonald D}, \textbf{Amir A}, \textbf{Ladau J}, \textbf{Locey
+KJ}, \textbf{Prill RJ}, \textbf{Tripathi A}, \textbf{Gibbons SM},
+\textbf{Ackermann G}, \textbf{Navas-Molina JA}, \textbf{Janssen S},
+\textbf{Kopylova E}, \textbf{Vázquez-Baeza Y}, \textbf{González A},
+\textbf{Morton JT}, \textbf{Mirarab S}, \textbf{Xu ZZ}, \textbf{Jiang
+L}, \textbf{Haroon MF}, \textbf{Kanbar J}, \textbf{Zhu Q}, \textbf{Song
+SJ}, \textbf{Kosciolek T}, \textbf{Bokulich NA}, \textbf{Lefler J},
 \textbf{Brislawn CJ}, \textbf{Humphrey G}, \textbf{Owens SM},
@@
 microbial diversity. Nature \textbf{551}:457--463.
-doi:\href{http://dx.doi.org/10.1038/nature24621}{10.1038/nature24621}.
+doi:\href{https://doi.org/10.1038/nature24621}{10.1038/nature24621}.}
 
-8. \textbf{McDonald D}, \textbf{Hyde E}, \textbf{Debelius JW},
-\textbf{Morton JT}, \textbf{Gonzalez A}, \textbf{Ackermann G},
+\leavevmode\hypertarget{ref-McDonald2018}{}%
+\CSLLeftMargin{8. }
+\CSLRightInline{\textbf{McDonald D}, \textbf{Hyde E}, \textbf{Debelius
+JW}, \textbf{Morton JT}, \textbf{Gonzalez A}, \textbf{Ackermann G},
 \textbf{Aksenov AA}, \textbf{Behsaz B}, \textbf{Brennan C}, \textbf{Chen
@@
 AD}, \textbf{Thackray VG}, \textbf{Thompson LR}, \textbf{Tripathi A},
-\textbf{V{á}zquez-Baeza Y}, \textbf{Vrbanac A}, \textbf{Wischmeyer P},
+\textbf{Vázquez-Baeza Y}, \textbf{Vrbanac A}, \textbf{Wischmeyer P},
 \textbf{Wolfe E}, \textbf{Zhu Q}, \textbf{Knight R}, \textbf{Mann AE},
@@
 \textbf{Nazarova E}, \textbf{Ball M}, \textbf{Gunderson B}. 2018.
-American gut: an open platform for citizen science microbiome research.
-mSystems \textbf{3}.
-doi:\href{http://dx.doi.org/10.1128/msystems.00031-18}{10.1128/msystems.00031-18}.
+American gut: An open platform for citizen science microbiome research.
+{mSystems} \textbf{3}.
+doi:\href{https://doi.org/10.1128/msystems.00031-18}{10.1128/msystems.00031-18}.}
 
-9. \textbf{Schloss PD}, \textbf{Gevers D}, \textbf{Westcott SL}. 2011.
-Reducing the effects of PCR amplification and sequencing artifacts on
-16S rRNA-based studies. PLoS ONE \textbf{6}:e27310.
-doi:\href{http://dx.doi.org/10.1371/journal.pone.0027310}{10.1371/journal.pone.0027310}.
+\leavevmode\hypertarget{ref-Schloss2011}{}%
+\CSLLeftMargin{9. }
+\CSLRightInline{\textbf{Schloss PD}, \textbf{Gevers D}, \textbf{Westcott
+SL}. 2011. Reducing the effects of {PCR} amplification and sequencing
+artifacts on 16S {rRNA}-based studies. {PLoS} {ONE} \textbf{6}:e27310.
+doi:\href{https://doi.org/10.1371/journal.pone.0027310}{10.1371/journal.pone.0027310}.}
 
-10. \textbf{Schloss PD}. 2018. Identifying and overcoming threats to
-reproducibility, replicability, robustness, and generalizability in
-microbiome research. mBio \textbf{9}.
-doi:\href{http://dx.doi.org/10.1128/mbio.00525-18}{10.1128/mbio.00525-18}.
+\leavevmode\hypertarget{ref-Schloss2018}{}%
+\CSLLeftMargin{10. }
+\CSLRightInline{\textbf{Schloss PD}. 2018. Identifying and overcoming
+threats to reproducibility, replicability, robustness, and
+generalizability in microbiome research. {mBio} \textbf{9}.
+doi:\href{https://doi.org/10.1128/mbio.00525-18}{10.1128/mbio.00525-18}.}
 
-11. \textbf{Sogin ML}, \textbf{Morrison HG}, \textbf{Huber JA},
-\textbf{Welch DM}, \textbf{Huse SM}, \textbf{Neal PR}, \textbf{Arrieta
-JM}, \textbf{Herndl GJ}. 2006. Microbial diversity in the deep sea and
-the underexplored ``rare biosphere''. Proceedings of the National
-Academy of Sciences \textbf{103}:12115--12120.
-doi:\href{http://dx.doi.org/10.1073/pnas.0605127103}{10.1073/pnas.0605127103}.
+\leavevmode\hypertarget{ref-Sogin2006}{}%
+\CSLLeftMargin{11. }
+\CSLRightInline{\textbf{Sogin ML}, \textbf{Morrison HG}, \textbf{Huber
+JA}, \textbf{Welch DM}, \textbf{Huse SM}, \textbf{Neal PR},
+\textbf{Arrieta JM}, \textbf{Herndl GJ}. 2006. Microbial diversity in
+the deep sea and the underexplored "rare biosphere". Proceedings of the
+National Academy of Sciences \textbf{103}:12115--12120.
+doi:\href{https://doi.org/10.1073/pnas.0605127103}{10.1073/pnas.0605127103}.}
 
-12. \textbf{Huse SM}, \textbf{Welch DM}, \textbf{Morrison HG},
-\textbf{Sogin ML}. 2010. Ironing out the wrinkles in the rare biosphere
-through improved OTU clustering. Environmental Microbiology
+\leavevmode\hypertarget{ref-Huse2010}{}%
+\CSLLeftMargin{12. }
+\CSLRightInline{\textbf{Huse SM}, \textbf{Welch DM}, \textbf{Morrison
+HG}, \textbf{Sogin ML}. 2010. Ironing out the wrinkles in the rare
+biosphere through improved {OTU} clustering. Environmental Microbiology
 \textbf{12}:1889--1898.
-doi:\href{http://dx.doi.org/10.1111/j.1462-2920.2010.02193.x}{10.1111/j.1462-2920.2010.02193.x}.
+doi:\href{https://doi.org/10.1111/j.1462-2920.2010.02193.x}{10.1111/j.1462-2920.2010.02193.x}.}
 
-13. \textbf{Reeder J}, \textbf{Knight R}. 2009. The rare biosphere: a
-reality check. Nature Methods \textbf{6}:636--637.
-doi:\href{http://dx.doi.org/10.1038/nmeth0909-636}{10.1038/nmeth0909-636}.
+\leavevmode\hypertarget{ref-Reeder2009}{}%
+\CSLLeftMargin{13. }
+\CSLRightInline{\textbf{Reeder J}, \textbf{Knight R}. 2009. The {{}}rare
+biosphere{{}}: A reality check. Nature Methods \textbf{6}:636--637.
+doi:\href{https://doi.org/10.1038/nmeth0909-636}{10.1038/nmeth0909-636}.}
 
-14. \textbf{Reid A}, \textbf{Buckley M}. 2011. The rare biosphere.
+\leavevmode\hypertarget{ref-Reid2011}{}%
+\CSLLeftMargin{14. }
+\CSLRightInline{\textbf{Reid A}, \textbf{Buckley M}. 2011. The rare
+biosphere.}
+
+\end{CSLReferences}
+
+\newpage
+
+\textbf{Table 1. Summary of studies used in the analysis.} For all
+studies, the number of sequences used from each study was rarefied to
+the smallest sample size. A graphical represenation of the distribution
+of sample sizes for each study and the samples that were removed from
+each study are provided in Figure S1.
+
+\begin{tabular}{lrrrrr}
+\toprule
+\textbf{Study} & \textbf{Samples} & \textbf{\makecell[c]{Total\\sequences}} & \textbf{\makecell[c]{Median\\sequences}} & \textbf{\makecell[c]{Range of\\sequences}} & \textbf{\makecell[c]{Fold-difference\\between largest\\and smallest sample}}\\
+\midrule
+Bioethanol & 95 & 3,972,943 & 16,015 & 3,688-356,136 & 96.6\\
+Human & 490 & 20,909,768 & 32,505 & 10,523-430,415 & 40.9\\
+Lake & 52 & 3,169,868 & 69,041 & 15,347-112,871 & 7.4\\
+Marine & 7 & 1,391,396 & 193,464 & 133,516-254,060 & 1.9\\
+Mice & 348 & 2,813,747 & 6,477 & 1,804-30,565 & 16.9\\
+Peromyscus & 111 & 1,555,545 & 12,446 & 4,464-33,644 & 7.5\\
+Rainforest & 69 & 946,295 & 11,561 & 4,932-37,767 & 7.7\\
+Rice & 490 & 22,591,168 & 43,216 & 2,776-193,464 & 69.7\\
+Seagrass & 286 & 4,130,454 & 13,567 & 1,803-45,191 & 25.1\\
+Sediment & 58 & 1,154,174 & 17,584 & 7,685-68,321 & 8.9\\
+Soil & 18 & 956,656 & 51,844 & 47,806-59,956 & 1.3\\
+Stream & 201 & 21,162,574 & 90,159 & 9,175-390,964 & 42.6\\
+\bottomrule
+\end{tabular}
+
+\newpage
+
+\includegraphics{figure_1.png}
+
+\textbf{Figure 1. Singletons are more common in samples with fewer
+seqeunces and tend to be shared with samples having more sequences.} For
+each of the 12 datasets, Spearman correlation coefficients were
+calcualted between the number of sequences in each sample and the number
+of singletons in the sample (A) and the fraction of its singletons that
+were shared with another sample (C). Those correlations that were not
+statisically significant had a P-value greater than 0.05. The faction of
+singletons shared across samples (B) were calculated for each dataset.
+The median value is shown with a solid circle and the 95\% confidence
+interval is indicated by the solid line.
+
+\newpage
+
+\includegraphics{figure_2.png}
+
+\textbf{Figure 2. Removing rare sequences from samples alters their
+representation of alpha-diversity using amplicon sequence variants
+(ASVs).} The average difference in the richness (A), Shannon diversity
+(B), and Kullback-Leiber divergence (C) for each sample within a dataset
+was calculated between the original community structures relative to
+applying different minimum abundance thresholds.
+
+\newpage
+
+\includegraphics{figure_3.png}
+
+\textbf{Figure 3. Removing rare sequences from samples increases the
+inter-sample variation for amplicon sequence variants (ASVs).} The
+coefficient of variation in richness (A), Shannon diversity (B), and
+Bray-Curtis distances (C) for each dataset was calculated using the null
+distributed samples for each dataset with varying minimum abundance
+thresholds.
+
+\newpage
+
+\includegraphics{figure_4.png}
+
+\textbf{Figure 4. Removing rare sequences from samples reduces the
+statistical power to detect differences between empirically generated
+treatment groups when using amplicon sequence variants (ASVs).} The
+fraction of significant tests comparing the richness (A) and Shannon
+diversity (B) using a Wilcox test and Bray-Curtis distances (C) using
+analysis of molecular variance for each dataset was calculated using
+empirically generated treatment groups containing equal numbers of
+samples for each dataset with varying minimum abundance thresholds. For
+each dataset and minimum abundance threshold, 100 randomizations were
+peformed.
+
+\newpage
+
+\includegraphics{figure_5.png}
+
+\textbf{Figure 5. Removing rare sequences does not impact the false
+detection rate unless the number of sequences per sample is confounded
+with the treatment groups when using amplicon sequence variants (ASVs).}
+The fraction of significant tests comparing the richness and Shannon
+diversity using a Wilcox test and Bray-Curtis distances using analysis
+of molecular variance for each dataset was calculated. Empirically
+generated treatment groups were generated containing equal numbers of
+samples where the samples represented a null distribution. In one
+simulation the samples were randomly assigned to a treatment group (A)
+and in the other the samples were assigned based on the number of
+sequences in each sample (B). For each dataset and minimum abundance
+threshold, 100 randomizations were peformed.
+
+\newpage
+
+\includegraphics{figure_s1.png}
+
+\textbf{Figure S1. Distribution of the number of sequences per sample in
+the 12 datasets included in this study.} A different minimum number of
+sequences per sample threshold was applied to each dataset based on
+identifying natural breaks in the distribution of the number of
+sequences per sample.
+
+\newpage
+
+\includegraphics{figure_s2.png}
+
+\textbf{Figure S2. Removing rare sequences from samples alters their
+representation of alpha-diversity using operational taxonomic units
+(OTUs).} The average difference in the richness (A), Shannon diversity
+(B), and Kullback-Leiber divergence (C) for each sample within a dataset
+was calculated between the original community structures relative to
+applying different minimum abundance thresholds.
+
+\newpage
+
+\includegraphics{figure_s3.png}
+
+\textbf{Figure S3. Removing rare sequences from samples alters their
+representation of alpha-diversity when regenerating samples using a null
+distribution for each dataset.} The average difference in the richness,
+Shannon diversity, and Kullback-Leiber divergence for each sample within
+a dataset was calculated between the original community structures
+relative to applying different minimum abundance thresholds. Some values
+of Kullback-Leiber divergence are missing because undefined values were
+calculated due to the removal of rare sequences. Data are shown for
+amplicon sequence variants (ASVs) and operational taxonomic units (OTUs)
+
+\newpage
+
+\includegraphics{figure_s4.png}
+
+\textbf{Figure S4. Removing rare sequences from samples increases the
+inter-sample variation for operational taxonomic units (OTUs).} The
+coefficient of variation in richness (A), Shannon diversity (B), and
+Bray-Curtis distances (C) for each dataset was calculated using the null
+distributed samples for each dataset with varying minimum abundance
+thresholds.
+
+\newpage
+
+\includegraphics{figure_s5.png}
+
+\textbf{Figure S5. Removing rare sequences from samples reduces the
+statistical power to detect differences between empirically generated
+treatment groups when using operational taxonomic units (OTUs).} The
+fraction of significant tests comparing the richness (A) and Shannon
+diversity (B) using a Wilcox test and Bray-Curtis distances (C) using
+analysis of molecular variance for each dataset was calculated using
+empirically generated treatment groups containing equal numbers of
+samples for each dataset with varying minimum abundance thresholds. For
+each dataset and minimum abundance threshold, 100 randomizations were
+peformed.
+
+\newpage
+
+\includegraphics{figure_s6.png}
+
+\textbf{Figure S6. Removing rare sequences does not impact the false
+detection rate unless the number of sequences per sample is confounded
+with the treatment groups when using operational taxonomic units
+(OTUs).} The fraction of significant tests comparing the richness and
+Shannon diversity using a Wilcox test and Bray-Curtis distances using
+analysis of molecular variance for each dataset was calculated.
+Empirically generated treatment groups were generated containing equal
+numbers of samples where the samples represented a null distribution. In
+one simulation the samples were randomly assigned to a treatment group
+(A) and in the other the samples were assigned based on the number of
+sequences in each sample (B). For each dataset and minimum abundance
+threshold, 100 randomizations were peformed.
 

### 2020-11-05 code/plot_loss_of_information_obs.R

@@ -34,7 +34,7 @@ plot_data <- function(subset_method, subset_string) {
       labels=c(
         paste0("Reduction in\nrichness (%)"),
         paste0("Reduction in\nShannon diversity (%)"),
-        paste0("Kullbac-Leibler\ndivergence"))
+        paste0("Kullback-Leibler\ndivergence"))
       )
     )
 

### 2020-11-05 code/plot_loss_of_information_random.R

@@ -32,7 +32,7 @@ subset_data <- loss %>%
     labels=c(
       paste0("Reduction in\nrichness (%)"),
       paste0("Reduction in\nShannon diversity (%)"),
-      paste0("Kullbac-Leibler\ndivergence"))
+      paste0("Kullback-Leibler\ndivergence"))
     )
   ) %>%
   mutate(method = factor(method,

### 2020-11-05 code/plot_type1.R

@@ -64,7 +64,7 @@ plot_data <- function(subset_method, subset_string) {
       dataset = factor(dataset, levels=studies, labels=nice_studies),
       simulation = factor(simulation, levels=c("null", "skew"),
             labels=c(
-              "Samples randomly assignedto groups\nwithout regard for sample size",
+              "Samples randomly assigned to groups\nwithout regard for sample size",
               "Samples randomly assigned to groups\nbased on sample size"
             )
           )

### 2020-11-05 submission/manuscript.Rmd
Acknowledgements section
Incorporate citations into Table 1

@@
 This region and sequencing platform were selected because previous work has shown that a standard data analysis pipeline in mothur results in a sequencing error rate below 0.02% [@Kozich2013].
-All sequence data were obtained from the Sequence Read Archive and processed using a standard mothur-based sequencing pipeline that resulted in ASVs as generated by the pre.cluster algorithm [@Kozich2013; @Schloss2009].
+All sequence data were obtained from the Sequence Read Archive (SRA) and processed using a standard mothur-based sequencing pipeline that resulted in ASVs as generated by the pre.cluster algorithm [@Kozich2013; @Schloss2009].
 After removing poor quality and chimeric ASVs and samples that had uncharacteristically low number of sequences for the dataset, these datasets included between `r oxford_comma(n_sample_range)` samples (Figure S1).
@@
 
-\newpage
-
 ## Materials and Methods
 
-* sequencing pipeline description
+Datasets were selected because they sequenced the V4 region of the 16S rRNA gene using Illumina's MiSeq platform using the v2 chemistry with paired 250 nt reads.
+I identified 12 datasets that satisfied these criteria and downloaded the data directly from the SRA (Table 1 and Figure S1).
+
+* Used the mothur pipeline to
+  - Assembly of paired 250 nt reads
+*
 
 ## Acknowledgements
+
+I endebted to the researchers who developed the 12 datasets used in this study for depositing their sequence data into the Sequence Read Archive.
+This work was supported in part by funding from the National Institutes of Health (U01AI124255, P30DK034933, R01CA215574).
 
@@
 
+\small
+
 ```{r}
@@
   arrange(nice_name) %>%
-  mutate(total_seqs=format(total_seqs, big.mark=",", trim=TRUE),
+  mutate(
+    nice_name = glue("{nice_name} {reference}"),
+    total_seqs=format(total_seqs, big.mark=",", trim=TRUE),
     median=format(as.integer(median), big.mark=",", trim=TRUE),
@@
     fold_difference = round(fold_difference, 1)) %>%
-  select(nice_name, n_samples, total_seqs, median, range, fold_difference) %>%
-  kable("latex", booktabs=TRUE, escape=F, align="lrrrrr", linesep="",
-		col.names = linebreak(c("Study", "Samples", "Total\nsequences", "Median\nsequences",
-			"Range of\nsequences", "Fold-difference\nbetween largest\nand smallest sample"), align="c")
-	) %>%
-	row_spec(0, bold=T)
+  select(nice_name, n_samples, total_seqs, median, range, fold_difference, sra_study) %>%
+  kableExtra::kable(format="markdown", booktabs=TRUE, escape=F, align="lrrrrr",
+    linesep="",
+		col.names = linebreak(
+      c(
+        "\\textbf{Study (Ref)}",
+        "\\textbf{Samples}",
+        "\\textbf{Total}\n\\textbf{sequences}",
+        "\\textbf{Median}\n\\textbf{sequences}",
+        "\\textbf{Range of}\n\\textbf{sequences}",
+        "\\textbf{Fold-difference}\n\\textbf{between largest}\n\\textbf{and smallest sample}",
+        "\\textbf{SRA study}\n\\textbf{accession}"
+      ),
+      align="c")
+	)
 ```
+\normalsize
 

### 2020-11-05 submission/manuscript.tex
Acknowledgements section
Incorporate citations into Table 1

@@
 \usepackage[margin=1.0in]{geometry}
+\usepackage{longtable,booktabs}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
 \usepackage{graphicx}
@@
 rate below 0.02\% (1). All sequence data were obtained from the Sequence
-Read Archive and processed using a standard mothur-based sequencing
-pipeline that resulted in ASVs as generated by the pre.cluster algorithm
-(1, 2). After removing poor quality and chimeric ASVs and samples that
-had uncharacteristically low number of sequences for the dataset, these
-datasets included between 7 and 490 samples (Figure S1). The median
-number of sequences for each dataset ranged between 6,477 and 193,464
-(Table 1). Strikingly, aside from the relatively small marine and soil
-datasets, the difference between the sample with the fewest sequences
-and the sample with the most sequences for each dataset varied by
-between 7.4 and 96.6-fold (Table 1).
+Read Archive (SRA) and processed using a standard mothur-based
+sequencing pipeline that resulted in ASVs as generated by the
+pre.cluster algorithm (1, 2). After removing poor quality and chimeric
+ASVs and samples that had uncharacteristically low number of sequences
+for the dataset, these datasets included between 7 and 490 samples
+(Figure S1). The median number of sequences for each dataset ranged
+between 6,477 and 193,464 (Table 1). Strikingly, aside from the
+relatively small marine and soil datasets, the difference between the
+sample with the fewest sequences and the sample with the most sequences
+for each dataset varied by between 7.4 and 96.6-fold (Table 1).
 
@@
 
-\newpage
-
 \hypertarget{materials-and-methods}{%
 \subsection{Materials and Methods}\label{materials-and-methods}}
+
+Datasets were selected because they sequenced the V4 region of the 16S
+rRNA gene using Illumina's MiSeq platform using the v2 chemistry with
+paired 250 nt reads. I identified 12 datasets that satisfied these
+criteria and downloaded the data directly from the SRA (Table 1 and
+Figure S1).
 
@@
 \item
-  sequencing pipeline description
+  Used the mothur pipeline to
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Assembly of paired 250 nt reads
+  \end{itemize}
+\item
 \end{itemize}
@@
 \subsection{Acknowledgements}\label{acknowledgements}}
+
+I endebted to the researchers who developed the 12 datasets used in this
+study for depositing their sequence data into the Sequence Read Archive.
+This work was supported in part by funding from the National Institutes
+of Health (U01AI124255, P30DK034933, R01CA215574).
 
@@
 
+\leavevmode\hypertarget{ref-Li2015}{}%
+\CSLLeftMargin{15. }
+\CSLRightInline{\textbf{Li Q}, \textbf{Heist EP}, \textbf{Moe LA}. 2015.
+Bacterial community structure and dynamics during corn-based bioethanol
+fermentation. Microbial Ecology \textbf{71}:409--421.
+doi:\href{https://doi.org/10.1007/s00248-015-0673-9}{10.1007/s00248-015-0673-9}.}
+
+\leavevmode\hypertarget{ref-Baxter2016}{}%
+\CSLLeftMargin{16. }
+\CSLRightInline{\textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers
+MAM}, \textbf{Schloss PD}. 2016. Microbiota-based model improves the
+sensitivity of fecal immunochemical test for detecting colonic lesions.
+Genome Medicine \textbf{8}.
+doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.}
+
+\leavevmode\hypertarget{ref-Beall2015}{}%
+\CSLLeftMargin{17. }
+\CSLRightInline{\textbf{Beall BFN}, \textbf{Twiss MR}, \textbf{Smith
+DE}, \textbf{Oyserman BO}, \textbf{Rozmarynowycz MJ}, \textbf{Binding
+CE}, \textbf{Bourbonniere RA}, \textbf{Bullerjahn GS}, \textbf{Palmer
+ME}, \textbf{Reavie ED}, \textbf{Waters LMK}, \textbf{Woityra LWC},
+\textbf{McKay RML}. 2015. Ice cover extent drives phytoplankton and
+bacterial community structure in a large north-temperate lake:
+Implications for a warming climate. Environmental Microbiology
+\textbf{18}:1704--1719.
+doi:\href{https://doi.org/10.1111/1462-2920.12819}{10.1111/1462-2920.12819}.}
+
+\leavevmode\hypertarget{ref-Henson2016}{}%
+\CSLLeftMargin{18. }
+\CSLRightInline{\textbf{Henson MW}, \textbf{Pitre DM}, \textbf{Weckhorst
+JL}, \textbf{Lanclos VC}, \textbf{Webber AT}, \textbf{Thrash JC}. 2016.
+Artificial seawater media facilitate cultivating members of the
+microbial majority from the gulf of mexico. {mSphere} \textbf{1}.
+doi:\href{https://doi.org/10.1128/msphere.00028-16}{10.1128/msphere.00028-16}.}
+
+\leavevmode\hypertarget{ref-Baxter2014}{}%
+\CSLLeftMargin{19. }
+\CSLRightInline{\textbf{Baxter NT}, \textbf{Wan JJ}, \textbf{Schubert
+AM}, \textbf{Jenior ML}, \textbf{Myers P}, \textbf{Schloss PD}. 2014.
+Intra- and interindividual variations mask interspecies variation in the
+microbiota of sympatric peromyscus populations. Applied and
+Environmental Microbiology \textbf{81}:396--404.
+doi:\href{https://doi.org/10.1128/aem.02303-14}{10.1128/aem.02303-14}.}
+
+\leavevmode\hypertarget{ref-LevyBooth2018}{}%
+\CSLLeftMargin{20. }
+\CSLRightInline{\textbf{Levy-Booth DJ}, \textbf{Giesbrecht IJW},
+\textbf{Kellogg CTE}, \textbf{Heger TJ}, \textbf{D'Amore DV},
+\textbf{Keeling PJ}, \textbf{Hallam SJ}, \textbf{Mohn WW}. 2018.
+Seasonal and ecohydrological regulation of active microbial populations
+involved in {DOC}, {CO}2, and {CH}4 fluxes in temperate rainforest soil.
+The {ISME} Journal \textbf{13}:950--963.
+doi:\href{https://doi.org/10.1038/s41396-018-0334-3}{10.1038/s41396-018-0334-3}.}
+
+\leavevmode\hypertarget{ref-Edwards2015}{}%
+\CSLLeftMargin{21. }
+\CSLRightInline{\textbf{Edwards J}, \textbf{Johnson C},
+\textbf{Santos-Medellín C}, \textbf{Lurie E}, \textbf{Podishetty NK},
+\textbf{Bhatnagar S}, \textbf{Eisen JA}, \textbf{Sundaresan V}. 2015.
+Structure, variation, and assembly of the root-associated microbiomes of
+rice. Proceedings of the National Academy of Sciences
+\textbf{112}:E911--E920.
+doi:\href{https://doi.org/10.1073/pnas.1414592112}{10.1073/pnas.1414592112}.}
+
+\leavevmode\hypertarget{ref-Ettinger2017}{}%
+\CSLLeftMargin{22. }
+\CSLRightInline{\textbf{Ettinger CL}, \textbf{Williams SL},
+\textbf{Abbott JM}, \textbf{Stachowicz JJ}, \textbf{Eisen JA}. 2017.
+Microbiome succession during ammonification in eelgrass bed sediments.
+{PeerJ} \textbf{5}:e3674.
+doi:\href{https://doi.org/10.7717/peerj.3674}{10.7717/peerj.3674}.}
+
+\leavevmode\hypertarget{ref-Graw2018}{}%
+\CSLLeftMargin{23. }
+\CSLRightInline{\textbf{Graw MF}, \textbf{DAngelo G}, \textbf{Borchers
+M}, \textbf{Thurber AR}, \textbf{Johnson JE}, \textbf{Zhang C},
+\textbf{Liu H}, \textbf{Colwell FS}. 2018. Energy gradients structure
+microbial communities across sediment horizons in deep marine sediments
+of the south china sea. Frontiers in Microbiology \textbf{9}.
+doi:\href{https://doi.org/10.3389/fmicb.2018.00729}{10.3389/fmicb.2018.00729}.}
+
+\leavevmode\hypertarget{ref-Johnston2016}{}%
+\CSLLeftMargin{24. }
+\CSLRightInline{\textbf{Johnston ER}, \textbf{Rodriguez-R LM},
+\textbf{Luo C}, \textbf{Yuan MM}, \textbf{Wu L}, \textbf{He Z},
+\textbf{Schuur EAG}, \textbf{Luo Y}, \textbf{Tiedje JM}, \textbf{Zhou
+J}, \textbf{Konstantinidis KT}. 2016. Metagenomics reveals pervasive
+bacterial populations and reduced community diversity across the alaska
+tundra ecosystem. Frontiers in Microbiology \textbf{7}.
+doi:\href{https://doi.org/10.3389/fmicb.2016.00579}{10.3389/fmicb.2016.00579}.}
+
+\leavevmode\hypertarget{ref-Hassell2018}{}%
+\CSLLeftMargin{25. }
+\CSLRightInline{\textbf{Hassell N}, \textbf{Tinker KA}, \textbf{Moore
+T}, \textbf{Ottesen EA}. 2018. Temporal and spatial dynamics in
+microbial community composition within a temperate stream network.
+Environmental Microbiology \textbf{20}:3560--3572.
+doi:\href{https://doi.org/10.1111/1462-2920.14311}{10.1111/1462-2920.14311}.}
+
 \end{CSLReferences}
@@
 
-\begin{tabular}{lrrrrr}
+\small
+
+\begin{longtable}[]{@{}lrrrrrl@{}}
 \toprule
-\textbf{Study} & \textbf{Samples} & \textbf{\makecell[c]{Total\\sequences}} & \textbf{\makecell[c]{Median\\sequences}} & \textbf{\makecell[c]{Range of\\sequences}} & \textbf{\makecell[c]{Fold-difference\\between largest\\and smallest sample}}\\
+\textbf{Study (Ref)} & \textbf{Samples} &
+\makecell[c]{\textbf{Total}\\\textbf{sequences}} &
+\makecell[c]{\textbf{Median}\\\textbf{sequences}} &
+\makecell[c]{\textbf{Range of}\\\textbf{sequences}} &
+\makecell[c]{\textbf{Fold-difference}\\\textbf{between largest}\\\textbf{and smallest sample}}
+& \makecell[c]{\textbf{SRA study}\\\textbf{accession}}\tabularnewline
 \midrule
-Bioethanol & 95 & 3,972,943 & 16,015 & 3,688-356,136 & 96.6\\
-Human & 490 & 20,909,768 & 32,505 & 10,523-430,415 & 40.9\\
-Lake & 52 & 3,169,868 & 69,041 & 15,347-112,871 & 7.4\\
-Marine & 7 & 1,391,396 & 193,464 & 133,516-254,060 & 1.9\\
-Mice & 348 & 2,813,747 & 6,477 & 1,804-30,565 & 16.9\\
-Peromyscus & 111 & 1,555,545 & 12,446 & 4,464-33,644 & 7.5\\
-Rainforest & 69 & 946,295 & 11,561 & 4,932-37,767 & 7.7\\
-Rice & 490 & 22,591,168 & 43,216 & 2,776-193,464 & 69.7\\
-Seagrass & 286 & 4,130,454 & 13,567 & 1,803-45,191 & 25.1\\
-Sediment & 58 & 1,154,174 & 17,584 & 7,685-68,321 & 8.9\\
-Soil & 18 & 956,656 & 51,844 & 47,806-59,956 & 1.3\\
-Stream & 201 & 21,162,574 & 90,159 & 9,175-390,964 & 42.6\\
+\endhead
+Bioethanol (15) & 95 & 3,972,943 & 16,015 & 3,688-356,136 & 96.6 &
+SRP055545\tabularnewline
+Human (16) & 490 & 20,909,768 & 32,505 & 10,523-430,415 & 40.9 &
+SRP062005\tabularnewline
+Lake (17) & 52 & 3,169,868 & 69,041 & 15,347-112,871 & 7.4 &
+SRP050963\tabularnewline
+Marine (18) & 7 & 1,391,396 & 193,464 & 133,516-254,060 & 1.9 &
+SRP068101\tabularnewline
+Mice (1) & 348 & 2,813,747 & 6,477 & 1,804-30,565 & 16.9 &
+SRP192323\tabularnewline
+Peromyscus (19) & 111 & 1,555,545 & 12,446 & 4,464-33,644 & 7.5 &
+SRP044050\tabularnewline
+Rainforest (20) & 69 & 946,295 & 11,561 & 4,932-37,767 & 7.7 &
+ERP023747\tabularnewline
+Rice (21) & 490 & 22,591,168 & 43,216 & 2,776-193,464 & 69.7 &
+SRP044745\tabularnewline
+Seagrass (22) & 286 & 4,130,454 & 13,567 & 1,803-45,191 & 25.1 &
+SRP092441\tabularnewline
+Sediment (23) & 58 & 1,154,174 & 17,584 & 7,685-68,321 & 8.9 &
+SRP097192\tabularnewline
+Soil (24) & 18 & 956,656 & 51,844 & 47,806-59,956 & 1.3 &
+ERP012016\tabularnewline
+Stream (25) & 201 & 21,162,574 & 90,159 & 9,175-390,964 & 42.6 &
+SRP075852\tabularnewline
 \bottomrule
-\end{tabular}
+\end{longtable}
+
+\normalsize
 

### 2020-11-05 code/get_sample_summary_statistics.R

@@ -9,6 +9,12 @@ summarize_data <- function(environment) {
 		enframe(., name=NULL, value="sample") %>%
 		mutate(sample = as.character(sample))
 
+	sra_study <- paste0("data/", environment, "/sra_info.tsv") %>%
+		read_tsv() %>%
+		pull("SRA_Study") %>%
+		unique()
+
+
 	paste0("data/", environment, "/data.count.summary") %>%
 		read_tsv(col_names=c("sample", "n_seqs"), col_types=cols(sample=col_character(), n_seqs=col_integer())) %>%
 		anti_join(., removed_samples, by="sample") %>%
@@ -16,18 +22,30 @@ summarize_data <- function(environment) {
 			min=min(n_seqs), l_quartile = quantile(n_seqs, 0.25), median=median(n_seqs),
 			u_quartile = quantile(n_seqs, 0.75), max=max(n_seqs),
 			n_removed = removed_samples %>% filter(sample != "") %>% nrow(),
-			fold_difference = max/min)
+			fold_difference = max/min) %>%
+		mutate(sra_study = sra_study)
 }
 
 
-tibble(directory=c("bioethanol", "human", "lake", "marine", "mice", "peromyscus",
-									"rainforest",	"rice", "seagrass", "sediment", "soil", "stream"),
-			nice_name=c("Bioethanol", "Human", "Lake", "Marine", "Mice", "Peromyscus",
-									"Rainforest", "Rice", "Seagrass", "Sediment", "Soil", "Stream"),
-			reference=rep("", 12)
-		) %>%
-		group_by(directory) %>%
-		mutate(summary_data = map(directory, summarize_data)) %>%
-		unnest(summary_data) %>%
-		ungroup() %>%
-		write_tsv("data/process/study_summary_statistics.tsv")
+tibble(
+		directory=c(
+			"bioethanol", "human", "lake", "marine",
+			"mice", "peromyscus", "rainforest",	"rice",
+			"seagrass", "sediment", "soil", "stream"
+		),
+		nice_name=c(
+			"Bioethanol", "Human", "Lake", "Marine",
+			"Mice", "Peromyscus", "Rainforest", "Rice",
+			"Seagrass", "Sediment", "Soil", "Stream"
+		),
+		reference=c(
+			"[@Li2015]", "[@Baxter2016]", "[@Beall2015]", "[@Henson2016]",
[…]

### 2020-12-03 submission/manuscript.Rmd
First draft of Methods section

@@
 **word choice: sample vs. community**
+**word choice: dataset vs. sample**
 
@@
 
-**EDIT: Two elements that are held in tension in the analysis of 16S rRNA gene sequence data is how to adequately remove PCR and sequencing artifacts and decrease the granularity of the taxonomic level that is used in the analysis.** Previous attempts have included screening for sequencing quality based on quality scores [Kozich/Edgar] followed by a polishing step based on the frequency of the sequences relative to similar sequences [Kozich/Edgar/DeBlur].
+Two elements that are held in tension in the analysis of 16S rRNA gene sequence data is how to adequately remove PCR and sequencing artifacts while decreasing the granularity of the taxonomic level that is used in the analysis.
+Previous attempts have included screening for sequencing quality based on quality scores [Kozich/Edgar] followed by a polishing step based on the frequency of the sequences relative to similar sequences [Kozich/Edgar/DeBlur].
 Other pipelines model the quality scores and types of errors to cluster sequences directly [Dada2].
@@
 **Datasets.** I collected `r nrow(table_1)` publicly available datasets that used the Illumina MiSeq platform to sequence the V4 region of the 16S rRNA gene from a variety of environments (Table 1).
-To insure the highest possible data quality, datasets were limited to those where the 500 cycle v2 MiSeq chemistry was used to sequence the amplicons.
-The paired 250 nt reads resulted in near complete 2-fold sequencing coverage of every nucleotide in the ca.
-250 nt-long region.
-This region and sequencing platform were selected because previous work has shown that a standard data analysis pipeline in mothur results in a sequencing error rate below 0.02% [@Kozich2013].
-All sequence data were obtained from the Sequence Read Archive (SRA) and processed using a standard mothur-based sequencing pipeline that resulted in ASVs as generated by the pre.cluster algorithm [@Kozich2013; @Schloss2009].
 After removing poor quality and chimeric ASVs and samples that had uncharacteristically low number of sequences for the dataset, these datasets included between `r oxford_comma(n_sample_range)` samples (Figure S1).
@@
 The effect of removing rare ASVs on the richness of OTUs and their diversity was similar (Figure S5AB).
-I used the Bray-Curtis dissimilarity index to compare the simulated communities within each dataset and calculated the power to detect differences between the two simulated treatment groups using the analysis of molecular variance (also called PERMANOVA) (Figure 4C and S5C).
+I used the Bray-Curtis dissimilarity index to compare the simulated communities within each dataset and calculated the power to detect differences between the two simulated treatment groups using the analysis of molecular variance (Figure 4C and S5C).
 Without removing rare sequences, the power to detect a difference between the two simulated treatment groups varied between `r oxford_comma(pull(b_base_beta, '1') %>% range(), 2)` (`r slice(b_base_beta, 1) %>% pull(dataset)` and `r oxford_comma(slice_max(b_base_beta, '1') %>% pull(dataset))`).
@@
 
-Datasets were selected because they sequenced the V4 region of the 16S rRNA gene using Illumina's MiSeq platform using the v2 chemistry with paired 250 nt reads.
-I identified 12 datasets that satisfied these criteria and downloaded the data directly from the SRA (Table 1 and Figure S1).
+**Data curation and analysis.** To insure the highest possible data quality, datasets were limited to those where the 500 cycle v2 MiSeq chemistry was used to sequence the amplicons.
+The paired 250 nt reads resulted in near complete 2-fold sequencing coverage of every nucleotide in the ca.
+250 nt-long region.
+This region and sequencing platform were selected because previous work has shown that a standard data analysis pipeline in mothur results in a sequencing error rate below 0.02% [@Kozich2013].
+All sequence data were obtained from the Sequence Read Archive (SRA) and processed using a standard mothur-based sequencing pipeline that resulted in ASVs as generated by the pre.cluster algorithm using a threshold of 2 nt [@Kozich2013; @Schloss2009].
+ASVs were assigned to OTUs using a 3% distance threshold using mothur's cluster function with the OptiClust algorithm [@Westcott2017].
+To minimize the effects of uneven sampling effort, samples were rarefied to the number of sequences in the smallest sample for each dataset.
+Because metrics of alpha diversity did not consistently follow a normal distribution, I used the non-parametric Wilcoxon rank test as implemented in R.
+For comparisons of Bray-Curtis distances, the amova function within mothur was used, which implements the analysis of molecular variance algorithm [@Anderson2001].
 
-* Used the mothur pipeline to
-  - Assembly of paired 250 nt reads
-*
+**Reproducibility.** All analyses were performed using mothur (version 1.44.1) and R (version `r paste(R.version$major, R.version$minor, sep='.')`) with the tidyverse (v.`r packageVersion("tidyverse")`), broom (v.`r packageVersion("broom")`), data.table (v.`r packageVersion("data.table")`), and cowplot (v.`r packageVersion("cowplot")`) packages.
+All of the code that was used in this analysis as well as the Makefile for running the analysis are available at https://github.com/SchlossLab/Schloss_Singletons_XXXXX_2019 along with the complete version history of the project.
 

### 2020-12-08 submission/manuscript.Rmd
First draft of Intro/Abstract/Importance sections

@@
 
+**word choice: sample vs. community**
+**word choice: dataset vs. sample**
+
 \newpage
+
 <!-- \linenumbers -->
@@
 
-**word choice: sample vs. community**
-**word choice: dataset vs. sample**
-
-250 words
+Methods for remediating PCR and sequencing artifacts in 16S rRNA gene sequence collections are in continuous development.
+The choices that researchers make have significant ramifications on the inferences that can be drawn from microbial ecology studies.
+One common approach is to remove rare sequences from datasets.
+But, the threshold that is applied is generally selected without regard for the number of sequences in the sample or the variation in sequencing depth across samples within a study.
+I analyzed the impact of removing rare sequences on metrics of alpha and beta diversity using samples collected across `r nrow(table_1)` published datasets.
+Removal of rare sequences significantly decreased the number of amplicon sequence variants and operational taxonomic units as well as their diversity.
+Furthermore, their removal increased the variation in community structure between samples.
+When a known effect size was simulated, removal of rare sequences reduced the statistical power to detect the effect relative to not removing rare sequences.
+Although removal of rare sequences did not affect the probability of falsely detetecting differences when samples were randomized to simulate a null model, the false detection rate increased when samples were assigned to simulated treatment groups according to their sequencing depth and rare sequences were removed.
+The false detection rate did not vary when rare sequences were retained.
+This analysis demonstrates the problems inherent in removing rare sequences.
+Researchers are encouraged to retain these sequences, to select approaches that minimize PCR and sequencing artifacts, and to use rarefaction to control for uneven sampling effort.
 
@@
 ## Importance
-150 words
+
+Removing rare sequences from 16S rRNA gene sequence collections is an approach that has grown in popularity for handling PCR and sequencing artifacts.
+Yet, it is unclear what impact an abundance-based filter has on downstream analyses.
+To investigate the effects of removing rare sequences, I analyzed the community distributions found in the samples of `r nrow(table_1)` published datasets.
+Analysis of these data and simulations based on them showed that removal of rare sequences distorts the representation of microbial communities.
+This has the effect of artificially making it more difficult to detect differences between treatment groups.
+Also of concern was the observation that if sequencing depth is confounded with treatment group, then the probability of falsely detecting a difference between the treatment groups increases with the removal of rare sequences.
+The practice of removing rare sequences should stop, lest researcher adversely affect the interpretation of their data.
 
@@
 
-Two elements that are held in tension in the analysis of 16S rRNA gene sequence data is how to adequately remove PCR and sequencing artifacts while decreasing the granularity of the taxonomic level that is used in the analysis.
-Previous attempts have included screening for sequencing quality based on quality scores [Kozich/Edgar] followed by a polishing step based on the frequency of the sequences relative to similar sequences [Kozich/Edgar/DeBlur].
-Other pipelines model the quality scores and types of errors to cluster sequences directly [Dada2].
-But, as a final step many pipelines advocate for removing rare sequences from each dataset prior to outputting the sequence data as amplicon sequence variants (ASVs) [Knight/Edgar/DeBlur/Dada2].
-ASVs are often clustered further to generate operational taxonomic units or phylotypes.
-Some pipelines remove all ASVs that appear once (i.e.
-singletons) [XXXX], XXXXXX [XXXXXX], XXXXX [XXXXXXX], or XXXXXX [XXXXXX] times prior to further clustering or making ecological comparisions.
-Notably, the mothur-based pipeline discourages the practice of removing rare sequences.
+## Introduction
+
+16S rRNA gene sequencing is a mainstay of microbial community analysis [@Knight2018].
+Two elements that are held in tension in the analysis of 16S rRNA gene sequence data are how to adequately remove PCR and sequencing artifacts while decreasing the granularity of the taxonomic level that is used in the analysis.
+When coarse taxonomic levels (e.g.
+genus level) are used, the effects of artifacts are minimized since the genetic breadth of the level is wider than the diversity of artifacts.
+Conversely, with fine taxonomic levels (e.g.
+amplicon sequence variants; ASVs) the effects of artifacts are significant since each artifact may represent a new ASV.
+
+Numerous studies have attempted to address the problem of removing or "denoising" artifacts from data generated using Illumina's MiSeq platform.
+In one approach, paired sequence reads are aligned and any discrepencies between the reads are resolved based on the difference in quality score for the position in question [@Kozich2013; @Edgar2016; @Masella2012]; quality scores are also used to curate single reads [@Bokulich2013].
+In addition, a polishing step is often used to identify ideal sequecnces and their related artifacts based on the frequency of the sequences and their similarity sequences [@Kozich2013; @Edgar2016; @Amir2017].
+In a second approach, the quality scores and types of errors to cluster sequences directly [@Callahan2016].
+Regardless of the approach, many pipelines advocate for abundance-based sceeening where rare sequences are removed from each dataset prior to outputting the sequence data as ASVs [@Edgar2016; @Amir2017; @Callahan2016].
+Some algorithms recommend removing all ASVs that appear one (i.e.
+singletons) [@Callahan2016], eight [@Edgar2016], or ten [@Amir2017] or fewer times; these pipelines also vary in whether the minimum abundance threshold should be applied to individual samples [@Callahan2016; @Amir2017] or the pool of samples in a study [@Edgar2016].
+A notable exception, the mothur-based pipeline discourages the practice of removing rare sequences [@Kozich2013].
+After assigning sequences to ASVs, they are often analyzed as a taxonomic unit or clustered to generate operational taxonomic units (OTUs) or phylotypes.
 
 The abundance-based screening approach assumes that rare ASVs are more likely to be artifacts than more abundant ASVs.
-Sequencing of mock communities confirms that artifacts tend to be rare.
-Proponents of abundance-based screening point to their ability to obtain the correct number of ASVs, OTUs, or phylotypes with data generated from sequencing mock communities when rare ASVs are removed.
-However, this approach effectively overfits the curation pipeline to data generated from a phylogenetically simple community with an atypical community distribution that is often sequenced to a depth that is not achieved with biological samples.
+Sequencing of mock communities confirms that artifacts tend to be rare [@Kozich2013; @Bokulich2013].
+Proponents of abundance-based screening point to their ability to obtain the correct number of ASVs, OTUs, or phylotypes with data generated from sequencing mock communities when rare ASVs are removed [@Bokulich2013].
+However, this approach effectively overfits the curation pipeline to data generated from a phylogenetically simple community with an atypical community distribution that is often sequenced to a coverage that is not achieved with biological samples.
 It is necessary to think more deeply about the practice of abundance-based screening.
 
-The minimum abundance thresholds that have been proscribed are developed and applied without regard for the total number of sequences generated from each sample.
-Ignored in their recommendations is the common experience that the number of sequences generated from each sample may vary by two or three orders of magnitude.
+The minimum abundance thresholds that have been proscribed were developed and applied without regard for the number of sequences generated from each sample.
+These recommendations appear to assume that the number of reads per sample is consistent within and between studies.
+However, it is common that the number of sequences generated from each sample may vary by two or three orders of magnitude (e.g.
+Figure S1).
 An ASV that appears once in a sample with 2,000 sequences is more trustworthy than an ASV that appears once in a sample with 100,000 sequences since it has a 50-fold higher relative abundance.
@@
 
-Experience sequencing biological samples demonstrates that there are good ASVs that may have an abundance below the proscribed threshold.
+Experience sequencing biological samples also demonstrates that there are bona fide ASVs that may have an abundance below the proscribed threshold.
 For example, the abundance of an ASV may be below the threshold in some samples or time points and above the threshold in others.
 However, rarity, both in terms of prevalence and incidence, is an important ecological concept.
-Removing rare ASVs likely hinders one's ability to ability to make inferences about the dynamics and nature of the populations that rare ASVs represent.
+Removing rare ASVs likely hinders one's to ability to make inferences about the dynamics and nature of the populations that rare ASVs represent.
 Furthermore, removing ASVs whose abundances are below the proscribed threshold also potentially biases the community structure of the samples.
 
-In the current study, I use published sequence data from `r nrow(table_1)` studies to investigate the nature of rare ASVs (i.e.
-those that appear 10 or fewer times) and the effects of removing them on downstream analysis of microbial communities.
-The analysis was also performed using traditional operational taxonomic units, where ASVs subjected to abundance-based screening were clustered such that the ASVs within an OTU were no more than 3% different from each other.
+In the current study, I used published sequence data from `r nrow(table_1)` studies to investigate the nature of rare ASVs (i.e.
+those that appear 10 or fewer times) and the effect that removing them has on downstream analysis of microbial communities.
+The analysis was also performed using traditional OTUs, where ASVs subjected to abundance-based screening were clustered such that the ASVs within an OTU were no more than 3% different from each other.
 The results reject the assumptions built into abundance-based screening and highlight the problems inherent in removing rare ASVs.

### 2020-12-08 submission/manuscript.tex
First draft of Intro/Abstract/Importance sections

@@
 
+\textbf{word choice: sample vs.~community} \textbf{word choice: dataset
+vs.~sample}
+
 \newpage
@@
 
-\textbf{word choice: sample vs.~community}
-
-250 words
+Methods for remediating PCR and sequencing artifacts in 16S rRNA gene
+sequence collections are in continuous development. The choices that
+researchers make have significant ramifications on the inferences that
+can be drawn from microbial ecology studies. One common approach is to
+remove rare sequences from datasets. But, the threshold that is applied
+is generally selected without regard for the number of sequences in the
+sample or the variation in sequencing depth across samples within a
+study. I analyzed the impact of removing rare sequences on metrics of
+alpha and beta diversity using samples collected across 12 published
+datasets. Removal of rare sequences significantly decreased the number
+of amplicon sequence variants and operational taxonomic units as well as
+their diversity. Furthermore, their removal increased the variation in
+community structure between samples. When a known effect size was
+simulated, removal of rare sequences reduced the statistical power to
+detect the effect relative to not removing rare sequences. Although
+removal of rare sequences did not affect the probability of falsely
+detetecting differences when samples were randomized to simulate a null
+model, the false detection rate increased when samples were assigned to
+simulated treatment groups according to their sequencing depth and rare
+sequences were removed. The false detection rate did not vary when rare
+sequences were retained. This analysis demonstrates the problems
+inherent in removing rare sequences. Researchers are encouraged to
+retain these sequences, to select approaches that minimize PCR and
+sequencing artifacts, and to use rarefaction to control for uneven
+sampling effort.
 
@@
 
-150 words
-
-\newpage
-
-\textbf{EDIT: Two elements that are held in tension in the analysis of
-16S rRNA gene sequence data is how to adequately remove PCR and
-sequencing artifacts and decrease the granularity of the taxonomic level
-that is used in the analysis.} Previous attempts have included screening
-for sequencing quality based on quality scores {[}Kozich/Edgar{]}
-followed by a polishing step based on the frequency of the sequences
-relative to similar sequences {[}Kozich/Edgar/DeBlur{]}. Other pipelines
-model the quality scores and types of errors to cluster sequences
-directly {[}Dada2{]}. But, as a final step many pipelines advocate for
-removing rare sequences from each dataset prior to outputting the
-sequence data as amplicon sequence variants (ASVs)
-{[}Knight/Edgar/DeBlur/Dada2{]}. ASVs are often clustered further to
-generate operational taxonomic units or phylotypes. Some pipelines
-remove all ASVs that appear once (i.e.~singletons) {[}XXXX{]}, XXXXXX
-{[}XXXXXX{]}, XXXXX {[}XXXXXXX{]}, or XXXXXX {[}XXXXXX{]} times prior to
-further clustering or making ecological comparisions. Notably, the
-mothur-based pipeline discourages the practice of removing rare
-sequences.
+Removing rare sequences from 16S rRNA gene sequence collections is an
+approach that has grown in popularity for handling PCR and sequencing
+artifacts. Yet, it is unclear what impact an abundance-based filter has
+on downstream analyses. To investigate the effects of removing rare
+sequences, I analyzed the community distributions found in the samples
+of 12 published datasets. Analysis of these data and simulations based
+on them showed that removal of rare sequences distorts the
+representation of microbial communities. This has the effect of
+artificially making it more difficult to detect differences between
+treatment groups. Also of concern was the observation that if sequencing
+depth is confounded with treatment group, then the probability of
+falsely detecting a difference between the treatment groups increases
+with the removal of rare sequences. The practice of removing rare
+sequences should stop, lest researcher adversely affect the
+interpretation of their data.
+
+\newpage
+
+\hypertarget{introduction}{%
+\subsection{Introduction}\label{introduction}}
+
+16S rRNA gene sequencing is a mainstay of microbial community analysis
+(1). Two elements that are held in tension in the analysis of 16S rRNA
+gene sequence data are how to adequately remove PCR and sequencing
+artifacts while decreasing the granularity of the taxonomic level that
+is used in the analysis. When coarse taxonomic levels (e.g.~genus level)
+are used, the effects of artifacts are minimized since the genetic
+breadth of the level is wider than the diversity of artifacts.
+Conversely, with fine taxonomic levels (e.g.~amplicon sequence variants;
+ASVs) the effects of artifacts are significant since each artifact may
+represent a new ASV.
+
+Numerous studies have attempted to address the problem of removing or
+``denoising'' artifacts from data generated using Illumina's MiSeq
+platform. In one approach, paired sequence reads are aligned and any
+discrepencies between the reads are resolved based on the difference in
+quality score for the position in question (2--4); quality scores are
+also used to curate single reads (5). In addition, a polishing step is
+often used to identify ideal sequecnces and their related artifacts
+based on the frequency of the sequences and their similarity sequences
+(2, 3, 6). In a second approach, the quality scores and types of errors
+to cluster sequences directly (7). Regardless of the approach, many
+pipelines advocate for abundance-based sceeening where rare sequences
+are removed from each dataset prior to outputting the sequence data as
+ASVs (3, 6, 7). Some algorithms recommend removing all ASVs that appear
+one (i.e.~singletons) (7), eight (3), or ten (6) or fewer times; these
+pipelines also vary in whether the minimum abundance threshold should be
+applied to individual samples (6, 7) or the pool of samples in a study
+(3). A notable exception, the mothur-based pipeline discourages the
+practice of removing rare sequences (2). After assigning sequences to
+ASVs, they are often analyzed as a taxonomic unit or clustered to
+generate operational taxonomic units (OTUs) or phylotypes.
 
@@
 likely to be artifacts than more abundant ASVs. Sequencing of mock
-communities confirms that artifacts tend to be rare. Proponents of
-abundance-based screening point to their ability to obtain the correct
-number of ASVs, OTUs, or phylotypes with data generated from sequencing
-mock communities when rare ASVs are removed. However, this approach
-effectively overfits the curation pipeline to data generated from a
-phylogenetically simple community with an atypical community
-distribution that is often sequenced to a depth that is not achieved
-with biological samples. It is necessary to think more deeply about the
-practice of abundance-based screening.
-
-The minimum abundance thresholds that have been proscribed are developed
-and applied without regard for the total number of sequences generated
-from each sample. Ignored in their recommendations is the common
-experience that the number of sequences generated from each sample may
-vary by two or three orders of magnitude. An ASV that appears once in a
-sample with 2,000 sequences is more trustworthy than an ASV that appears
-once in a sample with 100,000 sequences since it has a 50-fold higher
-relative abundance. But, according to the pipeline recommendations, they
-are treated as being equally trustworthy. Rather than removing rare
-ASVs, the approach taken by the mothur pipeline applies the classical
-ecological approach of rarefaction. Each sample is rarefied to the same
-sequencing depth so that the number of artifacts that appears in each
-sample is controlled.
-
-Experience sequencing biological samples demonstrates that there are
-good ASVs that may have an abundance below the proscribed threshold. For
-example, the abundance of an ASV may be below the threshold in some
-samples or time points and above the threshold in others. However,
-rarity, both in terms of prevalence and incidence, is an important
-ecological concept. Removing rare ASVs likely hinders one's ability to
-ability to make inferences about the dynamics and nature of the
+communities confirms that artifacts tend to be rare (2, 5). Proponents
+of abundance-based screening point to their ability to obtain the
+correct number of ASVs, OTUs, or phylotypes with data generated from
+sequencing mock communities when rare ASVs are removed (5). However,
+this approach effectively overfits the curation pipeline to data
+generated from a phylogenetically simple community with an atypical
+community distribution that is often sequenced to a coverage that is not
+achieved with biological samples. It is necessary to think more deeply
+about the practice of abundance-based screening.
+
+The minimum abundance thresholds that have been proscribed were
+developed and applied without regard for the number of sequences
+generated from each sample. These recommendations appear to assume that
+the number of reads per sample is consistent within and between studies.
+However, it is common that the number of sequences generated from each
+sample may vary by two or three orders of magnitude (e.g.~Figure S1). An
+ASV that appears once in a sample with 2,000 sequences is more
+trustworthy than an ASV that appears once in a sample with 100,000
+sequences since it has a 50-fold higher relative abundance. But,
+according to the pipeline recommendations, they are treated as being
+equally trustworthy. Rather than removing rare ASVs, the approach taken
+by the mothur pipeline applies the classical ecological approach of
+rarefaction. Each sample is rarefied to the same sequencing depth so
+that the number of artifacts that appears in each sample is controlled.
+
+Experience sequencing biological samples also demonstrates that there
+are bona fide ASVs that may have an abundance below the proscribed
+threshold. For example, the abundance of an ASV may be below the
+threshold in some samples or time points and above the threshold in
+others. However, rarity, both in terms of prevalence and incidence, is
+an important ecological concept. Removing rare ASVs likely hinders one's
+to ability to make inferences about the dynamics and nature of the
 populations that rare ASVs represent. Furthermore, removing ASVs whose
@@
 
-In the current study, I use published sequence data from 12 studies to
+In the current study, I used published sequence data from 12 studies to
 investigate the nature of rare ASVs (i.e.~those that appear 10 or fewer
-times) and the effects of removing them on downstream analysis of
+times) and the effect that removing them has on downstream analysis of
 microbial communities. The analysis was also performed using traditional
-operational taxonomic units, where ASVs subjected to abundance-based
-screening were clustered such that the ASVs within an OTU were no more
-than 3\% different from each other. The results reject the assumptions
-built into abundance-based screening and highlight the problems inherent
-in removing rare ASVs.
+OTUs, where ASVs subjected to abundance-based screening were clustered
+such that the ASVs within an OTU were no more than 3\% different from
+each other. The results reject the assumptions built into
+abundance-based screening and highlight the problems inherent in
+removing rare ASVs.
 
@@
 the Illumina MiSeq platform to sequence the V4 region of the 16S rRNA
-gene from a variety of environments (Table 1). To insure the highest
-possible data quality, datasets were limited to those where the 500
-cycle v2 MiSeq chemistry was used to sequence the amplicons. The paired
-250 nt reads resulted in near complete 2-fold sequencing coverage of
-every nucleotide in the ca. 250 nt-long region. This region and
-sequencing platform were selected because previous work has shown that a
-standard data analysis pipeline in mothur results in a sequencing error
-rate below 0.02\% (1). All sequence data were obtained from the Sequence
-Read Archive (SRA) and processed using a standard mothur-based
-sequencing pipeline that resulted in ASVs as generated by the
-pre.cluster algorithm (1, 2). After removing poor quality and chimeric
-ASVs and samples that had uncharacteristically low number of sequences
-for the dataset, these datasets included between 7 and 490 samples
-(Figure S1). The median number of sequences for each dataset ranged
-between 6,477 and 193,464 (Table 1). Strikingly, aside from the
-relatively small marine and soil datasets, the difference between the
-sample with the fewest sequences and the sample with the most sequences
-for each dataset varied by between 7.4 and 96.6-fold (Table 1).
+gene from a variety of environments (Table 1). After removing poor
+quality and chimeric ASVs and samples that had uncharacteristically low
+number of sequences for the dataset, these datasets included between 7
+and 490 samples (Figure S1). The median number of sequences for each
+dataset ranged between 6,477 and 193,464 (Table 1). Strikingly, aside
+from the relatively small marine and soil datasets, the difference
+between the sample with the fewest sequences and the sample with the
+most sequences for each dataset varied by between 7.4 and 96.6-fold
+(Table 1).
 
@@
 detect differences between the two simulated treatment groups using the
-analysis of molecular variance (also called PERMANOVA) (Figure 4C and
-S5C). Without removing rare sequences, the power to detect a difference
-between the two simulated treatment groups varied between 0.41 and 1.00
-(rainforest and rainforest). Aside from the bioethanol, human, and mice
-datasets, the power to detect differences dropped by between 6.5 and
-64.0\% (soil and rice) when singletons were removed. However, when ASVs
-that occurred 10 or fewer times were removed from each sample, the power
-to detect differences dropped by 12.0 and 97.2\% (human and peromyscus);
-similar results were observed when ASVs were clustered into OTUs.
-Removing rare ASVs reduced the ability to detect simulated treatment
-effects using metrics commonly used to compare microbial communities.
+analysis of molecular variance (Figure 4C and S5C). Without removing
+rare sequences, the power to detect a difference between the two
+simulated treatment groups varied between 0.41 and 1.00 (rainforest and
+rainforest). Aside from the bioethanol, human, and mice datasets, the
+power to detect differences dropped by between 6.5 and 64.0\% (soil and
+rice) when singletons were removed. However, when ASVs that occurred 10
+or fewer times were removed from each sample, the power to detect
+differences dropped by 12.0 and 97.2\% (human and peromyscus); similar
+results were observed when ASVs were clustered into OTUs. Removing rare
+ASVs reduced the ability to detect simulated treatment effects using
+metrics commonly used to compare microbial communities.
 
@@
 relative to those in the oral cavity leading to a significant difference
-in sequencing depth between treatment groups (3). When rare sequences
+in sequencing depth between treatment groups (8). When rare sequences
 were not removed, the fraction of significant tests did not differ from
@@
 response to researchers prioritizing the number of reads and length of
-sequences over their quality (4). Previous work has shown that assembly
+sequences over their quality (5). Previous work has shown that assembly
 of fully overlapping sequence reads results in the lowest sequencing
-error rates (1). The studies highlighted in this analysis sequenced the
+error rates (2). The studies highlighted in this analysis sequenced the
 V4 region of the 16S rRNA gene using the Illumina MiSeq sequencing
@@
 there has been a movement to sequence longer regions with similar
-chemistry resulting in less overlap between the sequencing reads (5, 6).
-Alternatively, others have prioritized increasing the number of
+chemistry resulting in less overlap between the sequencing reads (9,
+10). Alternatively, others have prioritized increasing the number of
 sequences per sample by sequencing the V4 region, but with paired or
-single reads that are 150 nt long (7, 8). Both practices result in a
+single reads that are 150 nt long (11, 12). Both practices result in a
 significantly higher error rate for the resulting assmebly. Instead,
@@
 Previous work sequencing mock communities has shown that the number of
-spurious sequences increases with sequencing depth (1, 9). By rarefying
+spurious sequences increases with sequencing depth (2, 13). By rarefying
 data to a common number of sequences per sample, the number of spurious
@@
 inferences that they make about rare sequences. This is a standard that
-should be applied regardless of their abundance (10).
+should be applied regardless of their abundance (14).
 
@@
 challenge since microbial ecologists transitioned away from Sanger
-sequencing of samples (11--13). Although the extent of the ``rare
+sequencing of samples (15--17). Although the extent of the ``rare
 biosphere'' is still an open question, it is important to appreciate the
@@
 role of such populations or the processes that drive their patchy
-distributions that are so common with microbial communities (14).
+distributions that are so common with microbial communities (18).
 
@@
 
-Datasets were selected because they sequenced the V4 region of the 16S
-rRNA gene using Illumina's MiSeq platform using the v2 chemistry with
-paired 250 nt reads. I identified 12 datasets that satisfied these
-criteria and downloaded the data directly from the SRA (Table 1 and
-Figure S1).
-
-\begin{itemize}
-\tightlist
-\item
-  Used the mothur pipeline to
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Assembly of paired 250 nt reads
-  \end{itemize}
-\item
-\end{itemize}
+\textbf{Data curation and analysis.} To insure the highest possible data
+quality, datasets were limited to those where the 500 cycle v2 MiSeq
+chemistry was used to sequence the amplicons. The paired 250 nt reads
+resulted in near complete 2-fold sequencing coverage of every nucleotide
+in the ca. 250 nt-long region. This region and sequencing platform were
+selected because previous work has shown that a standard data analysis
+pipeline in mothur results in a sequencing error rate below 0.02\% (2).
+All sequence data were obtained from the Sequence Read Archive (SRA) and
+processed using a standard mothur-based sequencing pipeline that
+resulted in ASVs as generated by the pre.cluster algorithm using a
+threshold of 2 nt (2, 19). ASVs were assigned to OTUs using a 3\%
+distance threshold using mothur's cluster function with the OptiClust
+algorithm (20). To minimize the effects of uneven sampling effort,
+samples were rarefied to the number of sequences in the smallest sample
+for each dataset. Because metrics of alpha diversity did not
+consistently follow a normal distribution, I used the non-parametric
+Wilcoxon rank test as implemented in R. For comparisons of Bray-Curtis
+distances, the amova function within mothur was used, which implements
+the analysis of molecular variance algorithm (21).
+
+\textbf{Reproducibility.} All analyses were performed using mothur
+(version 1.44.1) and R (version 4.0.2) with the tidyverse (v.130), broom
+(v.070), data.table (v. 113 4), and cowplot (v.110) packages. All of the
+code that was used in this analysis as well as the Makefile for running
+the analysis are available at
+\url{https://github.com/SchlossLab/Schloss_Singletons_XXXXX_2019} along
+with the complete version history of the project.
 
@@
 \begin{CSLReferences}{0}{0}
+\leavevmode\hypertarget{ref-Knight2018}{}%
+\CSLLeftMargin{1. }
+\CSLRightInline{\textbf{Knight R}, \textbf{Vrbanac A}, \textbf{Taylor
+BC}, \textbf{Aksenov A}, \textbf{Callewaert C}, \textbf{Debelius J},
+\textbf{Gonzalez A}, \textbf{Kosciolek T}, \textbf{McCall L-I},
+\textbf{McDonald D}, \textbf{Melnik AV}, \textbf{Morton JT},
+\textbf{Navas J}, \textbf{Quinn RA}, \textbf{Sanders JG},
+\textbf{Swafford AD}, \textbf{Thompson LR}, \textbf{Tripathi A},
+\textbf{Xu ZZ}, \textbf{Zaneveld JR}, \textbf{Zhu Q}, \textbf{Caporaso
+JG}, \textbf{Dorrestein PC}. 2018. Best practices for analysing
+microbiomes. Nature Reviews Microbiology \textbf{16}:410--422.
+doi:\href{https://doi.org/10.1038/s41579-018-0029-9}{10.1038/s41579-018-0029-9}.}
+
 \leavevmode\hypertarget{ref-Kozich2013}{}%
-\CSLLeftMargin{1. }
+\CSLLeftMargin{2. }
 \CSLRightInline{\textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter
@@
 
-\leavevmode\hypertarget{ref-Schloss2009}{}%
-\CSLLeftMargin{2. }
-\CSLRightInline{\textbf{Schloss PD}, \textbf{Westcott SL},
-\textbf{Ryabin T}, \textbf{Hall JR}, \textbf{Hartmann M},
-\textbf{Hollister EB}, \textbf{Lesniewski RA}, \textbf{Oakley BB},
-\textbf{Parks DH}, \textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres
-B}, \textbf{Thallinger GG}, \textbf{Horn DJV}, \textbf{Weber CF}. 2009.
-Introducing mothur: Open-source, platform-independent,
-community-supported software for describing and comparing microbial
-communities. Applied and Environmental Microbiology
-\textbf{75}:7537--7541.
-doi:\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.}
+\leavevmode\hypertarget{ref-Edgar2016}{}%
+\CSLLeftMargin{3. }
+\CSLRightInline{\textbf{Edgar RC}. 2016. {UNOISE}2: Improved
+error-correction for illumina 16S and {ITS} amplicon sequencing.
+bio{R}xiv. doi:\href{https://doi.org/10.1101/081257}{10.1101/081257}.}
+
+\leavevmode\hypertarget{ref-Masella2012}{}%
+\CSLLeftMargin{4. }
+\CSLRightInline{\textbf{Masella AP}, \textbf{Bartram AK},
+\textbf{Truszkowski JM}, \textbf{Brown DG}, \textbf{Neufeld JD}. 2012.
+{PANDAseq}: Paired-end assembler for illumina sequences. {BMC}
+Bioinformatics \textbf{13}:31.
+doi:\href{https://doi.org/10.1186/1471-2105-13-31}{10.1186/1471-2105-13-31}.}
+
+\leavevmode\hypertarget{ref-Bokulich2013}{}%
+\CSLLeftMargin{5. }
+\CSLRightInline{\textbf{Bokulich NA}, \textbf{Subramanian S},
+\textbf{Faith JJ}, \textbf{Gevers D}, \textbf{Gordon JI}, \textbf{Knight
+R}, \textbf{Mills DA}, \textbf{Caporaso JG}. 2013. Quality-filtering
+vastly improves diversity estimates from illumina amplicon sequencing.
+Nature Methods \textbf{10}:57--59.
+doi:\href{https://doi.org/10.1038/nmeth.2276}{10.1038/nmeth.2276}.}
+
+\leavevmode\hypertarget{ref-Amir2017}{}%
+\CSLLeftMargin{6. }
+\CSLRightInline{\textbf{Amir A}, \textbf{McDonald D},
+\textbf{Navas-Molina JA}, \textbf{Kopylova E}, \textbf{Morton JT},
+\textbf{Xu ZZ}, \textbf{Kightley EP}, \textbf{Thompson LR}, \textbf{Hyde
+ER}, \textbf{Gonzalez A}, \textbf{Knight R}. 2017. Deblur rapidly
+resolves single-nucleotide community sequence patterns. {mSystems}
+\textbf{2}.
+doi:\href{https://doi.org/10.1128/msystems.00191-16}{10.1128/msystems.00191-16}.}
+
+\leavevmode\hypertarget{ref-Callahan2016}{}%
+\CSLLeftMargin{7. }
+\CSLRightInline{\textbf{Callahan BJ}, \textbf{McMurdie PJ},
+\textbf{Rosen MJ}, \textbf{Han AW}, \textbf{Johnson AJA}, \textbf{Holmes
+SP}. 2016. {DADA}2: High-resolution sample inference from illumina
+amplicon data. Nature Methods \textbf{13}:581--583.
+doi:\href{https://doi.org/10.1038/nmeth.3869}{10.1038/nmeth.3869}.}
 
 \leavevmode\hypertarget{ref-Morris2013}{}%
-\CSLLeftMargin{3. }
+\CSLLeftMargin{8. }
 \CSLRightInline{\textbf{Morris A}, \textbf{Beck JM}, \textbf{Schloss
@@
 
-\leavevmode\hypertarget{ref-Bokulich2013}{}%
-\CSLLeftMargin{4. }
-\CSLRightInline{\textbf{Bokulich NA}, \textbf{Subramanian S},
-\textbf{Faith JJ}, \textbf{Gevers D}, \textbf{Gordon JI}, \textbf{Knight
-R}, \textbf{Mills DA}, \textbf{Caporaso JG}. 2013. Quality-filtering
-vastly improves diversity estimates from illumina amplicon sequencing.
-Nature Methods \textbf{10}:57--59.
-doi:\href{https://doi.org/10.1038/nmeth.2276}{10.1038/nmeth.2276}.}
-
 \leavevmode\hypertarget{ref-Fadrosh2014}{}%
-\CSLLeftMargin{5. }
+\CSLLeftMargin{9. }
 \CSLRightInline{\textbf{Fadrosh DW}, \textbf{Ma B}, \textbf{Gajer P},
@@
 \leavevmode\hypertarget{ref-Holm2019}{}%
-\CSLLeftMargin{6. }
+\CSLLeftMargin{10. }
 \CSLRightInline{\textbf{Holm JB}, \textbf{Humphrys MS}, \textbf{Robinson
@@
 \leavevmode\hypertarget{ref-Thompson2017}{}%
-\CSLLeftMargin{7. }
+\CSLLeftMargin{11. }
 \CSLRightInline{\textbf{Thompson LR}, \textbf{Jon G. Sanders and},
@@
 \leavevmode\hypertarget{ref-McDonald2018}{}%
-\CSLLeftMargin{8. }
+\CSLLeftMargin{12. }
 \CSLRightInline{\textbf{McDonald D}, \textbf{Hyde E}, \textbf{Debelius
@@
 \leavevmode\hypertarget{ref-Schloss2011}{}%
-\CSLLeftMargin{9. }
+\CSLLeftMargin{13. }
 \CSLRightInline{\textbf{Schloss PD}, \textbf{Gevers D}, \textbf{Westcott
@@
 \leavevmode\hypertarget{ref-Schloss2018}{}%
-\CSLLeftMargin{10. }
+\CSLLeftMargin{14. }
 \CSLRightInline{\textbf{Schloss PD}. 2018. Identifying and overcoming
@@
 \leavevmode\hypertarget{ref-Sogin2006}{}%
-\CSLLeftMargin{11. }
+\CSLLeftMargin{15. }
 \CSLRightInline{\textbf{Sogin ML}, \textbf{Morrison HG}, \textbf{Huber
@@
 \leavevmode\hypertarget{ref-Huse2010}{}%
-\CSLLeftMargin{12. }
+\CSLLeftMargin{16. }
 \CSLRightInline{\textbf{Huse SM}, \textbf{Welch DM}, \textbf{Morrison
@@
 \leavevmode\hypertarget{ref-Reeder2009}{}%
-\CSLLeftMargin{13. }
+\CSLLeftMargin{17. }
 \CSLRightInline{\textbf{Reeder J}, \textbf{Knight R}. 2009. The {{}}rare
@@
 \leavevmode\hypertarget{ref-Reid2011}{}%
-\CSLLeftMargin{14. }
+\CSLLeftMargin{18. }
 \CSLRightInline{\textbf{Reid A}, \textbf{Buckley M}. 2011. The rare
@@
 
+\leavevmode\hypertarget{ref-Schloss2009}{}%
+\CSLLeftMargin{19. }
+\CSLRightInline{\textbf{Schloss PD}, \textbf{Westcott SL},
+\textbf{Ryabin T}, \textbf{Hall JR}, \textbf{Hartmann M},
+\textbf{Hollister EB}, \textbf{Lesniewski RA}, \textbf{Oakley BB},
+\textbf{Parks DH}, \textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres
+B}, \textbf{Thallinger GG}, \textbf{Horn DJV}, \textbf{Weber CF}. 2009.
+Introducing mothur: Open-source, platform-independent,
+community-supported software for describing and comparing microbial
+communities. Applied and Environmental Microbiology
+\textbf{75}:7537--7541.
+doi:\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.}
+
+\leavevmode\hypertarget{ref-Westcott2017}{}%
+\CSLLeftMargin{20. }
+\CSLRightInline{\textbf{Westcott SL}, \textbf{Schloss PD}. 2017.
+{OptiClust}, an improved method for assigning amplicon-based sequence
+data to operational taxonomic units. {mSphere} \textbf{2}.
+doi:\href{https://doi.org/10.1128/mspheredirect.00073-17}{10.1128/mspheredirect.00073-17}.}
+
+\leavevmode\hypertarget{ref-Anderson2001}{}%
+\CSLLeftMargin{21. }
+\CSLRightInline{\textbf{Anderson MJ}. 2001. A new method for
+non-parametric multivariate analysis of variance. Austral Ecology
+\textbf{26}:32--46.
+doi:\href{https://doi.org/10.1111/j.1442-9993.2001.01070.pp.x}{10.1111/j.1442-9993.2001.01070.pp.x}.}
+
 \leavevmode\hypertarget{ref-Li2015}{}%
-\CSLLeftMargin{15. }
+\CSLLeftMargin{22. }
 \CSLRightInline{\textbf{Li Q}, \textbf{Heist EP}, \textbf{Moe LA}. 2015.
@@
 \leavevmode\hypertarget{ref-Baxter2016}{}%
-\CSLLeftMargin{16. }
+\CSLLeftMargin{23. }
 \CSLRightInline{\textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers
@@
 \leavevmode\hypertarget{ref-Beall2015}{}%
-\CSLLeftMargin{17. }
+\CSLLeftMargin{24. }
 \CSLRightInline{\textbf{Beall BFN}, \textbf{Twiss MR}, \textbf{Smith
@@
 \leavevmode\hypertarget{ref-Henson2016}{}%
-\CSLLeftMargin{18. }
+\CSLLeftMargin{25. }
 \CSLRightInline{\textbf{Henson MW}, \textbf{Pitre DM}, \textbf{Weckhorst
@@
 \leavevmode\hypertarget{ref-Baxter2014}{}%
-\CSLLeftMargin{19. }
+\CSLLeftMargin{26. }
 \CSLRightInline{\textbf{Baxter NT}, \textbf{Wan JJ}, \textbf{Schubert
@@
 \leavevmode\hypertarget{ref-LevyBooth2018}{}%
-\CSLLeftMargin{20. }
+\CSLLeftMargin{27. }
 \CSLRightInline{\textbf{Levy-Booth DJ}, \textbf{Giesbrecht IJW},
@@
 \leavevmode\hypertarget{ref-Edwards2015}{}%
-\CSLLeftMargin{21. }
+\CSLLeftMargin{28. }
 \CSLRightInline{\textbf{Edwards J}, \textbf{Johnson C},
@@
 \leavevmode\hypertarget{ref-Ettinger2017}{}%
-\CSLLeftMargin{22. }
+\CSLLeftMargin{29. }
 \CSLRightInline{\textbf{Ettinger CL}, \textbf{Williams SL},
@@
 \leavevmode\hypertarget{ref-Graw2018}{}%
-\CSLLeftMargin{23. }
+\CSLLeftMargin{30. }
 \CSLRightInline{\textbf{Graw MF}, \textbf{DAngelo G}, \textbf{Borchers
@@
 \leavevmode\hypertarget{ref-Johnston2016}{}%
-\CSLLeftMargin{24. }
+\CSLLeftMargin{31. }
 \CSLRightInline{\textbf{Johnston ER}, \textbf{Rodriguez-R LM},
@@
 \leavevmode\hypertarget{ref-Hassell2018}{}%
-\CSLLeftMargin{25. }
+\CSLLeftMargin{32. }
 \CSLRightInline{\textbf{Hassell N}, \textbf{Tinker KA}, \textbf{Moore
@@
 \endhead
-Bioethanol (15) & 95 & 3,972,943 & 16,015 & 3,688-356,136 & 96.6 &
+Bioethanol (22) & 95 & 3,972,943 & 16,015 & 3,688-356,136 & 96.6 &
 SRP055545\tabularnewline
-Human (16) & 490 & 20,909,768 & 32,505 & 10,523-430,415 & 40.9 &
+Human (23) & 490 & 20,909,768 & 32,505 & 10,523-430,415 & 40.9 &
 SRP062005\tabularnewline
-Lake (17) & 52 & 3,169,868 & 69,041 & 15,347-112,871 & 7.4 &
+Lake (24) & 52 & 3,169,868 & 69,041 & 15,347-112,871 & 7.4 &
 SRP050963\tabularnewline
-Marine (18) & 7 & 1,391,396 & 193,464 & 133,516-254,060 & 1.9 &
+Marine (25) & 7 & 1,391,396 & 193,464 & 133,516-254,060 & 1.9 &
 SRP068101\tabularnewline
-Mice (1) & 348 & 2,813,747 & 6,477 & 1,804-30,565 & 16.9 &
+Mice (2) & 348 & 2,813,747 & 6,477 & 1,804-30,565 & 16.9 &
 SRP192323\tabularnewline
-Peromyscus (19) & 111 & 1,555,545 & 12,446 & 4,464-33,644 & 7.5 &
+Peromyscus (26) & 111 & 1,555,545 & 12,446 & 4,464-33,644 & 7.5 &
 SRP044050\tabularnewline
-Rainforest (20) & 69 & 946,295 & 11,561 & 4,932-37,767 & 7.7 &
+Rainforest (27) & 69 & 946,295 & 11,561 & 4,932-37,767 & 7.7 &
 ERP023747\tabularnewline
-Rice (21) & 490 & 22,591,168 & 43,216 & 2,776-193,464 & 69.7 &
+Rice (28) & 490 & 22,591,168 & 43,216 & 2,776-193,464 & 69.7 &
 SRP044745\tabularnewline
-Seagrass (22) & 286 & 4,130,454 & 13,567 & 1,803-45,191 & 25.1 &
+Seagrass (29) & 286 & 4,130,454 & 13,567 & 1,803-45,191 & 25.1 &
 SRP092441\tabularnewline
-Sediment (23) & 58 & 1,154,174 & 17,584 & 7,685-68,321 & 8.9 &
+Sediment (30) & 58 & 1,154,174 & 17,584 & 7,685-68,321 & 8.9 &
 SRP097192\tabularnewline
-Soil (24) & 18 & 956,656 & 51,844 & 47,806-59,956 & 1.3 &
+Soil (31) & 18 & 956,656 & 51,844 & 47,806-59,956 & 1.3 &
 ERP012016\tabularnewline
-Stream (25) & 201 & 21,162,574 & 90,159 & 9,175-390,964 & 42.6 &
+Stream (32) & 201 & 21,162,574 & 90,159 & 9,175-390,964 & 42.6 &
 SRP075852\tabularnewline

### 2020-12-10 submission/manuscript.Rmd
Final edits

@@
 
+package_version <- function(package){
+
+  paste(unlist(packageVersion(package)), collapse='.')
+
+}
+
 oxford_comma <- function(x, digits=2) {
@@
 
-# Removal of rare sequences from 16S rRNA gene sequence surveys biases the interpretation of community structure data
+# Removal of rare amplicon sequence variants from 16S rRNA gene sequence surveys biases the interpretation of community structure data
 
@@
 ## Research article format
-
-**word choice: sample vs. community**
-**word choice: dataset vs. sample**
 
@@
 
-Methods for remediating PCR and sequencing artifacts in 16S rRNA gene sequence collections are in continuous development.
-The choices that researchers make have significant ramifications on the inferences that can be drawn from microbial ecology studies.
-One common approach is to remove rare sequences from datasets.
-But, the threshold that is applied is generally selected without regard for the number of sequences in the sample or the variation in sequencing depth across samples within a study.
-I analyzed the impact of removing rare sequences on metrics of alpha and beta diversity using samples collected across `r nrow(table_1)` published datasets.
-Removal of rare sequences significantly decreased the number of amplicon sequence variants and operational taxonomic units as well as their diversity.
+Methods for remediating PCR and sequencing artifacts in 16S rRNA gene sequence collections are in continuous development and have significant ramifications on the inferences that can be drawn.
+A common approach is to remove rare amplcon sequence variants (ASVs) from datasets.
+But, the definition of rarity is generally selected without regard for the number of sequences in the samples or the variation in sequencing depth across samples within a study.
+I analyzed the impact of removing rare ASVs on metrics of alpha and beta diversity using samples collected across `r nrow(table_1)` published datasets.
+Removal of rare ASVs significantly decreased the number of ASVs and operational taxonomic units as well as their diversity.
 Furthermore, their removal increased the variation in community structure between samples.
-When a known effect size was simulated, removal of rare sequences reduced the statistical power to detect the effect relative to not removing rare sequences.
-Although removal of rare sequences did not affect the probability of falsely detetecting differences when samples were randomized to simulate a null model, the false detection rate increased when samples were assigned to simulated treatment groups according to their sequencing depth and rare sequences were removed.
-The false detection rate did not vary when rare sequences were retained.
-This analysis demonstrates the problems inherent in removing rare sequences.
-Researchers are encouraged to retain these sequences, to select approaches that minimize PCR and sequencing artifacts, and to use rarefaction to control for uneven sampling effort.
+When simulating a known effect size, removal of rare ASVs reduced the power to detect the effect relative to not removing rare ASVs.
+Removal of rare ASVs did not affect the false detection rate when samples were randomized to simulate a null model.
+However, the false detection rate increased when rare ASVs were removed using a null distribution and assignment of samples to simulated treatment groups according to their sequencing depth.
+The false detection rate did not vary when rare ASVs were retained.
+This analysis demonstrates the problems inherent in removing rare ASVs.
+Researchers are encouraged to retain rare ASVs, to select approaches that minimize PCR and sequencing artifacts, and to use rarefaction to control for uneven sequencing effort.
 
@@
 
-Removing rare sequences from 16S rRNA gene sequence collections is an approach that has grown in popularity for handling PCR and sequencing artifacts.
+Removing rare amplicon sequence variants (ASVs) from 16S rRNA gene sequence collections is an approach that has grown in popularity for limiting PCR and sequencing artifacts.
 Yet, it is unclear what impact an abundance-based filter has on downstream analyses.
-To investigate the effects of removing rare sequences, I analyzed the community distributions found in the samples of `r nrow(table_1)` published datasets.
-Analysis of these data and simulations based on them showed that removal of rare sequences distorts the representation of microbial communities.
+To investigate the effects of removing rare ASVs, I analyzed the community distributions found in the samples of `r nrow(table_1)` published datasets.
+Analysis of these data and simulations based on them showed that removal of rare ASVs distorts the representation of microbial communities.
 This has the effect of artificially making it more difficult to detect differences between treatment groups.
-Also of concern was the observation that if sequencing depth is confounded with treatment group, then the probability of falsely detecting a difference between the treatment groups increases with the removal of rare sequences.
-The practice of removing rare sequences should stop, lest researcher adversely affect the interpretation of their data.
+Also of concern was the observation that if sequencing depth is confounded with the treatment, then the probability of falsely detecting a difference between the treatment groups increased with the removal of rare ASVs.
+The practice of removing rare ASVs should stop, lest researcher adversely affect the interpretation of their data.
 
@@
 In one approach, paired sequence reads are aligned and any discrepencies between the reads are resolved based on the difference in quality score for the position in question [@Kozich2013; @Edgar2016; @Masella2012]; quality scores are also used to curate single reads [@Bokulich2013].
-In addition, a polishing step is often used to identify ideal sequecnces and their related artifacts based on the frequency of the sequences and their similarity sequences [@Kozich2013; @Edgar2016; @Amir2017].
-In a second approach, the quality scores and types of errors to cluster sequences directly [@Callahan2016].
-Regardless of the approach, many pipelines advocate for abundance-based sceeening where rare sequences are removed from each dataset prior to outputting the sequence data as ASVs [@Edgar2016; @Amir2017; @Callahan2016].
+In addition, a polishing step is often used to identify ASVs based on the frequency and similarity of sequences [@Kozich2013; @Edgar2016; @Amir2017].
+In a second approach, the quality scores and types of errors are modelled to cluster sequence reads directly into ASVs [@Callahan2016].
+Regardless of the approach, many pipelines advocate for abundance-based screening where rare sequences are removed from each dataset prior to outputting the sequence data as ASVs [@Edgar2016; @Amir2017; @Callahan2016].
 Some algorithms recommend removing all ASVs that appear one (i.e.
@@
 A notable exception, the mothur-based pipeline discourages the practice of removing rare sequences [@Kozich2013].
-After assigning sequences to ASVs, they are often analyzed as a taxonomic unit or clustered to generate operational taxonomic units (OTUs) or phylotypes.
+After assigning sequences to ASVs, ASVs are often analyzed as a taxonomic unit or clustered to generate operational taxonomic units (OTUs) or phylotypes.
 
@@
 However, it is common that the number of sequences generated from each sample may vary by two or three orders of magnitude (e.g.
-Figure S1).
+Table 1 and Figure S1).
 An ASV that appears once in a sample with 2,000 sequences is more trustworthy than an ASV that appears once in a sample with 100,000 sequences since it has a 50-fold higher relative abundance.
@@
 For example, the abundance of an ASV may be below the threshold in some samples or time points and above the threshold in others.
-However, rarity, both in terms of prevalence and incidence, is an important ecological concept.
+However, rarity, both in terms of prevalence and incidence, is an important ecological concept [@Reid2011].
 Removing rare ASVs likely hinders one's to ability to make inferences about the dynamics and nature of the populations that rare ASVs represent.
@@
 
-**Datasets.** I collected `r nrow(table_1)` publicly available datasets that used the Illumina MiSeq platform to sequence the V4 region of the 16S rRNA gene from a variety of environments (Table 1).
-After removing poor quality and chimeric ASVs and samples that had uncharacteristically low number of sequences for the dataset, these datasets included between `r oxford_comma(n_sample_range)` samples (Figure S1).
-The median number of sequences for each dataset ranged between `r oxford_comma(n_med_sequence_range)` (Table 1).
-Strikingly, aside from the relatively small `r oxford_comma(low_fold_samples)` datasets, the difference between the sample with the fewest sequences and the sample with the most sequences for each dataset varied by between `r oxford_comma(high_fold_range, digits=1)`-fold (Table 1).
+**Datasets.** I collected `r nrow(table_1)` publicly available datasets that used the Illumina MiSeq platform to sequence the V4 region of the 16S rRNA gene from a variety of environments (Table 1; Figure S1).
+After removing poor quality and chimeric ASVs and samples that had uncharacteristically low number of sequences for the dataset, these datasets included between `r oxford_comma(n_sample_range)` samples.
+The median number of sequences for each dataset ranged between `r oxford_comma(n_med_sequence_range)`.
+Strikingly, aside from the relatively small `r oxford_comma(low_fold_samples)` datasets, the difference between the sample with the fewest sequences and the sample with the most sequences for each dataset varied by between `r oxford_comma(high_fold_range, digits=1)`-fold.
 
@@
 This effectively made every community in a study a statistical sample of the study-wide composite community distribution.
-For example, after this procedure, the 490 samples from the human dataset would be expected to have the same richness and diversity of ASVs and one would not expect to find treatment-based effects between the samples.
+For example, after this procedure, each of the 490 samples from the human dataset would be expected to have the same richness and diversity of ASVs and one would not expect to find treatment-based effects between the samples.
 Because of the risk of bias if only one representation of the null distribution was generated, I generated 100 randomized datasets for each study.
-The trends between removing rare ASVs and the richness, diversity, and information loss that were identified using the observed community community distribution data were also identified with the data from the null distribution; however, the losses were larger when using the null distribution data (Figure S3).
+The trends between removing rare ASVs and the richness, diversity, and information loss that were identified using the observed community distribution data were also identified with the data from the null distribution; however, the losses were larger when using the null distribution data (Figure S3).
 The null distribution data were used in the remainder of the study to minimize the risk of bias.
@@
 Among the remaining datasets, the power to detect a difference in the richness of ASVs ranged between `r oxford_comma(pull(b_base_pc_sobs,'1'), 2)` (`r oxford_comma(pull(b_base_pc_sobs,dataset))`) and between `r oxford_comma(pull(b_base_pc_shannon, '1'), 2)` (`r oxford_comma(pull(b_base_pc_shannon, dataset))`) to detect a difference in diversity when using a Wilcox test (Figure 4A).
-When singleton ASVs were removed, the power to detect a difference in the diversity of ASVs dropped by between `r oxford_comma(pull(b_pc_sobs, percentage_drop), 1)`% (`r oxford_comma(pull(b_pc_sobs, dataset))`) and by between `r oxford_comma(pull(b_pc_shannon, percentage_drop), 1)`% (`r oxford_comma(pull(b_pc_shannon, dataset))`; Figure 4B).
+When singleton ASVs were removed, the power to detect a difference in the richness of ASVs dropped by between `r oxford_comma(pull(b_pc_sobs, percentage_drop), 1)`% (`r oxford_comma(pull(b_pc_sobs, dataset))`) and by between `r oxford_comma(pull(b_pc_shannon, percentage_drop), 1)`% for their diversity(`r oxford_comma(pull(b_pc_shannon, dataset))`; Figure 4B).
 The effect of removing rare ASVs on the richness of OTUs and their diversity was similar (Figure S5AB).
@@
 Previous work has shown that assembly of fully overlapping sequence reads results in the lowest sequencing error rates [@Kozich2013].
-The studies highlighted in this analysis sequenced the V4 region of the 16S rRNA gene using the Illumina MiSeq sequencing platform with their v2 chemistry.
-The resulting data consists of two 250 nt reads that span a region that is about 250 nt long.
+The studies highlighted in this analysis sequenced the V4 region of the 16S rRNA gene using the Illumina MiSeq sequencing platform with their version 2 chemistry.
+The resulting data consists of two 250 nt reads that span a region that is about 253 nt long.
 In contrast, there has been a movement to sequence longer regions with similar chemistry resulting in less overlap between the sequencing reads [@Fadrosh2014; @Holm2019].
@@
 Instead, researchers should prioritize the quality over the quantity and length of their data.
-For these reasons, this analysis did not use lower quality data generated by alternative methods
+For these reasons, this analysis did not use lower quality data generated by alternative methods.
 
 The impacts of removing rare sequences on the representation of communities are caused by the uneven impact of applying a single abundance threshold across all samples.
-The number of sequences per sample in the datasets highlighted in this analysis varied by 100-fold (Figure S1); I suspect the range is actually larger since researchers may have opted against depositing samples with fewer reads into the SRA.
+The number of sequences per sample in the datasets highlighted in this analysis varied by as much as 100-fold (Table 1; Figure S1).
+In practice, I suspect this range is actually larger since researchers may have opted against depositing samples with fewer reads into the SRA.
 If the range in read coverage for a study is 100-fold, a singleton in a small sample would have had a comparable relative abundance as a sequence that appeared 100 times in a more densely sequenced sample.
@@
 
-**Data curation and analysis.** To insure the highest possible data quality, datasets were limited to those where the 500 cycle v2 MiSeq chemistry was used to sequence the amplicons.
+**Data curation and analysis.** To insure the highest possible data quality, datasets were limited to those where the 500 cycle version 2 MiSeq chemistry was used to sequence the amplicons.
 The paired 250 nt reads resulted in near complete 2-fold sequencing coverage of every nucleotide in the ca.
-250 nt-long region.
+253 nt-long region.
 This region and sequencing platform were selected because previous work has shown that a standard data analysis pipeline in mothur results in a sequencing error rate below 0.02% [@Kozich2013].
@@
 
-**Reproducibility.** All analyses were performed using mothur (version 1.44.1) and R (version `r paste(R.version$major, R.version$minor, sep='.')`) with the tidyverse (v.`r packageVersion("tidyverse")`), broom (v.`r packageVersion("broom")`), data.table (v.`r packageVersion("data.table")`), and cowplot (v.`r packageVersion("cowplot")`) packages.
+**Reproducibility.** All analyses were performed using mothur (version 1.44.1) and R (version `r paste(R.version$major, R.version$minor, sep='.')`) with the tidyverse (version `r package_version("tidyverse")`), broom (version `r package_version("broom")`), data.table (version `r package_version("data.table")`), and cowplot (version `r package_version("cowplot")`) packages.
 All of the code that was used in this analysis as well as the Makefile for running the analysis are available at https://github.com/SchlossLab/Schloss_Singletons_XXXXX_2019 along with the complete version history of the project.

### 2020-12-10 submission/manuscript.tex
Final edits

@@
 
-\hypertarget{removal-of-rare-sequences-from-16s-rrna-gene-sequence-surveys-biases-the-interpretation-of-community-structure-data}{%
-\section{Removal of rare sequences from 16S rRNA gene sequence surveys
-biases the interpretation of community structure
-data}\label{removal-of-rare-sequences-from-16s-rrna-gene-sequence-surveys-biases-the-interpretation-of-community-structure-data}}
+\hypertarget{removal-of-rare-amplicon-sequence-variants-from-16s-rrna-gene-sequence-surveys-biases-the-interpretation-of-community-structure-data}{%
+\section{Removal of rare amplicon sequence variants from 16S rRNA gene
+sequence surveys biases the interpretation of community structure
+data}\label{removal-of-rare-amplicon-sequence-variants-from-16s-rrna-gene-sequence-surveys-biases-the-interpretation-of-community-structure-data}}
 
@@
 
-\textbf{word choice: sample vs.~community} \textbf{word choice: dataset
-vs.~sample}
-
 \newpage
@@
 Methods for remediating PCR and sequencing artifacts in 16S rRNA gene
-sequence collections are in continuous development. The choices that
-researchers make have significant ramifications on the inferences that
-can be drawn from microbial ecology studies. One common approach is to
-remove rare sequences from datasets. But, the threshold that is applied
-is generally selected without regard for the number of sequences in the
-sample or the variation in sequencing depth across samples within a
-study. I analyzed the impact of removing rare sequences on metrics of
-alpha and beta diversity using samples collected across 12 published
-datasets. Removal of rare sequences significantly decreased the number
-of amplicon sequence variants and operational taxonomic units as well as
-their diversity. Furthermore, their removal increased the variation in
-community structure between samples. When a known effect size was
-simulated, removal of rare sequences reduced the statistical power to
-detect the effect relative to not removing rare sequences. Although
-removal of rare sequences did not affect the probability of falsely
-detetecting differences when samples were randomized to simulate a null
-model, the false detection rate increased when samples were assigned to
-simulated treatment groups according to their sequencing depth and rare
-sequences were removed. The false detection rate did not vary when rare
-sequences were retained. This analysis demonstrates the problems
-inherent in removing rare sequences. Researchers are encouraged to
-retain these sequences, to select approaches that minimize PCR and
-sequencing artifacts, and to use rarefaction to control for uneven
-sampling effort.
+sequence collections are in continuous development and have significant
+ramifications on the inferences that can be drawn. A common approach is
+to remove rare amplcon sequence variants (ASVs) from datasets. But, the
+definition of rarity is generally selected without regard for the number
+of sequences in the samples or the variation in sequencing depth across
+samples within a study. I analyzed the impact of removing rare ASVs on
+metrics of alpha and beta diversity using samples collected across 12
+published datasets. Removal of rare ASVs significantly decreased the
+number of ASVs and operational taxonomic units as well as their
+diversity. Furthermore, their removal increased the variation in
+community structure between samples. When simulating a known effect
+size, removal of rare ASVs reduced the power to detect the effect
+relative to not removing rare ASVs. Removal of rare ASVs did not affect
+the false detection rate when samples were randomized to simulate a null
+model. However, the false detection rate increased when rare ASVs were
+removed using a null distribution and assignment of samples to simulated
+treatment groups according to their sequencing depth. The false
+detection rate did not vary when rare ASVs were retained. This analysis
+demonstrates the problems inherent in removing rare ASVs. Researchers
+are encouraged to retain rare ASVs, to select approaches that minimize
+PCR and sequencing artifacts, and to use rarefaction to control for
+uneven sequencing effort.
 
@@
 
-Removing rare sequences from 16S rRNA gene sequence collections is an
-approach that has grown in popularity for handling PCR and sequencing
-artifacts. Yet, it is unclear what impact an abundance-based filter has
-on downstream analyses. To investigate the effects of removing rare
-sequences, I analyzed the community distributions found in the samples
-of 12 published datasets. Analysis of these data and simulations based
-on them showed that removal of rare sequences distorts the
-representation of microbial communities. This has the effect of
+Removing rare amplicon sequence variants (ASVs) from 16S rRNA gene
+sequence collections is an approach that has grown in popularity for
+limiting PCR and sequencing artifacts. Yet, it is unclear what impact an
+abundance-based filter has on downstream analyses. To investigate the
+effects of removing rare ASVs, I analyzed the community distributions
+found in the samples of 12 published datasets. Analysis of these data
+and simulations based on them showed that removal of rare ASVs distorts
+the representation of microbial communities. This has the effect of
 artificially making it more difficult to detect differences between
 treatment groups. Also of concern was the observation that if sequencing
-depth is confounded with treatment group, then the probability of
-falsely detecting a difference between the treatment groups increases
-with the removal of rare sequences. The practice of removing rare
-sequences should stop, lest researcher adversely affect the
-interpretation of their data.
+depth is confounded with the treatment, then the probability of falsely
+detecting a difference between the treatment groups increased with the
+removal of rare ASVs. The practice of removing rare ASVs should stop,
+lest researcher adversely affect the interpretation of their data.
 
@@
 also used to curate single reads (5). In addition, a polishing step is
-often used to identify ideal sequecnces and their related artifacts
-based on the frequency of the sequences and their similarity sequences
-(2, 3, 6). In a second approach, the quality scores and types of errors
-to cluster sequences directly (7). Regardless of the approach, many
-pipelines advocate for abundance-based sceeening where rare sequences
-are removed from each dataset prior to outputting the sequence data as
-ASVs (3, 6, 7). Some algorithms recommend removing all ASVs that appear
-one (i.e.~singletons) (7), eight (3), or ten (6) or fewer times; these
-pipelines also vary in whether the minimum abundance threshold should be
-applied to individual samples (6, 7) or the pool of samples in a study
-(3). A notable exception, the mothur-based pipeline discourages the
-practice of removing rare sequences (2). After assigning sequences to
-ASVs, they are often analyzed as a taxonomic unit or clustered to
-generate operational taxonomic units (OTUs) or phylotypes.
+often used to identify ASVs based on the frequency and similarity of
+sequences (2, 3, 6). In a second approach, the quality scores and types
+of errors are modelled to cluster sequence reads directly into ASVs (7).
+Regardless of the approach, many pipelines advocate for abundance-based
+screening where rare sequences are removed from each dataset prior to
+outputting the sequence data as ASVs (3, 6, 7). Some algorithms
+recommend removing all ASVs that appear one (i.e.~singletons) (7), eight
+(3), or ten (6) or fewer times; these pipelines also vary in whether the
+minimum abundance threshold should be applied to individual samples (6,
+7) or the pool of samples in a study (3). A notable exception, the
+mothur-based pipeline discourages the practice of removing rare
+sequences (2). After assigning sequences to ASVs, ASVs are often
+analyzed as a taxonomic unit or clustered to generate operational
+taxonomic units (OTUs) or phylotypes.
 
@@
 However, it is common that the number of sequences generated from each
-sample may vary by two or three orders of magnitude (e.g.~Figure S1). An
-ASV that appears once in a sample with 2,000 sequences is more
-trustworthy than an ASV that appears once in a sample with 100,000
+sample may vary by two or three orders of magnitude (e.g.~Table 1 and
+Figure S1). An ASV that appears once in a sample with 2,000 sequences is
+more trustworthy than an ASV that appears once in a sample with 100,000
 sequences since it has a 50-fold higher relative abundance. But,
@@
 others. However, rarity, both in terms of prevalence and incidence, is
-an important ecological concept. Removing rare ASVs likely hinders one's
-to ability to make inferences about the dynamics and nature of the
+an important ecological concept (8). Removing rare ASVs likely hinders
+one's to ability to make inferences about the dynamics and nature of the
 populations that rare ASVs represent. Furthermore, removing ASVs whose
@@
 the Illumina MiSeq platform to sequence the V4 region of the 16S rRNA
-gene from a variety of environments (Table 1). After removing poor
-quality and chimeric ASVs and samples that had uncharacteristically low
-number of sequences for the dataset, these datasets included between 7
-and 490 samples (Figure S1). The median number of sequences for each
-dataset ranged between 6,477 and 193,464 (Table 1). Strikingly, aside
-from the relatively small marine and soil datasets, the difference
-between the sample with the fewest sequences and the sample with the
-most sequences for each dataset varied by between 7.4 and 96.6-fold
-(Table 1).
+gene from a variety of environments (Table 1; Figure S1). After removing
+poor quality and chimeric ASVs and samples that had uncharacteristically
+low number of sequences for the dataset, these datasets included between
+7 and 490 samples. The median number of sequences for each dataset
+ranged between 6,477 and 193,464. Strikingly, aside from the relatively
+small marine and soil datasets, the difference between the sample with
+the fewest sequences and the sample with the most sequences for each
+dataset varied by between 7.4 and 96.6-fold.
 
@@
 every community in a study a statistical sample of the study-wide
-composite community distribution. For example, after this procedure, the
-490 samples from the human dataset would be expected to have the same
-richness and diversity of ASVs and one would not expect to find
+composite community distribution. For example, after this procedure,
+each of the 490 samples from the human dataset would be expected to have
+the same richness and diversity of ASVs and one would not expect to find
 treatment-based effects between the samples. Because of the risk of bias
@@
 removing rare ASVs and the richness, diversity, and information loss
-that were identified using the observed community community distribution
-data were also identified with the data from the null distribution;
-however, the losses were larger when using the null distribution data
-(Figure S3). The null distribution data were used in the remainder of
-the study to minimize the risk of bias.
+that were identified using the observed community distribution data were
+also identified with the data from the null distribution; however, the
+losses were larger when using the null distribution data (Figure S3).
+The null distribution data were used in the remainder of the study to
+minimize the risk of bias.
 
@@
 diversity when using a Wilcox test (Figure 4A). When singleton ASVs were
-removed, the power to detect a difference in the diversity of ASVs
+removed, the power to detect a difference in the richness of ASVs
 dropped by between 27.3 and 92.9\% (bioethanol and soil) and by between
-40.0 and 93.3\% (rainforest and soil; Figure 4B). The effect of removing
-rare ASVs on the richness of OTUs and their diversity was similar
-(Figure S5AB). I used the Bray-Curtis dissimilarity index to compare the
-simulated communities within each dataset and calculated the power to
-detect differences between the two simulated treatment groups using the
-analysis of molecular variance (Figure 4C and S5C). Without removing
-rare sequences, the power to detect a difference between the two
-simulated treatment groups varied between 0.41 and 1.00 (rainforest and
-rainforest). Aside from the bioethanol, human, and mice datasets, the
-power to detect differences dropped by between 6.5 and 64.0\% (soil and
-rice) when singletons were removed. However, when ASVs that occurred 10
-or fewer times were removed from each sample, the power to detect
+40.0 and 93.3\% for their diversity(rainforest and soil; Figure 4B). The
+effect of removing rare ASVs on the richness of OTUs and their diversity
+was similar (Figure S5AB). I used the Bray-Curtis dissimilarity index to
+compare the simulated communities within each dataset and calculated the
+power to detect differences between the two simulated treatment groups
+using the analysis of molecular variance (Figure 4C and S5C). Without
+removing rare sequences, the power to detect a difference between the
+two simulated treatment groups varied between 0.41 and 1.00 (rainforest
+and rainforest). Aside from the bioethanol, human, and mice datasets,
+the power to detect differences dropped by between 6.5 and 64.0\% (soil
+and rice) when singletons were removed. However, when ASVs that occurred
+10 or fewer times were removed from each sample, the power to detect
 differences dropped by 12.0 and 97.2\% (human and peromyscus); similar
@@
 relative to those in the oral cavity leading to a significant difference
-in sequencing depth between treatment groups (8). When rare sequences
+in sequencing depth between treatment groups (9). When rare sequences
 were not removed, the fraction of significant tests did not differ from
@@
 V4 region of the 16S rRNA gene using the Illumina MiSeq sequencing
-platform with their v2 chemistry. The resulting data consists of two 250
-nt reads that span a region that is about 250 nt long. In contrast,
-there has been a movement to sequence longer regions with similar
-chemistry resulting in less overlap between the sequencing reads (9,
-10). Alternatively, others have prioritized increasing the number of
-sequences per sample by sequencing the V4 region, but with paired or
-single reads that are 150 nt long (11, 12). Both practices result in a
+platform with their version 2 chemistry. The resulting data consists of
+two 250 nt reads that span a region that is about 253 nt long. In
+contrast, there has been a movement to sequence longer regions with
+similar chemistry resulting in less overlap between the sequencing reads
+(10, 11). Alternatively, others have prioritized increasing the number
+of sequences per sample by sequencing the V4 region, but with paired or
+single reads that are 150 nt long (12, 13). Both practices result in a
 significantly higher error rate for the resulting assmebly. Instead,
@@
 of their data. For these reasons, this analysis did not use lower
-quality data generated by alternative methods
+quality data generated by alternative methods.
 
@@
 abundance threshold across all samples. The number of sequences per
-sample in the datasets highlighted in this analysis varied by 100-fold
-(Figure S1); I suspect the range is actually larger since researchers
-may have opted against depositing samples with fewer reads into the SRA.
-If the range in read coverage for a study is 100-fold, a singleton in a
-small sample would have had a comparable relative abundance as a
-sequence that appeared 100 times in a more densely sequenced sample.
-However, it would have been removed from the smaller sample and not the
-larger sample. Thus, applying a single abundance threshold
-disproportionately impacted the samples with fewer sequences. This seems
-to contradict the purpose of removing rare sequences since a singleton
-in the smaller sample is more reliable than the singleton in the larger
-sample. A superior approach to removing rare sequences is to use
-rarefaction to conrol for uneven sampling.
+sample in the datasets highlighted in this analysis varied by as much as
+100-fold (Table 1; Figure S1). In practice, I suspect this range is
+actually larger since researchers may have opted against depositing
+samples with fewer reads into the SRA. If the range in read coverage for
+a study is 100-fold, a singleton in a small sample would have had a
+comparable relative abundance as a sequence that appeared 100 times in a
+more densely sequenced sample. However, it would have been removed from
+the smaller sample and not the larger sample. Thus, applying a single
+abundance threshold disproportionately impacted the samples with fewer
+sequences. This seems to contradict the purpose of removing rare
+sequences since a singleton in the smaller sample is more reliable than
+the singleton in the larger sample. A superior approach to removing rare
+sequences is to use rarefaction to conrol for uneven sampling.
 
@@
 Previous work sequencing mock communities has shown that the number of
-spurious sequences increases with sequencing depth (2, 13). By rarefying
+spurious sequences increases with sequencing depth (2, 14). By rarefying
 data to a common number of sequences per sample, the number of spurious
@@
 inferences that they make about rare sequences. This is a standard that
-should be applied regardless of their abundance (14).
+should be applied regardless of their abundance (15).
 
@@
 challenge since microbial ecologists transitioned away from Sanger
-sequencing of samples (15--17). Although the extent of the ``rare
+sequencing of samples (16--18). Although the extent of the ``rare
 biosphere'' is still an open question, it is important to appreciate the
@@
 role of such populations or the processes that drive their patchy
-distributions that are so common with microbial communities (18).
+distributions that are so common with microbial communities (8).
 
@@
 \textbf{Data curation and analysis.} To insure the highest possible data
-quality, datasets were limited to those where the 500 cycle v2 MiSeq
-chemistry was used to sequence the amplicons. The paired 250 nt reads
-resulted in near complete 2-fold sequencing coverage of every nucleotide
-in the ca. 250 nt-long region. This region and sequencing platform were
-selected because previous work has shown that a standard data analysis
-pipeline in mothur results in a sequencing error rate below 0.02\% (2).
-All sequence data were obtained from the Sequence Read Archive (SRA) and
-processed using a standard mothur-based sequencing pipeline that
-resulted in ASVs as generated by the pre.cluster algorithm using a
-threshold of 2 nt (2, 19). ASVs were assigned to OTUs using a 3\%
-distance threshold using mothur's cluster function with the OptiClust
-algorithm (20). To minimize the effects of uneven sampling effort,
-samples were rarefied to the number of sequences in the smallest sample
-for each dataset. Because metrics of alpha diversity did not
+quality, datasets were limited to those where the 500 cycle version 2
+MiSeq chemistry was used to sequence the amplicons. The paired 250 nt
+reads resulted in near complete 2-fold sequencing coverage of every
+nucleotide in the ca. 253 nt-long region. This region and sequencing
+platform were selected because previous work has shown that a standard
+data analysis pipeline in mothur results in a sequencing error rate
+below 0.02\% (2). All sequence data were obtained from the Sequence Read
+Archive (SRA) and processed using a standard mothur-based sequencing
+pipeline that resulted in ASVs as generated by the pre.cluster algorithm
+using a threshold of 2 nt (2, 19). ASVs were assigned to OTUs using a
+3\% distance threshold using mothur's cluster function with the
+OptiClust algorithm (20). To minimize the effects of uneven sampling
+effort, samples were rarefied to the number of sequences in the smallest
+sample for each dataset. Because metrics of alpha diversity did not
 consistently follow a normal distribution, I used the non-parametric
@@
 \textbf{Reproducibility.} All analyses were performed using mothur
-(version 1.44.1) and R (version 4.0.2) with the tidyverse (v.130), broom
-(v.070), data.table (v. 113 4), and cowplot (v.110) packages. All of the
-code that was used in this analysis as well as the Makefile for running
-the analysis are available at
+(version 1.44.1) and R (version 4.0.2) with the tidyverse (version
+1.3.0), broom (version 0.7.0), data.table (version 1.13.4), and cowplot
+(version 1.1.0) packages. All of the code that was used in this analysis
+as well as the Makefile for running the analysis are available at
 \url{https://github.com/SchlossLab/Schloss_Singletons_XXXXX_2019} along
@@
 
+\leavevmode\hypertarget{ref-Reid2011}{}%
+\CSLLeftMargin{8. }
+\CSLRightInline{\textbf{Reid A}, \textbf{Buckley M}. 2011. The rare
+biosphere.}
+
 \leavevmode\hypertarget{ref-Morris2013}{}%
-\CSLLeftMargin{8. }
+\CSLLeftMargin{9. }
 \CSLRightInline{\textbf{Morris A}, \textbf{Beck JM}, \textbf{Schloss
@@
 \leavevmode\hypertarget{ref-Fadrosh2014}{}%
-\CSLLeftMargin{9. }
+\CSLLeftMargin{10. }
 \CSLRightInline{\textbf{Fadrosh DW}, \textbf{Ma B}, \textbf{Gajer P},
@@
 \leavevmode\hypertarget{ref-Holm2019}{}%
-\CSLLeftMargin{10. }
+\CSLLeftMargin{11. }
 \CSLRightInline{\textbf{Holm JB}, \textbf{Humphrys MS}, \textbf{Robinson
@@
 \leavevmode\hypertarget{ref-Thompson2017}{}%
-\CSLLeftMargin{11. }
+\CSLLeftMargin{12. }
 \CSLRightInline{\textbf{Thompson LR}, \textbf{Jon G. Sanders and},
@@
 \leavevmode\hypertarget{ref-McDonald2018}{}%
-\CSLLeftMargin{12. }
+\CSLLeftMargin{13. }
 \CSLRightInline{\textbf{McDonald D}, \textbf{Hyde E}, \textbf{Debelius
@@
 \leavevmode\hypertarget{ref-Schloss2011}{}%
-\CSLLeftMargin{13. }
+\CSLLeftMargin{14. }
 \CSLRightInline{\textbf{Schloss PD}, \textbf{Gevers D}, \textbf{Westcott
@@
 \leavevmode\hypertarget{ref-Schloss2018}{}%
-\CSLLeftMargin{14. }
+\CSLLeftMargin{15. }
 \CSLRightInline{\textbf{Schloss PD}. 2018. Identifying and overcoming
@@
 \leavevmode\hypertarget{ref-Sogin2006}{}%
-\CSLLeftMargin{15. }
+\CSLLeftMargin{16. }
 \CSLRightInline{\textbf{Sogin ML}, \textbf{Morrison HG}, \textbf{Huber
@@
 \leavevmode\hypertarget{ref-Huse2010}{}%
-\CSLLeftMargin{16. }
+\CSLLeftMargin{17. }
 \CSLRightInline{\textbf{Huse SM}, \textbf{Welch DM}, \textbf{Morrison
@@
 \leavevmode\hypertarget{ref-Reeder2009}{}%
-\CSLLeftMargin{17. }
+\CSLLeftMargin{18. }
 \CSLRightInline{\textbf{Reeder J}, \textbf{Knight R}. 2009. The {{}}rare
@@
 doi:\href{https://doi.org/10.1038/nmeth0909-636}{10.1038/nmeth0909-636}.}
-
-\leavevmode\hypertarget{ref-Reid2011}{}%
-\CSLLeftMargin{18. }
-\CSLRightInline{\textbf{Reid A}, \textbf{Buckley M}. 2011. The rare
-biosphere.}
 

### 2021-10-06 submission/manuscript.tex
Add running title

@@
 
-\vspace{35mm}
+\vspace{10mm}
+
+\textbf{Running title:} Removal of rare ASVs biases community analyses
+
+\vspace{25mm}
 
@@
 \textbf{Reproducibility.} All analyses were performed using mothur
-(version 1.44.1) and R (version 4.0.2) with the tidyverse (version
-1.3.0), broom (version 0.7.0), data.table (version 1.13.4), and cowplot
+(version 1.44.1) and R (version 4.0.3) with the tidyverse (version
+1.3.1), broom (version 0.7.8), data.table (version 1.13.4), and cowplot
 (version 1.1.0) packages. All of the code that was used in this analysis

### 2021-10-06 code/plot_correlation_coverage.R
Remove samples with too few sequences

@@ -23,14 +23,14 @@ correlations <- bind_rows(
 		cor_coverage=read_tsv('data/process/sequence_coverage_table_cor.tsv'),
 		.id="comparison"
 	) %>%
- 	filter(min_freq == 1) %>%
+ 	filter(freq_removed == 1) %>%
   mutate(sig = if_else(cor_n_p.value < 0.05, NA_character_, "not\nsignificant"),
     value = if_else(is.na(sig), cor_n_estimate, NA_real_)
 	) %>%
 	select(comparison, dataset, value, sig)
 
 coverage <- read_tsv('data/process/sequence_coverage_table_raw.tsv') %>%
-	filter(min_freq == 1) %>%
+	filter(freq_removed == 1) %>%
 	group_by(dataset) %>%
 	summarize(
 		value=median(coverage),

### 2021-10-06 code/quantify_correlation_with_sample_size.R

@@ -7,16 +7,16 @@ input_file <- input[1] # e.g. input_file <- "data/process/sequence_loss_table_ra
 output_file <- str_replace(input_file, "raw", "cor")
 
 read_tsv(input_file) %>%
-	group_by(dataset, min_freq) %>%
+	group_by(dataset, freq_removed) %>%
 	nest() %>%
 	mutate(cor_n = map(data, ~tidy(cor.test(.x$n, .x[[4]], method="spearman", exact=F))),
 				cor_shannon = map(data, ~tidy(cor.test(.x$shannon, .x[[4]], method="spearman", exact=F))),
 				cor_nshannon = map(data, ~tidy(cor.test(.x$shannon, .x$n, method="spearman", exact=F)))
 		) %>%
-	select(dataset, min_freq, cor_n, cor_shannon, cor_nshannon) %>%
+	select(dataset, freq_removed, cor_n, cor_shannon, cor_nshannon) %>%
 	ungroup() %>%
 	unnest(c(cor_n, cor_shannon, cor_nshannon), names_sep="_") %>%
-	select(dataset, min_freq,
+	select(dataset, freq_removed,
 					cor_n_estimate, cor_n_p.value,
 					cor_shannon_estimate, cor_shannon_p.value,
 					cor_nshannon_estimate, cor_nshannon_p.value) %>%

### 2021-10-06 code/quantify_sample_coverage.R

@@ -1,8 +1,8 @@
 library(tidyverse)
 library(data.table)
 
-datasets <- c("bioethanol", "human", "lake", "marine", "mice", "peromyscus", "rainforest", "rice",
-	"seagrass", "sediment", "soil", "stream")
+datasets <- c("bioethanol", "human", "lake", "marine", "mice", "peromyscus",
+	"rainforest", "rice", "seagrass", "sediment", "soil", "stream")
 
 get_shannon <- function(x) {
 
@@ -13,10 +13,10 @@ get_shannon <- function(x) {
 }
 
 
-get_coverage <- function(test_sample, all_counts, min_freq = 1){
+get_coverage <- function(test_sample, all_counts, freq_removed = 1){
 
 	sample_ntons <- all_counts %>%
-		filter(sample == test_sample, n_seqs == min_freq) %>%
+		filter(sample == test_sample, n_seqs == freq_removed) %>%
 		select(sequences)
 
 	all_counts %>%
@@ -31,34 +31,38 @@ get_singleton_coverage <- function(dataset){
 
 	print(dataset)
 
+	to_remove <- scan(paste0("data/", dataset, "/data.remove_accnos"), what=character(), quiet=TRUE) %>%
+		str_split(., pattern="-") %>%
+		unlist()
+
 	count_data <- paste0("data/", dataset, "/data.count_table") %>%
 		fread(., colClasses=c(Representative_Sequence="character")) %>%
 		melt(id.vars=c("Representative_Sequence"), variable.name="sample", variable.factor=F, value.name="n_seqs") %>%
 		filter(sample != "total") %>%
 		rename(sequences=Representative_Sequence) %>%
-		filter(n_seqs != 0)
+		filter(n_seqs != 0 & !sample %in% to_remove) %>% as_tibble()
 
 	count_data %>%
 		group_by(sample) %>%
 		nest() %>%
-		mutate(coverage_1 = map(sample, ~get_coverage(., all_counts = count_data, min_freq=1)),
-					coverage_2 = map(sample, ~get_coverage(., all_counts = count_data, min_freq=2)),
-					coverage_3 = map(sample, ~get_coverage(., all_counts = count_data, min_freq=3)),
-					coverage_4 = map(sample, ~get_coverage(., all_counts = count_data, min_freq=4)),
-					coverage_5 = map(sample, ~get_coverage(., all_counts = count_data, min_freq=5)),
[…]

### 2021-10-06 code/quantify_sequence_loss.R

@@ -18,17 +18,20 @@ get_shannon <- function(x) {
 }
 
 
-
 get_loss_table <- function(dataset){
 
 	print(dataset)
+
+	to_remove <- scan(paste0("data/", dataset, "/data.remove_accnos"), what=character(), quiet=TRUE) %>%
+		str_split(., pattern="-") %>%
+		unlist()
 
 	paste0("data/", dataset, "/data.count_table") %>%
 		fread(., colClasses=c(Representative_Sequence="character")) %>%
 		melt(id.vars=c("Representative_Sequence"), variable.name="sample", variable.factor=F, value.name="n_seqs") %>%
 	  filter(sample != "total") %>%
 		rename(sequences=Representative_Sequence) %>%
-		filter(n_seqs != 0) %>%
+		filter(n_seqs != 0 & !sample %in% to_remove) %>%
 		group_by(sample) %>%
 		summarize(n = sum(n_seqs),
 			shannon = get_shannon(n_seqs),
@@ -41,11 +44,12 @@ get_loss_table <- function(dataset){
 			n_7 = frac_n(n_seqs, 7),
 			n_8 = frac_n(n_seqs, 8),
 			n_9 = frac_n(n_seqs, 9),
-			n_10 = frac_n(n_seqs, 10)
+			n_10 = frac_n(n_seqs, 10),
+			.groups = "drop"
 		) %>%
-		pivot_longer(starts_with("n_"), names_to="min_freq", values_to="fraction_lost") %>%
+		pivot_longer(starts_with("n_"), names_to="freq_removed", values_to="fraction_lost") %>%
 		mutate(dataset = dataset,
-					min_freq = str_replace(min_freq, "n_", "")) %>%
+					freq_removed = str_replace(freq_removed, "n_", "")) %>%
 		select(dataset, sample, everything())
 
 }

### 2021-10-06 submission/response_to_reviewers.md
Rough in response to reviewers

**I read the manuscript  "Removal of rare amplicon sequence variants from 16S rRNA gene sequence surveys biases the interpretation of community structure data" from Dr. Schloss with great interest. There are a diversity of opinions about whether ASV or OTU tables should or should not be filtered or quality controlled prior to subsequent statistical investigation and an even greater diversity about the specific approaches that should be applied. This may hold important implications for whether and which patterns are observed in the analysis of microbiome data, yet there has been surprisingly little work to measure what the effects of these decisions actually are. Dr. Schloss' effort to evaluate the effect of these decisions in a data-driven way thus holds broad implications for the entire field of microbiome data analysis. In particular, Dr. Schloss analyzed a variety of publicly available datasets to determine how the removal of rare ASVs or OTUs affects the patterns present in microbiome data. This extensive analysis seeks to clarify whether the common approach of removing rare variants - which are often presumed to represent spurious sequences (i.e., artifacts) - from an analysis is meritorious. Dr. Schloss' analysis points towards two big conclusions: (1) rare variants add information to microbiome studies and should be retained and (2) rarefaction techniques should be used to control sequencing depth and thus the number of artifacts present in an analysis.**

**While these are important questions to address, I do not walk away from the manuscript - in its current form - convinced by its conclusions, especially given some of the strong language that is invoked. In particular, I think the manuscript could make a better case by improving its clarity in some areas, especially with respect to methodology and rationale, and by revising or adding analyses in other places to solidify conclusions. Moreover, some of the conclusions may require hedging given caveats that are not currently considered in the present analysis. Given that this manuscript upends long-standing practice and expectations, it seems especially important that it fill in these gaps.**

**Below, I itemize my criticisms with the manuscript. I hope Dr. Schloss sees these as constructive comments designed to help him improve this manuscript, which I ultimately think will play an important role in the advancement of microbiome data analytics.**

* Grateful that the reviewer sees value in this work and the questions I have raised.

* A significant challenge of this type of work is that we do not know the *true* answer for any of these communities. Any analysis that has been done regarding thresholds for removing sequences has been based on mock communities which are very much not realistic

* Point out that aside from those problematic mock community analyses there has been little to no work demonstrating that we *should* be removing singletons. That decision was made by fiat rather than based on data.

**1. It is currently challenging to evaluate the work because the methodological details used to conduct it are frequently sparsely described. The materials and methods are in effect integrated into the results section. While I do not have a problem with this per se, it seems to have come at the cost of not providing specific details about various analytical steps that were implemented as part of the analysis as well as the rationale for implementing these techniques. This ranges from relatively straightforward aspects of methods, such as the reason for why these specific data sets were selected (from among the others that fit the same criterion of targeting the V4 locus with the same sequencing technology), to whether multiple tests correction was applied to statistical tests. In other places, the methodology is confusing as described, especially regarding the various null model sampling approaches (for example, see my specific comment about line 200 below). While I appreciate that Dr. Schloss has made all code available to readers, I think it's important to ensure that an understanding of the approach by a reader does not depend upon their ability to read code. I therefore recommend that the technical approach be more explicitly described in the manuscript.**

* More justification for why studies were selected
* Have removed description of the methods from the Results section
* Statistical tests
* More detailed model descriptions

**2. I'm not sure that the manuscript provides evidence that rare sequences should be retained in an investigation, but rather documents the sensitivity of commonly applied analytical approaches to the decisions made about whether and how rare sequences should be incorporated. The former statement requires demonstrating that the inclusion of rare variants improves analytical accuracy. While the present analysis suggests that including rare variants impacts alpha- and beta-diversity metrics as well as the sensitivity of tests that utilize these metrics (according to the null model approach), it is not evident if the filtered or unfiltered data better reflects the underlying ecology. As a brief aside, Figure 5 offers the clearest insight into the potential impact of filtering on analytical accuracy, but it is not clear from the methodology how treatment groups were established and thus it is hard to follow Dr. Schloss' interpretation here.**

* Problem with knowing *truth*
* All models/simulations are worthless
* Figures 4 and 5 show that removing rare sequences causes problems for statistical power (Figure 4) and false detection of differences (Figure 5B)

**In general, much of the interpretation of the results appears predicated on the idea that the rare variants represent real taxa. However, there is only limited evidence in support of this view. In the section "The nature of singletons", Dr. Schloss finds that singletons are more prevalent in more shallowly sequenced samples and that singletons in a sample are often also present in another sample. Dr. Schloss concludes that these singletons are thus likely real variants and the rest of the investigation seems somewhat predicated upon this interpretation. However, I am left with several questions:**

**(A) Are the correlations meaningfully representative of a trend across all samples, or are they driven by a small number of outlier samples? Providing support of these trends by plotting the scatterplots for each study would help assuage this concern.**

* Used Spearman correlation coefficients, which are robust to the effects of outliers
* Create a new Figure S2 that shows the requested plots for the data in Fig 1AC

**(B) Despite these observations, couldn't these rare taxa include artifacts nonetheless and, if so, do real rare variants and artifacts contribute uniform information in all subsequent analyses? An effort to directly measure the impact of spurious sequences on analytical accuracy would help offset this concern.**

* Lengthy discussion of the fact that there are likely spurious sequences in the rare pool
* Value of rarefaction to control for the higher likelihood of crummy sequences with greater sequencing depth
* Make comparisons based on relative differences in diversity rather than absolute quantities

**(C) Since the processes underlying the generation of artifacts are not explicitly described, the reader is left unsure about the expectation that the formation of singleton artificats is driven by sequencing depth.**

* Hopefully the revised text will clarify these issues.

**3. I'm not sure that the description of variant filtering accurately reflects common practice. For example, while many groups appear to apply count thresholds to determine which taxa should be retained for analysis, these thresholds are typically applied across all samples and are often cross referenced with prevalence filtering. As a result, taxa that are singletons in one sample but present in others are typically included in an analysis despite otherwise being below the count threshold. In Dr. Schloss' study, however, such taxa are considered candidates for filtration, which raises questions about the applicability of the results. Moreover, frequent practice thresholds on relative abundance rather than direct counts to offset the potential for sequencing depths to drive singleton formation effects. Dr. Schloss may consider discussing or even experimentally measuring the effect of these alternative strategies on the outcomes of his analyses.**

* As outlined in the Introduction, sample-by-sample abundance filtering is the norm for current approaches to identifying ASVs
* Prevalence filtering...
* Relative abundance is effectively the same as abundance filtering. The relative abundance would indicate the average number of counts for the OTU across a large number of resamplings

**4. Related to the above, it is not clear if the results presented here are generalizable to other investigations. For example, it is possible that the use of the high-quality data sets included in this analysis or the specific bioinformatic techniques applied here results in a lower than typical rate of artifact formation compared to what many readers may encounter in their own analyses. For example, studies that do not apply library prep approaches that produce two overlapping V4 reads, but that rather amplify a longer amplicon target and thus have lower sequence accuracy, may have an increased rate of forming artifacts in their data. Likewise, ASVs in this study are sequences that tolerate a 2 bp difference, which may differ from the set of ASVs produced using another approach. Since the study never directly measures the cost of including artifactual sequences (2B above), it is not clear if the recommendations made here are meaningful for investigations that involve data subject to a greater rate of artifact formation.**

* I do not believe that authors have an excuse to generate data with low quality data. As outlined in the Discussion, this is part of the problem with the field currently.

**5. Conclusions around rarefaction need more justification or support. There are several strongly worded statements in the manuscript about the use of rarefaction techniques, but there appears to be little direct assessment of the impact of rarefaction on study outcomes. Additionally, there is no comparison with alternative rarefaction-free normalization techniques, including simply filtering based on relative abundance. Given the on-going debate in the literature about the value of rarefaction, it seems important that the manuscript solidify this point if the language is to be included as is.**

* Even Holmes's study says to rarefy alpha and beta diversity metrics
* Add citations to rarefaction yes papers
* Analysis of rarefaction is beyond the scope of this paper

**6. The rationale underlying some of the analyses, especially with respect to the null models, needs support to help readers connect the results with the conclusions. For example, for the section on line 122 ("The impact of removing rare ASVs on the information represented in each sample.") it seems self-evident that removing rare taxa will affect the estimate of alpha-diversity, especially when considering the well documented rank abundance distribution and long-tail of low abundance taxa that typically comprises microbial communities. Dr. Schloss may get more mileage out of this section by clarifying why this analysis is noteworthy, or by focusing of metrics that are less directly sensitive to the number of taxa (perhaps evenness).**

* Sobs is definitely obvious
* Bigger point in Figure 3 is that the coefficient of variation between samples goes up with removal of rare sequences reducing power to detect differences in communities
* Will also caculate Shannon evenness (keep in mind that this is dependent on richness)

**Additionally, I'm not sure that the treatment effect simulation analysis is as meaningful as indicated by the text (though again, I'm not sure I follow the approach properly). If the ASV sampling procedure to determine which ASVs are increased in abundance is simply selecting from a list of ASVs, then the long-tail of rare ASVs will tend to be selected versus those that tend to abundant. Consequently, when the procedure amplifies the abundance of these (and only these) selected ASVs, again which are biased towards rare taxa, then it will be rare taxa driving the differences between simulated treatments. Abundance-based filtering of said taxa will obviously have an effect on the analysis. Is this akin to what happens in microbial communities in nature? Well, maybe in some instances, but lots of studies find that it's the abundant taxa that differentiate treatments. Such studies may not need the singleton ASVs to observe the effect. Could those ASVs help clarify the effect? Perhaps. But inclusion of artifact sequences (which again are not directly tested here) could theoretically increase the error of the analysis in such cases at a greater rate. I would recommend considering a parallel analysis that simulates ASVs based on their abundance to help clarify whether the interpretation drawn from this section is relevant to communities that behave along the lines described above. Alternatively, hedge the conclusion by clarifying the above caveat, but that this analysis may yet reveal insights into how rare ASVs could reveal differences currently being missed in communities that do not substantially change with respect to the abundant fraction of community membership.**

* As the reviewer points out, we don't know. There will certainly be times where differences between communities are driven by rare populations and times where it splits by abundant populations. We really don't know. Those that are driven by abundant populations will be obviously different already so it's a moot point for that case
* High quality data reduces risk of artifacts and rarefaction allows us to control for inclusion of artifacts

**Minor comments:**

**Line 94: What is meant by "relatively small" when referring to the marine and soil data sets? Based on table 1, it seems to be the number of samples. But you can help readers by being more explicit here especially given that the surrounding context of discussion pertains to sequencing depth.**

* Yes number of samples. Clarify text

**Line 102: "In addition, with the exception of the samples from the marine and sediment datasets (Spearman correlation, P>0.05)..." Indicate parenthetically here that these samples were non-significant (as opposed to the immediate assumption that they are positively associated) and move the parenthetical pointer to spearman and the pvalue to later in the sentence.**

* Clarify text

**Line 148: "For example, after this procedure, each of the 490 samples from the human dataset would be expected to have the same richness  and diversity of ASVs and one would not expect to find treatment-based effects between the samples." It would be more compelling to demonstrate that this expectation was met in the resulting distributions. Can you offer graphical or statistical evidence in support of this?**

* That is shown in Figure 3 when rare sequences were not removed

**Line 175: How many samples are being selected for each simulated treatment in this analysis?  At first I thought all samples in a study were selected, but it looks like this isn't the case, as an undescribed number of "communities were randomly sampled from the null distribution..."**

* Clarify that the samples were sampled into one of two equally sized treatment groups

**Line 179: What p-value threshold is being applied to "detect a difference" and quantify the "fraction of significant tests"?**

* Clarify that it is P<0.05

**Line 200: It's a little challenging to determine what happened here: "First, I sampled sequences from the 200 null distribution for each dataset and randomly assigned each sample to one of two treatment groups and determined the richness and diversity of ASVs and OTUs". For example, what is meant by "sequences" here? ASVs? How did you sample the sequences? Are you saying that you subsampled the data somehow? I recommend you clarify.**

* Clarify text

**Line 204: What is meant by "meaningfully vary"? Was this quantified in some manner?**

* Indicate a range? Seems clear in Fig 5A and S6A

**Line 215: "However, when rare taxa of any frequency were removed, the probability of falsely detecing a difference as signifiant increased with the definition of rarity". But Fig 5B seems to show that the rate only tends to increase when singletons are removed, such that the rate is flat as more abundant rare sequences are removed.**

* Yes it's flat, but the rate increased when any rare sequences were removed. That's the problem. The value is the lowest when all sequences were retained.

**Line 236: "The practice of removing rare sequences from samples seems to be a response to researchers prioritizing the number of reads and length of sequences over their quality." I'm not sure this is entirely accurate. Some of it stems from this, but it also appears to have emerged as a way to control for overclustering, which often results in spurious OTUs. This methodology may have been a holdover in the era of ASV analysis, where it may not be as critical.**

* I stand by this comment

**Line 250: "In practice, I suspect this range is actually larger since researchers may have opted against depositing samples with fewer reads into the SRA". Perhaps. But this would imply that such samples were not (nor would be) included in the analysis and not subject to the bias being discussed here.**

* Sentence removed

**Line 255: I think I follow your intention, but note that your results of finding that rare sequences are higher in shallow depth samples seems initially at odds with the subsequent part of the discussion that observes that "number of spurious sequences increases with sequencing depth". You can avoid some confusion by using this observation is the expectation and then discussing how your results differ from this expectation, reinforcing the interpretation that the rare sequences observed in your analysis are real variants. This may be your intention in the current narrative; I simply recommend a modest restructure of this logic to improve comprehension.**

* Clarify text

**Line 287: Typo. Change insure -> ensure**

* Corrected

### 2021-10-21 submission/response_to_reviewers.md
Show raw data that went into correlations

@@
 
-* Used Spearman correlation coefficients, which are robust to the effects of outliers
-* Create a new Figure S2 that shows the requested plots for the data in Fig 1AC
+I appreciate the reviewer's concern that the correlations presented in Figure 1AC could be driven by outliers.
+It is worth noting that I used Spearman correlation coefficients, which are robust to the effects of outliers since it is a ranked-based test.
+As suggested, by the reviewer, I created two new figures, Figures S2 and S3, which shows the requested plots for the data in Fig 1AC.
+I think it's pretty clear that the trend is robust to outliers.
 

### 2021-10-21 submission/manuscript.Rmd
Add running title
Show raw data that went into correlations

@@
 
-\vspace{35mm}
+\vspace{10mm}
+
+**Running title:** Removal of rare ASVs biases community analyses
+
+\vspace{25mm}
 
@@
 
-```{r}
+```{r results_a}
 
 frac_lost <- read_tsv(here('data/process/sequence_loss_table_raw.tsv')) %>%
-	filter(min_freq == 1) %>%
+	filter(freq_removed == 1) %>%
 	group_by(dataset) %>%
@@
 cor_n <- read_tsv(here('data/process/sequence_loss_table_cor.tsv')) %>%
- 	filter(min_freq == 1) %>%
+ 	filter(freq_removed == 1) %>%
 	select(dataset, cor_n_estimate, cor_n_p.value) %>%
@@
 frac_coverage <- read_tsv(here('data/process/sequence_coverage_table_raw.tsv')) %>%
-	filter(min_freq == 1) %>%
+	filter(freq_removed == 1) %>%
 	group_by(dataset) %>%
@@
 coverage_n <- read_tsv(here('data/process/sequence_coverage_table_cor.tsv')) %>%
-	filter(min_freq == 1) %>%
+	filter(freq_removed == 1) %>%
 	select(dataset, cor_n_estimate, cor_n_p.value) %>%
@@
 Contrary to this expectation, the median percentage of sequences that were discarded when singleton ASVs were removed from each dataset varied between `r oxford_comma(percent_singletons_range_vals)`% (`r oxford_comma(percent_singletons_range_names)`).
-In addition, with the exception of the samples from the `r oxford_comma(cor_n_nosig)` datasets (Spearman correlation, P>0.05), the fraction of singleton ASVs in samples was negatively correlated with the number of sequences in each sample with a range between `r oxford_comma(cor_n_sig_values)` (`r oxford_comma(cor_n_sig_names)`) (Figure 1A).
+In addition, with the exception of the samples from the `r oxford_comma(cor_n_nosig)` datasets (Spearman correlation, P>0.05), the fraction of singleton ASVs in samples was negatively correlated with the number of sequences in each sample with a range between `r oxford_comma(cor_n_sig_values)` (`r oxford_comma(cor_n_sig_names)`) (Figures 1A and S2).
 This showed that with additional sequencing, the probability of seeing singleton ASVs in multiple samples was greater than the probability of generating an artifact.
@@
 Considering the likelihood of finding an ASV duplicated in another sample is confounded by the number of samples and inter-sample diversity, the high coverage of singleton ASVs in these datasets was remarkable.
-The correlation between the number of sequences in a sample and the fraction of that sample's singleton ASVs that were covered by another sample in the dataset was significant and negative for `r nrow(coverage_n) - length(coverage_n_nosig)` of the datasets ranging between `r oxford_comma(coverage_n_sig_values)` for the `r oxford_comma(coverage_n_sig_names)` datasets, respectively (Figure 1C).
+The correlation between the number of sequences in a sample and the fraction of that sample's singleton ASVs that were covered by another sample in the dataset was significant and negative for `r nrow(coverage_n) - length(coverage_n_nosig)` of the datasets ranging between `r oxford_comma(coverage_n_sig_values)` for the `r oxford_comma(coverage_n_sig_names)` datasets, respectively ((Figures 1C and S3)).
 The negative correlation indicated that the singleton ASVs in the smaller samples were more likely to be covered by ASVs in the larger samples.
@@
 
-```{r}
+```{r  results_b}
 ointra <- read_tsv(here("data/process/ointra_analysis.tsv"),
@@
 Next, I assigned the ASVs to OTUs, which were defined as a group of ASVs that were more than 97% similar to each other to assess the impact of removing rare ASVs on higher level taxonomic groupings that are commonly used in microbial ecology studies.
-Although pooling similar ASVs into OTUs reduced the impact of removing the rare ASVs relative to the ASV-based analysis, the minimum abundance threshold still decreased the richness of OTUs and the diversity decreased relative to the full community (Figure S2AB).
+Although pooling similar ASVs into OTUs reduced the impact of removing the rare ASVs relative to the ASV-based analysis, the minimum abundance threshold still decreased the richness of OTUs and the diversity decreased relative to the full community (Figure S4AB).
 In contrast to the richness and diversity measurements, the Kullback–Leibler divergence compares the relative abundance of specific ASVs or OTUs between representations of the community.
 I calculated the Kullback–Leibler divergence between the full communities and those where rare ASVs were removed.
-As the threshold for removing ASVs increased, the amount of information lost also increased for both ASVs and OTUs (Figure 2C and Figure S2C).
+As the threshold for removing ASVs increased, the amount of information lost also increased for both ASVs and OTUs (Figure 2C and Figure S4C).
 The relative loss of information was generally smaller for OTUs than than it was for ASVs.
@@
 Because of the risk of bias if only one representation of the null distribution was generated, I generated 100 randomized datasets for each study.
-The trends between removing rare ASVs and the richness, diversity, and information loss that were identified using the observed community distribution data were also identified with the data from the null distribution; however, the losses were larger when using the null distribution data (Figure S3).
+The trends between removing rare ASVs and the richness, diversity, and information loss that were identified using the observed community distribution data were also identified with the data from the null distribution; however, the losses were larger when using the null distribution data (Figure S5).
 The null distribution data were used in the remainder of the study to minimize the risk of bias.
 
-```{r}
+```{r results_c}
 r_cov_alpha <- read_tsv(here("data/process/ralpha_analysis.tsv"),
@@
 The COVs between Bray-Curtis distances within a study when singletons were removed was between `r oxford_comma(pull(pc_cov_bc, cv_diff), 1)`-times larger than when they were not removed (`r oxford_comma(pull(pc_cov_bc, dataset))`; Figure 3C).
-When ASVs were clustered into OTUs the difference in COVs was less than it was for the ASVs (Figure S4).
+When ASVs were clustered into OTUs the difference in COVs was less than it was for the ASVs (Figure S6).
 These results indicate that removing rare ASVs increases the dissimilarity between samples, which could have a significant impact on the statistical power to detect differences between treatment groups.
 
-```{r}
+```{r results_d}
 b_alpha <- read_tsv(here("data/process/bffect_alpha_analysis.tsv"),
@@
 When singleton ASVs were removed, the power to detect a difference in the richness of ASVs dropped by between `r oxford_comma(pull(b_pc_sobs, percentage_drop), 1)`% (`r oxford_comma(pull(b_pc_sobs, dataset))`) and by between `r oxford_comma(pull(b_pc_shannon, percentage_drop), 1)`% for their diversity(`r oxford_comma(pull(b_pc_shannon, dataset))`; Figure 4B).
-The effect of removing rare ASVs on the richness of OTUs and their diversity was similar (Figure S5AB).
-I used the Bray-Curtis dissimilarity index to compare the simulated communities within each dataset and calculated the power to detect differences between the two simulated treatment groups using the analysis of molecular variance (Figure 4C and S5C).
+The effect of removing rare ASVs on the richness of OTUs and their diversity was similar (Figure S7AB).
+I used the Bray-Curtis dissimilarity index to compare the simulated communities within each dataset and calculated the power to detect differences between the two simulated treatment groups using the analysis of molecular variance (Figure 4C and S7C).
 Without removing rare sequences, the power to detect a difference between the two simulated treatment groups varied between `r oxford_comma(pull(b_base_beta, '1') %>% range(), 2)` (`r slice(b_base_beta, 1) %>% pull(dataset)` and `r oxford_comma(slice_max(b_base_beta, '1') %>% pull(dataset))`).
@@
 
-```{r}
+```{r results_e}
 type_one_null_alpha_summary <-
@@
 Indeed, there was no evidence that removing rare ASVs resulted in an inflated experiment-wise error rate.
-The average fraction of significant tests did not meaningfully vary from 0.05 across the minimum abundance threshold, dataset, metric of describing sample alpha-diversity, or whether the abundance of ASVs or OTUs were used (Figure 5A and S6A).
-Similarly, the average fraction of significant tests did not meaningfully vary from 0.05 when using analysis of molecular variance to compare communities using Bray-Curtis distances (Figure 5A and S6A).
+The average fraction of significant tests did not meaningfully vary from 0.05 across the minimum abundance threshold, dataset, metric of describing sample alpha-diversity, or whether the abundance of ASVs or OTUs were used (Figure 5A and S8A).
+Similarly, the average fraction of significant tests did not meaningfully vary from 0.05 when using analysis of molecular variance to compare communities using Bray-Curtis distances (Figure 5A and S8A).
 Second, I again sampled sequences from the null distribution, but assigned samples to one of two treatment groups based on the number of sequences in each sample.
@@
 This exaggerated bias has been observed in comparisons of the lung and oral microbiota because of the larger number of non-specific amplicons that can be sequenced from lung samples relative to those in the oral cavity leading to a significant difference in sequencing depth between treatment groups [@Morris2013].
-When rare sequences were not removed, the fraction of significant tests did not differ from 5% for comparing the richness, their diversity, or Bray-Curtis distances (Figure 5B and S6B).
-However, when rare taxa of any frequency were removed, the probability of falsely detecing a difference as signifiant increased with the definition of rarity (Figure 5B and S6B).
+When rare sequences were not removed, the fraction of significant tests did not differ from 5% for comparing the richness, their diversity, or Bray-Curtis distances (Figure 5B and S8B).
+However, when rare taxa of any frequency were removed, the probability of falsely detecing a difference as signifiant increased with the definition of rarity (Figure 5B and S8B).
 Not including the small marine dataset, the average fraction the average fraction of falsely detecting a difference across datasets when only singletons were removed was `r 100*min_ave_p`%.
@@
 By rarefying data to a common number of sequences per sample, the number of spurious sequences can be controlled.
-As shown in the data I presented, which was rarefied to a common number of reads per sample within a dataset, when rare sequences were not removed the power to detect differences was the highest and the false discovery rate was the expected 5% (Figures 5 and S6).
+As shown in the data I presented, which was rarefied to a common number of reads per sample within a dataset, when rare sequences were not removed the power to detect differences was the highest and the false discovery rate was the expected 5% (Figures 5 and S8).
 
@@
 
-```{r}
+```{r results_f}
 table_1 %>%
@@
 The median value is shown with a solid circle and the 95% confidence interval is indicated by the solid line.
+The raw data for A and C are presented in Figures S2 and S3, respectively.
 
@@
 **Figure S2.
-Removing rare sequences from samples alters their representation of alpha-diversity using operational taxonomic units (OTUs).** The average difference in the richness (A), Shannon diversity (B), and Kullback-Leiber divergence (C) for each sample within a dataset was calculated between the original community structures relative to applying different minimum abundance thresholds.
+Singletons are more common in samples with fewer seqeunces.** Spearman correlation coefficients ($\rho$) were calcualted between the total number of sequences and the number of singletons in the sample.
+The Spearman correlation coefficients for the Marine and Sediment datasets were not significantly different from 0.00 (P>0.05).
+The blue line indicates the linear trend in the data.
+These data are summarized in Figure 1A.
 
@@
 **Figure S3.
+Singletons tend to be shared with samples having more sequences.** Spearman correlation coefficients ($\rho$) were calcualted between the total number of sequences and the number of singletons in a sample that were shared with other samples from the same deataset.
+The Spearman correlation coefficients for the Marine, Soil, and Stream datasets were not significantly different from 0.00 (P>0.05).
+The blue line indicates the linear trend in the data.
+These data are summarized in Figure 1C.
+
+\newpage
+
+\includegraphics{figure_s4.png}
+
+**Figure S4.
+Removing rare sequences from samples alters their representation of alpha-diversity using operational taxonomic units (OTUs).** The average difference in the richness (A), Shannon diversity (B), and Kullback-Leiber divergence (C) for each sample within a dataset was calculated between the original community structures relative to applying different minimum abundance thresholds.
+
+\newpage
+
+\includegraphics{figure_s5.png}
+
+**Figure S5.
 Removing rare sequences from samples alters their representation of alpha-diversity when regenerating samples using a null distribution for each dataset.** The average difference in the richness, Shannon diversity, and Kullback-Leiber divergence for each sample within a dataset was calculated between the original community structures relative to applying different minimum abundance thresholds.
@@
 
-\includegraphics{figure_s4.png}
+\includegraphics{figure_s6.png}
 
-**Figure S4.
+**Figure S6.
 Removing rare sequences from samples increases the inter-sample variation for operational taxonomic units (OTUs).** The coefficient of variation in richness (A), Shannon diversity (B), and Bray-Curtis distances (C) for each dataset was calculated using the null distributed samples for each dataset with varying minimum abundance thresholds.
@@
 
-\includegraphics{figure_s5.png}
+\includegraphics{figure_s7.png}
 
-**Figure S5.
+**Figure S7.
 Removing rare sequences from samples reduces the statistical power to detect differences between empirically generated treatment groups when using operational taxonomic units (OTUs).** The fraction of significant tests comparing the richness (A) and Shannon diversity (B) using a Wilcox test and Bray-Curtis distances (C) using analysis of molecular variance for each dataset was calculated using empirically generated treatment groups containing equal numbers of samples for each dataset with varying minimum abundance thresholds.
@@
 
-\includegraphics{figure_s6.png}
+\includegraphics{figure_s8.png}
 
-**Figure S6.
+**Figure S8.
 Removing rare sequences does not impact the false detection rate unless the number of sequences per sample is confounded with the treatment groups when using operational taxonomic units (OTUs).** The fraction of significant tests comparing the richness and Shannon diversity using a Wilcox test and Bray-Curtis distances using analysis of molecular variance for each dataset was calculated.

### 2021-10-21 submission/manuscript.tex
Show raw data that went into correlations

@@
 median percentage of sequences that were discarded when singleton ASVs
-were removed from each dataset varied between 0.42 and 22.23\%
+were removed from each dataset varied between 0.42 and 22.20\%
 (bioethanol and seagrass). In addition, with the exception of the
@@
 negatively correlated with the number of sequences in each sample with a
-range between -0.27 and -0.87 (rice and bioethanol) (Figure 1A). This
-showed that with additional sequencing, the probability of seeing
+range between -0.26 and -0.87 (rice and bioethanol) (Figures 1A and S2).
+This showed that with additional sequencing, the probability of seeing
 singleton ASVs in multiple samples was greater than the probability of
@@
 covered by another sample in the dataset was significant and negative
-for 9 of the datasets ranging between -0.31 and -0.84 for the rice and
-seagrass datasets, respectively (Figure 1C). The negative correlation
-indicated that the singleton ASVs in the smaller samples were more
-likely to be covered by ASVs in the larger samples. Among the three
-datasets without a significant correlation (Spearman correlation,
+for 9 of the datasets ranging between -0.33 and -0.84 for the rice and
+seagrass datasets, respectively ((Figures 1C and S3)). The negative
+correlation indicated that the singleton ASVs in the smaller samples
+were more likely to be covered by ASVs in the larger samples. Among the
+three datasets without a significant correlation (Spearman correlation,
 P\textgreater0.05), the marine and soil datasets had the fewest samples
@@
 richness of OTUs and the diversity decreased relative to the full
-community (Figure S2AB). In contrast to the richness and diversity
+community (Figure S4AB). In contrast to the richness and diversity
 measurements, the Kullback--Leibler divergence compares the relative
@@
 threshold for removing ASVs increased, the amount of information lost
-also increased for both ASVs and OTUs (Figure 2C and Figure S2C). The
+also increased for both ASVs and OTUs (Figure 2C and Figure S4C). The
 relative loss of information was generally smaller for OTUs than than it
@@
 also identified with the data from the null distribution; however, the
-losses were larger when using the null distribution data (Figure S3).
+losses were larger when using the null distribution data (Figure S5).
 The null distribution data were used in the remainder of the study to
@@
 OTUs the difference in COVs was less than it was for the ASVs (Figure
-S4). These results indicate that removing rare ASVs increases the
+S6). These results indicate that removing rare ASVs increases the
 dissimilarity between samples, which could have a significant impact on
@@
 effect of removing rare ASVs on the richness of OTUs and their diversity
-was similar (Figure S5AB). I used the Bray-Curtis dissimilarity index to
+was similar (Figure S7AB). I used the Bray-Curtis dissimilarity index to
 compare the simulated communities within each dataset and calculated the
 power to detect differences between the two simulated treatment groups
-using the analysis of molecular variance (Figure 4C and S5C). Without
+using the analysis of molecular variance (Figure 4C and S7C). Without
 removing rare sequences, the power to detect a difference between the
@@
 dataset, metric of describing sample alpha-diversity, or whether the
-abundance of ASVs or OTUs were used (Figure 5A and S6A). Similarly, the
+abundance of ASVs or OTUs were used (Figure 5A and S8A). Similarly, the
 average fraction of significant tests did not meaningfully vary from
 0.05 when using analysis of molecular variance to compare communities
-using Bray-Curtis distances (Figure 5A and S6A). Second, I again sampled
+using Bray-Curtis distances (Figure 5A and S8A). Second, I again sampled
 sequences from the null distribution, but assigned samples to one of two
@@
 5\% for comparing the richness, their diversity, or Bray-Curtis
-distances (Figure 5B and S6B). However, when rare taxa of any frequency
+distances (Figure 5B and S8B). However, when rare taxa of any frequency
 were removed, the probability of falsely detecing a difference as
-signifiant increased with the definition of rarity (Figure 5B and S6B).
+signifiant increased with the definition of rarity (Figure 5B and S8B).
 Not including the small marine dataset, the average fraction the average
@@
 highest and the false discovery rate was the expected 5\% (Figures 5 and
-S6).
+S8).
 
@@
 The median value is shown with a solid circle and the 95\% confidence
-interval is indicated by the solid line.
+interval is indicated by the solid line. The raw data for A and C are
+presented in Figures S2 and S3, respectively.
 
@@
 
-\textbf{Figure S2. Removing rare sequences from samples alters their
+\textbf{Figure S2. Singletons are more common in samples with fewer
+seqeunces.} Spearman correlation coefficients (\(\rho\)) were calcualted
+between the total number of sequences and the number of singletons in
+the sample. The Spearman correlation coefficients for the Marine and
+Sediment datasets were not significantly different from 0.00
+(P\textgreater0.05). The blue line indicates the linear trend in the
+data. These data are summarized in Figure 1A.
+
+\newpage
+
+\includegraphics{figure_s3.png}
+
+\textbf{Figure S3. Singletons tend to be shared with samples having more
+sequences.} Spearman correlation coefficients (\(\rho\)) were calcualted
+between the total number of sequences and the number of singletons in a
+sample that were shared with other samples from the same deataset. The
+Spearman correlation coefficients for the Marine, Soil, and Stream
+datasets were not significantly different from 0.00 (P\textgreater0.05).
+The blue line indicates the linear trend in the data. These data are
+summarized in Figure 1C.
+
+\newpage
+
+\includegraphics{figure_s4.png}
+
+\textbf{Figure S4. Removing rare sequences from samples alters their
 representation of alpha-diversity using operational taxonomic units
@@
 
-\includegraphics{figure_s3.png}
-
-\textbf{Figure S3. Removing rare sequences from samples alters their
+\includegraphics{figure_s5.png}
+
+\textbf{Figure S5. Removing rare sequences from samples alters their
 representation of alpha-diversity when regenerating samples using a null
@@
 
-\includegraphics{figure_s4.png}
-
-\textbf{Figure S4. Removing rare sequences from samples increases the
+\includegraphics{figure_s6.png}
+
+\textbf{Figure S6. Removing rare sequences from samples increases the
 inter-sample variation for operational taxonomic units (OTUs).} The
@@
 
-\includegraphics{figure_s5.png}
-
-\textbf{Figure S5. Removing rare sequences from samples reduces the
+\includegraphics{figure_s7.png}
+
+\textbf{Figure S7. Removing rare sequences from samples reduces the
 statistical power to detect differences between empirically generated
@@
 
-\includegraphics{figure_s6.png}
-
-\textbf{Figure S6. Removing rare sequences does not impact the false
+\includegraphics{figure_s8.png}
+
+\textbf{Figure S8. Removing rare sequences does not impact the false
 detection rate unless the number of sequences per sample is confounded

### 2021-10-21 code/plot_correlation_supp.R

@@ -1,0 +1,56 @@
+library(tidyverse)
+library(ggtext)
+library(glue)
+library(broom)
+
+input <- commandArgs(trailingOnly=TRUE)
+input_file <- input[1] # e.g. input_file <- "data/process/sequence_coverage_table_raw.tsv"
+
+tag <- case_when(str_detect(input_file, "loss") ~ "loss",
+					str_detect(input_file, "coverage") ~ "coverage")
+
+output_file <- glue("results/figures/supp_sequence_{tag}_cor.tiff")
+
+y_axis_label <- if_else(tag == "loss",
+		"Fraction of sequences in sample as singletons",
+		"Fraction of singletons shared across samples"
+)
+
+scientific_10 <- function(x) {
+  str_replace(scales::label_scientific(digits=2)(x), "e\\+0", "x10\\^")
+}
+
+
+data <- read_tsv(input_file) %>%
+	filter(freq_removed == 1)
+
+name_cor <- data %>%
+	nest(subset=-dataset) %>%
+	mutate(corr = map(subset,
+			~cor.test(.x[[5]], .x$n, method="spearman", exact=FALSE) %>% tidy())
+	) %>%
+ 	unnest(corr) %>%
+	mutate(rho = if_else(p.value < 0.05,
+						format(round(estimate, 2), nsmall=2L),
+						"NA"),
+				pretty = glue("{str_to_title(dataset)} (\u03C1={rho})")) %>%
+	select(dataset, pretty)
+
+nice_studies <- pull(name_cor, pretty)
+names(nice_studies) <- pull(name_cor, dataset)
+
+data %>%
+	ggplot(aes(x=n, y=.[[6]])) +
+		geom_point(alpha=0.2) +
+		geom_smooth(method="lm", se=FALSE, formula="y~log(x)") +
+		facet_wrap(~dataset, nrow=3, scales="free", labeller=as_labeller(nice_studies)) +
+		scale_x_log10(n.breaks=4, labels=scientific_10) +
+		scale_y_continuous(n.breaks=5, lim=c(0,NA)) +
+		labs(x = "Number of sequences per sample", y=y_axis_label) +
+		theme_classic() +
+		theme(
+			axis.text.x = element_markdown(size=8),
+			strip.background = element_blank()
+		)
+
+ggsave(filename=output_file, width=8, height=7, unit="in")

### 2021-10-21 submission/response_to_reviewers.md
Address reviewer's comments

@@
+I am grateful that the reviewer sees value in this work and the questions I have raised.
+I apologize for taking so long to resubmit a revised version of the manuscript.
+In part because of the reviewers' comments, I paused work on this project to take on the question of whether the use of rarefaction was valid for alpha and beta diversity metrics.
+Two manuscripts: (1) a direct response to the original McMurdie and Holmes study and (2) an approach using a different simulation framework have now been posted as preprints to bioRxiv and submitted for review also at *mSphere* (see the new references **XX** and **XX**).
+Although the Editor's comments indicated that it was not necessary to take on the issue of rarefaction, I felt it would overcome many of the concerns that the reviewer and others have raised.
+Beyond the issue of rarefaction, I hope that the reviewer will find that I have addressed their comments throughout the manuscript and taken to revise the text to make my methods more clear and results more convincing.
+The line numbers included below correspond to the "track changes" version of the manuscript.
+
 **I read the manuscript "Removal of rare amplicon sequence variants from 16S rRNA gene sequence surveys biases the interpretation of community structure data" from Dr.
@@
 
-* Grateful that the reviewer sees value in this work and the questions I have raised.
+I appreciate the reviewers constructive comments and am confident that the manuscript is much improved because of their feedback.
+A significant challenge of this type of work is that we do not know the *true* answer for any of the original communities.
+One could use mock communities and risk over-fitting parameters to a highly artificial construct, which is my critique of the Bokulich et al.
+study (**XX**).
+Alternatively, one could try many methods on multiple studies and look for common results as was recently done by Nearing et al.
+(**XX**); however, the analyses cannot be easily controlled and the conclusions are unsatisfying.
+Finally, the approach I took was to design a thought experiment that asked what the data would look like if rare sequences were artifacts.
+I expected there to be more rare sequences when we had more sequences and there to be little overlap of rare sequences between different samples.
+As outlined in the section **The nature of singletons**, I found the opposite.
+The fraction of singleton ASVs decreased with increasing sequencing depth and there was considerable overlap of rare sequences in a poorly sequenced sample with the same sequences at greater abundance in more thoroughly sequenced samples.
 
-* A significant challenge of this type of work is that we do not know the *true* answer for any of these communities.
-Any analysis that has been done regarding thresholds for removing sequences has been based on mock communities which are very much not realistic
-
-* Point out that aside from those problematic mock community analyses there has been little to no work demonstrating that we *should* be removing singletons.
-That decision was made by fiat rather than based on data.
+There has been little to no rigorous work demonstrating that we *should* be removing low abundance or patchy sequences.
+The decision to employ abundance or prevalence based filtering has made by fiat rather than based on data.
+For example, McMurdie and Holmes decried removing samples and sequence data by rarefaction, but thought nothing of removing rare OTUs in their analysis (**XX**).
+Except for Bokulich et al.
+who justified abundance filtering because of low quality data, none of the papers describing methods for assigning sequences to ASVs experimentally justified their removal of rare sequences.
+Thus, I feel that I am left arguing against a "long-standing practice" with little objective and experimental support.
 
@@
 
-* More justification for why studies were selected
-* Have removed description of the methods from the Results section
-* Statistical tests
-* More detailed model descriptions
+I appreciate the reviewer's concern about the mixing of methods throughout the Results section.
+Clearly, I needed to do a better job of describing those connections and I appreciate the reviewer for pointing out places that could be improved.
+The revised manuscript moves the details of the simulations to the Methods section and replaces that text in the Results section with more rationale for the simulation design.
+I have also tried to improve the description of the statistical tests and thresholds I used with the simulations.
+All the comparisons were binary within each study and did not require use of corrections for multiple comparisons.
+
+There are certainly more datasets that could fit my criteria, but I felt that 12 diverse datasets were sufficient.
+I would be happy to include additional datasets that the Editor or reviewer recommend.
+Ultimately, the actual datasets that are used is not critical as I am only using the datasets to simulate various community structures.
+I feel that the description of why datasets were selected is sufficiently clear in the "Datasets" subsection of the Results section and the "Data curation and analysis" subsection of the Materials and Methods section.
 
@@
 
-* Problem with knowing *truth*
-* All models/simulations are worthless
-* Figures 4 and 5 show that removing rare sequences causes problems for statistical power (Figure 4) and false detection of differences (Figure 5B)
+Inclusion of rare variants definitely improves analytical accuracy.
+The **The nature of singletons** subsection of the Results section highlights why I think that the rare sequences are largely robust.
+However, even if that were not the case, I still think that their removal would be problematic.
+I showed that the inter-sample variation as measured by the coefficient of variation increases with removal of rare sequences (**LXX**) and this impacts the ability to detect differences between samples (**LXX**).
+Furthermore, when sample sizes are confounded with treatment group removal of rare sequences exacerbates the false detection rate (**LXX**).
+
+As mentioned above, a consistent challenge in studies such as mine is our lack of knowledge of the true structure of microbial communities or their underlying ecology.
+Although Mock community data can tell us what is spurious and what is real, their community structure is highly artificial and unlike anything we see in nature.
+Analysis of the original studies also possible but would leave the interpretation of what better reflects the "underlying ecology" to the investigator or an interpretative mess.
+Therefore, we are left to use model communities where we know their true structure and perform simulations based on those models.
+Of course, "all models are wrong, but some are useful" (George Box) and we are left to ask whether the models I've created here illuminate concerns that we should have with removing rare sequences.
+The models presented here do not incorporate sequencing errors per se, rather rare ASVs that may be good or spurious.
+What we find (and the reviewer notes) in Figures 4 and 5 is that regardless of the quality of the sequences, removing rare sequences causes problems for statistical power (Figure 4) and false detection of differences (Figure 5B).
+My models argue that retaining rare ASVs provides a better sense of the ecology of the study because it makes it easier to detect real differences (i.e., power) and harder to detect false differences.
 
@@
 As suggested, by the reviewer, I created two new figures, Figures S2 and S3, which shows the requested plots for the data in Fig 1AC.
-I think it's pretty clear that the trend is robust to outliers.
+There are no obvious outliers in these figures that are skewing the Spearman correlation values.
 
@@
 
-* Lengthy discussion of the fact that there are likely spurious sequences in the rare pool
-* Value of rarefaction to control for the higher likelihood of crummy sequences with greater sequencing depth
-* Make comparisons based on relative differences in diversity rather than absolute quantities
+Rare taxa could certainly be artifacts.
+At L264-267 I discuss how even under the best of circumstances PCR artifacts like chimeras and sequencing errors will persist.
+As part of this discussion I described two strategies for dealing with lingering artifacts: rarefaction (L268-272) and making comparisons between treatment based on relative differences in diversity rather than absolute quantities (L273-283).
+Of course, ideally researchers would generate the best possible sequence data where reads fully overlap with each other.
+Using rarefaction gives every sample an equal chance to have artifacts.
+As I emphasize in the Discussion, by keeping interpretation to relative differences and not inflating the importance of rare taxa we can further mitigate against problems caused by artifacts.
 
-**(C) Since the processes underlying the generation of artifacts are not explicitly described, the reader is left unsure about the expectation that the formation of singleton artificats is driven by sequencing depth.**
+**(C) Since the processes underlying the generation of artifacts are not explicitly described, the reader is left unsure about the expectation that the formation of singleton artifacts is driven by sequencing depth.**
 
-* Hopefully the revised text will clarify these issues.
+I hope that the revised text at **LXX** (2nd paragraph of Discussion) addresses the reviewer's concerns.
 
@@
 
-* As outlined in the Introduction, sample-by-sample abundance filtering is the norm for current approaches to identifying ASVs
-* Prevalence filtering...
-* Relative abundance is effectively the same as abundance filtering. The relative abundance would indicate the average number of counts for the OTU across a large number of resamplings
+* As outlined in the Introduction, sample-by-sample abundance filtering is the norm for dada2, Deblur, and the Bokulich pipeline whereas study-wide abundance filtering is used by UNOISE2.
+ Closed reference clustering as a filtration method...
+
+*   [ simulation? ]
+  - study-wide abundance filtering 
+  - counting the number of samples that have a sample
+
+* The reviewer suggests using relative abundance thresholds as an alternative to abundance thresholds.
+Relative abundance thresholds are effectively the same as rarefied OTU counts since, mathematically, relative abundances are the average number of counts for the OTU across a large number of resamplings divided by the total number of sequences in a sample.
+Unfortunately, filtering by relative abundances do not control for the total number of sequences in a dataset and so any ASV below the threshold will effectively have its abundances redistributed across the other ASVs.
+This is the same problem I encountered when removing rare ASVs.
+[ simulation?
+]
+* Fully overlapping read pairs have previously been recommended for 16S rRNA analyses, since the increase in sequencing errors in nonoverlapping regions can result in spurious OTU calls (11).
+Related to this suggestion, there has also been some concern about the base-quality performance of Illumina’s v3 kit chemistry toward the ends of the 300-nucleotide (nt) reads (20).
+These are valid concerns, and we have tweaked our quality-filtering steps to help minimize these problems.
+The distribution of quality scores in forward and reverse reads is shown in Fig.
+3A and B to help visualize this issue.
+For both forward and reverse reads, there is a decrease in quality near the 3′ ends.
+After stitching these 300-nt reads together (Fig.
+3C), there is a clear increase in quality in the overlapping region (from ~150 to 300 bp).
+Notably, there is a large degree of variation in quality, especially toward the end of the reads.
+After quality filtering, there is much less variation (Fig.
+3D).
+Despite this improvement in quality score distributions, spurious OTU calls still occur, as shown by the massive number of raw OTUs originally called (139,253).
+After eliminating OTUs that are called by <0.1% of reads (our “remove low-confidence OTUs” step), which is the maximal expected bleed-through between MiSeq runs according to Illumina (21), we retained 4,504 OTUs.
+This 97% removal of (most probably) spurious OTUs demonstrates, along with the mock community results in Fig.
+2, that reasonable OTU estimates can be achieved by using proper quality control.
+journals.asm.org/doi/full/10.1128/mSystems.00127-16
+* UPARSE removes singletons?
 
@@
 
-* I do not believe that authors have an excuse to generate data with low quality data. As outlined in the Discussion, this is part of the problem with the field currently.
+I think the reviewer may be putting too much emphasis on the importance of the individual datasets.
+These datasets were selected to help me simulate communities with different properties, number sequences per sample, and number of samples per study.
+At the same time, I would resist any attempt to normalize the generation of data that we know has a high error rate.
+Why shouldn't researchers be expected to generate the best data they can?
 
@@
 
-* Even Holmes's study says to rarefy alpha and beta diversity metrics
-* Add citations to rarefaction yes papers
-* Analysis of rarefaction is beyond the scope of this paper
+As I mentioned in the preamble to my response to the reviewer's comments, since receiving their comments, I have written two manuscripts using different simulation approaches to demonstrate that rarefaction is critical for drawing proper inferences from microbiome studies.
+As outlined in the response to McMurdie and Holmes's study, their approach was effectively designed to make look rarefaction look bad - this was principally achieved through (1) removing 15% of the samples and then penalizing the method for the loss of those samples and (2) basing the analysis on a cluster analysis of samples that used a clustering algorithm that performed the worst of the available options.
+In the second manuscript, I developed a different simulation strategy that is similar to the one used in this study and employed additional normalization approaches.
+Both studies show that rarefaction is critical for controlling for uneven sampling effort when pursuing alpha and beta diversity analyses.
+I contend that the results I have shown regarding coefficients of variation and type 1 and 2 errors support the use of rarefaction.
+Finally, the reviewer encourages the use of relative abundance, which I show in my rarefaction papers can lead to problematic interpretations.
 
@@
 * Bigger point in Figure 3 is that the coefficient of variation between samples goes up with removal of rare sequences reducing power to detect differences in communities
-* Will also caculate Shannon evenness (keep in mind that this is dependent on richness)
+* Will also calculate Shannon evenness (keep in mind that this is dependent on richness)
 
@@
 * High quality data reduces risk of artifacts and rarefaction allows us to control for inclusion of artifacts
+* Not sure if reviweer is asking about differential abundance, which isn't my concern here
 
@@
 
-* Yes number of samples. Clarify text
+The reviewer is correct that I was referring to the number of samples. I have clarified this sentence.
 
@@
 
-* Clarify text
+The sentence has been edited as suggested by the reviewer.
 
@@
 
-* That is shown in Figure 3 when rare sequences were not removed
+This was shown in Figure 3 when rare sequences were not removed (i.e.
+when the smallest number of sequences per ASV was equal to 1).
+In this case the coefficients of variation were nearly zero.
+Furthermore, in Figure 5A and 5B the Type 1 errors were near the expected 5% (i.e., P=0.05) for the case when the smallest number of sequences was equal to 1.
 
@@
 
-* Clarify that the samples were sampled into one of two equally sized treatment groups
+The simulations did indeed use all of the samples. I have clarified the text to indicate that all of the samples were randomly assigned into one of two treatment groups
 
@@
 
-* Clarify that it is P<0.05
+I have clarified this sentence to indicate that a significant test was one where the P-value threshold was 0.05.
 
-**Line 200: It's a little challenging to determine what happened here: "First, I sampled sequences from the 200 null distribution for each dataset and randomly assigned each sample to one of two treatment groups and determined the richness and diversity of ASVs and OTUs".
+**Line 200: It's a little challenging to determine what happened here: "First, I sampled sequences from the null distribution for each dataset and randomly assigned each sample to one of two treatment groups and determined the richness and diversity of ASVs and OTUs".
 For example, what is meant by "sequences" here?
@@
 
-* Clarify text
+I have restructured the text here to make it more clear how the simulations were created.
 
@@
 
-* Indicate a range? Seems clear in Fig 5A and S6A
+I have inserted the median, minimum, and maximum p-values that were observed across the range of conditions that are listed in these sentences.
 
@@
 
-* Yes it's flat, but the rate increased when any rare sequences were removed. That's the problem. The value is the lowest when all sequences were retained.
+The reviewer is correct that the curve is flat; however, the curve flattens at a very high level when any rare sequences were removed.
+That's the problem.
+The value is the lowest when all sequences were retained.
+The reviewer has correctly identified the problem: if any sequences are removed, the false detection rate increases dramatically.
 
@@
 
-* I stand by this comment
+This sentence has been modified to reflect that the perception of overclustering is a product of prioritizing the number of read depth and length of sequences.
+Again, we don't know what is too many OTUs if we don't know the true number of OTUs in natural samples
 
@@
 
-* Sentence removed
+This sentence has been removed in the revised manuscript
 
@@
 
-* Clarify text
+As suggested by the reviewer, this text has been revised to make it more clear.
 
@@
 
-* Corrected
+This has been corrected in the revised manuscript

### 2021-10-21 submission/manuscript.Rmd
Address reviewer's comments

@@
 
-```{r knitr_settings, tidy=TRUE, eval=TRUE, echo=FALSE, cache=FALSE}
+```{r knitr_settings, tidy=TRUE, eval=TRUE, echo=FALSE, cache=FALSE, warning=FALSE}
 options(tidyverse.quiet = TRUE)
@@
 Regardless of the approach, many pipelines advocate for abundance-based screening where rare sequences are removed from each dataset prior to outputting the sequence data as ASVs [@Edgar2016; @Amir2017; @Callahan2016].
-Some algorithms recommend removing all ASVs that appear one (i.e.
-singletons) [@Callahan2016], eight [@Edgar2016], or ten [@Amir2017] or fewer times; these pipelines also vary in whether the minimum abundance threshold should be applied to individual samples [@Callahan2016; @Amir2017] or the pool of samples in a study [@Edgar2016].
+Popular algorithms employ abundance-based filtering of sequences that appear one [i.e.
+singletons, DADA2, @Callahan2016], four [UNOISE2, @Edgar2016], or ten [Deblur, @Amir2017] or fewer times; these pipelines also vary in whether the minimum abundance threshold should be applied to individual samples [@Callahan2016; @Amir2017] or the pool of counts from across the samples [@Edgar2016].
+Beyond these abundance-based approaches, researchers will also often employ a prevalence-based filter.
+For example, Nearing et al.
+[@Nearing2022] used Deblur to produce ASVs [@Amir2017], which removes any sequences that appear 10 or fewer times in each sample and required that an ASV appear in 10% of samples seemingly regardless of their abundance in those samples.
 A notable exception, the mothur-based pipeline discourages the practice of removing rare sequences [@Kozich2013].
-After assigning sequences to ASVs, ASVs are often analyzed as a taxonomic unit or clustered to generate operational taxonomic units (OTUs) or phylotypes.
+Once sequences are assigned to ASVs, ASVs are often analyzed as their own taxonomic unit or clustered to generate operational taxonomic units (OTUs) or phylotypes.
 
-The abundance-based screening approach assumes that rare ASVs are more likely to be artifacts than more abundant ASVs.
+The abundance and prevalence-based screening approach assumes that rare ASVs are more likely to be artifacts than more abundant ASVs.
 Sequencing of mock communities confirms that artifacts tend to be rare [@Kozich2013; @Bokulich2013].
 Proponents of abundance-based screening point to their ability to obtain the correct number of ASVs, OTUs, or phylotypes with data generated from sequencing mock communities when rare ASVs are removed [@Bokulich2013].
-However, this approach effectively overfits the curation pipeline to data generated from a phylogenetically simple community with an atypical community distribution that is often sequenced to a coverage that is not achieved with biological samples.
-It is necessary to think more deeply about the practice of abundance-based screening.
+However, this approach risks overfitting the curation pipeline to data generated from a phylogenetically simple community with an atypical community distribution that is often sequenced to a coverage that is not achieved with biological samples.
+It is necessary to think more deeply about the practice of abundance-based screening and question whether all rare or patchily-distributed sequences are artifacts.
+Rarity and patchiness are important ecological concept that appear to make microbial ecologists uncomfortable.
 
-The minimum abundance thresholds that have been proscribed were developed and applied without regard for the number of sequences generated from each sample.
+The minimum abundance thresholds that have been proscribed were largely developed and applied without regard for the number of sequences generated from each sample.
 These recommendations appear to assume that the number of reads per sample is consistent within and between studies.
@@
 But, according to the pipeline recommendations, they are treated as being equally trustworthy.
-Rather than removing rare ASVs, the approach taken by the mothur pipeline applies the classical ecological approach of rarefaction.
-Each sample is rarefied to the same sequencing depth so that the number of artifacts that appears in each sample is controlled.
+One exception to the use of threshold counts was proposed by Bokulich et al.
+[@Bokulich2013] who suggested a 0.005% relative abundance threshold; however, they cautioned that this threshold was dependent on the quality of the data and should be reassessed with mock community data.
+Rather than removing rare or patchy ASVs, the approach taken by the mothur pipeline applies the classical ecological approach of rarefaction.
+Each sample is rarefied to the same sequencing depth so that the number of artifacts that appears in each sample is controlled and problems inherent in using relative abundance, scaling, and other normalization procedures are avoided [@Schloss2023rare; @Schloss2023wnwn].
 
-Experience sequencing biological samples also demonstrates that there are bona fide ASVs that may have an abundance below the proscribed threshold.
+Experience sequencing biological samples also demonstrates that there are *bona fide* ASVs that may have an abundance below the proscribed threshold.
 For example, the abundance of an ASV may be below the threshold in some samples or time points and above the threshold in others.
-However, rarity, both in terms of prevalence and incidence, is an important ecological concept [@Reid2011].
-Removing rare ASVs likely hinders one's to ability to make inferences about the dynamics and nature of the populations that rare ASVs represent.
+For example, my research group frequently encounters cases where it is possible to culture *Clostridiodes difficile* from human and mouse fecal samples, but not find their 16S rRNA gene sequences or to find their sequences early in colonization but not once the organism has been cleared [e.g.
+@Lesniak2022; @Schubert2014]; naturally, we only find their sequences when the plating density is the highest.
+This seeming patchiness is a product of their ecology and not an artifact.
+Rarity, both in terms of abundance and prevalence, is an important ecological concept [@Reid2011].
+Removing rare ASVs likely hinders one's to ability to make biological inferences about the dynamics and nature of the populations that rare ASVs represent.
 Furthermore, removing ASVs whose abundances are below the proscribed threshold also potentially biases the community structure of the samples.
 
-In the current study, I used published sequence data from `r nrow(table_1)` studies to investigate the nature of rare ASVs (i.e.
-those that appear 10 or fewer times) and the effect that removing them has on downstream analysis of microbial communities.
+In the current study, I used sequence data from `r nrow(table_1)` published studies to investigate the nature of rare ASVs (i.e.
+those that appear 10 or fewer times) and the effect of removing them on downstream analysis of microbial communities.
 The analysis was also performed using traditional OTUs, where ASVs subjected to abundance-based screening were clustered such that the ASVs within an OTU were no more than 3% different from each other.
@@
 The median number of sequences for each dataset ranged between `r oxford_comma(n_med_sequence_range)`.
-Strikingly, aside from the relatively small `r oxford_comma(low_fold_samples)` datasets, the difference between the sample with the fewest sequences and the sample with the most sequences for each dataset varied by between `r oxford_comma(high_fold_range, digits=1)`-fold.
+Strikingly, aside from the `r oxford_comma(low_fold_samples)` datasets, which had a relatively small number of samples, the difference between the sample with the fewest sequences and the sample with the most sequences for each dataset varied by between `r oxford_comma(high_fold_range, digits=1)`-fold.
 
@@
 **The nature of singletons.** Removal of rare ASVs is commonly justified as a method of removing ASVs that are artifacts.
-If such ASVs are artifacts, then one would expect the number of singleton ASVs to accumulate with sequencing depth.
-Contrary to this expectation, the median percentage of sequences that were discarded when singleton ASVs were removed from each dataset varied between `r oxford_comma(percent_singletons_range_vals)`% (`r oxford_comma(percent_singletons_range_names)`).
-In addition, with the exception of the samples from the `r oxford_comma(cor_n_nosig)` datasets (Spearman correlation, P>0.05), the fraction of singleton ASVs in samples was negatively correlated with the number of sequences in each sample with a range between `r oxford_comma(cor_n_sig_values)` (`r oxford_comma(cor_n_sig_names)`) (Figures 1A and S2).
+The median percentage of sequences that were discarded when singleton ASVs were removed from each dataset varied between `r oxford_comma(percent_singletons_range_vals)`% (`r oxford_comma(percent_singletons_range_names)`).
+If these singleton ASVs were artifacts, then one would expect the number of singleton ASVs to accumulate with sequencing depth.
+Contrary to this expectation, with the exception of the samples from the `r oxford_comma(cor_n_nosig)` datasets whose Spearman correlation was not significantly different from 0.00 (P>0.05), the fraction of singleton ASVs in samples was negatively correlated with the number of sequences in each sample with a range between `r oxford_comma(cor_n_sig_values)` (`r oxford_comma(cor_n_sig_names)`) (Figures 1A and S2).
 This showed that with additional sequencing, the probability of seeing singleton ASVs in multiple samples was greater than the probability of generating an artifact.
@@
 
-**Removing treatment group effects from community data.** Because treatment effects often affect a sample's diversity and inter-sample variation, I generated null distributions for each study by randomizing, without replacement, the number of times each ASV was observed in each sample such that the total number of sequences in each sample and the total number of times each ASV was observed across all samples in the study was the same as was originally observed.
-This effectively made every community in a study a statistical sample of the study-wide composite community distribution.
-For example, after this procedure, each of the 490 samples from the human dataset would be expected to have the same richness and diversity of ASVs and one would not expect to find treatment-based effects between the samples.
-Because of the risk of bias if only one representation of the null distribution was generated, I generated 100 randomized datasets for each study.
-The trends between removing rare ASVs and the richness, diversity, and information loss that were identified using the observed community distribution data were also identified with the data from the null distribution; however, the losses were larger when using the null distribution data (Figure S5).
-The null distribution data were used in the remainder of the study to minimize the risk of bias.
+**Removing treatment group effects from community data.** One challenge with using the observed ASV and OTU frequency data to investigate the effects of removing rare ASVs is that we do not know the true difference between treatment groups to know whether the false detection rate or statistical power have been altered by their removal.
+Therefore, I used the observed frequency data to simulate samples where I could specify the true difference between the treatment groups.
+To minimize the effects of the random number generator seed, I replicated the analysis 100 times.
+The first simulation I created was a null model where all of the samples in a dataset were drawn from the same sampling distribution, without replacement.
+When I repeated the analysis investigating the effects of removing rare ASVs on richness, diversity, and information loss the losses were larger than were detected with the observed data (Figure S5).
+This result was likely due to the reduced inter-sample variation and patchiness in the datasets.
 
@@
 **The impact of removing rare ASVs on the information represented between samples.** Considering the loss of richness, diversity, and information when a community has its rarest ASVs removed, it seemed likely that the relationship between communities would also be altered.
-To assess the impact of removing rare ASVs on measures of alpha diversity between samples I calculated the coefficients of variation (COVs, i.e.
+To assess the impact of removing rare ASVs on measures of alpha diversity between samples I used the null community distribution and calculated the coefficients of variation (COVs, i.e.
 the standard deviation divided by the mean) for richness and diversity for each study at multiple abundance thresholds.
+I expected COVs to increase with the removal of rare ASVs because their removal would increase the dissimilarity between the samples.
 The COVs for the richness of ASVs across the studies after removing singletons were between `r oxford_comma(pull(pc_cov_sobs, cv_sobs), 1)`-times larger than they were without removing singleton ASVs (`r oxford_comma(pull(pc_cov_sobs, dataset))`; Figure 3A).
@@
 The COVs between Bray-Curtis distances within a study when singletons were removed was between `r oxford_comma(pull(pc_cov_bc, cv_diff), 1)`-times larger than when they were not removed (`r oxford_comma(pull(pc_cov_bc, dataset))`; Figure 3C).
-When ASVs were clustered into OTUs the difference in COVs was less than it was for the ASVs (Figure S6).
-These results indicate that removing rare ASVs increases the dissimilarity between samples, which could have a significant impact on the statistical power to detect differences between treatment groups.
+When ASVs were clustered into OTUs the difference in COVs was less than it was for the ASVs, but qualitatively similar (Figure S6).
+These results supported my prediction that removing rare ASVs would increase the dissimilarity between samples.
+This increased variation in the data could have a significant impact on the statistical power to detect differences between treatment groups.
 
@@
 
-b_base_pc_sobs <- b_alpha %>% filter(metric == "sobs", method == "pc") %>% arrange(`1`) %>% slice(1, nrow(.))
-b_base_pc_shannon <- b_alpha %>% filter(metric == "shannon", method == "pc") %>% arrange(`1`) %>% slice(1, nrow(.))
+b_base_pc_sobs <- b_alpha %>% filter(metric == "sobs", method == "pc") %>% arrange(`1`) %>% slice(1, nrow(.)) %>%
+	rename('one' = `1`)
+
+b_base_pc_shannon <- b_alpha %>% filter(metric == "shannon", method == "pc") %>% arrange(`1`) %>% slice(1, nrow(.)) %>%
+	rename('one' = `1`)
+
 b_pc_sobs <- b_alpha %>% filter(metric == "sobs", method == "pc") %>% slice(1, n())
@@
 
-b_base_beta <- b_beta %>% filter(clustering == "pc") %>% arrange(`1`)
+b_base_beta <- b_beta %>% filter(clustering == "pc") %>% arrange(`1`) %>%
+	rename('one' = `1`)
 
@@
 
-**The impact of removing rare ASVs on the ability to detect statistically significant differences between treatment groups.** To test the effect of increased inter-sample variation, I randomly assigned samples to one of two treatment groups.
-In the first treatment group, communities were randomly sampled from the null distribution as described above.
-For the second treatment group, I randomly selected 10% of the ASVs in the pooled study distribution to increase their abundance by 5%.
-I randomly generated 100 simulated sets of treatment groups and samples.
+**The impact of removing rare ASVs on the ability to detect statistically significant differences between treatment groups.** To test the effect of increased inter-sample variation, I simulated two treatment groups using ASV data from each study.
+Half of the samples were randomly assigned to the first treatment group, which followed a null model.
+The other samples were assigned to a second treatment group where 10% of the ASVs in the null model were randomly selected to have their abundance increased  by 5%.
+To minimize the effects of the choice of the random number generator seed, each dataset was replicated 100 times.
 I then tested the ability to detect a difference between the two treatment groups using alpha and beta diversity metrics.
-The fraction of significant tests was a measurement of the statistical power to detect the difference between the treatment groups.
+The fraction of significant tests (i.e., P<0.05) was a measurement of the statistical power to detect the difference between the treatment groups.
 When considering the differences in richness and diversity, the marine dataset yielded no simulated sets that were statistically significant, which was likely due to the small number of samples in the study (N=7).
-Among the remaining datasets, the power to detect a difference in the richness of ASVs ranged between `r oxford_comma(pull(b_base_pc_sobs,'1'), 2)` (`r oxford_comma(pull(b_base_pc_sobs,dataset))`) and between `r oxford_comma(pull(b_base_pc_shannon, '1'), 2)` (`r oxford_comma(pull(b_base_pc_shannon, dataset))`) to detect a difference in diversity when using a Wilcox test (Figure 4A).
+Among the remaining datasets, the power to detect a difference in the richness of ASVs ranged between `r oxford_comma(pull(b_base_pc_sobs, one), 2)` (`r oxford_comma(pull(b_base_pc_sobs,dataset))`) and between `r oxford_comma(pull(b_base_pc_shannon, one), 2)` (`r oxford_comma(pull(b_base_pc_shannon, dataset))`) to detect a difference in diversity when using a Wilcoxon test (Figure 4A).
 When singleton ASVs were removed, the power to detect a difference in the richness of ASVs dropped by between `r oxford_comma(pull(b_pc_sobs, percentage_drop), 1)`% (`r oxford_comma(pull(b_pc_sobs, dataset))`) and by between `r oxford_comma(pull(b_pc_shannon, percentage_drop), 1)`% for their diversity(`r oxford_comma(pull(b_pc_shannon, dataset))`; Figure 4B).
@@
 I used the Bray-Curtis dissimilarity index to compare the simulated communities within each dataset and calculated the power to detect differences between the two simulated treatment groups using the analysis of molecular variance (Figure 4C and S7C).
-Without removing rare sequences, the power to detect a difference between the two simulated treatment groups varied between `r oxford_comma(pull(b_base_beta, '1') %>% range(), 2)` (`r slice(b_base_beta, 1) %>% pull(dataset)` and `r oxford_comma(slice_max(b_base_beta, '1') %>% pull(dataset))`).
+Without removing rare sequences, the power to detect a difference between the two simulated treatment groups varied between `r oxford_comma(pull(b_base_beta, one) %>% range(), 2)` (`r slice(b_base_beta, 1) %>% pull(dataset)` and `r oxford_comma(slice_max(b_base_beta, order_by = one) %>% pull(dataset))`).
 Aside from the `r oxford_comma(unfazed_singletons)` datasets, the power to detect differences dropped by between `r oxford_comma(pull(fazed_singletons,p_drop_singletons), 1)`% (`r oxford_comma(pull(fazed_singletons, dataset))`) when singletons were removed.
@@
 	group_by(prune, method, metric) %>%
-	summarize(p=mean(frac_sig), p_sd = sd(frac_sig)) %>%
-	ungroup()
+	summarize(p=mean(frac_sig), p_sd = sd(frac_sig), .groups="drop") %>%
+	summarize(median = median(p),
+						min = min(p),
+						max = max(p)
+						)
 
@@
 	group_by(prune, clustering) %>%
-	summarize(p=mean(frac_sig), p_sd = sd(frac_sig)) %>%
-  ungroup()
+	summarize(p=mean(frac_sig), p_sd = sd(frac_sig), .groups="drop") %>%
+	summarize(median = median(p),
+						min = min(p),
+						max = max(p)
+						)
 
@@
 **The impact of removing rare ASVs on the probability of falsely detecting a difference between treatment groups.** I next asked whether removing rare ASVs could lead to falsely claiming that a treatment effect had a significant effect on community diversity and structure.
-First, I sampled sequences from the null distribution for each dataset and randomly assigned each sample to one of two treatment groups and determined the richness and diversity of ASVs and OTUs.
-Testing at an experiment-wise error rate of 0.05, I expected 5% of the iterations for each dataset to yield a significant test result.
+First, for each dataset I randomly assigned samples to one of two treatment groups.
+I then determined the richness and diversity of ASVs and OTUs in each of the samples.
+Testing at an experiment-wise error rate of 0.05, I expected 5% of the iterations (i.e., ~5 iterations) for each dataset to yield a significant test result.
 Indeed, there was no evidence that removing rare ASVs resulted in an inflated experiment-wise error rate.
-The average fraction of significant tests did not meaningfully vary from 0.05 across the minimum abundance threshold, dataset, metric of describing sample alpha-diversity, or whether the abundance of ASVs or OTUs were used (Figure 5A and S8A).
-Similarly, the average fraction of significant tests did not meaningfully vary from 0.05 when using analysis of molecular variance to compare communities using Bray-Curtis distances (Figure 5A and S8A).
+The average fraction of significant tests did not meaningfully vary from 0.05 across the minimum abundance threshold, dataset, metric of describing sample alpha-diversity, or whether the abundance of ASVs or OTUs were used (median: `r type_one_null_alpha_summary$median`, minimum: `r type_one_null_alpha_summary$min`, maximum: `r type_one_null_alpha_summary$max`; Figure 5A and S8A).
+Similarly, the average fraction of significant tests did not meaningfully vary from 0.05 when using analysis of molecular variance to compare communities using Bray-Curtis distances (median: `r type_one_null_beta_summary$median`, minimum: `r type_one_null_beta_summary$min`, maximum: `r type_one_null_beta_summary$max`; Figure 5A and S8A).
 Second, I again sampled sequences from the null distribution, but assigned samples to one of two treatment groups based on the number of sequences in each sample.
@@
 
-The practice of removing rare sequences from samples seems to be a response to researchers prioritizing the number of reads and length of sequences over their quality [@Bokulich2013].
+The practice of removing rare sequences from samples seems to be a response to researchers prioritizing the number of reads and length of sequences over their quality and needing a mechanism to reduce the inflated number of spurious OTUs [@Bokulich2013].
 Previous work has shown that assembly of fully overlapping sequence reads results in the lowest sequencing error rates [@Kozich2013].
@@
 In contrast, there has been a movement to sequence longer regions with similar chemistry resulting in less overlap between the sequencing reads [@Fadrosh2014; @Holm2019].
-Alternatively, others have prioritized increasing the number of sequences per sample by sequencing the V4 region, but with paired or single reads that are 150 nt long [@Thompson2017; @McDonald2018].
-Both practices result in a significantly higher error rate for the resulting assmebly.
-Instead, researchers should prioritize the quality over the quantity and length of their data.
+Alternatively, others have prioritized increasing the number of sequences per sample by sequencing the V4 region, but with single reads or with reads that are only 150 nt long [@Thompson2017; @McDonald2018].
+These practices result in a significantly higher error rate for the resulting assembly since each read has an average per base error rate near 1%.
+With the approach used here, the error rate of the polished sequences for the V4 region is on the order of 0.01%, but it is at least 10-fold higher when sequencing regions that not fully overlap [@Kozich2013].
+Another source of artifacts are chimeras, which are formed during PCR [@Sze2019; @Haas2011].
+Although lab-based strategies are available to minimize their formation, most pipelines use bioinformatic approaches such as UCHIME to remove these artifacts [@Edgar2011].
+Researchers should prioritize the quality over the quantity and length of their data.
 For these reasons, this analysis did not use lower quality data generated by alternative methods.
-
-The impacts of removing rare sequences on the representation of communities are caused by the uneven impact of applying a single abundance threshold across all samples.
-The number of sequences per sample in the datasets highlighted in this analysis varied by as much as 100-fold (Table 1; Figure S1).
-In practice, I suspect this range is actually larger since researchers may have opted against depositing samples with fewer reads into the SRA.
-If the range in read coverage for a study is 100-fold, a singleton in a small sample would have had a comparable relative abundance as a sequence that appeared 100 times in a more densely sequenced sample.
-However, it would have been removed from the smaller sample and not the larger sample.
-Thus, applying a single abundance threshold disproportionately impacted the samples with fewer sequences.
-This seems to contradict the purpose of removing rare sequences since a singleton in the smaller sample is more reliable than the singleton in the larger sample.
-A superior approach to removing rare sequences is to use rarefaction to conrol for uneven sampling.
 
@@
 Researchers are encouraged to use rarefaction to control for uneven sampling and the presence of spurious sequences.
-Previous work sequencing mock communities has shown that the number of spurious sequences increases with sequencing depth [@Kozich2013; @Schloss2011].
 By rarefying data to a common number of sequences per sample, the number of spurious sequences can be controlled.
-As shown in the data I presented, which was rarefied to a common number of reads per sample within a dataset, when rare sequences were not removed the power to detect differences was the highest and the false discovery rate was the expected 5% (Figures 5 and S8).
+As shown in the data I presented, which was rarefied to a common number of reads per sample within a dataset, when rare sequences were not removed the power to detect differences was the highest (Figures 4 and S7) and the false discovery rate was the expected 5% (Figures 5 and S8).
+Although using a relative abundance-based filter has been suggested [@Bokulich2013], relative abundance normalization has been shown inferior to rarefaction for alpha and beta diversity analyses [@Schloss2023wnwn; @Schloss2023rare].
 
@@
 Therefore, researchers must approach rare sequences with more skepticism than more abundant sequences.
-Reseachers should seek out other methods to confirm inferences that they make about rare sequences.
+Researchers should seek out other methods to confirm inferences that they make about rare sequences.
 This is a standard that should be applied regardless of their abundance [@Schloss2018].
@@
 
-**Data curation and analysis.** To insure the highest possible data quality, datasets were limited to those where the 500 cycle version 2 MiSeq chemistry was used to sequence the amplicons.
+**Data curation and analysis.** To ensure the highest possible data quality, datasets were limited to those where the 500 cycle version 2 MiSeq chemistry was used to sequence the amplicons.
 The paired 250 nt reads resulted in near complete 2-fold sequencing coverage of every nucleotide in the ca.
@@
 For comparisons of Bray-Curtis distances, the amova function within mothur was used, which implements the analysis of molecular variance algorithm [@Anderson2001].
+
+Richness
+Shannon
+Kullback–Leibler divergence
+Bray-Curtis
+rarefaction
+
+Generated community wiht singletons and that was used for all filtering
+
+**Null model.**
+I generated null distributions for each study by randomizing, without replacement, the number of times each ASV was observed in each sample such that the total number of sequences in each sample and the total number of times each ASV was observed across all samples in the study was the same as was originally observed.
+This effectively made every community in a study a statistical sample of the study-wide composite community distribution.
+For example, after this procedure, each of the 490 samples from the human dataset would be expected to have the same richness and diversity of ASVs and one would not expect to find treatment-based effects between the samples.
+Because of the risk of bias if only one representation of the null distribution was generated, I generated 100 randomized datasets for each study.
+
+**Other model.**
+randomly assigned samples to one of two treatment groups.
+In the first treatment group, samples were regenerated by randomly sampling ASVs from the null distribution as described above such that each samples had the same number of sequences as it did in the original study.
+For the second treatment group, I randomly selected 10% of the ASVs in the pooled study distribution to increase their abundance by 5%.
+I randomly generated 100 simulated sets of treatment groups and samples.
 
@@
 
-\includegraphics{figure_1.png}
-
 **Figure 1.
@@
 
-\newpage
-
-\includegraphics{figure_2.png}
-
 **Figure 2.
 Removing rare sequences from samples alters their representation of alpha-diversity using amplicon sequence variants (ASVs).** The average difference in the richness (A), Shannon diversity (B), and Kullback-Leiber divergence (C) for each sample within a dataset was calculated between the original community structures relative to applying different minimum abundance thresholds.
-
-\newpage
-
-\includegraphics{figure_3.png}
 
@@
 
-\newpage
-
-\includegraphics{figure_4.png}
-
 **Figure 4.
-Removing rare sequences from samples reduces the statistical power to detect differences between empirically generated treatment groups when using amplicon sequence variants (ASVs).** The fraction of significant tests comparing the richness (A) and Shannon diversity (B) using a Wilcox test and Bray-Curtis distances (C) using analysis of molecular variance for each dataset was calculated using empirically generated treatment groups containing equal numbers of samples for each dataset with varying minimum abundance thresholds.
+Removing rare sequences from samples reduces the statistical power to detect differences between empirically generated treatment groups when using amplicon sequence variants (ASVs).** The fraction of significant tests comparing the richness (A) and Shannon diversity (B) using a Wilcoxon test and Bray-Curtis distances (C) using analysis of molecular variance for each dataset was calculated using empirically generated treatment groups containing equal numbers of samples for each dataset with varying minimum abundance thresholds.
 For each dataset and minimum abundance threshold, 100 randomizations were peformed.
 
-\newpage
-
-\includegraphics{figure_5.png}
-
 **Figure 5.
-Removing rare sequences does not impact the false detection rate unless the number of sequences per sample is confounded with the treatment groups when using amplicon sequence variants (ASVs).** The fraction of significant tests comparing the richness and Shannon diversity using a Wilcox test and Bray-Curtis distances using analysis of molecular variance for each dataset was calculated.
+Removing rare sequences does not impact the false detection rate unless the number of sequences per sample is confounded with the treatment groups when using amplicon sequence variants (ASVs).** The fraction of significant tests comparing the richness and Shannon diversity using a Wilcoxon test and Bray-Curtis distances using analysis of molecular variance for each dataset was calculated.
 Empirically generated treatment groups were generated containing equal numbers of samples where the samples represented a null distribution.
@@
 
-\includegraphics{figure_s1.png}
-
 **Figure S1.
 Distribution of the number of sequences per sample in the 12 datasets included in this study.** A different minimum number of sequences per sample threshold was applied to each dataset based on identifying natural breaks in the distribution of the number of sequences per sample.
-
-\newpage
-
-\includegraphics{figure_s2.png}
 
@@
 
-\newpage
-
-\includegraphics{figure_s3.png}
-
 **Figure S3.
@@
 
-\newpage
-
-\includegraphics{figure_s4.png}
-
 **Figure S4.
 Removing rare sequences from samples alters their representation of alpha-diversity using operational taxonomic units (OTUs).** The average difference in the richness (A), Shannon diversity (B), and Kullback-Leiber divergence (C) for each sample within a dataset was calculated between the original community structures relative to applying different minimum abundance thresholds.
-
-\newpage
-
-\includegraphics{figure_s5.png}
 
@@
 
-\newpage
-
-\includegraphics{figure_s6.png}
-
 **Figure S6.
@@
 
-\newpage
-
-\includegraphics{figure_s7.png}
-
 **Figure S7.
-Removing rare sequences from samples reduces the statistical power to detect differences between empirically generated treatment groups when using operational taxonomic units (OTUs).** The fraction of significant tests comparing the richness (A) and Shannon diversity (B) using a Wilcox test and Bray-Curtis distances (C) using analysis of molecular variance for each dataset was calculated using empirically generated treatment groups containing equal numbers of samples for each dataset with varying minimum abundance thresholds.
+Removing rare sequences from samples reduces the statistical power to detect differences between empirically generated treatment groups when using operational taxonomic units (OTUs).** The fraction of significant tests comparing the richness (A) and Shannon diversity (B) using a Wilcoxon test and Bray-Curtis distances (C) using analysis of molecular variance for each dataset was calculated using empirically generated treatment groups containing equal numbers of samples for each dataset with varying minimum abundance thresholds.
 For each dataset and minimum abundance threshold, 100 randomizations were peformed.
 
-\newpage
-
-\includegraphics{figure_s8.png}
-
 **Figure S8.
-Removing rare sequences does not impact the false detection rate unless the number of sequences per sample is confounded with the treatment groups when using operational taxonomic units (OTUs).** The fraction of significant tests comparing the richness and Shannon diversity using a Wilcox test and Bray-Curtis distances using analysis of molecular variance for each dataset was calculated.
+Removing rare sequences does not impact the false detection rate unless the number of sequences per sample is confounded with the treatment groups when using operational taxonomic units (OTUs).** The fraction of significant tests comparing the richness and Shannon diversity using a Wilcoxon test and Bray-Curtis distances using analysis of molecular variance for each dataset was calculated.
 Empirically generated treatment groups were generated containing equal numbers of samples where the samples represented a null distribution.

### 2021-10-21 submission/manuscript.tex
Address reviewer's comments

@@
 ]{article}
+\usepackage{amsmath,amssymb}
 \usepackage{lmodern}
-\usepackage{amsmath}
-\usepackage{ifxetex,ifluatex}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+\usepackage{iftex}
+\ifPDFTeX
   \usepackage[T1]{fontenc}
@@
   \usepackage{textcomp} % provide euro and other symbols
-  \usepackage{amssymb}
 \else % if luatex or xetex
@@
 \usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
 \usepackage[margin=1.0in]{geometry}
-\usepackage{longtable,booktabs}
+\usepackage{longtable,booktabs,array}
 \usepackage{calc} % for calculating minipage widths
@@
 \setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
+  \fi
+  % set entry spacing
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 \usepackage{booktabs}
@@
 \usepackage{makecell}
-\ifluatex
+\usepackage{xcolor}
+\ifLuaTeX
   \usepackage{selnolig}  % disable illegal ligatures
 \fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[3] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same} % disable monospaced font for URLs
+\hypersetup{
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
 
@@
 recommend removing all ASVs that appear one (i.e.~singletons) (7), eight
-(3), or ten (6) or fewer times; these pipelines also vary in whether the
-minimum abundance threshold should be applied to individual samples (6,
-7) or the pool of samples in a study (3). A notable exception, the
-mothur-based pipeline discourages the practice of removing rare
-sequences (2). After assigning sequences to ASVs, ASVs are often
-analyzed as a taxonomic unit or clustered to generate operational
+(3), four (3), or ten (6) or fewer times; these pipelines also vary in
+whether the minimum abundance threshold should be applied to individual
+samples (6, 7) or the pool of samples in a study (3). A notable
+exception, the mothur-based pipeline discourages the practice of
+removing rare sequences (2). After assigning sequences to ASVs, ASVs are
+often analyzed as a taxonomic unit or clustered to generate operational
 taxonomic units (OTUs) or phylotypes.
@@
 the Illumina MiSeq platform to sequence the V4 region of the 16S rRNA
-gene from a variety of environments (Table 1; Figure S1). After removing
-poor quality and chimeric ASVs and samples that had uncharacteristically
-low number of sequences for the dataset, these datasets included between
-7 and 490 samples. The median number of sequences for each dataset
-ranged between 6,477 and 193,464. Strikingly, aside from the relatively
-small marine and soil datasets, the difference between the sample with
-the fewest sequences and the sample with the most sequences for each
-dataset varied by between 7.4 and 96.6-fold.
+gene from a variety of environments (Table 1; Figure S1).
+
+After removing poor quality and chimeric ASVs and samples that had
+uncharacteristically low number of sequences for the dataset, these
+datasets included between 7 and 490 samples. The median number of
+sequences for each dataset ranged between 6,477 and 193,464. Strikingly,
+aside from the marine and soil datasets, which had a relatively small
+number of samples, the difference between the sample with the fewest
+sequences and the sample with the most sequences for each dataset varied
+by between 7.4 and 96.6-fold.
 
@@
 (bioethanol and seagrass). In addition, with the exception of the
-samples from the marine and sediment datasets (Spearman correlation,
-P\textgreater0.05), the fraction of singleton ASVs in samples was
-negatively correlated with the number of sequences in each sample with a
-range between -0.26 and -0.87 (rice and bioethanol) (Figures 1A and S2).
-This showed that with additional sequencing, the probability of seeing
-singleton ASVs in multiple samples was greater than the probability of
-generating an artifact. This suggests that the singleton ASVs are not as
-likely to be artifacts as previously thought. Furthermore, if singleton
-ASVs were artifacts, then one would not expect to find them in other
-samples from the same dataset. In fact, singleton ASVs from samples with
-fewer sequences were often found in samples with more sequences. At
-least 50\% of the singleton ASVs found in the samples from the mice,
-rice, seagrass, and stream datasets were found in another sample from
-the same dataset (Figure 1B). Considering the likelihood of finding an
-ASV duplicated in another sample is confounded by the number of samples
-and inter-sample diversity, the high coverage of singleton ASVs in these
+samples from the marine and sediment datasets whose Spearman correlation
+was not significantly different from 0.00 (P\textgreater0.05), the
+fraction of singleton ASVs in samples was negatively correlated with the
+number of sequences in each sample with a range between -0.26 and -0.87
+(rice and bioethanol) (Figures 1A and S2). This showed that with
+additional sequencing, the probability of seeing singleton ASVs in
+multiple samples was greater than the probability of generating an
+artifact. This suggests that the singleton ASVs are not as likely to be
+artifacts as previously thought. Furthermore, if singleton ASVs were
+artifacts, then one would not expect to find them in other samples from
+the same dataset. In fact, singleton ASVs from samples with fewer
+sequences were often found in samples with more sequences. At least 50\%
+of the singleton ASVs found in the samples from the mice, rice,
+seagrass, and stream datasets were found in another sample from the same
+dataset (Figure 1B). Considering the likelihood of finding an ASV
+duplicated in another sample is confounded by the number of samples and
+inter-sample diversity, the high coverage of singleton ASVs in these
 datasets was remarkable. The correlation between the number of sequences
@@
 samples to one of two treatment groups. In the first treatment group,
-communities were randomly sampled from the null distribution as
-described above. For the second treatment group, I randomly selected
-10\% of the ASVs in the pooled study distribution to increase their
-abundance by 5\%. I randomly generated 100 simulated sets of treatment
-groups and samples. I then tested the ability to detect a difference
-between the two treatment groups using alpha and beta diversity metrics.
-The fraction of significant tests was a measurement of the statistical
-power to detect the difference between the treatment groups. When
-considering the differences in richness and diversity, the marine
+samples were regenerated by randomly sampling ASVs from the null
+distribution as described above such that each samples had the same
+number of sequences as it did in the original study. For the second
+treatment group, I randomly selected 10\% of the ASVs in the pooled
+study distribution to increase their abundance by 5\%. I randomly
+generated 100 simulated sets of treatment groups and samples. I then
+tested the ability to detect a difference between the two treatment
+groups using alpha and beta diversity metrics. The fraction of
+significant tests (i.e., P\textless0.05) was a measurement of the
+statistical power to detect the difference between the treatment groups.
+When considering the differences in richness and diversity, the marine
 dataset yielded no simulated sets that were statistically significant,
@@
 two simulated treatment groups varied between 0.41 and 1.00 (rainforest
-and rainforest). Aside from the bioethanol, human, and mice datasets,
-the power to detect differences dropped by between 6.5 and 64.0\% (soil
-and rice) when singletons were removed. However, when ASVs that occurred
-10 or fewer times were removed from each sample, the power to detect
-differences dropped by 12.0 and 97.2\% (human and peromyscus); similar
-results were observed when ASVs were clustered into OTUs. Removing rare
-ASVs reduced the ability to detect simulated treatment effects using
-metrics commonly used to compare microbial communities.
+and human, stream, and rice). Aside from the bioethanol, human, and mice
+datasets, the power to detect differences dropped by between 6.5 and
+64.0\% (soil and rice) when singletons were removed. However, when ASVs
+that occurred 10 or fewer times were removed from each sample, the power
+to detect differences dropped by 12.0 and 97.2\% (human and peromyscus);
+similar results were observed when ASVs were clustered into OTUs.
+Removing rare ASVs reduced the ability to detect simulated treatment
+effects using metrics commonly used to compare microbial communities.
 
@@
 effect had a significant effect on community diversity and structure.
-First, I sampled sequences from the null distribution for each dataset
-and randomly assigned each sample to one of two treatment groups and
-determined the richness and diversity of ASVs and OTUs. Testing at an
+First, I sampled individual sequences from the null distribution for
+each dataset so that each sample had the same number of sequences as it
+did in the original dataset. I then determined the richness and
+diversity of ASVs and OTUs in each of the samples and randomly assigned
+each sample to one of two treatment groups. Testing at an
 experiment-wise error rate of 0.05, I expected 5\% of the iterations for
@@
 dataset, metric of describing sample alpha-diversity, or whether the
-abundance of ASVs or OTUs were used (Figure 5A and S8A). Similarly, the
-average fraction of significant tests did not meaningfully vary from
-0.05 when using analysis of molecular variance to compare communities
-using Bray-Curtis distances (Figure 5A and S8A). Second, I again sampled
-sequences from the null distribution, but assigned samples to one of two
-treatment groups based on the number of sequences in each sample. The
-samples with fewer than the median number of sequences for the dataset
-were assigned to one group and those with more than the median were
-assigned to the other. This exaggerated bias has been observed in
-comparisons of the lung and oral microbiota because of the larger number
-of non-specific amplicons that can be sequenced from lung samples
-relative to those in the oral cavity leading to a significant difference
-in sequencing depth between treatment groups (9). When rare sequences
-were not removed, the fraction of significant tests did not differ from
-5\% for comparing the richness, their diversity, or Bray-Curtis
-distances (Figure 5B and S8B). However, when rare taxa of any frequency
-were removed, the probability of falsely detecing a difference as
-signifiant increased with the definition of rarity (Figure 5B and S8B).
-Not including the small marine dataset, the average fraction the average
-fraction of falsely detecting a difference across datasets when only
-singletons were removed was 92.45\%. If there is any relationship
-between the number of sequences and the treatment group, the risk of
-falsely rejecting the null hypothesis is inflated when researchers use
-the strategy of removing rare sequences. The most conservative approach
-is to not remove low abundance sequences.
+abundance of ASVs or OTUs were used (median: 0.049, minimum: 0.041,
+maximum: 0.062; Figure 5A and S8A). Similarly, the average fraction of
+significant tests did not meaningfully vary from 0.05 when using
+analysis of molecular variance to compare communities using Bray-Curtis
+distances (median: 0.05, minimum: 0.047, maximum: 0.055; Figure 5A and
+S8A). Second, I again sampled sequences from the null distribution, but
+assigned samples to one of two treatment groups based on the number of
+sequences in each sample. The samples with fewer than the median number
+of sequences for the dataset were assigned to one group and those with
+more than the median were assigned to the other. This exaggerated bias
+has been observed in comparisons of the lung and oral microbiota because
+of the larger number of non-specific amplicons that can be sequenced
+from lung samples relative to those in the oral cavity leading to a
+significant difference in sequencing depth between treatment groups (9).
+When rare sequences were not removed, the fraction of significant tests
+did not differ from 5\% for comparing the richness, their diversity, or
+Bray-Curtis distances (Figure 5B and S8B). However, when rare taxa of
+any frequency were removed, the probability of falsely detecing a
+difference as signifiant increased with the definition of rarity (Figure
+5B and S8B). Not including the small marine dataset, the average
+fraction the average fraction of falsely detecting a difference across
+datasets when only singletons were removed was 92.45\%. If there is any
+relationship between the number of sequences and the treatment group,
+the risk of falsely rejecting the null hypothesis is inflated when
+researchers use the strategy of removing rare sequences. The most
+conservative approach is to not remove low abundance sequences.
 
@@
 response to researchers prioritizing the number of reads and length of
-sequences over their quality (5). Previous work has shown that assembly
-of fully overlapping sequence reads results in the lowest sequencing
-error rates (2). The studies highlighted in this analysis sequenced the
-V4 region of the 16S rRNA gene using the Illumina MiSeq sequencing
-platform with their version 2 chemistry. The resulting data consists of
-two 250 nt reads that span a region that is about 253 nt long. In
-contrast, there has been a movement to sequence longer regions with
-similar chemistry resulting in less overlap between the sequencing reads
-(10, 11). Alternatively, others have prioritized increasing the number
-of sequences per sample by sequencing the V4 region, but with paired or
-single reads that are 150 nt long (12, 13). Both practices result in a
-significantly higher error rate for the resulting assmebly. Instead,
-researchers should prioritize the quality over the quantity and length
-of their data. For these reasons, this analysis did not use lower
+sequences over their quality and needing a mechanism to reduce the
+inflated number of spurious OTUs (5). Previous work has shown that
+assembly of fully overlapping sequence reads results in the lowest
+sequencing error rates (2). The studies highlighted in this analysis
+sequenced the V4 region of the 16S rRNA gene using the Illumina MiSeq
+sequencing platform with their version 2 chemistry. The resulting data
+consists of two 250 nt reads that span a region that is about 253 nt
+long. In contrast, there has been a movement to sequence longer regions
+with similar chemistry resulting in less overlap between the sequencing
+reads (10, 11). Alternatively, others have prioritized increasing the
+number of sequences per sample by sequencing the V4 region, but with
+paired or single reads that are 150 nt long (12, 13). Both practices
+result in a significantly higher error rate for the resulting assmebly.
+Instead, researchers should prioritize the quality over the quantity and
+length of their data. For these reasons, this analysis did not use lower
 quality data generated by alternative methods.
@@
 sample in the datasets highlighted in this analysis varied by as much as
-100-fold (Table 1; Figure S1). In practice, I suspect this range is
-actually larger since researchers may have opted against depositing
-samples with fewer reads into the SRA. If the range in read coverage for
-a study is 100-fold, a singleton in a small sample would have had a
-comparable relative abundance as a sequence that appeared 100 times in a
-more densely sequenced sample. However, it would have been removed from
-the smaller sample and not the larger sample. Thus, applying a single
+100-fold (Table 1; Figure S1). If the range in read coverage for a study
+is 100-fold, a singleton in a small sample would have had a comparable
+relative abundance as a sequence that appeared 100 times in a more
+densely sequenced sample. However, it would have been removed from the
+smaller sample and not the larger sample. Thus, applying a single
 abundance threshold disproportionately impacted the samples with fewer
@@
 or contained errors. Researchers are encouraged to use rarefaction to
-control for uneven sampling and the presence of spurious sequences.
-Previous work sequencing mock communities has shown that the number of
-spurious sequences increases with sequencing depth (2, 14). By rarefying
-data to a common number of sequences per sample, the number of spurious
-sequences can be controlled. As shown in the data I presented, which was
-rarefied to a common number of reads per sample within a dataset, when
-rare sequences were not removed the power to detect differences was the
-highest and the false discovery rate was the expected 5\% (Figures 5 and
-S8).
+control for uneven sampling and the presence of spurious sequences. By
+rarefying data to a common number of sequences per sample, the number of
+spurious sequences can be controlled. As shown in the data I presented,
+which was rarefied to a common number of reads per sample within a
+dataset, when rare sequences were not removed the power to detect
+differences was the highest and the false discovery rate was the
+expected 5\% (Figures 5 and S8).
 
@@
 inferences that they make about rare sequences. This is a standard that
-should be applied regardless of their abundance (15).
+should be applied regardless of their abundance (14).
 
@@
 challenge since microbial ecologists transitioned away from Sanger
-sequencing of samples (16--18). Although the extent of the ``rare
+sequencing of samples (15--17). Although the extent of the ``rare
 biosphere'' is still an open question, it is important to appreciate the
@@
 
-\textbf{Data curation and analysis.} To insure the highest possible data
+\textbf{Data curation and analysis.} To ensure the highest possible data
 quality, datasets were limited to those where the 500 cycle version 2
@@
 pipeline that resulted in ASVs as generated by the pre.cluster algorithm
-using a threshold of 2 nt (2, 19). ASVs were assigned to OTUs using a
+using a threshold of 2 nt (2, 18). ASVs were assigned to OTUs using a
 3\% distance threshold using mothur's cluster function with the
-OptiClust algorithm (20). To minimize the effects of uneven sampling
+OptiClust algorithm (19). To minimize the effects of uneven sampling
 effort, samples were rarefied to the number of sequences in the smallest
@@
 distances, the amova function within mothur was used, which implements
-the analysis of molecular variance algorithm (21).
+the analysis of molecular variance algorithm (20).
 
@@
 \hypertarget{refs}{}
-\begin{CSLReferences}{0}{0}
-\leavevmode\hypertarget{ref-Knight2018}{}%
-\CSLLeftMargin{1. }
+\begin{CSLReferences}{0}{1}
+\leavevmode\vadjust pre{\hypertarget{ref-Knight2018}{}}%
+\CSLLeftMargin{1. }%
 \CSLRightInline{\textbf{Knight R}, \textbf{Vrbanac A}, \textbf{Taylor
@@
 
-\leavevmode\hypertarget{ref-Kozich2013}{}%
-\CSLLeftMargin{2. }
+\leavevmode\vadjust pre{\hypertarget{ref-Kozich2013}{}}%
+\CSLLeftMargin{2. }%
 \CSLRightInline{\textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter
@@
 
-\leavevmode\hypertarget{ref-Edgar2016}{}%
-\CSLLeftMargin{3. }
+\leavevmode\vadjust pre{\hypertarget{ref-Edgar2016}{}}%
+\CSLLeftMargin{3. }%
 \CSLRightInline{\textbf{Edgar RC}. 2016. {UNOISE}2: Improved
@@
 
-\leavevmode\hypertarget{ref-Masella2012}{}%
-\CSLLeftMargin{4. }
+\leavevmode\vadjust pre{\hypertarget{ref-Masella2012}{}}%
+\CSLLeftMargin{4. }%
 \CSLRightInline{\textbf{Masella AP}, \textbf{Bartram AK},
@@
 
-\leavevmode\hypertarget{ref-Bokulich2013}{}%
-\CSLLeftMargin{5. }
+\leavevmode\vadjust pre{\hypertarget{ref-Bokulich2013}{}}%
+\CSLLeftMargin{5. }%
 \CSLRightInline{\textbf{Bokulich NA}, \textbf{Subramanian S},
@@
 
-\leavevmode\hypertarget{ref-Amir2017}{}%
-\CSLLeftMargin{6. }
+\leavevmode\vadjust pre{\hypertarget{ref-Amir2017}{}}%
+\CSLLeftMargin{6. }%
 \CSLRightInline{\textbf{Amir A}, \textbf{McDonald D},
@@
 
-\leavevmode\hypertarget{ref-Callahan2016}{}%
-\CSLLeftMargin{7. }
+\leavevmode\vadjust pre{\hypertarget{ref-Callahan2016}{}}%
+\CSLLeftMargin{7. }%
 \CSLRightInline{\textbf{Callahan BJ}, \textbf{McMurdie PJ},
@@
 
-\leavevmode\hypertarget{ref-Reid2011}{}%
-\CSLLeftMargin{8. }
+\leavevmode\vadjust pre{\hypertarget{ref-Reid2011}{}}%
+\CSLLeftMargin{8. }%
 \CSLRightInline{\textbf{Reid A}, \textbf{Buckley M}. 2011. The rare
@@
 
-\leavevmode\hypertarget{ref-Morris2013}{}%
-\CSLLeftMargin{9. }
+\leavevmode\vadjust pre{\hypertarget{ref-Morris2013}{}}%
+\CSLLeftMargin{9. }%
 \CSLRightInline{\textbf{Morris A}, \textbf{Beck JM}, \textbf{Schloss
@@
 
-\leavevmode\hypertarget{ref-Fadrosh2014}{}%
-\CSLLeftMargin{10. }
+\leavevmode\vadjust pre{\hypertarget{ref-Fadrosh2014}{}}%
+\CSLLeftMargin{10. }%
 \CSLRightInline{\textbf{Fadrosh DW}, \textbf{Ma B}, \textbf{Gajer P},
@@
 
-\leavevmode\hypertarget{ref-Holm2019}{}%
-\CSLLeftMargin{11. }
+\leavevmode\vadjust pre{\hypertarget{ref-Holm2019}{}}%
+\CSLLeftMargin{11. }%
 \CSLRightInline{\textbf{Holm JB}, \textbf{Humphrys MS}, \textbf{Robinson
@@
 
-\leavevmode\hypertarget{ref-Thompson2017}{}%
-\CSLLeftMargin{12. }
+\leavevmode\vadjust pre{\hypertarget{ref-Thompson2017}{}}%
+\CSLLeftMargin{12. }%
 \CSLRightInline{\textbf{Thompson LR}, \textbf{Jon G. Sanders and},
@@
 
-\leavevmode\hypertarget{ref-McDonald2018}{}%
-\CSLLeftMargin{13. }
+\leavevmode\vadjust pre{\hypertarget{ref-McDonald2018}{}}%
+\CSLLeftMargin{13. }%
 \CSLRightInline{\textbf{McDonald D}, \textbf{Hyde E}, \textbf{Debelius
@@
 
-\leavevmode\hypertarget{ref-Schloss2011}{}%
-\CSLLeftMargin{14. }
-\CSLRightInline{\textbf{Schloss PD}, \textbf{Gevers D}, \textbf{Westcott
-SL}. 2011. Reducing the effects of {PCR} amplification and sequencing
-artifacts on 16S {rRNA}-based studies. {PLoS} {ONE} \textbf{6}:e27310.
-doi:\href{https://doi.org/10.1371/journal.pone.0027310}{10.1371/journal.pone.0027310}.}
-
-\leavevmode\hypertarget{ref-Schloss2018}{}%
-\CSLLeftMargin{15. }
+\leavevmode\vadjust pre{\hypertarget{ref-Schloss2018}{}}%
+\CSLLeftMargin{14. }%
 \CSLRightInline{\textbf{Schloss PD}. 2018. Identifying and overcoming
@@
 
-\leavevmode\hypertarget{ref-Sogin2006}{}%
-\CSLLeftMargin{16. }
+\leavevmode\vadjust pre{\hypertarget{ref-Sogin2006}{}}%
+\CSLLeftMargin{15. }%
 \CSLRightInline{\textbf{Sogin ML}, \textbf{Morrison HG}, \textbf{Huber
@@
 
-\leavevmode\hypertarget{ref-Huse2010}{}%
-\CSLLeftMargin{17. }
+\leavevmode\vadjust pre{\hypertarget{ref-Huse2010}{}}%
+\CSLLeftMargin{16. }%
 \CSLRightInline{\textbf{Huse SM}, \textbf{Welch DM}, \textbf{Morrison
@@
 
-\leavevmode\hypertarget{ref-Reeder2009}{}%
-\CSLLeftMargin{18. }
-\CSLRightInline{\textbf{Reeder J}, \textbf{Knight R}. 2009. The {{}}rare
-biosphere{{}}: A reality check. Nature Methods \textbf{6}:636--637.
+\leavevmode\vadjust pre{\hypertarget{ref-Reeder2009}{}}%
+\CSLLeftMargin{17. }%
+\CSLRightInline{\textbf{Reeder J}, \textbf{Knight R}. 2009. The
+{\textquotesingle}rare biosphere{\textquotesingle}: A reality check.
+Nature Methods \textbf{6}:636--637.
 doi:\href{https://doi.org/10.1038/nmeth0909-636}{10.1038/nmeth0909-636}.}
 
-\leavevmode\hypertarget{ref-Schloss2009}{}%
-\CSLLeftMargin{19. }
+\leavevmode\vadjust pre{\hypertarget{ref-Schloss2009}{}}%
+\CSLLeftMargin{18. }%
 \CSLRightInline{\textbf{Schloss PD}, \textbf{Westcott SL},
@@
 
-\leavevmode\hypertarget{ref-Westcott2017}{}%
-\CSLLeftMargin{20. }
+\leavevmode\vadjust pre{\hypertarget{ref-Westcott2017}{}}%
+\CSLLeftMargin{19. }%
 \CSLRightInline{\textbf{Westcott SL}, \textbf{Schloss PD}. 2017.
@@
 
-\leavevmode\hypertarget{ref-Anderson2001}{}%
-\CSLLeftMargin{21. }
+\leavevmode\vadjust pre{\hypertarget{ref-Anderson2001}{}}%
+\CSLLeftMargin{20. }%
 \CSLRightInline{\textbf{Anderson MJ}. 2001. A new method for
@@
 
-\leavevmode\hypertarget{ref-Li2015}{}%
-\CSLLeftMargin{22. }
+\leavevmode\vadjust pre{\hypertarget{ref-Li2015}{}}%
+\CSLLeftMargin{21. }%
 \CSLRightInline{\textbf{Li Q}, \textbf{Heist EP}, \textbf{Moe LA}. 2015.
@@
 
-\leavevmode\hypertarget{ref-Baxter2016}{}%
-\CSLLeftMargin{23. }
+\leavevmode\vadjust pre{\hypertarget{ref-Baxter2016}{}}%
+\CSLLeftMargin{22. }%
 \CSLRightInline{\textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers
@@
 
-\leavevmode\hypertarget{ref-Beall2015}{}%
-\CSLLeftMargin{24. }
+\leavevmode\vadjust pre{\hypertarget{ref-Beall2015}{}}%
+\CSLLeftMargin{23. }%
 \CSLRightInline{\textbf{Beall BFN}, \textbf{Twiss MR}, \textbf{Smith
@@
 
-\leavevmode\hypertarget{ref-Henson2016}{}%
-\CSLLeftMargin{25. }
+\leavevmode\vadjust pre{\hypertarget{ref-Henson2016}{}}%
+\CSLLeftMargin{24. }%
 \CSLRightInline{\textbf{Henson MW}, \textbf{Pitre DM}, \textbf{Weckhorst
@@
 
-\leavevmode\hypertarget{ref-Baxter2014}{}%
-\CSLLeftMargin{26. }
+\leavevmode\vadjust pre{\hypertarget{ref-Baxter2014}{}}%
+\CSLLeftMargin{25. }%
 \CSLRightInline{\textbf{Baxter NT}, \textbf{Wan JJ}, \textbf{Schubert
@@
 
-\leavevmode\hypertarget{ref-LevyBooth2018}{}%
-\CSLLeftMargin{27. }
+\leavevmode\vadjust pre{\hypertarget{ref-LevyBooth2018}{}}%
+\CSLLeftMargin{26. }%
 \CSLRightInline{\textbf{Levy-Booth DJ}, \textbf{Giesbrecht IJW},
@@
 
-\leavevmode\hypertarget{ref-Edwards2015}{}%
-\CSLLeftMargin{28. }
+\leavevmode\vadjust pre{\hypertarget{ref-Edwards2015}{}}%
+\CSLLeftMargin{27. }%
 \CSLRightInline{\textbf{Edwards J}, \textbf{Johnson C},
@@
 
-\leavevmode\hypertarget{ref-Ettinger2017}{}%
-\CSLLeftMargin{29. }
+\leavevmode\vadjust pre{\hypertarget{ref-Ettinger2017}{}}%
+\CSLLeftMargin{28. }%
 \CSLRightInline{\textbf{Ettinger CL}, \textbf{Williams SL},
@@
 
-\leavevmode\hypertarget{ref-Graw2018}{}%
-\CSLLeftMargin{30. }
+\leavevmode\vadjust pre{\hypertarget{ref-Graw2018}{}}%
+\CSLLeftMargin{29. }%
 \CSLRightInline{\textbf{Graw MF}, \textbf{DAngelo G}, \textbf{Borchers
@@
 
-\leavevmode\hypertarget{ref-Johnston2016}{}%
-\CSLLeftMargin{31. }
+\leavevmode\vadjust pre{\hypertarget{ref-Johnston2016}{}}%
+\CSLLeftMargin{30. }%
 \CSLRightInline{\textbf{Johnston ER}, \textbf{Rodriguez-R LM},
@@
 
-\leavevmode\hypertarget{ref-Hassell2018}{}%
-\CSLLeftMargin{32. }
+\leavevmode\vadjust pre{\hypertarget{ref-Hassell2018}{}}%
+\CSLLeftMargin{31. }%
 \CSLRightInline{\textbf{Hassell N}, \textbf{Tinker KA}, \textbf{Moore
@@
 
-\begin{longtable}[]{@{}lrrrrrl@{}}
-\toprule
-\textbf{Study (Ref)} & \textbf{Samples} &
-\makecell[c]{\textbf{Total}\\\textbf{sequences}} &
-\makecell[c]{\textbf{Median}\\\textbf{sequences}} &
-\makecell[c]{\textbf{Range of}\\\textbf{sequences}} &
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0814}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0494}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1424}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1453}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1512}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.2762}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1541}}@{}}
+\toprule()
+\begin{minipage}[b]{\linewidth}\raggedright
+\textbf{Study (Ref)}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+\textbf{Samples}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+\makecell[c]{\textbf{Total}\\\textbf{sequences}}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+\makecell[c]{\textbf{Median}\\\textbf{sequences}}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+\makecell[c]{\textbf{Range of}\\\textbf{sequences}}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
 \makecell[c]{\textbf{Fold-difference}\\\textbf{between largest}\\\textbf{and smallest sample}}
-& \makecell[c]{\textbf{SRA study}\\\textbf{accession}}\tabularnewline
-\midrule
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\makecell[c]{\textbf{SRA study}\\\textbf{accession}}
+\end{minipage} \\
+\midrule()
 \endhead
-Bioethanol (22) & 95 & 3,972,943 & 16,015 & 3,688-356,136 & 96.6 &
-SRP055545\tabularnewline
-Human (23) & 490 & 20,909,768 & 32,505 & 10,523-430,415 & 40.9 &
-SRP062005\tabularnewline
-Lake (24) & 52 & 3,169,868 & 69,041 & 15,347-112,871 & 7.4 &
-SRP050963\tabularnewline
-Marine (25) & 7 & 1,391,396 & 193,464 & 133,516-254,060 & 1.9 &
-SRP068101\tabularnewline
-Mice (2) & 348 & 2,813,747 & 6,477 & 1,804-30,565 & 16.9 &
-SRP192323\tabularnewline
-Peromyscus (26) & 111 & 1,555,545 & 12,446 & 4,464-33,644 & 7.5 &
-SRP044050\tabularnewline
-Rainforest (27) & 69 & 946,295 & 11,561 & 4,932-37,767 & 7.7 &
-ERP023747\tabularnewline
-Rice (28) & 490 & 22,591,168 & 43,216 & 2,776-193,464 & 69.7 &
-SRP044745\tabularnewline
-Seagrass (29) & 286 & 4,130,454 & 13,567 & 1,803-45,191 & 25.1 &
-SRP092441\tabularnewline
-Sediment (30) & 58 & 1,154,174 & 17,584 & 7,685-68,321 & 8.9 &
-SRP097192\tabularnewline
-Soil (31) & 18 & 956,656 & 51,844 & 47,806-59,956 & 1.3 &
-ERP012016\tabularnewline
-Stream (32) & 201 & 21,162,574 & 90,159 & 9,175-390,964 & 42.6 &
-SRP075852\tabularnewline
-\bottomrule
+Bioethanol (21) & 95 & 3,972,943 & 16,015 & 3,688-356,136 & 96.6 &
+SRP055545 \\
+Human (22) & 490 & 20,909,768 & 32,505 & 10,523-430,415 & 40.9 &
+SRP062005 \\
+Lake (23) & 52 & 3,169,868 & 69,041 & 15,347-112,871 & 7.4 &
+SRP050963 \\
+Marine (24) & 7 & 1,391,396 & 193,464 & 133,516-254,060 & 1.9 &
+SRP068101 \\
+Mice (2) & 348 & 2,813,747 & 6,477 & 1,804-30,565 & 16.9 & SRP192323 \\
+Peromyscus (25) & 111 & 1,555,545 & 12,446 & 4,464-33,644 & 7.5 &
+SRP044050 \\
+Rainforest (26) & 69 & 946,295 & 11,561 & 4,932-37,767 & 7.7 &
+ERP023747 \\
+Rice (27) & 490 & 22,591,168 & 43,216 & 2,776-193,464 & 69.7 &
+SRP044745 \\
+Seagrass (28) & 286 & 4,130,454 & 13,567 & 1,803-45,191 & 25.1 &
+SRP092441 \\
+Sediment (29) & 58 & 1,154,174 & 17,584 & 7,685-68,321 & 8.9 &
+SRP097192 \\
+Soil (30) & 18 & 956,656 & 51,844 & 47,806-59,956 & 1.3 & ERP012016 \\
+Stream (31) & 201 & 21,162,574 & 90,159 & 9,175-390,964 & 42.6 &
+SRP075852 \\
+\bottomrule()
 \end{longtable}
@@
 \newpage
-
-\includegraphics{figure_1.png}
 
@@
 
-\newpage
-
-\includegraphics{figure_2.png}
-
 \textbf{Figure 2. Removing rare sequences from samples alters their
@@
 
-\newpage
-
-\includegraphics{figure_3.png}
-
 \textbf{Figure 3. Removing rare sequences from samples increases the
@@
 thresholds.
-
-\newpage
-
-\includegraphics{figure_4.png}
 
@@
 peformed.
-
-\newpage
-
-\includegraphics{figure_5.png}
 
@@
 
-\includegraphics{figure_s1.png}
-
 \textbf{Figure S1. Distribution of the number of sequences per sample in
@@
 sequences per sample.
-
-\newpage
-
-\includegraphics{figure_s2.png}
 
@@
 data. These data are summarized in Figure 1A.
-
-\newpage
-
-\includegraphics{figure_s3.png}
 
@@
 
-\newpage
-
-\includegraphics{figure_s4.png}
-
 \textbf{Figure S4. Removing rare sequences from samples alters their
@@
 applying different minimum abundance thresholds.
-
-\newpage
-
-\includegraphics{figure_s5.png}
 
@@
 
-\newpage
-
-\includegraphics{figure_s6.png}
-
 \textbf{Figure S6. Removing rare sequences from samples increases the
@@
 thresholds.
-
-\newpage
-
-\includegraphics{figure_s7.png}
 
@@
 peformed.
-
-\newpage
-
-\includegraphics{figure_s8.png}
 

### 2021-10-21 code/pool_intra_analysis.R

@@ -8,7 +8,7 @@ output_file <- arguments[length(arguments)]
 
 read_intra_analysis_file <- function(intra_analysis_file){
 
-	read_tsv(intra_analysis_file, col_type=cols(group=col_character())) %>%
+	read_tsv(intra_analysis_file, col_type=cols(group=col_character(), d_kl=col_double())) %>%
 		group_by(group, min_class) %>%
 		summarize(n_seqs_no=median(n_seqs.no),
 			n_seqs_yes=median(n_seqs.yes), n_seqs_yes_l=quantile(n_seqs.yes, 0.025),
@@ -19,7 +19,7 @@ read_intra_analysis_file <- function(intra_analysis_file){
 			h_yes=median(h.yes), h_yes_l=quantile(h.yes, 0.025), h_yes_u=quantile(h.yes, 0.975),
 			j_no=median(even.no), j_no_l=quantile(even.no, 0.025), j_no_u=quantile(even.no, 0.975),
 			j_yes=median(even.yes), j_yes_l=quantile(even.yes, 0.025), j_yes_u=quantile(even.yes, 0.975),
-			kl=median(d_kl), kl_l=quantile(d_kl, 0.025), kl_u=quantile(d_kl, 0.975),
+			kl=median(d_kl, na.rm=T), kl_l=quantile(d_kl, 0.025, na.rm=T), kl_u=quantile(d_kl, 0.975, na.rm=T),
 			rho=median(spearman), rho_l=quantile(spearman, 0.025), rho_u=quantile(spearman, 0.975),
 			diff=median(count_diff), diff_l=quantile(count_diff, 0.025), diff_u=quantile(count_diff, 0.975)
 			) %>%
